### 2017-06-13 paper.tex
First commit

\documentclass[twocolumn,superscriptaddress,aps]{revtex4-1}

\usepackage[utf8]{inputenc}

\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

\usepackage{bm}
\usepackage{cancel}
\usepackage{bbold}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}

\newcommand{\glnote}[1]{\textcolor{red}{[GL: #1]}}
\newcommand{\kcnote}[1]{\textcolor{red}{[KC: #1]}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}

\begin{document}

% ==============================================================================

\title{\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}}
\vspace{1cm}
\author{\small{\bf Gilles Louppe}\thanks{\texttt{g.louppe@nyu.edu}}}
\affiliation{New York University}
\author{\small{\bf Kyle Cranmer}\thanks{\texttt{kyle.cranmer@nyu.edu}}}
\affiliation{New York University}

\begin{abstract}

In this note, ... \glnote{todo.}

\end{abstract}

\maketitle

% ==============================================================================

\section{Introduction}

\glnote{Prescribed vs. implicit. See case of non-diff models in Balaji et al.}

% ==============================================================================

\section{Problem statement}

Consider a family of conditional densities $p(\mathbf{x}|\mathbf{\theta})$
defined implicitly through the simulation of a stochastic generative
process, where $\mathbf{x} \in \mathbb{R}^d$ is the observed data and
$\mathbf{\theta}$ are the parameters of interest. The simulation may involve
some complicated latent process, such that
\begin{equation}\label{eqn:p_x}
    p(\mathbf{x}|\mathbf{\theta}) = \int p(\mathbf{x}|\mathbf{z},\mathbf{\theta}) p(\mathbf{z}) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in \mathbb{R}^m$ is a latent variable providing an external source
of randomness.

We assume that we already have an accurate simulation of the stochastic process
underlying $p(\mathbf{x}|\mathbf{z},\mathbf{\theta})$, specified as a function $g(\cdot;
\theta) : \mathbb{R}^m \to \mathbb{R}^d$. That is,
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\mathbf{\theta}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\mathbf{\theta}) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
\end{equation}
In particular, the simulator $g$ is assumed to be a non-invertible function, that can
only be used to generate data in forward mode.
For this reason, evaluating the integral in Eqn.~\ref{eqn:p_x_sim} is intractable.
Importantly, and as increasingly found in science, we consider the additional constraint that $g$ is a non-differentiable
model, e.g. as specified by computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$, our goal is the inference
of the parameters of interest $\mathbf{\theta}^*$ that maximize the (marginal) log-likelihood of the observations:
\begin{equation}
    \mathbf{\theta}^* = \arg \max_\mathbf{\theta} \sum_i \log p(\mathbf{x}_i | \mathbf{\theta}).
\end{equation}
\glnote{Or rather, define the goal as the minimization of a divergence?}

% ==============================================================================

\section{Method}

\subsection{Variational optimization}

\subsection{Generative adversarial networks}

\subsection{Adversarial variational optimization}

% ==============================================================================

\section{Experiments}

\subsection{Toy problem}

\subsection{Physics example}

% ==============================================================================

\section{Related works}

\glnote{Implicit generative models.}
\glnote{ABC.}
\glnote{carl~\citep{cranmer2015approximating}.}
\glnote{Wood's papers.}
\glnote{CMA-ES.}

% ==============================================================================

\section{Summary}

% ==============================================================================

\section*{Acknowledgments}

KC and GL are both supported through NSF ACI-1450310, additionally KC is
supported through PHY-1505463 and PHY-1205376.

% ==============================================================================

\bibliographystyle{acm}
\bibliography{bibliography.bib}

\end{document}

### 2017-06-13 paper.tex
GANs

@@
 
-Consider a family of conditional densities $p(\mathbf{x}|\mathbf{\theta})$
+We consider a family of conditional densities $p(\mathbf{x}|\mathbf{\theta})$
 defined implicitly through the simulation of a stochastic generative
@@
 
-We assume that we already have an accurate simulation of the stochastic process
-underlying $p(\mathbf{x}|\mathbf{z},\mathbf{\theta})$, specified as a function $g(\cdot;
+We assume that we already have an accurate simulation of the stochastic generative process
+$p(\mathbf{x}|\mathbf{z},\mathbf{\theta})$, as specified through a deterministic function $g(\cdot;
 \theta) : \mathbb{R}^m \to \mathbb{R}^d$. That is,
@@
 \end{equation}
-In particular, the simulator $g$ is assumed to be a non-invertible function, that can
+The simulator $g$ is assumed to be a non-invertible function, that can
 only be used to generate data in forward mode.
@@
 Importantly, and as increasingly found in science, we consider the additional constraint that $g$ is a non-differentiable
-model, e.g. as specified by computer program.
+model, e.g. as when specified as a computer program.
 
@@
 \end{equation}
-\glnote{Or rather, define the goal as the minimization of a divergence?}
+\glnote{Redefine the goal as the minimization of a divergence between the true and the induced distribution?}
 
@@
 
-\section{Method}
+\section{Background}
+
+\subsection{Generative adversarial networks}
+
+Generative adversarial networks were first proposed by
+\cite{goodfellow2014generative} as a way to build an implicit generative model
+capable of producing samples from random noise $\mathbf{z}$. More specifically,
+a generative model $g(\cdot; \mathbf{\theta})$ is pit against an adversarial
+network $d$ whose antigonistic objective is to recognize real data $\mathbf{x}$
+from generated data $g(\mathbf{z}; \mathbf{\theta})$. Both models $g$ and $d$
+are trained simultaneously, in such a way that $g$ learns to maximally confuse
+its adversary $d$ (which happens when $g$ produces samples comparable to the
+observed data), while $d$ continuously adapts to changes in $g$. At the
+equilibrium and assuming enough capacity in the networks, it can be shown that
+$g$ induces a distribution  that is indistinguishable from the distribution of
+the observed data $\mathbf{x}$.
+
+\glnote{Explain WGAN, loss and optimum.}
 
@@
 
-\subsection{Generative adversarial networks}
+% ==============================================================================
 
-\subsection{Adversarial variational optimization}
+\section{Adversarial variational optimization}
 

### 2017-06-13 paper.tex
notations

@@
 
-We consider a family of conditional densities $p(\mathbf{x}|\mathbf{\theta})$
+We consider a family of parameterized densities $p_\mathbf{\theta}(\mathbf{x})$
 defined implicitly through the simulation of a stochastic generative
-process, where $\mathbf{x} \in \mathbb{R}^d$ is the observed data and
+process, where $\mathbf{x} \in \mathbb{R}^d$ is the data and
 $\mathbf{\theta}$ are the parameters of interest. The simulation may involve
@@
 \begin{equation}\label{eqn:p_x}
-    p(\mathbf{x}|\mathbf{\theta}) = \int p(\mathbf{x}|\mathbf{z},\mathbf{\theta}) p(\mathbf{z}) d\mathbf{z}
+    p_\mathbf{\theta}(\mathbf{x}) = \int p_\mathbf{\theta}(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) d\mathbf{z}
 \end{equation}
@@
 We assume that we already have an accurate simulation of the stochastic generative process
-$p(\mathbf{x}|\mathbf{z},\mathbf{\theta})$, as specified through a deterministic function $g(\cdot;
+that defines $p_\mathbf{\theta}(\mathbf{x}|\mathbf{z})$, as specified through a deterministic function $g(\cdot;
 \theta) : \mathbb{R}^m \to \mathbb{R}^d$. That is,
 \begin{equation}\label{eqn:p_x_sim}
-    p(\mathbf{x}|\mathbf{\theta}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\mathbf{\theta}) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
+    p_\mathbf{\theta}(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\mathbf{\theta}) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
 \end{equation}
@@
 
-Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$, our goal is the inference
-of the parameters of interest $\mathbf{\theta}^*$ that maximize the (marginal) log-likelihood of the observations:
+Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the (unknown) true distribution $p_r$, our goal is the inference
+of the parameters of interest $\mathbf{\theta}^*$ that minimize the divergence
+between $p_r$ and the modeled
+data distribution $p_\mathbf{\theta}$ induced by $g(\cdot; \mathbf{\theta})$. That is,
 \begin{equation}
-    \mathbf{\theta}^* = \arg \max_\mathbf{\theta} \sum_i \log p(\mathbf{x}_i | \mathbf{\theta}).
+    \mathbf{\theta}^* = \arg \min_\mathbf{\theta} \rho(p_r, p_\mathbf{\theta}),
 \end{equation}
-\glnote{Redefine the goal as the minimization of a divergence between the true and the induced distribution?}
+where $\rho$ is some divergence.
 
@@
 
-Generative adversarial networks were first proposed by
+Generative adversarial networks (GAN) were first proposed by
 \cite{goodfellow2014generative} as a way to build an implicit generative model
@@
 the observed data $\mathbf{x}$.
+
+More recently, ...
 

### 2017-06-13 paper.tex
save

@@
 Importantly, and as increasingly found in science, we consider the additional constraint that $g$ is a non-differentiable
-model, e.g. as when specified as a computer program.
+model, e.g. when specified as a computer program.
 
@@
 \end{equation}
-where $\rho$ is some divergence.
+where $\rho$ is some distance or divergence.
 
@@
 its adversary $d$ (which happens when $g$ produces samples comparable to the
-observed data), while $d$ continuously adapts to changes in $g$. At the
-equilibrium and assuming enough capacity in the networks, it can be shown that
-$g$ induces a distribution  that is indistinguishable from the distribution of
-the observed data $\mathbf{x}$.
+observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
+trained to optimality before each parameter update of the generator, it can
+be shown that the original adversarial learning procedure amounts to minimizing
+the Jensen-Shannon divergence between $p_r$ and $p_\theta$.
 
@@
 
-KC and GL are both supported through NSF ACI-1450310, additionally KC is
+GL and KL are both supported through NSF ACI-1450310, additionally KC is
 supported through PHY-1505463 and PHY-1205376.

### 2017-06-14 paper.tex
GAN section

@@
 between $p_r$ and the modeled
-data distribution $p_\mathbf{\theta}$ induced by $g(\cdot; \mathbf{\theta})$. That is,
+data distribution $p_\mathbf{\theta}$ induced by $g(\cdot; \mathbf{\theta})$ over $\mathbf{z}$. That is,
 \begin{equation}
@@
 
-Generative adversarial networks (GAN) were first proposed by
+Generative adversarial networks (GANs) were first proposed by
 \cite{goodfellow2014generative} as a way to build an implicit generative model
@@
 a generative model $g(\cdot; \mathbf{\theta})$ is pit against an adversarial
-network $d$ whose antigonistic objective is to recognize real data $\mathbf{x}$
+classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ whose antagonistic objective is to recognize real data $\mathbf{x}$
 from generated data $g(\mathbf{z}; \mathbf{\theta})$. Both models $g$ and $d$
@@
 be shown that the original adversarial learning procedure amounts to minimizing
-the Jensen-Shannon divergence between $p_r$ and $p_\theta$.
+the Jensen-Shannon divergence $\text{JSD}(p_r \parallel p_\theta)$ between $p_r$ and $p_\theta$.
 
-More recently, ...
-
-\glnote{Explain WGAN, loss and optimum.}
+As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
+difficult to train because of vanishing gradients as $d$ saturates, or because of
+unreliable updates when the training procedure is relaxed. As a remedy,
+Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
+setup in order to minimize the Wasserstein-1 distance $W(p_r, p_\theta)$ by
+replacing the adversarial classifier with a 1-Lipschitz adversarial critic
+$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. Under the WGAN-GP formulation of \cite{2017arXiv170400028G},
+training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize
+\begin{align}
+    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r} [d(\mathbf{x};\phi)]  \nonumber \\
+                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p_{\hat{\mathbf{x}}}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
+    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)]
+\end{align}
+where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} + (1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim p_r$ and $\tilde{\mathbf{x}} \sim p_\theta$.
 

### 2017-06-14 paper.tex
variational optimization

@@
 We consider a family of parameterized densities $p_\mathbf{\theta}(\mathbf{x})$
-defined implicitly through the simulation of a stochastic generative
-process, where $\mathbf{x} \in \mathbb{R}^d$ is the data and
-$\mathbf{\theta}$ are the parameters of interest. The simulation may involve
-some complicated latent process, such that
+defined implicitly through the simulation of a stochastic generative process,
+where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\mathbf{\theta}$ are the
+parameters of interest. The simulation may involve some complicated latent
+process, such that
 \begin{equation}\label{eqn:p_x}
@@
 
-We assume that we already have an accurate simulation of the stochastic generative process
-that defines $p_\mathbf{\theta}(\mathbf{x}|\mathbf{z})$, as specified through a deterministic function $g(\cdot;
-\theta) : \mathbb{R}^m \to \mathbb{R}^d$. That is,
+We assume that we already have an accurate simulation of the stochastic
+generative process that defines $p_\mathbf{\theta}(\mathbf{x}|\mathbf{z})$, as
+specified through a deterministic function $g(\cdot; \theta) : \mathbb{R}^m \to
+\mathbb{R}^d$. That is
 \begin{equation}\label{eqn:p_x_sim}
@@
 \end{equation}
-The simulator $g$ is assumed to be a non-invertible function, that can
-only be used to generate data in forward mode.
-For this reason, evaluating the integral in Eqn.~\ref{eqn:p_x_sim} is intractable.
-Importantly, and as increasingly found in science, we consider the additional constraint that $g$ is a non-differentiable
-model, e.g. when specified as a computer program.
+The simulator $g$ is assumed to be a non-invertible function, that can only be
+used to generate data in forward mode. For this reason, evaluating the integral
+in Eqn.~\ref{eqn:p_x_sim} is intractable. Importantly, and as increasingly found
+in science, we consider the additional constraint that $g$ is a
+non-differentiable model, e.g. when specified as a computer program.
 
-Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the (unknown) true distribution $p_r$, our goal is the inference
-of the parameters of interest $\mathbf{\theta}^*$ that minimize the divergence
-between $p_r$ and the modeled
-data distribution $p_\mathbf{\theta}$ induced by $g(\cdot; \mathbf{\theta})$ over $\mathbf{z}$. That is,
+Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
+(unknown) true distribution $p_r$, our goal is the inference of the parameters
+of interest $\mathbf{\theta}^*$ that minimize the divergence between $p_r$ and
+the modeled data distribution $p_\mathbf{\theta}$ induced by $g(\cdot;
+\mathbf{\theta})$ over $\mathbf{z}$. That is,
 \begin{equation}
@@
 a generative model $g(\cdot; \mathbf{\theta})$ is pit against an adversarial
-classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ whose antagonistic objective is to recognize real data $\mathbf{x}$
+classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
 from generated data $g(\mathbf{z}; \mathbf{\theta})$. Both models $g$ and $d$
@@
 replacing the adversarial classifier with a 1-Lipschitz adversarial critic
-$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. Under the WGAN-GP formulation of \cite{2017arXiv170400028G},
+$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
+for stabilizing the optimization procedure,
 training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize
@@
 \subsection{Variational optimization}
+
+Following \citep{2012arXiv1212.4507S}, variational optimization (VO) (also known as the search gradient algorithm~\citep{2011arXiv1106.4487W}) is a general
+optimization technique that can be used to form a differentiable bound
+on the optima of a non-differentiable function. Given a function $f$ to minimize, VO
+is based on the simple fact that
+\begin{equation}
+    \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c}) \leq \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] = U(\psi),
+\end{equation}
+where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\mathbf{c}$.
+That is, the minimum of a set of function values is always less than or equal
+to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
+can be updated to place its mass arbitrarily tight around the optimum $\mathbf{c}^* = \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c})$.
+
+Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
+\begin{align}\label{eqn:vo-grad}
+    \nabla_\psi U(\psi) &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
+    &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
+    &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
+    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c} \nabla_\psi \log q_\psi(\mathbf{c}))]
+\end{align}
+Effectively, this means that provided that the score function $\nabla_\psi \log q_\psi(\mathbf{c})$ of the proposal
+is known and that one can evaluate $f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then
+one can construct empirical estimates of Eqn.~\ref{eqn:vo-grad}, which can
+in turn be used to perform stochastic gradient descent (or a variant thereof)
+in order to minimize $U(\psi)$.
 

### 2017-06-15 paper.tex
save

@@
 used to generate data in forward mode. For this reason, evaluating the integral
-in Eqn.~\ref{eqn:p_x_sim} is intractable. Importantly, and as increasingly found
-in science, we consider the additional constraint that $g$ is a
-non-differentiable model, e.g. when specified as a computer program.
+in Eqn.~\ref{eqn:p_x_sim} is intractable. Importantly, and as commonly found
+in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$, e.g. when specified as a computer program.
 
@@
 \begin{align}\label{eqn:vo-grad}
-    \nabla_\psi U(\psi) &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
+    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
+    &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
     &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
     &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
-    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c} \nabla_\psi \log q_\psi(\mathbf{c}))]
+    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c})]
 \end{align}
@@
 \section{Adversarial variational optimization}
+
+\glnote{Naive approach by using VO on ${\cal L}_g$.}
+\glnote{Smarter approach by exploiting the fact that $\partial d / \partial x$ is known exactly.}
 

### 2017-06-16 paper.tex
save

@@
 \section{Problem statement}
+\label{sec:problem}
 
-We consider a family of parameterized densities $p_\mathbf{\theta}(\mathbf{x})$
+We consider a family of parameterized densities $p_\theta(\mathbf{x})$
 defined implicitly through the simulation of a stochastic generative process,
-where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\mathbf{\theta}$ are the
+where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
 parameters of interest. The simulation may involve some complicated latent
@@
 \begin{equation}\label{eqn:p_x}
-    p_\mathbf{\theta}(\mathbf{x}) = \int p_\mathbf{\theta}(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) d\mathbf{z}
+    p_\theta(\mathbf{x}) = \int p_\theta(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) d\mathbf{z}
 \end{equation}
@@
 We assume that we already have an accurate simulation of the stochastic
-generative process that defines $p_\mathbf{\theta}(\mathbf{x}|\mathbf{z})$, as
+generative process that defines $p_\theta(\mathbf{x}|\mathbf{z})$, as
 specified through a deterministic function $g(\cdot; \theta) : \mathbb{R}^m \to
-\mathbb{R}^d$. That is
+\mathbb{R}^d$. That is, we consider
+\begin{equation}\label{eqn:p_theta}
+    \mathbf{x} \sim p_\theta \equiv \mathbf{z} \sim p_z, \mathbf{x} = g(\mathbf{z}; \theta)
+\end{equation}
+such that the likelihood $p_\theta(\mathbf{x})$ can be rewritten as
 \begin{equation}\label{eqn:p_x_sim}
-    p_\mathbf{\theta}(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\mathbf{\theta}) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
+    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
 \end{equation}
-The simulator $g$ is assumed to be a non-invertible function, that can only be
+Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
 used to generate data in forward mode. For this reason, evaluating the integral
-in Eqn.~\ref{eqn:p_x_sim} is intractable. Importantly, and as commonly found
-in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$, e.g. when specified as a computer program.
+in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
+in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$,
+e.g. as when $g$ is specified as a computer program.
 
@@
 (unknown) true distribution $p_r$, our goal is the inference of the parameters
-of interest $\mathbf{\theta}^*$ that minimize the divergence between $p_r$ and
-the modeled data distribution $p_\mathbf{\theta}$ induced by $g(\cdot;
-\mathbf{\theta})$ over $\mathbf{z}$. That is,
+of interest $\theta^*$ that minimize the divergence between $p_r$ and
+the modeled data distribution $p_\theta$ induced by $g(\cdot;
+\theta)$ over $\mathbf{z}$. That is,
 \begin{equation}
-    \mathbf{\theta}^* = \arg \min_\mathbf{\theta} \rho(p_r, p_\mathbf{\theta}),
+    \theta^* = \arg \min_\theta \rho(p_r, p_\theta),
 \end{equation}
@@
 capable of producing samples from random noise $\mathbf{z}$. More specifically,
-a generative model $g(\cdot; \mathbf{\theta})$ is pit against an adversarial
+a generative model $g(\cdot; \theta)$ is pit against an adversarial
 classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
-from generated data $g(\mathbf{z}; \mathbf{\theta})$. Both models $g$ and $d$
+from generated data $\tilde{\mathbf{x}} = g(\mathbf{z}; \theta)$. Both models $g$ and $d$
 are trained simultaneously, in such a way that $g$ learns to maximally confuse
@@
 
-Following \citep{2012arXiv1212.4507S}, variational optimization (VO) (also known as the search gradient algorithm~\citep{2011arXiv1106.4487W}) is a general
+Following \citep{2012arXiv1212.4507S}, variational optimization (VO) (also known as the search gradient algorithm~\citep{2011arXiv1106.4487W} related to evolution strategies) is a general
 optimization technique that can be used to form a differentiable bound
@@
 
-\glnote{Naive approach by using VO on ${\cal L}_g$.}
+The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
+GANs inherently assume that the generator $g$ is a differentiable function. In
+the setting where we are not interested in learning an implicit model but are
+rather interested in the inference of parameters of a fixed
+non-differentiable simulator (as outlined in Section~\ref{sec:problem}),
+gradients $\nabla_\theta g$ either do not exist or cannot be accessed. As a
+result, gradients $\nabla_\theta {\cal L}_g$ cannot be constructed and the
+optimization procedure cannot be carried out.
+
+In this work, we propose to perform variational optimization
+on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
+of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
+over the parameters of $g$ and minimize in alternance the variational upper bounds
+\begin{align}
+    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
+    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
+\end{align}
+respectively over $\phi$ and $\psi$.
+When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
+obtained by sampling mini-batches of true and generated data, as ordinarily
+done in stochastic gradient descent. When updating $g$, gradient estimates
+of $\nabla_\psi U_g$ can be derived as described in Eqn.~\ref{eqn:vo-grad},
+which gives
+\begin{equation}
+    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)].
+\end{equation}
 \glnote{Smarter approach by exploiting the fact that $\partial d / \partial x$ is known exactly.}
+
+Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
+have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
+a distribution parameterized in terms of $\psi$:
+\begin{equation}\label{eqn:p_psi}
+    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
+\end{equation}
+Intuitively, this corresponds to a family of simulators, each configured
+with randomly sampled parameters $\theta \sim q_\psi$, whose collection
+of generated samples is optimized to approach the real data distribution $p_r$.
 

### 2017-06-16 paper.tex
save

@@
 
-In this note, ... \glnote{todo.}
+\glnote{
+In this note, lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Pellentesque justo arcu, facilisis id convallis in, vestibulum cursus leo.
+Pellentesque non nisl quis dolor dignissim elementum non ut orci.
+Mauris lobortis convallis elit, eget vulputate erat placerat id.
+Aliquam aliquet posuere eros, a molestie odio molestie ac.
+Morbi orci sem, consectetur non tincidunt sed, viverra non eros.
+Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
+Nulla maximus, ex et malesuada hendrerit, est eros cursus velit, sed scelerisque libero turpis nec dolor.
+Fusce vitae metus aliquet ipsum mollis aliquet in a velit.
+Nullam mi ante, dictum at urna quis, suscipit tempor nulla.
+Vivamus at suscipit lectus, a tristique leo.
+Sed et cursus lacus.
+}
 
@@
 done in stochastic gradient descent. When updating $g$, gradient estimates
-of $\nabla_\psi U_g$ can be derived as described in Eqn.~\ref{eqn:vo-grad},
-which gives
+of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
+That is,
 \begin{equation}
-    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)].
+    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
 \end{equation}
-\glnote{Smarter approach by exploiting the fact that $\partial d / \partial x$ is known exactly.}
+which we can approximate by sampling mini-batches of generated data
+\begin{equation}
+    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
+\end{equation}
+for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_z$.
+
+\glnote{
+While the above gradient estimate $\tilde{\nabla}_\psi U_g$ can readily be plugged
+into the adversarial training procedure, it may also exhibit a sampling
+variance that is larger than necessary, hence making the optimization unstable.
+However, one can exploit the fact that $-d(g(\mathbf{z}; \theta); \phi)$
+is a composition where $-d(\tilde{\mathbf{x}})$ is
+known and differentiable with respect to $\tilde{\mathbf{x}}$.
+In practice, using
+\begin{equation}
+    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
+\end{equation}
+works much better ite seems, but I am not sure to precisely understand why, nor whether this is correct...
+It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
+}
 
@@
 of generated samples is optimized to approach the real data distribution $p_r$.
+In particular, this formulation does not necessarily guarantee that the
+proposal distribution $q_\psi$ will place its mass arbitrarily tight
+around the parameters of interest $\theta^*$. \glnote{Describe regularization
+penalty on the parameters of the proposal to enforce that.}
 

### 2017-06-16 paper.tex
save

@@
 When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
-obtained by sampling mini-batches of true and generated data, as ordinarily
+obtained by evaluating the exact and known gradient of $U_d$ over
+mini-batches of true and generated data, as ordinarily
 done in stochastic gradient descent. When updating $g$, gradient estimates
@@
 \end{equation}
-which we can approximate by sampling mini-batches of generated data
+which we can approximate with mini-batches of generated data
 \begin{equation}
@@
 \end{equation}
-works much better ite seems, but I am not sure to precisely understand why, nor whether this is correct...
+works much better it seems, but I am not sure to precisely understand why, nor whether this is correct...
 It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
@@
 Intuitively, this corresponds to a family of simulators, each configured
-with randomly sampled parameters $\theta \sim q_\psi$, whose collection
+with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
 of generated samples is optimized to approach the real data distribution $p_r$.

### 2017-06-19 paper.tex
save

--- 2017-06-16 paper.tex
@@
 
-In this note, ... \glnote{todo.}
+Complex computer simulators are increasingly used across fields of science to
+describe generative models tying parameters of an underlying theory to
+experimental observations. Inference in this likelihood-free setup is often
+intractable, as simulators rarely provide a way to evaluate the likehood
+function for a given observation. In this note, we develop a likelihood-free
+inference algorithm for fitting a forward non-differentiable generative model to
+observed data. We adapt the adversarial training procedure of generative
+adversarial networks by replacing the implicit generative network with a
+domain-based scientific simulator, and solve the resulting non-differentiable
+minimax problem with variational optimization. Effectively, the adversarial
+variational optimization procedure results in learning an arbitrarily tight
+proposal distribution over simulator parameters, such that the respective
+hierarchical model matches the observed data.
+\glnote{Mention experimental results.}
+\glnote{Add 'so what?' conclusion.}
 
@@
 
-\glnote{Prescribed vs. implicit. See case of non-diff models in Balaji et al.}
+In many fields of science, computer simulators are used to describe complex
+processes that tie parameters of an underlying theory to high dimensional
+observations. In most cases, these implicit generative
+models~\citep{2016arXiv161003483M} are specified as forward stochastic processes
+that generate data. Because it is usually computationally intractable,
+simulators often do not provide a way to directly evaluate the likelihood
+function for a given observation,  which motivates a class of algorithms for
+likelihood-free inference.
+
+\glnote{State contributions clearly.}
 
@@
 When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
-obtained by sampling mini-batches of true and generated data, as ordinarily
+obtained by evaluating the exact and known gradient of $U_d$ over
+mini-batches of true and generated data, as ordinarily
 done in stochastic gradient descent. When updating $g$, gradient estimates
-of $\nabla_\psi U_g$ can be derived as described in Eqn.~\ref{eqn:vo-grad},
-which gives
+of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
+That is,
 \begin{equation}
-    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)].
+    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
 \end{equation}
-\glnote{Smarter approach by exploiting the fact that $\partial d / \partial x$ is known exactly.}
+which we can approximate with mini-batches of generated data
+\begin{equation}
+    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
+\end{equation}
+for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_z$.
+\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?}
 
@@
 Intuitively, this corresponds to a family of simulators, each configured
-with randomly sampled parameters $\theta \sim q_\psi$, whose collection
-of generated samples is optimized to approach the real data distribution $p_r$.
+with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
+of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
+However, this formulation does not necessarily guarantee that the
+proposal distribution $q_\psi$ will place its mass arbitrarily tight
+around the parameters of interest $\theta^*$. For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
+with a regularization term corresponding to the differential entropy $H$ of
+the proposal distribution. That is,
+\begin{equation}
+    U_g = \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] + \lambda H(q_\psi)
+\end{equation}
+where $\lambda \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
+between the generator objective and the tightness of the proposal distribution.
+
+% \glnote{
+% While the above gradient estimate $\tilde{\nabla}_\psi U_g$ can readily be plugged
+% into the adversarial training procedure, it may also exhibit a sampling
+% variance that is larger than necessary, hence making the optimization unstable.
+% However, one can exploit the fact that $-d(g(\mathbf{z}; \theta); \phi)$
+% is a composition where $-d(\tilde{\mathbf{x}})$ is
+% known and differentiable with respect to $\tilde{\mathbf{x}}$.
+% In practice, using
+% \begin{equation}
+%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
+% \end{equation}
+% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
+% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
+% With large mini-batches, both approaches seem to work fine, though the former seems to be less stable.
+% }
 
@@
 \glnote{Implicit generative models.}
-\glnote{ABC.}
+\glnote{classifier ABC.}
 \glnote{carl~\citep{cranmer2015approximating}.}
-\glnote{Wood's papers.}
 \glnote{CMA-ES.}

### 2017-06-19 paper.tex
save

@@
 describe generative models tying parameters of an underlying theory to
-experimental observations. Inference in this likelihood-free setup is often
+experimental observations. However, inference in this likelihood-free setup is often
 intractable, as simulators rarely provide a way to evaluate the likehood
@@
 observations. In most cases, these implicit generative
-models~\citep{2016arXiv161003483M} are specified as forward stochastic processes
-that generate data. Because it is usually computationally intractable,
-simulators often do not provide a way to directly evaluate the likelihood
-function for a given observation,  which motivates a class of algorithms for
-likelihood-free inference.
+models~\citep{2016arXiv161003483M} are specified as imperative implementations
+of forward stochastic processes that generate data. Because it is usually
+computationally intractable, simulators often do not provide a way to directly
+evaluate the likelihood function for a given observation,  which motivates a
+class of algorithms for likelihood-free inference.
 
@@
 The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
-GANs inherently assume that the generator $g$ is a differentiable function. In
-the setting where we are not interested in learning an implicit model but are
+GANs inherently assumes that the generator $g$ is a differentiable function. In
+the setting where we are not interested in learning the implicit model itself but are
 rather interested in the inference of parameters of a fixed

### 2017-06-19 paper.tex
save

@@
 between the generator objective and the tightness of the proposal distribution.
+For large values of $\lambda$, the procedure is constrained to fit a proposal
+distribution with low entropy, which has the effect of concentrating its density
+tightly around one or a few $\theta$ values. On the other hand, for small values of $\lambda$,
+proposal distributions with larger entropy are not penalized, which may result
+in learning a smeared variation of the original simulator.
 

### 2017-06-20 paper.tex
save

@@
 describe generative models tying parameters of an underlying theory to
-experimental observations. However, inference in this likelihood-free setup is often
-intractable, as simulators rarely provide a way to evaluate the likehood
+experimental observations. Inference in this setup is often
+difficult, as simulators rarely provide a way to evaluate the likehood
 function for a given observation. In this note, we develop a likelihood-free
@@
 
-In many fields of science, computer simulators are used to describe complex
-processes that tie parameters of an underlying theory to high dimensional
-observations. In most cases, these implicit generative
-models~\citep{2016arXiv161003483M} are specified as imperative implementations
-of forward stochastic processes that generate data. Because it is usually
-computationally intractable, simulators often do not provide a way to directly
-evaluate the likelihood function for a given observation,  which motivates a
-class of algorithms for likelihood-free inference.
+In many fields of science such as particle physics, climatology or population
+genetics, computer simulators are used to describe complex processes that tie
+parameters of an underlying theory to high dimensional observations. In most
+cases, these implicit generative models~\citep{2016arXiv161003483M} are
+specified as imperative implementations of forward stochastic processes that
+generate data. Because it is usually computationally intractable, simulators
+often do not provide a way to directly evaluate the likelihood function for a
+given observation, thereby making inference difficult.
 
-\glnote{State contributions clearly.}
+In this note, we develop a likelihood-free inference algorithm for fitting a
+forward non-differentiable generative model to observed data. We propose to
+adapt the adversarial training procedure of generative adversarial
+networks~\cite{goodfellow2014generative} by replacing the implicit generative
+network with a domain-based scientific simulator, and solve the resulting
+non-differentiable minimax problem with variational
+optimization~\citep{2012arXiv1212.4507S}. The adversarial variational
+optimization procedure results in learning a proposal distribution over
+simulator parameters, hence producing an abitrarily tight family of models whose
+joint collection of generated samples matches the observed data.
 
@@
 where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} + (1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim p_r$ and $\tilde{\mathbf{x}} \sim p_\theta$.
+\glnote{Mention that we use WGAN-GP in this note, but that the VO procedure can be coupled other vairants of GANs.}
 
@@
 
+\glnote{add algorithm}
+
 % \glnote{
@@
 \glnote{Implicit generative models.}
-\glnote{classifier ABC.}
+\glnote{gan + classifier ABC.}
+
+\glnote{likelihood free inference, density ratio by classification}
 \glnote{carl~\citep{cranmer2015approximating}.}
-\glnote{CMA-ES.}
 

### 2017-06-20 paper.tex
save

@@
 experimental observations. Inference in this setup is often
-difficult, as simulators rarely provide a way to evaluate the likehood
+difficult, as simulators rarely provide a way to directly evaluate the likehood
 function for a given observation. In this note, we develop a likelihood-free
@@
 specified as imperative implementations of forward stochastic processes that
-generate data. Because it is usually computationally intractable, simulators
-often do not provide a way to directly evaluate the likelihood function for a
+generate data. Because it is usually computationally intractable, most simulators
+do not provide a way to directly evaluate the likelihood function for a
 given observation, thereby making inference difficult.
@@
 \subsection{Generative adversarial networks}
+\label{sec:gans}
 
@@
 where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} + (1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim p_r$ and $\tilde{\mathbf{x}} \sim p_\theta$.
-\glnote{Mention that we use WGAN-GP in this note, but that the VO procedure can be coupled other vairants of GANs.}
 
@@
 
+\begin{figure*}
+    \begin{minipage}{\linewidth}
+    \begin{algorithm}[H]
+    \caption{Adversarial variational optimization.}
+
+    \begin{flushleft}
+        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N$, simulator $g$;\\
+        {\it Outputs:} proposal distribution $q_\psi(\theta)$;\\
+        {\it Hyper-parameters:} Number $T$ of training iterations,
+                                Number $K$ of gradient steps to update $d$,
+                                Size $M$ of the minibatches.
+    \end{flushleft}
+
+    \label{alg:avo}
+    \begin{algorithmic}[1]
+        \State{\glnote{todo!}}
+        \For{$t=1$ to $T$}
+            % \For{$k=1$ to $K$} \Comment{Update $r$}
+            %     \State{Sample minibatch $\{x_m, z_m, s_m = f(x_m;\theta_f) \}_{m=1}^M$} of size $M$;
+            %     \State{With $\theta_f$ fixed, update $r$ by ascending its stochastic gradient $\nabla_{\theta_r} E(\theta_f, \theta_r) :=$
+            %     $$\nabla_{\theta_r} \sum_{m=1}^M \log p_{\theta_r}(z_m|s_m)  ;$$}
+            % \EndFor
+            % \State{Sample minibatch $\{x_m, y_m, z_m, s_m = f(x_m;\theta_f)  \}_{m=1}^M$} of size $M$; \Comment{Update $f$}
+            % \State{With $\theta_r$ fixed, update $f$ by descending its stochastic gradient $\nabla_{\theta_f} E(\theta_f, \theta_r) :=$
+            % $$\nabla_{\theta_f}  \sum_{m=1}^M \left[ -\log p_{\theta_f}(y_m|x_m)  +\log p_{\theta_r}(z_m|s_m)  \right],$$
+            % \indent where $p_{\theta_f}(y_m|x_m)$ denotes $\mathbf{1}(y_m=0)(1-s_m) + \mathbf{1}(y_m=1)s_m$;}
+        \EndFor
+    \end{algorithmic}
+    \end{algorithm}
+    \end{minipage}
+\end{figure*}
+
 The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
-GANs inherently assumes that the generator $g$ is a differentiable function. In
+GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
 the setting where we are not interested in learning the implicit model itself but are
 rather interested in the inference of parameters of a fixed
-non-differentiable simulator (as outlined in Section~\ref{sec:problem}),
+non-differentiable simulator (Section~\ref{sec:problem}),
 gradients $\nabla_\theta g$ either do not exist or cannot be accessed. As a
@@
 
-\glnote{add algorithm}
+For completeness, Algorithm~\ref{alg:avo} finally outlines the adversarial variational
+optimization procedure when coupled with WGAN-GP.
 

### 2017-06-20 paper.tex
save

@@
 for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_z$.
-\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?}
+\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
+In practice, using
+\begin{equation}
+    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
+\end{equation}
+works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
+It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
+}
 
@@
 optimization procedure when coupled with WGAN-GP.
-
-% \glnote{
-% While the above gradient estimate $\tilde{\nabla}_\psi U_g$ can readily be plugged
-% into the adversarial training procedure, it may also exhibit a sampling
-% variance that is larger than necessary, hence making the optimization unstable.
-% However, one can exploit the fact that $-d(g(\mathbf{z}; \theta); \phi)$
-% is a composition where $-d(\tilde{\mathbf{x}})$ is
-% known and differentiable with respect to $\tilde{\mathbf{x}}$.
-% In practice, using
-% \begin{equation}
-%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
-% \end{equation}
-% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
-% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
-% With large mini-batches, both approaches seem to work fine, though the former seems to be less stable.
-% }
 

### 2017-06-20 paper.tex
add algo

@@
         {\it Outputs:} proposal distribution $q_\psi(\theta)$;\\
-        {\it Hyper-parameters:} Number $T$ of training iterations,
-                                Number $K$ of gradient steps to update $d$,
-                                Size $M$ of the minibatches.
+        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$. 
     \end{flushleft}
@@
     \begin{algorithmic}[1]
-        \State{\glnote{todo!}}
-        \For{$t=1$ to $T$}
-            % \For{$k=1$ to $K$} \Comment{Update $r$}
-            %     \State{Sample minibatch $\{x_m, z_m, s_m = f(x_m;\theta_f) \}_{m=1}^M$} of size $M$;
-            %     \State{With $\theta_f$ fixed, update $r$ by ascending its stochastic gradient $\nabla_{\theta_r} E(\theta_f, \theta_r) :=$
-            %     $$\nabla_{\theta_r} \sum_{m=1}^M \log p_{\theta_r}(z_m|s_m)  ;$$}
-            % \EndFor
-            % \State{Sample minibatch $\{x_m, y_m, z_m, s_m = f(x_m;\theta_f)  \}_{m=1}^M$} of size $M$; \Comment{Update $f$}
-            % \State{With $\theta_r$ fixed, update $f$ by descending its stochastic gradient $\nabla_{\theta_f} E(\theta_f, \theta_r) :=$
-            % $$\nabla_{\theta_f}  \sum_{m=1}^M \left[ -\log p_{\theta_f}(y_m|x_m)  +\log p_{\theta_r}(z_m|s_m)  \right],$$
-            % \indent where $p_{\theta_f}(y_m|x_m)$ denotes $\mathbf{1}(y_m=0)(1-s_m) + \mathbf{1}(y_m=1)s_m$;}
+        \While{$\psi$ has not converged}
+            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
+                \State{Sample a mini-batch $\{\mathbf{x}_m \}_{m=1}^M \sim p_r$, $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$, $\{\epsilon_m \}_{m=1}^M \sim U[0,1]$.}
+                \For{$m=1$ to $M$}
+                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
+                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
+                    \State{$L^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                 \EndFor
+                \State{$\phi \leftarrow \text{RMSProp}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M L^{(m)})$}
+            \EndFor
+            \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
+            \State{$\psi \leftarrow \text{RMSProp}(\tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m) + \nabla_\psi \gamma q_\psi(\theta_m) \log q_\psi(\theta_m))$}
+        \EndWhile
     \end{algorithmic}
@@
 \begin{equation}
-    U_g = \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] + \lambda H(q_\psi)
+    U_g = \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] + \gamma H(q_\psi)
 \end{equation}
-where $\lambda \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
+where $\gamma \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
 between the generator objective and the tightness of the proposal distribution.
-For large values of $\lambda$, the procedure is constrained to fit a proposal
+For large values of $\gamma$, the procedure is constrained to fit a proposal
 distribution with low entropy, which has the effect of concentrating its density
-tightly around one or a few $\theta$ values. On the other hand, for small values of $\lambda$,
+tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
 proposal distributions with larger entropy are not penalized, which may result

### 2017-06-21 paper.tex
save

@@
     \begin{flushleft}
-        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N$, simulator $g$;\\
-        {\it Outputs:} proposal distribution $q_\psi(\theta)$;\\
+        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N \sim p_r$, simulator $g$.\\
+        {\it Outputs:} proposal distribution $q_\psi(\theta)$, such that $p_r \approx p_\psi$.\\
         {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
@@
     \begin{algorithmic}[1]
+        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known likelihood)}$}
         \While{$\psi$ has not converged}
@@
             \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
-            \State{$\psi \leftarrow \text{RMSProp}(\tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m) + \nabla_\psi \gamma q_\psi(\theta_m) \log q_\psi(\theta_m))$}
+            \State{$\psi \leftarrow \text{RMSProp}(\tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m) + \gamma  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m))$}
         \EndWhile
@@
 \end{equation}
-for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_z$.
+for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
 \glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
@@
 
-For completeness, Algorithm~\ref{alg:avo} finally outlines the adversarial variational
-optimization procedure when coupled with WGAN-GP.
+For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
+optimization procedure, as built on top of WGAN-GP.
+Let us note however that the variational relaxation could similarly be coupled with
+other variants of GANs. \glnote{add note on rmsprop vs. adam}
 

### 2017-06-21 paper.tex
save

@@
 
-Following \citep{2012arXiv1212.4507S}, variational optimization (VO) (also known as the search gradient algorithm~\citep{2011arXiv1106.4487W} related to evolution strategies) is a general
+Following \citep{2012arXiv1212.4507S}, variational optimization (VO) is a general
 optimization technique that can be used to form a differentiable bound
@@
 in order to minimize $U(\psi)$.
+\glnote{Consider instead natural evolution strategies, which should be more stable?}
+\glnote{Rewrite in terms of NES!!}
 

### 2017-06-21 paper.tex
natural evolution strategies

--- 2017-06-21 paper.tex
@@
 domain-based scientific simulator, and solve the resulting non-differentiable
-minimax problem with variational optimization. Effectively, the adversarial
-variational optimization procedure results in learning an arbitrarily tight
-proposal distribution over simulator parameters, such that the respective
-hierarchical model matches the observed data.
+minimax problem by minimizing variationial upper bounds of the adversarial objectives.
+Effectively, the procedure results in learning an arbitrarily tight
+proposal distribution over simulator parameters, such that the corresponding
+marginal distribution  matches the observed data.
 \glnote{Mention experimental results.}
@@
 network with a domain-based scientific simulator, and solve the resulting
-non-differentiable minimax problem with variational
-optimization~\citep{2012arXiv1212.4507S}. The adversarial variational
-optimization procedure results in learning a proposal distribution over
+non-differentiable minimax problem by minimizing variational upper bounds of the adversarial objectives
+respectively with stochastic gradient descent and  natural evolution strategies~\citep{2011arXiv1106.4487W}.
+The procedure results in learning a proposal distribution over
 simulator parameters, hence producing an abitrarily tight family of models whose
@@
 
-\subsection{Variational optimization}
+\subsection{Natural evolution strategies}
 
-Following \citep{2012arXiv1212.4507S}, variational optimization (VO) (also known as the search gradient algorithm~\citep{2011arXiv1106.4487W} related to evolution strategies) is a general
-optimization technique that can be used to form a differentiable bound
-on the optima of a non-differentiable function. Given a function $f$ to minimize, VO
-is based on the simple fact that
+Evolution strategies~\citep{2011arXiv1106.4487W} and variational optimization~\cite{2012arXiv1212.4507S} are general
+optimization techniques that can be used to form a differentiable bound
+on the optima of a non-differentiable function. Given a function $f$ to minimize,
+these techniques are based on the simple fact that
 \begin{equation}
@@
 Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
-\begin{align}\label{eqn:vo-grad}
+\begin{align}\label{eqn:approx-grad}
     \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
@@
 is known and that one can evaluate $f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then
-one can construct empirical estimates of Eqn.~\ref{eqn:vo-grad}, which can
+one can construct empirical estimates of Eqn.~\ref{eqn:approx-grad}, which can
 in turn be used to perform stochastic gradient descent (or a variant thereof)
 in order to minimize $U(\psi)$.
+
+In this simple form, stochastic gradient descent based on
+Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
+of scale invariance. To avoid the dependence on the parameterization of the
+proposal distribution, natural gradients~\citep{amari1998natural} can be used
+instead to point towards the steepest descent in the space of realizable
+proposal distributions, rather than in the space of their parameters. Effectively,
+the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
+corrects the gradient of $U$ according to the local curvature of the KL-divergence
+surface of proposal distributions, which results in forming
+\begin{equation}\label{eqn:approx-grad-nat}
+    \tilde{\nabla}_\psi U(\psi) = \mathbf{F}(\psi)^{-1} \nabla_\psi U(\psi),
+\end{equation}
+where $\mathbf{F}(\psi)$ is the Fisher information matrix
+\begin{equation}
+    \mathbf{F}(\psi) = \mathbb{E}_{\theta \sim q_\phi}[ \nabla_\psi \log q_\psi(\theta) \nabla_\psi \log q_\psi(\theta)^{\top} ]
+\end{equation}
+and can be estimated from samples, in particular by reusing the derivatives $\nabla_\psi \log q_\psi(\theta)$ already
+evaluated to form $\nabla_\psi U(\psi)$.
 
@@
             \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
-            \State{$\psi \leftarrow \text{RMSProp}(\tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m) + \gamma  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m))$}
+            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
+            \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
+            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$)}
+            \State{$\psi \leftarrow \text{SGD}(\mathbf{F}(\psi)^{-1} \nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi)$}
         \EndWhile
@@
 
-In this work, we propose to perform variational optimization
-on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
-of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
-over the parameters of $g$ and minimize in alternance the variational upper bounds
+In this work, we propose to rely on evolution strategies to minimize ${\cal
+L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
+specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
+parameters of $g$ and minimize in alternance the variational upper bounds
 \begin{align}
@@
     U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
-\end{align}
-respectively over $\phi$ and $\psi$.
-When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
-obtained by evaluating the exact and known gradient of $U_d$ over
-mini-batches of true and generated data, as ordinarily
-done in stochastic gradient descent. When updating $g$, gradient estimates
-of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
-That is,
-\begin{equation}
-    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
-\end{equation}
-which we can approximate with mini-batches of generated data
-\begin{equation}
-    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
-\end{equation}
-for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
-\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$ is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
-In practice, using
-\begin{equation}
-    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
-\end{equation}
-works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
-It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
+\end{align} respectively over $\phi$ and $\psi$.
+When updating
+$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
+evaluating the exact and known gradient of $U_d$ over mini-batches of true and
+generated data, as ordinarily done in stochastic gradient descent. When updating
+$g$, estimates of $\nabla_\psi U_g$ and $\tilde{\nabla}_\psi U_g$ can be derived with forward
+simulations, as described in the previous section.
+% That is,
+% \begin{equation}
+%     \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
+% \end{equation} which we can approximate with mini-batches of
+% generated data
+% \begin{equation}
+%     \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
+% \end{equation}
+% for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
+\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
+is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
+% In practice, using
+% \begin{equation}
+%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
+% \end{equation}
+% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
+% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
 }
 
+\glnote{Bayesian vs. frequentist view.}
 Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
@@
 For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
-optimization procedure, as built on top of WGAN-GP.
+optimization procedure, as built on top of WGAN-GP and NES.
 Let us note however that the variational relaxation could similarly be coupled with
-other variants of GANs. \glnote{add note on rmsprop vs. adam}
+other variants of GANs and/or of evolution strategies.
 

### 2017-06-22 paper.tex
save

@@
 proposal distribution over simulator parameters, such that the corresponding
-marginal distribution  matches the observed data.
+marginal distribution of the generated data matches the observations.
 \glnote{Mention experimental results.}
@@
 network with a domain-based scientific simulator, and solve the resulting
-non-differentiable minimax problem by minimizing variational upper bounds of the adversarial objectives
-respectively with stochastic gradient descent and  natural evolution strategies~\citep{2011arXiv1106.4487W}.
-The procedure results in learning a proposal distribution over
+non-differentiable minimax problem by minimizing variational upper bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the
+adversarial objectives. The procedure results in learning a proposal distribution over
 simulator parameters, hence producing an abitrarily tight family of models whose
@@
 
-\subsection{Natural evolution strategies}
+\subsection{Variational optimization}
 
@@
 one can construct empirical estimates of Eqn.~\ref{eqn:approx-grad}, which can
-in turn be used to perform stochastic gradient descent (or a variant thereof)
-in order to minimize $U(\psi)$.
+in turn be used to minimize $U(\psi)$ with stochastic gradient descent (or a variant thereof, like Adam~\cite{2014arXiv1412.6980K} to account for scaling invariance and robustness to noisy gradients).
 
-In this simple form, stochastic gradient descent based on
-Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
-of scale invariance. To avoid the dependence on the parameterization of the
-proposal distribution, natural gradients~\citep{amari1998natural} can be used
-instead to point towards the steepest descent in the space of realizable
-proposal distributions, rather than in the space of their parameters. Effectively,
-the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
-corrects the gradient of $U$ according to the local curvature of the KL-divergence
-surface of proposal distributions, which results in forming
-\begin{equation}\label{eqn:approx-grad-nat}
-    \tilde{\nabla}_\psi U(\psi) = \mathbf{F}(\psi)^{-1} \nabla_\psi U(\psi),
-\end{equation}
-where $\mathbf{F}(\psi)$ is the Fisher information matrix
-\begin{equation}
-    \mathbf{F}(\psi) = \mathbb{E}_{\theta \sim q_\phi}[ \nabla_\psi \log q_\psi(\theta) \nabla_\psi \log q_\psi(\theta)^{\top} ]
-\end{equation}
-and can be estimated from samples, in particular by reusing the derivatives $\nabla_\psi \log q_\psi(\theta)$ already
-evaluated to form $\nabla_\psi U(\psi)$.
+% In this simple form, stochastic gradient descent based on
+% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
+% of scale invariance. To avoid the dependence on the parameterization of the
+% proposal distribution, natural gradients~\citep{amari1998natural} can be used
+% instead to point towards the steepest descent in the space of realizable
+% proposal distributions, rather than in the space of their parameters. Effectively,
+% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
+% corrects the gradient of $U$ according to the local curvature of the KL-divergence
+% surface of proposal distributions, which results in forming
+% \begin{equation}\label{eqn:approx-grad-nat}
+%     \tilde{\nabla}_\psi U(\psi) = \mathbf{F}(\psi)^{-1} \nabla_\psi U(\psi),
+% \end{equation}
+% where $\mathbf{F}(\psi)$ is the Fisher information matrix
+% \begin{equation}
+%     \mathbf{F}(\psi) = \mathbb{E}_{\theta \sim q_\phi}[ \nabla_\psi \log q_\psi(\theta) \nabla_\psi \log q_\psi(\theta)^{\top} ]
+% \end{equation}
+% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\psi \log q_\psi(\theta)$ already
+% evaluated to form $\nabla_\psi U(\psi)$.
 
@@
                     \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
-                    \State{$L^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
+                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                 \EndFor
-                \State{$\phi \leftarrow \text{RMSProp}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M L^{(m)})$}
+                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
             \EndFor
@@
             \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
-            \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
-            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$)}
-            \State{$\psi \leftarrow \text{SGD}(\mathbf{F}(\psi)^{-1} \nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi)$}
+            % \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
+            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$}
+            \State{$\psi \leftarrow \text{Adam}(\nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi))$}
         \EndWhile
@@
 
-In this work, we propose to rely on evolution strategies to minimize ${\cal
+In this work, we propose to rely on variational optimization to minimize ${\cal
 L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
@@
 For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
-optimization procedure, as built on top of WGAN-GP and NES.
+optimization procedure, as built on top of WGAN-GP.
 Let us note however that the variational relaxation could similarly be coupled with

### 2017-06-22 paper.tex
save

@@
 \end{align}
-Effectively, this means that provided that the score function $\nabla_\psi \log q_\psi(\mathbf{c})$ of the proposal
-is known and that one can evaluate $f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then
-one can construct empirical estimates of Eqn.~\ref{eqn:approx-grad}, which can
-in turn be used to minimize $U(\psi)$ with stochastic gradient descent (or a variant thereof, like Adam~\cite{2014arXiv1412.6980K} to account for scaling invariance and robustness to noisy gradients).
+Effectively, this means that provided that the score function $\nabla_\psi \log
+q_\psi(\mathbf{c})$ of the proposal is known and that one can evaluate
+$f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then one can construct empirical
+estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
+$U(\psi)$ with stochastic gradient descent (or a variant thereof, like
+Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
+algorithm~\citep{2011arXiv1106.4487W}, to account for scaling invariance and
+robustness to noisy gradients).
 
@@
 generated data, as ordinarily done in stochastic gradient descent. When updating
-$g$, estimates of $\nabla_\psi U_g$ and $\tilde{\nabla}_\psi U_g$ can be derived with forward
+$g$, estimates of $\nabla_\psi U_g$ can be derived with forward
 simulations, as described in the previous section.
@@
 optimization procedure, as built on top of WGAN-GP.
-Let us note however that the variational relaxation could similarly be coupled with
+Let us note that the variational relaxation could similarly be coupled with
 other variants of GANs and/or of evolution strategies.

### 2017-06-22 paper.tex
save

@@
 Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
-algorithm~\citep{2011arXiv1106.4487W}, to account for scaling invariance and
+algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
 robustness to noisy gradients).
@@
 simulations, as described in the previous section.
-% That is,
-% \begin{equation}
-%     \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
-% \end{equation} which we can approximate with mini-batches of
-% generated data
-% \begin{equation}
-%     \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
-% \end{equation}
-% for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
+That is,
+\begin{equation}
+    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
+\end{equation} which we can approximate with mini-batches of
+generated data
+\begin{equation}
+    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
+\end{equation}
+for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
 \glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
@@
 }
+For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
+optimization procedure, as built on top of WGAN-GP.
+Obviously, the variational relaxation could similarly be coupled with
+other variants of GANs and/or of evolution strategies.
 
-\glnote{Bayesian vs. frequentist view.}
+\glnote{Discuss Bayesian vs. frequentist view; See Kyle comments on Slack.}
 Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
@@
 in learning a smeared variation of the original simulator.
-
-For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
-optimization procedure, as built on top of WGAN-GP.
-Let us note that the variational relaxation could similarly be coupled with
-other variants of GANs and/or of evolution strategies.
 

### 2017-06-22 paper.tex
save

@@
 
-\glnote{Discuss Bayesian vs. frequentist view; See Kyle comments on Slack.}
 Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
@@
 of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
-However, this formulation does not necessarily guarantee that the
+More formally, the learned model $p_\psi(\mathbf{x})$ therefore corresponds to the marginal distribution
+$\int q_\psi(\theta) p_\theta(\mathbf{x}) d\theta$ of the generated data.
+
+In consequence, the proposed inference algorithm does not necessarily guarantee that the
 proposal distribution $q_\psi$ will place its mass arbitrarily tight
-around the parameters of interest $\theta^*$. For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
+around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
+For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
 with a regularization term corresponding to the differential entropy $H$ of

### 2017-06-22 paper.tex
save

@@
 
-\glnote{Implicit generative models.}
-\glnote{gan + classifier ABC.}
-
-\glnote{likelihood free inference, density ratio by classification}
-\glnote{carl~\citep{cranmer2015approximating}.}
+As reviewed in \cite{2016arXiv161003483M}, likelihood-free inference is
+intimately tied to a class of algorithms that can be framed as density
+estimation-by-comparison. In most cases, these algorithms are formulated as an
+iterative two-step process where the model distribution is first compared to the
+true data distribution and then updated to make it more comparable to the
+latter. Closest to our work are inference
+algorithms~\citep{Neal:2007zz,gutmann2012noise,goodfellow2014generative,cranmer2015approximating,2016arXiv161110242D}
+that rely on a classifier between generated and true data to estimate the
+density ratio between the two distributions and make parameter updates.
+\glnote{ABC + Classifier ABC.}
+\glnote{Explain how this work is different.}
 

### 2017-06-22 paper.tex
kyle's comments (sections I and II)

@@
 
-In this note, we develop a likelihood-free inference algorithm for fitting a
-forward non-differentiable generative model to observed data. We propose to
-adapt the adversarial training procedure of generative adversarial
+In this note, we develop a likelihood-free inference algorithm for the point
+estimation from observed data of the parameters of a forward non-differentiable
+generative model. We propose to adapt the adversarial
+training procedure of generative adversarial
 networks~\cite{goodfellow2014generative} by replacing the implicit generative
 network with a domain-based scientific simulator, and solve the resulting
-non-differentiable minimax problem by minimizing variational upper bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the
-adversarial objectives. The procedure results in learning a proposal distribution over
+non-differentiable minimax problem by minimizing variational upper
+bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
+objectives. The procedure results in learning a proposal distribution over
 simulator parameters, hence producing an abitrarily tight family of models whose
@@
 \begin{equation}\label{eqn:p_x}
-    p_\theta(\mathbf{x}) = \int p_\theta(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) d\mathbf{z}
+    p_\theta(\mathbf{x}) = \int p_\theta(\mathbf{x}|\mathbf{z}) p_{z(\theta)}(\mathbf{z}) d\mathbf{z}
 \end{equation}
-where $\mathbf{z} \in \mathbb{R}^m$ is a latent variable providing an external source
-of randomness.
+where $\mathbf{z} \in {\cal Z}$ is a latent variable providing an external
+source of randomness. In particular, $\mathbf{z}$ is not necessarily assumed to
+be a fixed-size vector (e.g., it can be a sequence of variable length) and its
+distribution $p_{z(\theta)}$ may itself depend on $\theta$ in some intricated way.
 
@@
 generative process that defines $p_\theta(\mathbf{x}|\mathbf{z})$, as
-specified through a deterministic function $g(\cdot; \theta) : \mathbb{R}^m \to
-\mathbb{R}^d$. That is, we consider
+specified through a highly regularized deterministic function $g(\cdot; \theta) : {\cal Z} \to
+\mathbb{R}^d$ with usually few parameters. That is, we consider
 \begin{equation}\label{eqn:p_theta}
-    \mathbf{x} \sim p_\theta \equiv \mathbf{z} \sim p_z, \mathbf{x} = g(\mathbf{z}; \theta)
+    \mathbf{x} \sim p_\theta \equiv \mathbf{z} \sim p_{z(\theta)}, \mathbf{x} = g(\mathbf{z}; \theta)
 \end{equation}
@@
 \begin{equation}\label{eqn:p_x_sim}
-    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
+    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p_{z(\theta)}(\mathbf{z}) d\mathbf{z}.
 \end{equation}
+\glnote{Is it correct to write $d\mathbf{z}$ given the undefined structure of ${\cal Z}$?}
 Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
@@
 $\int q_\psi(\theta) p_\theta(\mathbf{x}) d\theta$ of the generated data.
+\glnote{Connection with variational bayes?}
 

### 2017-06-23 paper.tex
take kyle's comments into account

@@
 \begin{equation}
-    \theta^* = \arg \min_\theta \rho(p_r, p_\theta),
+    \theta^* = \arg \min_{\theta \in \Theta} \rho(p_r, p_\theta),
 \end{equation}
@@
 from generated data $\tilde{\mathbf{x}} = g(\mathbf{z}; \theta)$. Both models $g$ and $d$
-are trained simultaneously, in such a way that $g$ learns to maximally confuse
+are trained simultaneously, in such a way that $g$ learns to fool
 its adversary $d$ (which happens when $g$ produces samples comparable to the
@@
 \begin{equation}
-    \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c}) \leq \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] = U(\psi),
+    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta)] = U(\psi),
 \end{equation}
-where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\mathbf{c}$.
+where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
 That is, the minimum of a set of function values is always less than or equal
 to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
-can be updated to place its mass arbitrarily tight around the optimum $\mathbf{c}^* = \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c})$.
+can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.
 
-Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
+Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
+$U(\psi)$ is differentiable with respect to $\psi$, and using the log-likelihood
+trick it comes:
 \begin{align}\label{eqn:approx-grad}
-    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
-    &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
-    &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
-    &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
-    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c})]
+    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta)] \nonumber \\
+    &= \nabla_\psi \int f(\theta)  q_\psi(\theta)  d\theta \nonumber \\
+    &= \int f(\theta) \nabla_\psi q_\psi(\theta)  d\theta \nonumber \\
+    &= \int \left[ f(\theta) \nabla_\psi \log q_\psi(\theta) \right]  q_\psi(\theta)  d\theta \nonumber \\
+    &= \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta) \nabla_\psi \log q_\psi(\theta)]
 \end{align}
 Effectively, this means that provided that the score function $\nabla_\psi \log
-q_\psi(\mathbf{c})$ of the proposal is known and that one can evaluate
-$f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then one can construct empirical
+q_\psi(\theta)$ of the proposal is known and that one can evaluate
+$f(\mathbf{\theta})$ for any $\theta$, then one can construct empirical
 estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
@@
         {\it Outputs:} proposal distribution $q_\psi(\theta)$, such that $p_r \approx p_\psi$.\\
-        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
+        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
     \end{flushleft}
@@
     \begin{algorithmic}[1]
-        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known likelihood)}$}
+        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
         \While{$\psi$ has not converged}
             \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
-                \State{Sample a mini-batch $\{\mathbf{x}_m \}_{m=1}^M \sim p_r$, $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$, $\{\epsilon_m \}_{m=1}^M \sim U[0,1]$.}
+                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r, \theta_m \sim q_\psi, \mathbf{z}_m \sim p_{z(\theta_m)}, \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                 \For{$m=1$ to $M$}
@@
             \EndFor
-            \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
+            \State{Sample a mini-batch $\{ \theta_m \sim q_\psi,  \mathbf{z}_m \sim p_{z(\theta_m)} \}_{m=1}^M$.}  \Comment{Update $q_\psi$}
             \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
@@
 specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
-parameters of $g$ and minimize in alternance the variational upper bounds
+parameters of $g$ and $p_\theta$ and minimize in alternation the variational upper bounds
 \begin{align}
@@
 \begin{equation}
-    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
+    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_{z(\theta)}}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
 \end{equation} which we can approximate with mini-batches of
@@
 \end{equation}
-for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
+for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_{z(\theta_m)}$ and .
 \glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
@@
 \begin{equation}\label{eqn:p_psi}
-    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
+    \mathbf{x} \sim p_\psi \equiv \theta \sim q_\psi, \mathbf{z} \sim p_{z(\theta)}, \mathbf{x} = g(\mathbf{z}; \theta).
 \end{equation}

### 2017-06-23 code/nn.py
add code

@@ -1,0 +1,134 @@
+import autograd.numpy as np
+from autograd.util import flatten_func
+import copy
+
+
+# Activations
+
+def sigmoid(x):
+    return 0.5 * (np.tanh(x) + 1.0)
+
+
+def relu(x, alpha=0.0):
+    if alpha == 0.0:
+        return 0.5 * (x + np.abs(x))
+    else:
+        f1 = 0.5 * (1 + alpha)
+        f2 = 0.5 * (1 - alpha)
+        return f1 * x + f2 * np.abs(x)
+
+
+def logsumexp(X):
+    max_X = np.max(X, axis=-1)[..., np.newaxis]
+    return max_X + np.log(np.sum(np.exp(X - max_X), axis=-1)[..., np.newaxis])
+
+
+def softmax(X):
+    return np.exp(X - logsumexp(X))
+
+
+# Initializations
+
+def glorot_uniform(fan_in, fan_out, rng, scale=0.1):
+    s = scale * np.sqrt(6. / (fan_in + fan_out))
+    if fan_out > 0:
+        return rng.rand(fan_in, fan_out) * 2 * s - s
+    else:
+        return rng.rand(fan_in) * 2 * s - s
+
+
+def orthogonal(shape, rng, scale=1.1):
+    # from Keras
+    a = rng.normal(0.0, 1.0, shape)
+    u, _, v = np.linalg.svd(a, full_matrices=False)
+    q = u if u.shape == shape else v
+    q = q.reshape(shape)
+    return scale * q
+
+
+# Training
+
+class AdamOptimizer:
+    def __init__(self, grad, init_params, callback=None,
+                 step_size=0.01, b1=0.9, b2=0.99, eps=10**-8):
+        self.grad = grad
+        self.init_params = copy.copy(init_params)
+        self.callback = callback
+        self.step_size = step_size
+        self.b1 = b1
+        self.b2 = b2
+        self.eps = eps
+
+        self.flattened_grad, self.unflatten, self.x = flatten_func(
+            self.grad, self.init_params)
+        self.reset()
+
+    def reset(self):
+        self.m = np.zeros(len(self.x))
+        self.v = np.zeros(len(self.x))
+        self.counter = 0
+
+    def step(self,  num_iters=1):
+        for i in range(num_iters):
+            g = self.flattened_grad(self.x, self.counter)
+
+            if self.callback:
+                self.callback(self.unflatten(self.x),
+                              self.counter,
[…]

### 2017-06-23 code/proposals.py

@@ -1,0 +1,45 @@
+import autograd as ag
+import autograd.numpy as np
+from scipy.stats import norm
+
+
+def make_proposal(n_parameters):
+    mu = np.zeros(n_parameters)
+    beta = np.zeros(n_parameters)
+    return {"mu": mu, "beta": beta}
+
+
+def proposal_draw(params, n_samples, random_state=None):
+    n_parameters = len(params["mu"])
+    mu = params["mu"]
+    sigma = np.exp(params["beta"])
+
+    thetas = np.zeros((n_samples, n_parameters))
+
+    for i in range(n_parameters):
+        thetas[:, i] = norm.rvs(size=n_samples,
+                                loc=mu[i],
+                                scale=sigma[i],
+                                random_state=random_state)
+
+    return thetas
+
+
+def proposal_logpdf(params, theta, eps=10e-8):
+    mu = params["mu"]
+    sigma = np.exp(params["beta"])
+
+    logp = -(np.log(sigma) +
+             np.log((2. * np.pi) ** 0.5) +
+             (theta - mu) ** 2 / (2. * sigma ** 2))
+
+    return np.sum(logp)
+
+
+def proposal_entropy(params, theta):
+    logp = proposal_logpdf(params, theta)
+    return -np.exp(logp) * logp
+
+
+grad_proposal_logpdf = ag.grad(proposal_logpdf)
+grad_proposal_entropy = ag.grad(proposal_entropy)

### 2017-06-23 code/toy.py

@@ -1,0 +1,187 @@
+import autograd as ag
+import autograd.numpy as np
+import matplotlib.pyplot as plt
+from scipy.stats import norm
+
+from nn import glorot_uniform
+from nn import relu
+from nn import AdamOptimizer
+
+from proposals import make_proposal
+from proposals import proposal_draw
+from proposals import proposal_logpdf
+from proposals import grad_proposal_logpdf
+from proposals import grad_proposal_entropy
+
+from sklearn.utils import check_random_state
+
+# Global params
+
+batch_size = 64 * 8
+true_theta = np.array([0.5])
+make_plots = True
+
+
+# Simulator
+
+def simulator(theta, n_samples, random_state=None):
+    return norm.rvs(size=n_samples,
+                    loc=theta[0],
+                    scale=0.5,
+                    random_state=random_state).reshape(-1, 1)
+
+X_obs = simulator(true_theta, 10000, random_state=123)
+n_params = len(true_theta)
+n_features = X_obs.shape[1]
+
+
+# Proposal distribution
+
+params_proposal = make_proposal(n_params)
+params_proposal["mu"][0] = 0.8
+params_proposal["beta"][0] = 0.0
+
+
+# Critic
+
+def make_critic(n_features, n_hidden, random_state=None):
+    rng = check_random_state(random_state)
+    params = {"W": [glorot_uniform(n_hidden, 1, rng),
+                    glorot_uniform(n_hidden, n_hidden, rng),
+                    glorot_uniform(n_hidden, 0, rng)],
+              "b": [np.zeros(n_hidden),
+                    np.zeros(n_hidden),
+                    np.zeros(1)]}
+    return params
+
+params_critic = make_critic(1, 10)
+
+
+def predict(X, params):
+    h = X
+    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
+    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.1)
+    h = np.dot(params["W"][2], h.T).T + params["b"][2]
+    return h
+
+grad_predict_critic = ag.elementwise_grad(predict)
+
+
+y_critic = np.zeros(batch_size)
+y_critic[:batch_size // 2] = 0.0  # 0 == fake
+y_critic[batch_size // 2:] = 1.0
+
[…]

### 2017-06-23 code/proposals.py
add more code

@@ -1,33 +1,37 @@
 import autograd as ag
 import autograd.numpy as np
-from scipy.stats import norm
+from autograd.scipy.special import gammaln
+
+from scipy.stats import norm as sp_norm
+from scipy.stats import beta as sp_beta
 
 
-def make_proposal(n_parameters):
-    mu = np.zeros(n_parameters)
-    beta = np.zeros(n_parameters)
-    return {"mu": mu, "beta": beta}
+# Gaussian proposal (for unbounded supports)
+
+def make_gaussian_proposal(n_parameters, mu=0.0, log_sigma=0.0):
+    return {"mu":  mu * np.ones(n_parameters),
+            "log_sigma": log_sigma * np.ones(n_parameters)}
 
 
-def proposal_draw(params, n_samples, random_state=None):
+def gaussian_draw(params, n_samples, random_state=None):
     n_parameters = len(params["mu"])
     mu = params["mu"]
-    sigma = np.exp(params["beta"])
+    sigma = np.exp(params["log_sigma"])
 
     thetas = np.zeros((n_samples, n_parameters))
 
     for i in range(n_parameters):
-        thetas[:, i] = norm.rvs(size=n_samples,
-                                loc=mu[i],
-                                scale=sigma[i],
-                                random_state=random_state)
+        thetas[:, i] = sp_norm.rvs(size=n_samples,
+                                   loc=mu[i],
+                                   scale=sigma[i],
+                                   random_state=random_state)
 
     return thetas
 
 
-def proposal_logpdf(params, theta, eps=10e-8):
+def gaussian_logpdf(params, theta):
     mu = params["mu"]
-    sigma = np.exp(params["beta"])
+    sigma = np.exp(params["log_sigma"])
 
     logp = -(np.log(sigma) +
              np.log((2. * np.pi) ** 0.5) +
@@ -36,10 +40,52 @@ def proposal_logpdf(params, theta, eps=10e-8):
     return np.sum(logp)
 
 
-def proposal_entropy(params, theta):
-    logp = proposal_logpdf(params, theta)
+def gaussian_entropy(params, theta):
+    logp = gaussian_logpdf(params, theta)
     return -np.exp(logp) * logp
 
 
-grad_proposal_logpdf = ag.grad(proposal_logpdf)
-grad_proposal_entropy = ag.grad(proposal_entropy)
[…]

### 2017-06-23 code/weinberg.py

@@ -1,34 +1,73 @@
+import copy
 import autograd as ag
 import autograd.numpy as np
 import matplotlib.pyplot as plt
-from scipy.stats import norm
 
 from nn import glorot_uniform
 from nn import relu
 from nn import AdamOptimizer
 
-from proposals import make_gaussian_proposal
-from proposals import gaussian_draw
-from proposals import gaussian_logpdf
-from proposals import grad_gaussian_logpdf
-from proposals import grad_gaussian_entropy
+from proposals import make_beta_proposal
+from proposals import beta_draw
+from proposals import beta_logpdf
+from proposals import grad_beta_logpdf
+from proposals import grad_beta_entropy
 
 from sklearn.utils import check_random_state
 
 # Global params
 
 batch_size = 64 * 8
-true_theta = np.array([0.5])
+true_theta = np.array([(42 - 40) / (50 - 40),
+                       (0.9 - 0.5) / (1.5 - 0.5)])
 make_plots = True
 
 
 # Simulator
 
+def a_fb(sqrtshalf, gf):
+    MZ = 90
+    GFNom = 1.0
+    sqrts = sqrtshalf * 2.
+    A_FB_EN = np.tanh((sqrts - MZ) / MZ * 10)
+    A_FB_GF = gf / GFNom
+
+    return 2 * A_FB_EN*A_FB_GF
+
+
+def diffxsec(costheta, sqrtshalf, gf):
+    norm = 2. * (1. + 1. / 3.)
+    return ((1 + costheta ** 2) + a_fb(sqrtshalf, gf) * costheta) / norm
+
+def rej_sample_costheta(n_samples, theta, rng):
+    sqrtshalf = theta[0]
+    gf = theta[1]
+
+    ntrials = 0
+    samples = []
+    x = np.linspace(-1, 1, num=1000)
+    maxval = np.max(diffxsec(x, sqrtshalf, gf))
+
+    while len(samples) < n_samples:
+        ntrials = ntrials+1
+        xprop = rng.uniform(-1, 1)
+        ycut = rng.rand()
+        yprop = diffxsec(xprop, sqrtshalf, gf)/maxval
+        if yprop/maxval < ycut:
+            continue
+        samples.append(xprop)
+
+    return np.array(samples)
+
 def simulator(theta, n_samples, random_state=None):
-    return norm.rvs(size=n_samples,
-                    loc=theta[0],
-                    scale=0.5,
[…]

### 2017-06-23 code/weinberg.py
code

@@ -76,7 +76,9 @@ n_features = X_obs.shape[1]
 
 # Proposal distribution
 
-params_proposal = make_beta_proposal(n_params)
+params_proposal = make_beta_proposal(n_params,
+                                     log_alpha=np.log(2.0),
+                                     log_beta=np.log(2.0))
 
 
 # Critic
@@ -85,8 +87,10 @@ def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
     params = {"W": [glorot_uniform(n_hidden, 1, rng),
                     glorot_uniform(n_hidden, n_hidden, rng),
+                    glorot_uniform(n_hidden, n_hidden, rng),
                     glorot_uniform(n_hidden, 0, rng)],
               "b": [np.zeros(n_hidden),
+                    np.zeros(n_hidden),
                     np.zeros(n_hidden),
                     np.zeros(1)]}
     return params
@@ -98,7 +102,8 @@ def predict(X, params):
     h = X
     h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
     h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.1)
-    h = np.dot(params["W"][2], h.T).T + params["b"][2]
+    h = relu(np.dot(params["W"][2], h.T).T + params["b"][2], alpha=0.1)
+    h = np.dot(params["W"][3], h.T).T + params["b"][3]
     return h
 
 grad_predict_critic = ag.elementwise_grad(predict)
@@ -131,14 +136,14 @@ def loss_critic(params_critic, i):
     norms = np.sum(grad_Dx ** 2, axis=1) ** 0.5
     l_gp = np.mean((norms - 1.0) ** 2.0)
 
-    return l_wgan + 0.025 * l_gp
+    return l_wgan + 0.01 * l_gp
 
 grad_loss_critic = ag.grad(loss_critic)
 
 
 # grad_psi E_theta~q_psi, z~p_z(theta) [ d(g(z, theta) ]
 
-def approx_grad_u(params_proposal, i, gamma=0.1):
+def approx_grad_u(params_proposal, i, gamma=0.05):
     rng = check_random_state(i)
     grad_u = make_beta_proposal(n_params)
     grad_ent = make_beta_proposal(n_params)
@@ -169,7 +174,7 @@ def approx_grad_u(params_proposal, i, gamma=0.1):
[…]

### 2017-06-23 tex/paper.tex
mu(dz)

--- 2017-06-23 paper.tex
@@
 \begin{equation}\label{eqn:p_x_sim}
-    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p_{z(\theta)}(\mathbf{z}) d\mathbf{z}.
+    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p_{z(\theta)}(\mathbf{z}) \mu(d\mathbf{z}),
 \end{equation}
-\glnote{Is it correct to write $d\mathbf{z}$ given the undefined structure of ${\cal Z}$?}
+where $\mu$ is a probability measure.
 Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
@@
 in learning a smeared variation of the original simulator.
+\glnote{Add note on minimum entropy, otherwise it degenerates back to a non-differentiable function.}
 

### 2017-06-23 code/weinberg.py

@@ -17,7 +17,7 @@ from sklearn.utils import check_random_state
 
 # Global params
 
-batch_size = 64 * 8
+batch_size = 64
 true_theta = np.array([(42 - 40) / (50 - 40),
                        (0.9 - 0.5) / (1.5 - 0.5)])
 make_plots = True

### 2017-06-24 tex/paper.tex
update notations

@@
 
-We consider a family of parameterized densities $p_\theta(\mathbf{x})$
+We consider a family of parameterized densities $p(\mathbf{x}|\theta)$
 defined implicitly through the simulation of a stochastic generative process,
@@
 \begin{equation}\label{eqn:p_x}
-    p_\theta(\mathbf{x}) = \int p_\theta(\mathbf{x}|\mathbf{z}) p_{z(\theta)}(\mathbf{z}) d\mathbf{z}
+    p(\mathbf{x}|\theta) = \int p(\mathbf{x}|\mathbf{z},\theta) p(\mathbf{z}|\theta) d\mathbf{z}
 \end{equation}
@@
 be a fixed-size vector (e.g., it can be a sequence of variable length) and its
-distribution $p_{z(\theta)}$ may itself depend on $\theta$ in some intricated way.
+distribution $p(\mathbf{z}|\theta)$ may itself depend on $\theta$ in some intricated way.
 
 We assume that we already have an accurate simulation of the stochastic
-generative process that defines $p_\theta(\mathbf{x}|\mathbf{z})$, as
+generative process that defines $p(\mathbf{x}|\mathbf{z},\theta)$, as
 specified through a highly regularized deterministic function $g(\cdot; \theta) : {\cal Z} \to
@@
 \begin{equation}\label{eqn:p_theta}
-    \mathbf{x} \sim p_\theta \equiv \mathbf{z} \sim p_{z(\theta)}, \mathbf{x} = g(\mathbf{z}; \theta)
+    \mathbf{x} \sim p(\mathbf{x}|\theta) \equiv \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta)
 \end{equation}
-such that the likelihood $p_\theta(\mathbf{x})$ can be rewritten as
+such that the likelihood $p(\mathbf{x}|\theta)$ can be rewritten as
 \begin{equation}\label{eqn:p_x_sim}
-    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p_{z(\theta)}(\mathbf{z}) \mu(d\mathbf{z}),
+    p(\mathbf{x}|\theta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p(\mathbf{z}|\theta) \mu(d\mathbf{z}),
 \end{equation}
@@
 Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
-(unknown) true distribution $p_r$, our goal is the inference of the parameters
-of interest $\theta^*$ that minimize the divergence between $p_r$ and
-the modeled data distribution $p_\theta$ induced by $g(\cdot;
+(unknown) true distribution $p_r(\mathbf{x})$, our goal is the inference of the parameters
+of interest $\theta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
+the modeled data distribution $p(\mathbf{x}|\theta)$ induced by $g(\cdot;
 \theta)$ over $\mathbf{z}$. That is,
 \begin{equation}
-    \theta^* = \arg \min_{\theta \in \Theta} \rho(p_r, p_\theta),
+    \theta^* = \arg \min_{\theta \in \Theta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\theta)),
 \end{equation}
@@
 be shown that the original adversarial learning procedure amounts to minimizing
-the Jensen-Shannon divergence $\text{JSD}(p_r \parallel p_\theta)$ between $p_r$ and $p_\theta$.
+the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\theta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\theta)$.
 
@@
 Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
-setup in order to minimize the Wasserstein-1 distance $W(p_r, p_\theta)$ by
+setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\theta))$ by
 replacing the adversarial classifier with a 1-Lipschitz adversarial critic
@@
 \begin{align}
-    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r} [d(\mathbf{x};\phi)]  \nonumber \\
-                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p_{\hat{\mathbf{x}}}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
-    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)]
+    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
+                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
+    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)]
 \end{align}
-where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} + (1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim p_r$ and $\tilde{\mathbf{x}} \sim p_\theta$.
+where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
+(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
+p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)$.
 
@@
 \begin{equation}
-    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta)] = U(\psi),
+    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] = U(\psi),
 \end{equation}
-where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
+where $q(\theta|\psi)$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
 That is, the minimum of a set of function values is always less than or equal
@@
 \begin{align}\label{eqn:approx-grad}
-    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta)] \nonumber \\
-    &= \nabla_\psi \int f(\theta)  q_\psi(\theta)  d\theta \nonumber \\
-    &= \int f(\theta) \nabla_\psi q_\psi(\theta)  d\theta \nonumber \\
-    &= \int \left[ f(\theta) \nabla_\psi \log q_\psi(\theta) \right]  q_\psi(\theta)  d\theta \nonumber \\
-    &= \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta) \nabla_\psi \log q_\psi(\theta)]
+    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] \nonumber \\
+    &= \nabla_\psi \int f(\theta)  q(\theta|\psi)  d\theta \nonumber \\
+    &= \int f(\theta) \nabla_\psi q(\theta|\psi)  d\theta \nonumber \\
+    &= \int \left[ f(\theta) \nabla_\psi \log q(\theta|\psi) \right]  q(\theta|\psi) d\theta \nonumber \\
+    &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta) \nabla_\psi \log q(\theta|\psi)]
 \end{align}
 Effectively, this means that provided that the score function $\nabla_\psi \log
-q_\psi(\theta)$ of the proposal is known and that one can evaluate
+q(\theta|\psi)$ of the proposal is known and that one can evaluate
 $f(\mathbf{\theta})$ for any $\theta$, then one can construct empirical
@@
     \begin{flushleft}
-        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N \sim p_r$, simulator $g$.\\
-        {\it Outputs:} proposal distribution $q_\psi(\theta)$, such that $p_r \approx p_\psi$.\\
+        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N \sim p_r(\mathbf{x})$, simulator $g$.\\
+        {\it Outputs:} proposal distribution $q(\theta|\psi)$, such that $p_r(\mathbf{x}) \approx p(\mathbf{x}|\psi)$.\\
         {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
@@
     \begin{algorithmic}[1]
-        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
+        \State{$q(\theta|\psi) \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
         \While{$\psi$ has not converged}
             \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
-                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r, \theta_m \sim q_\psi, \mathbf{z}_m \sim p_{z(\theta_m)}, \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
+                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \theta_m \sim q(\theta|\psi), \mathbf{z}_m \sim p(\mathbf{z}|\theta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                 \For{$m=1$ to $M$}
@@
             \EndFor
-            \State{Sample a mini-batch $\{ \theta_m \sim q_\psi,  \mathbf{z}_m \sim p_{z(\theta_m)} \}_{m=1}^M$.}  \Comment{Update $q_\psi$}
+            \State{Sample a mini-batch $\{ \theta_m \sim q(\theta|\psi),  \mathbf{z}_m \sim p(\mathbf{z}|\theta_m) \}_{m=1}^M$.}  \Comment{Update $q_\psi$}
             \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
@@
 L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
-specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
-parameters of $g$ and $p_\theta$ and minimize in alternation the variational upper bounds
+specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
+parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
 \begin{align}
-    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
-    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
+    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
+    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
 \end{align} respectively over $\phi$ and $\psi$.
@@
 \begin{equation}
-    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_{z(\theta)}}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
+    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
 \end{equation} which we can approximate with mini-batches of
@@
 \begin{equation}
-    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
+    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
 \end{equation}
-for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_{z(\theta_m)}$ and .
+for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$ and .
 \glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
@@
 \begin{equation}\label{eqn:p_psi}
-    \mathbf{x} \sim p_\psi \equiv \theta \sim q_\psi, \mathbf{z} \sim p_{z(\theta)}, \mathbf{x} = g(\mathbf{z}; \theta).
+    \mathbf{x} \sim p(\mathbf{x}|\psi) \equiv \theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta).
 \end{equation}
 Intuitively, this corresponds to a family of simulators, each configured
-with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
-of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
-More formally, the learned model $p_\psi(\mathbf{x})$ therefore corresponds to the marginal distribution
-$\int q_\psi(\theta) p_\theta(\mathbf{x}) d\theta$ of the generated data.
+with randomly sampled parameters $\theta \sim q(\theta|\psi)$, whose joint collection
+of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
+More formally, the learned model $p(\mathbf{x}|\psi)$ therefore corresponds to the marginal distribution
+$\int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta$ of the generated data.
 \glnote{Connection with variational bayes?}
@@
 In consequence, the proposed inference algorithm does not necessarily guarantee that the
-proposal distribution $q_\psi$ will place its mass arbitrarily tight
+proposal distribution $q(\theta|\psi)i$ will place its mass arbitrarily tight
 around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
@@
 \begin{equation}
-    U_g = \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] + \gamma H(q_\psi)
+    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
 \end{equation}

### 2017-06-26 tex/paper.tex
save

@@
 cases, these implicit generative models~\citep{2016arXiv161003483M} are
-specified as imperative implementations of forward stochastic processes that
+specified as procedural implementations of forward stochastic processes that
 generate data. Because it is usually computationally intractable, most simulators
@@
 given observation, thereby making inference difficult.
+In addition, most scientific simulators are written using opaque low-level programming
+languages, hence raising raising the bar for likelihood-free inference
+algorithms relying e.g. on the simulator being a function with computable derivatives~\citep{2016arXiv160507826G}
+or being a controllable probabilistic program~\citep{2016arXiv161009900L}.
 
@@
     \begin{flushleft}
-        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N \sim p_r(\mathbf{x})$, simulator $g$.\\
+        {\it Inputs:} observed data $\{ \mathbf{x}_i \sim p_r(\mathbf{x}) \}_{i=1}^N$, simulator $g$.\\
         {\it Outputs:} proposal distribution $q(\theta|\psi)$, such that $p_r(\mathbf{x}) \approx p(\mathbf{x}|\psi)$.\\
@@
 
+\glnote{Compare with gradient-free optimization.}
+
+\glnote{Wood -> assume that the simulator is probabilistic program.}
+\glnote{Storkey -> assume access to derivatives.}
+
 % ==============================================================================

### 2017-06-26 code/nn.py

@@ -71,6 +71,7 @@ class AdamOptimizer:
     def step(self,  num_iters=1):
         for i in range(num_iters):
             g = self.flattened_grad(self.x, self.counter)
+            # XXX add clipping
 
             if self.callback:
                 self.callback(self.unflatten(self.x),

### 2017-06-26 code/proposals.py

@@ -1,12 +1,13 @@
 import autograd as ag
 import autograd.numpy as np
+from autograd.scipy.special import digamma
 from autograd.scipy.special import gammaln
 
 from scipy.stats import norm as sp_norm
 from scipy.stats import beta as sp_beta
 
 
-# Gaussian proposal (for unbounded supports)
+# Gaussian proposal (for unbounded support)
 
 def make_gaussian_proposal(n_parameters, mu=0.0, log_sigma=0.0):
     return {"mu":  mu * np.ones(n_parameters),
@@ -40,13 +41,18 @@ def gaussian_logpdf(params, theta):
     return np.sum(logp)
 
 
-def gaussian_entropy(params, theta):
-    logp = gaussian_logpdf(params, theta)
-    return -np.exp(logp) * logp
+def gaussian_entropy(params):
+    sigma = np.exp(params["log_sigma"])
+    return np.sum(np.log(sigma * (2. * np.pi * np.e) ** 0.5))
 
 
 grad_gaussian_logpdf = ag.grad(gaussian_logpdf)
 grad_gaussian_entropy = ag.grad(gaussian_entropy)
+
+
+# Truncated gaussian  (for continuous in [0,1])
+
+# XXX
 
 
 # Beta proposal (for continuous in [0,1])
@@ -82,9 +88,17 @@ def beta_logpdf(params, theta):
     return np.sum(logp)
 
 
-def beta_entropy(params, theta):
-    logp = beta_logpdf(params, theta)
-    return -np.exp(logp) * logp
+def betaln(alpha, beta):
+    return gammaln(alpha) + gammaln(beta) - gammaln(alpha + beta)
+
+
+def beta_entropy(params):
+    alpha = np.exp(params["log_alpha"])
+    beta = np.exp(params["log_beta"])
+
+    return np.sum(betaln(alpha, beta) -
+                  (alpha - 1.0) * (digamma(alpha) - digamma(alpha + beta)) -
+                  (beta - 1.0) * (digamma(beta) - digamma(alpha + beta)))
 
 
 grad_beta_logpdf = ag.grad(beta_logpdf)

### 2017-06-26 code/toy.py

@@ -17,7 +17,7 @@ from sklearn.utils import check_random_state
 
 # Global params
 
-batch_size = 64 * 8
+batch_size = 64
 true_theta = np.array([0.5])
 make_plots = True
 
@@ -99,7 +99,7 @@ grad_loss_critic = ag.grad(loss_critic)
 
 # grad_psi E_theta~q_psi, z~p_z(theta) [ d(g(z, theta) ]
 
-def approx_grad_u(params_proposal, i, gamma=0.1):
+def approx_grad_u(params_proposal, i, gamma=0.5):
     rng = check_random_state(i)
     grad_u = make_gaussian_proposal(n_params)
     grad_ent = make_gaussian_proposal(n_params)
@@ -113,14 +113,14 @@ def approx_grad_u(params_proposal, i, gamma=0.1):
         for k, v in grad_q.items():
             grad_u[k] += -dx * v
 
-        grad_entropy = grad_gaussian_entropy(params_proposal, theta)
-        for k, v in grad_entropy.items():
-            grad_ent[k] += v
+    grad_entropy = grad_gaussian_entropy(params_proposal)
+    for k, v in grad_entropy.items():
+        grad_ent[k] += v
 
     M = len(thetas)
 
     for k in grad_u:
-        grad_u[k] = 1. / M * (grad_u[k] + gamma * grad_ent[k])
+        grad_u[k] = 1. / M * grad_u[k] + gamma * grad_ent[k]
 
     return grad_u
 

### 2017-06-26 code/weinberg.py

@@ -143,7 +143,7 @@ grad_loss_critic = ag.grad(loss_critic)
 
 # grad_psi E_theta~q_psi, z~p_z(theta) [ d(g(z, theta) ]
 
-def approx_grad_u(params_proposal, i, gamma=0.05):
+def approx_grad_u(params_proposal, i, gamma=0.5):
     rng = check_random_state(i)
     grad_u = make_beta_proposal(n_params)
     grad_ent = make_beta_proposal(n_params)
@@ -157,14 +157,14 @@ def approx_grad_u(params_proposal, i, gamma=0.05):
         for k, v in grad_q.items():
             grad_u[k] += -dx * v
 
-        grad_entropy = grad_beta_entropy(params_proposal, theta)
-        for k, v in grad_entropy.items():
-            grad_ent[k] += v
+    grad_entropy = grad_beta_entropy(params_proposal)
+    for k, v in grad_entropy.items():
+        grad_ent[k] += v
 
     M = len(thetas)
 
     for k in grad_u:
-        grad_u[k] = 1. / M * (grad_u[k] + gamma * grad_ent[k])
+        grad_u[k] = 1. / M * grad_u[k] + gamma * grad_ent[k]
 
     return grad_u
 
@@ -174,12 +174,12 @@ def approx_grad_u(params_proposal, i, gamma=0.05):
 opt_critic = AdamOptimizer(grad_loss_critic, params_critic,
                            step_size=0.01, b1=0.5, b2=0.5)
 opt_proposal = AdamOptimizer(approx_grad_u, params_proposal,
-                             step_size=0.005, b1=0.1, b2=0.1)
+                             step_size=0.01, b1=0.1, b2=0.1) # was 0.005
 
 opt_critic.step(100)
 opt_critic.move_to(params_critic)
 
-for i in range(201):
+for i in range(501):
     print(params_proposal)
 
     # fit simulator

### 2017-06-26 tex/paper.tex
save

@@
 In addition, most scientific simulators are written using opaque low-level programming
-languages, hence raising raising the bar for likelihood-free inference
+languages, hence raising the bar for likelihood-free inference
 algorithms relying e.g. on the simulator being a function with computable derivatives~\citep{2016arXiv160507826G}
@@
 
-\subsection{Toy problem}
+\subsection{Gaussian mixture model}
+
+\subsection{Plinko}
 
@@
 intimately tied to a class of algorithms that can be framed as density
-estimation-by-comparison. In most cases, these algorithms are formulated as an
-iterative two-step process where the model distribution is first compared to the
-true data distribution and then updated to make it more comparable to the
-latter. Closest to our work are inference
-algorithms~\citep{Neal:2007zz,gutmann2012noise,goodfellow2014generative,cranmer2015approximating,2016arXiv161110242D}
-that rely on a classifier between generated and true data to estimate the
-density ratio between the two distributions and make parameter updates.
-\glnote{ABC + Classifier ABC.}
-\glnote{Explain how this work is different.}
-
-\glnote{Compare with gradient-free optimization.}
-
-\glnote{Wood -> assume that the simulator is probabilistic program.}
-\glnote{Storkey -> assume access to derivatives.}
+estimation-by-comparison. In most cases, these inference algorithms are
+formulated as an iterative two-step process where the model distribution is
+first compared to the true data distribution and then updated to make it more
+comparable to the latter. Closest to our work are procedures that rely on a
+classifier to estimate the discrepancy between the true and the generated data
+distributions. For example, \citep{gutmann2012noise} uses non linear logistic
+regression for fitting unnormalized differentiable statistical models, while
+\citep{goodfellow2014generative} exploits an adversarial neural network for
+learning a differentiable implicit generative model. In the likelihood-free
+setup, \citep{cranmer2015approximating,2016arXiv161110242D} estimate likelihood
+ratios through supervised classification, which can in turn be used for
+parameter inference in combination with a gradient-free optimization algorithm.
+Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
+a summary statistics for approximate Bayesian computation.
+\glnote{Explain what's new about the proposed work.}
 

### 2017-06-27 tex/paper.tex
save

@@
 experimental observations. Inference in this setup is often
-difficult, as simulators rarely provide a way to directly evaluate the likehood
+difficult, as simulators rarely provide a way to directly evaluate the likelihood
 function for a given observation. In this note, we develop a likelihood-free
@@
 domain-based scientific simulator, and solve the resulting non-differentiable
-minimax problem by minimizing variationial upper bounds of the adversarial objectives.
+minimax problem by minimizing variational upper bounds of the adversarial objectives.
 Effectively, the procedure results in learning an arbitrarily tight
@@
 objectives. The procedure results in learning a proposal distribution over
-simulator parameters, hence producing an abitrarily tight family of models whose
+simulator parameters, hence producing an arbitrarily tight family of models whose
 joint collection of generated samples matches the observed data.
@@
 be a fixed-size vector (e.g., it can be a sequence of variable length) and its
-distribution $p(\mathbf{z}|\theta)$ may itself depend on $\theta$ in some intricated way.
+distribution $p(\mathbf{z}|\theta)$ may itself depend on $\theta$ in some intricate way.
 
@@
             \EndFor
-            \State{Sample a mini-batch $\{ \theta_m \sim q(\theta|\psi),  \mathbf{z}_m \sim p(\mathbf{z}|\theta_m) \}_{m=1}^M$.}  \Comment{Update $q_\psi$}
+            \State{Sample a mini-batch $\{ \theta_m \sim q(\theta|\psi),  \mathbf{z}_m \sim p(\mathbf{z}|\theta_m) \}_{m=1}^M$.}  \Comment{Update $q(\theta|\psi)$}
             \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
@@
 $\int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta$ of the generated data.
-\glnote{Connection with variational bayes?}
 
@@
 \end{equation}
-where $\gamma \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
+where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
 between the generator objective and the tightness of the proposal distribution.
@@
 
-\subsection{Gaussian mixture model}
+\subsection{Toy problems}
+
+As illustrative experiments, we evaluate in this section adversarial variational
+optimization on well-known distributions for continuous and discrete data.
+
+We first consider inference for a univariate Gaussian with unknown mean and
+unknown variance. The observed data is sampled from a univariate Gaussian with
+mean $\mu = 1$ and variance $\sigma^2 = 0.5^2$. Algorithm~\ref{alg:avo} is run
+for 300 epochs with mini-batches of size $M=64$ and the following configuration.
+For the critic, we use a 3-layer MLP with 10 hidden nodes per layer and ReLU
+activations. At each epoch, Adam is run for $n_\text{critic}=100$ iterations
+with a step size $\alpha=0.01$, decay rates $\beta_1=\beta_2=0.5$ and its inner
+first and second moment vectors reset at each epoch in order to avoid building
+momentum in staled directions.  For modeling $\mu$ and $\sigma^2$, we use
+univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
+\sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
+parameters $\psi$ are updated by making one Adam step, with $\alpha=0.01$ and
+$\beta_1=\beta_2=0.1$. Penalty coefficients are set to $\lambda=0.025$ and
+$\gamma=5$.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
+    \caption{\glnote{todo}}\label{fig:gaussian}
+\end{figure}
+
+The upper plot of Figure~\ref{fig:gaussian} illustrates the resulting model
+after the optimization procedure. As expected from adversarial training, the
+model density $p(\mathbf{x}|\psi)$ closely matches the true density
+$p_r(\mathbf{x})$, as shown by the aligned histograms in blue and orange. Under
+the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we also observe
+that the proposal distributions have correctly concentrated their mass around
+$\mu=1$ and $\log \sigma=-0.69$. The bottom plot of Figure~\ref{fig:gaussian}
+shows an empirical estimate of $-U_d$ with respect to the epoch. We observe
+that it quickly falls towards $0$, which indicates that
+$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$
+and that the critic cannot distinguish between true and model data.
+
+Similarly, ... \glnote{discrete case with a Poisson distribution.}
 
 \subsection{Plinko}
-
-\subsection{Physics example}
 
@@
 first compared to the true data distribution and then updated to make it more
-comparable to the latter. Closest to our work are procedures that rely on a
-classifier to estimate the discrepancy between the true and the generated data
-distributions. For example, \citep{gutmann2012noise} uses non linear logistic
-regression for fitting unnormalized differentiable statistical models, while
+comparable to the latter.
+
+Closest to our work are procedures that rely on a classifier to estimate the
+discrepancy between the true and the model distributions. For example,
+\citep{gutmann2012noise} uses non linear logistic regression for fitting
+unnormalized differentiable statistical models, while
 \citep{goodfellow2014generative} exploits an adversarial neural network for
@@
 a summary statistics for approximate Bayesian computation.
-\glnote{Explain what's new about the proposed work.}
+
+In this context, the proposed method can be considered as a direct adaptation of
+generative adversarial networks~\citep{goodfellow2014generative} to
+non-differentiable simulators. It also constitutes an approximate gradient
+descent alternative to bayesian optimization based on density
+ratios~\citep{cranmer2015approximating} or to classifier
+ABC~\citep{gutmann2017likelihood}.
 

### 2017-06-27 tex/paper.tex
save

@@
 
-As illustrative experiments, we evaluate in this section adversarial variational
+As illustrative experiments, we evaluate adversarial variational
 optimization on well-known distributions for continuous and discrete data.
+We consider these distributions as parameterized simulators, from which
+we can only generate data.
 
-We first consider inference for a univariate Gaussian with unknown mean and
+For continuous data, we first consider inference for
+a univariate Gaussian with unknown mean and
 unknown variance. The observed data is sampled from a univariate Gaussian with
@@
     \centering
+    \vspace{-0.8cm}
     \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
-    \caption{\glnote{todo}}\label{fig:gaussian}
+    \vspace{-1cm}
+    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.}\label{fig:gaussian}
+\end{figure}
+\begin{figure}
+    \centering
+    \vspace{-0.8cm}
+    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
+    \vspace{-1cm}
+    \caption{Toy problem on discrete data: Poisson with unknown scale.}\label{fig:poisson}
 \end{figure}
@@
 
-Similarly, ... \glnote{discrete case with a Poisson distribution.}
+Similarly, we evaluate adversarial variational optimization on discrete data,
+for the inference of the scale parameter $\lambda$ of a Poisson distribution.
+The observed data is sampled from a univariate Poisson with scale $\lambda=7$.
+We use a univariate Gaussian proposal distribution $q(\log \lambda|\psi)$
+initialized with a mean at $5$ and unit variance. Hyper-parameters are otherwise
+left unchanged with respect to the previous setup. Despite the discreteness
+and the non-differentiability of the underlying generator, Figure~\ref{fig:poisson}
+shows that inference with adversarial variational optimization works.
+The upper plot in the figure shows that $p(\mathbf{x}|\psi)$ closely
+matches $p_r(\mathbf{x})$, with the mass of the proposal distribution
+concentrated at $\log 7 = 1.94$.
 

### 2017-06-27 code/proposals.py

@@ -6,6 +6,8 @@ from autograd.scipy.special import gammaln
 from scipy.stats import norm as sp_norm
 from scipy.stats import beta as sp_beta
 
+from sklearn.utils import check_random_state
+
 
 # Gaussian proposal (for unbounded support)
 
@@ -30,7 +32,7 @@ def gaussian_draw(params, n_samples, random_state=None):
     return thetas
 
 
-def gaussian_logpdf(params, theta):
+def gaussian_logpdf(params, theta, to_scalar=True):
     mu = params["mu"]
     sigma = np.exp(params["log_sigma"])
 
@@ -38,7 +40,10 @@ def gaussian_logpdf(params, theta):
              np.log((2. * np.pi) ** 0.5) +
              (theta - mu) ** 2 / (2. * sigma ** 2))
 
-    return np.sum(logp)
+    if to_scalar:
+        return np.sum(logp)
+    else:
+        return logp
 
 
 def gaussian_entropy(params):

### 2017-06-27 code/toy-gaussian.py

@@ -1,88 +1,47 @@
-import copy
 import autograd as ag
 import autograd.numpy as np
 import matplotlib.pyplot as plt
+from scipy.stats import norm
 
 from nn import glorot_uniform
 from nn import relu
 from nn import AdamOptimizer
 
-from proposals import make_beta_proposal
-from proposals import beta_draw
-from proposals import beta_logpdf
-from proposals import grad_beta_logpdf
-from proposals import grad_beta_entropy
+from proposals import make_gaussian_proposal
+from proposals import gaussian_draw
+from proposals import gaussian_logpdf
+from proposals import grad_gaussian_logpdf
+from proposals import grad_gaussian_entropy
 
 from sklearn.utils import check_random_state
 
 # Global params
 
 batch_size = 64
-n_epochs = 200+1
-lambda_gp = 0.01
-gamma = 0.5
+n_epochs = 300+1
+lambda_gp = 0.025
+gamma = 5.0
 
-true_theta = np.array([(42 - 40) / (50 - 40),
-                       (0.9 - 0.5) / (1.5 - 0.5)])
+true_theta = np.array([0.5, np.log(0.5)])
 make_plots = True
 
 
 # Simulator
 
-def a_fb(sqrtshalf, gf):
-    MZ = 90
-    GFNom = 1.0
-    sqrts = sqrtshalf * 2.
-    A_FB_EN = np.tanh((sqrts - MZ) / MZ * 10)
-    A_FB_GF = gf / GFNom
-
-    return 2 * A_FB_EN*A_FB_GF
-
-
-def diffxsec(costheta, sqrtshalf, gf):
-    norm = 2. * (1. + 1. / 3.)
-    return ((1 + costheta ** 2) + a_fb(sqrtshalf, gf) * costheta) / norm
-
-def rej_sample_costheta(n_samples, theta, rng):
-    sqrtshalf = theta[0]
-    gf = theta[1]
-
-    ntrials = 0
-    samples = []
-    x = np.linspace(-1, 1, num=1000)
-    maxval = np.max(diffxsec(x, sqrtshalf, gf))
-
-    while len(samples) < n_samples:
-        ntrials = ntrials+1
-        xprop = rng.uniform(-1, 1)
-        ycut = rng.rand()
-        yprop = diffxsec(xprop, sqrtshalf, gf)/maxval
-        if yprop/maxval < ycut:
-            continue
-        samples.append(xprop)
-
-    return np.array(samples)
-
[…]

### 2017-06-27 code/toy-poisson.py

@@ -1,7 +1,7 @@
 import autograd as ag
 import autograd.numpy as np
 import matplotlib.pyplot as plt
-from scipy.stats import norm
+from scipy.stats import poisson
 
 from nn import glorot_uniform
 from nn import relu
@@ -18,26 +18,29 @@ from sklearn.utils import check_random_state
 # Global params
 
 batch_size = 64
-true_theta = np.array([0.5])
+n_epochs = 300+1
+lambda_gp = 0.025
+gamma = 5.0
+
+true_theta = np.array([np.log(7)])
 make_plots = True
 
 
 # Simulator
 
 def simulator(theta, n_samples, random_state=None):
-    return norm.rvs(size=n_samples,
-                    loc=theta[0],
-                    scale=0.5,
-                    random_state=random_state).reshape(-1, 1)
+    return poisson.rvs(np.exp(theta[0]),
+                       size=n_samples,
+                       random_state=random_state).reshape(-1, 1)
 
-X_obs = simulator(true_theta, 10000, random_state=123)
+X_obs = simulator(true_theta, 20000, random_state=123)
 n_params = len(true_theta)
 n_features = X_obs.shape[1]
 
 
 # Proposal distribution
 
-params_proposal = make_gaussian_proposal(n_params, mu=0.8, log_sigma=0.0)
+params_proposal = make_gaussian_proposal(n_params, mu=np.log(5.0))
 
 
 # Critic
@@ -65,11 +68,11 @@ def predict(X, params):
 grad_predict_critic = ag.elementwise_grad(predict)
 
 
-y_critic = np.zeros(batch_size)
-y_critic[:batch_size // 2] = 0.0  # 0 == fake
-y_critic[batch_size // 2:] = 1.0
+def loss_critic(params_critic, i, lambda_gp=lambda_gp, batch_size=batch_size):
+    y_critic = np.zeros(batch_size)
+    # y_critic[:batch_size // 2] = 0.0  # 0 == fake
+    y_critic[batch_size // 2:] = 1.0
 
-def loss_critic(params_critic, i):
     rng = check_random_state(i)
 
     # WGAN loss
@@ -92,17 +95,17 @@ def loss_critic(params_critic, i):
     norms = np.sum(grad_Dx ** 2, axis=1) ** 0.5
     l_gp = np.mean((norms - 1.0) ** 2.0)
 
-    return l_wgan + 0.025 * l_gp
+    return l_wgan + lambda_gp * l_gp
 
[…]

### 2017-06-27 tex/paper.tex
save

@@
 
-As illustrative experiments, we evaluate adversarial variational
-optimization on well-known distributions for continuous and discrete data.
-We consider these distributions as parameterized simulators, from which
+As illustrative experiments, we evaluate adversarial variational optimization on
+well-known distributions for continuous and discrete data. We artificially
+consider these distributions as parameterized black box simulators, from which
 we can only generate data.
@@
 momentum in staled directions.  For modeling $\mu$ and $\sigma^2$, we use
-univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
+factored univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
 \sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
@@
     \centering
-    \vspace{-0.8cm}
+    \vspace{-0.5cm}
     \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
@@
 The upper plot of Figure~\ref{fig:gaussian} illustrates the resulting model
-after the optimization procedure. As expected from adversarial training, the
+after the optimization procedure. As theoretically expected from adversarial training, the
 model density $p(\mathbf{x}|\psi)$ closely matches the true density

### 2017-06-28 tex/paper.tex
save

@@
 In this note, we develop a likelihood-free inference algorithm for the point
-estimation from observed data of the parameters of a forward non-differentiable
-generative model. We propose to adapt the adversarial
+estimation of the parameters of a forward non-differentiable
+generative model. We adapt the adversarial
 training procedure of generative adversarial
@@
 
-Evolution strategies~\citep{2011arXiv1106.4487W} and variational optimization~\cite{2012arXiv1212.4507S} are general
+Variational optimization~\cite{2012arXiv1212.4507S} and evolution strategies~\citep{2011arXiv1106.4487W} are general
 optimization techniques that can be used to form a differentiable bound
@@
 When updating
-$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
+$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
 evaluating the exact and known gradient of $U_d$ over mini-batches of true and
 generated data, as ordinarily done in stochastic gradient descent. When updating
-$g$, estimates of $\nabla_\psi U_g$ can be derived with forward
+$\psi$, estimates of $\nabla_\psi U_g$ can be derived with forward
 simulations, as described in the previous section.
@@
 between the generator objective and the tightness of the proposal distribution.
-For large values of $\gamma$, the procedure is constrained to fit a proposal
+For small values of $\gamma$,
+proposal distributions with large entropy are not penalized, which may result
+in learning a smeared variation of the original simulator. On the other hand,
+for large values of $\gamma$, the procedure is constrained to fit a proposal
 distribution with low entropy, which has the effect of concentrating its density
-tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
-proposal distributions with larger entropy are not penalized, which may result
-in learning a smeared variation of the original simulator.
-\glnote{Add note on minimum entropy, otherwise it degenerates back to a non-differentiable function.}
+tightly around one or a few $\theta$ values. A too large penalty may however
+eventually make the optimization unstable, as the variance of $\nabla_\psi \log q(\theta_m|\psi)$
+typically increases as the entropy of the proposal decreases.
 
@@
 unknown variance. The observed data is sampled from a univariate Gaussian with
-mean $\mu = 1$ and variance $\sigma^2 = 0.5^2$. Algorithm~\ref{alg:avo} is run
+mean $\mu^* = 1$ and standard deviation ${\sigma^*} = 0.5$. Algorithm~\ref{alg:avo} is run
 for 300 epochs with mini-batches of size $M=64$ and the following configuration.
@@
 first and second moment vectors reset at each epoch in order to avoid building
-momentum in staled directions.  For modeling $\mu$ and $\sigma^2$, we use
+momentum in staled directions.  For estimating $\theta^* = (\mu^*, \sigma^*)$, we use
 factored univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
@@
     \vspace{-1cm}
-    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.}\label{fig:gaussian}
+    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.
+             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distributions $q(\mu|\psi)$ and $q(\log
+             \sigma|\psi)$ over the generator parameters.
+             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic. }\label{fig:gaussian}
 \end{figure}
@@
     \vspace{-1cm}
-    \caption{Toy problem on discrete data: Poisson with unknown scale.}\label{fig:poisson}
+    \caption{Toy problem on discrete data: Poisson with unknown scale.
+             This example illustrates the effectiveness of adversarial variational optimization on discrete non-differentiable generators.}\label{fig:poisson}
 \end{figure}
@@
 The upper plot of Figure~\ref{fig:gaussian} illustrates the resulting model
-after the optimization procedure. As theoretically expected from adversarial training, the
-model density $p(\mathbf{x}|\psi)$ closely matches the true density
-$p_r(\mathbf{x})$, as shown by the aligned histograms in blue and orange. Under
-the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we also observe
-that the proposal distributions have correctly concentrated their mass around
-$\mu=1$ and $\log \sigma=-0.69$. The bottom plot of Figure~\ref{fig:gaussian}
-shows an empirical estimate of $-U_d$ with respect to the epoch. We observe
-that it quickly falls towards $0$, which indicates that
-$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$
-and that the critic cannot distinguish between true and model data.
+after the optimization procedure. As theoretically expected from adversarial
+training, the model density $p(\mathbf{x}|\psi)$ closely matches the true
+density $p_r(\mathbf{x})$, as shown by the aligned histograms in blue and
+orange. Under the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we
+also observe that the proposal distributions have correctly concentrated their
+mass around $\mu^*=1$ and $\log \sigma^*=-0.69$. The bottom plot of
+Figure~\ref{fig:gaussian} shows an empirical estimate of $-U_d$ with respect to
+the epoch. We observe that it quickly falls towards $0$, which indicates that
+$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
+[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
+p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
+between true and model data.
 
@@
 for the inference of the scale parameter $\lambda$ of a Poisson distribution.
-The observed data is sampled from a univariate Poisson with scale $\lambda=7$.
+The observed data is sampled from a univariate Poisson with scale $\lambda^*=7$.
 We use a univariate Gaussian proposal distribution $q(\log \lambda|\psi)$
-initialized with a mean at $5$ and unit variance. Hyper-parameters are otherwise
-left unchanged with respect to the previous setup. Despite the discreteness
-and the non-differentiability of the underlying generator, Figure~\ref{fig:poisson}
-shows that inference with adversarial variational optimization works.
-The upper plot in the figure shows that $p(\mathbf{x}|\psi)$ closely
-matches $p_r(\mathbf{x})$, with the mass of the proposal distribution
-concentrated at $\log 7 = 1.94$.
+initialized with a mean at $\log 5$ and unit variance. The optimization setup is
+otherwise left unchanged. Despite the discreteness and the non-differentiability
+of the underlying generator, Figure~\ref{fig:poisson} shows that inference with
+adversarial variational optimization works. The upper plot in the figure shows
+that $p(\mathbf{x}|\psi)$ closely matches $p_r(\mathbf{x})$, with the mass of
+the proposal distribution concentrated at $\log 7 = 1.94$.
 
@@
 
+\glnote{avo = extension of gans to non-differentiable generator}
+\glnote{nice for discrete data}
+
 % ==============================================================================

### 2017-06-28 tex/paper.tex
save

@@
 \end{equation}
-for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$ and .
-\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
-is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
-% In practice, using
-% \begin{equation}
-%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
-% \end{equation}
-% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
-% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
-}
+for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$.
 For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
@@
 
-For continuous data, we first consider inference for
+{\it Continuous data.}
+We first consider inference for
 a univariate Gaussian with unknown mean and
@@
     \caption{Toy problem on discrete data: Poisson with unknown scale.
-             This example illustrates the effectiveness of adversarial variational optimization on discrete non-differentiable generators.}\label{fig:poisson}
+             This example illustrates adversarial variational optimization on discrete non-differentiable generators.}\label{fig:poisson}
 \end{figure}
@@
 
+{\it Discrete data.}
 Similarly, we evaluate adversarial variational optimization on discrete data,
@@
 initialized with a mean at $\log 5$ and unit variance. The optimization setup is
-otherwise left unchanged. Despite the discreteness and the non-differentiability
+otherwise the same as for the continuous case. Despite the discreteness and the non-differentiability
 of the underlying generator, Figure~\ref{fig:poisson} shows that inference with
@@
 
-\subsection{Plinko}
+\subsection{Galton box}
 
@@
 \glnote{nice for discrete data}
+\glnote{next version = try on an actual particle physics simulator.}
+
+\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
+is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?}
 

### 2017-06-28 tex/paper.tex
save

@@
 joint collection of generated samples matches the observed data.
+\glnote{add note on discrete data}
 
@@
 
-\subsection{Toy problems}
+\subsection{Univariate discrete data}
 
-As illustrative experiments, we evaluate adversarial variational optimization on
-well-known distributions for continuous and discrete data. We artificially
-consider these distributions as parameterized black box simulators, from which
-we can only generate data.
+\glnote{contour plots of q(theta|psi) + histograms + learning curves}
+\glnote{contours maybe with $\gamma=0$ and $\gamma>0$ contours}
+\glnote{insist that we dont learn the simulator}
 
-{\it Continuous data.}
-We first consider inference for
-a univariate Gaussian with unknown mean and
-unknown variance. The observed data is sampled from a univariate Gaussian with
-mean $\mu^* = 1$ and standard deviation ${\sigma^*} = 0.5$. Algorithm~\ref{alg:avo} is run
-for 300 epochs with mini-batches of size $M=64$ and the following configuration.
-For the critic, we use a 3-layer MLP with 10 hidden nodes per layer and ReLU
-activations. At each epoch, Adam is run for $n_\text{critic}=100$ iterations
-with a step size $\alpha=0.01$, decay rates $\beta_1=\beta_2=0.5$ and its inner
-first and second moment vectors reset at each epoch in order to avoid building
-momentum in staled directions.  For estimating $\theta^* = (\mu^*, \sigma^*)$, we use
-factored univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
-\sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
-parameters $\psi$ are updated by making one Adam step, with $\alpha=0.01$ and
-$\beta_1=\beta_2=0.1$. Penalty coefficients are set to $\lambda=0.025$ and
-$\gamma=5$.
+As a first illustrative experiment, we evaluate adversarial variational
+optimization on a Poisson distribution with unknown scale parameter $\lambda$.
+We artificially consider the distribution as a parameterized black box
+simulator, from which we can only generate (discrete) data.
 
-\begin{figure}
-    \centering
-    \vspace{-0.8cm}
-    \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
-    \vspace{-1cm}
-    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.
-             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distributions $q(\mu|\psi)$ and $q(\log
-             \sigma|\psi)$ over the generator parameters.
-             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic. }\label{fig:gaussian}
-\end{figure}
+The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
+Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
+and the following configuration. For the critic, we use a 3-layer MLP with 10
+hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
+$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
+$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
+each epoch in order to avoid building momentum in staled directions.  For
+estimating $\lambda^*$, we use a univariate Gaussian proposal distribution
+$q(\log \lambda|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
+each epoch, parameters $\psi$ are updated by making one Adam step, with
+$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. Penalty coefficients are set to
+$\lambda=0.025$ and $\gamma=5$.
+
 \begin{figure}
@@
     \caption{Toy problem on discrete data: Poisson with unknown scale.
-             This example illustrates adversarial variational optimization on discrete non-differentiable generators.}\label{fig:poisson}
+             This example illustrates adversarial variational optimization for a discrete non-differentiable generator.
+             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distribution $q(\lambda|\psi)$.
+             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic.
+             }\label{fig:poisson}
 \end{figure}
 
-The upper plot of Figure~\ref{fig:gaussian} illustrates the resulting model
+The upper plot of Figure~\ref{fig:poisson} illustrates the resulting model
 after the optimization procedure. As theoretically expected from adversarial
@@
 orange. Under the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we
-also observe that the proposal distributions have correctly concentrated their
-mass around $\mu^*=1$ and $\log \sigma^*=-0.69$. The bottom plot of
-Figure~\ref{fig:gaussian} shows an empirical estimate of $-U_d$ with respect to
+also observe that the proposal distribution has correctly concentrated its
+mass around $\log(7) = 1.94$. The bottom plot of
+Figure~\ref{fig:poisson} shows an empirical estimate of $-U_d$ with respect to
 the epoch. We observe that it quickly falls towards $0$, which indicates that
@@
 p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
-between true and model data.
+between true and model data. Despite the discreteness and the non-differentiability
+of the underlying generator, this confirms that inference with
+adversarial variational optimization works.
 
-{\it Discrete data.}
-Similarly, we evaluate adversarial variational optimization on discrete data,
-for the inference of the scale parameter $\lambda$ of a Poisson distribution.
-The observed data is sampled from a univariate Poisson with scale $\lambda^*=7$.
-We use a univariate Gaussian proposal distribution $q(\log \lambda|\psi)$
-initialized with a mean at $\log 5$ and unit variance. The optimization setup is
-otherwise the same as for the continuous case. Despite the discreteness and the non-differentiability
-of the underlying generator, Figure~\ref{fig:poisson} shows that inference with
-adversarial variational optimization works. The upper plot in the figure shows
-that $p(\mathbf{x}|\psi)$ closely matches $p_r(\mathbf{x})$, with the mass of
-the proposal distribution concentrated at $\log 7 = 1.94$.
+\subsection{Multidimensional continuous data}
 
-\subsection{Galton box}
+\glnote{use 5d example from carl}
+\glnote{contour plots}
+
+\subsection{Weinberg}
 

### 2017-06-29 tex/paper.tex
<:(((>< example

@@
 
-\glnote{contour plots of q(theta|psi) + histograms + learning curves}
-\glnote{contours maybe with $\gamma=0$ and $\gamma>0$ contours}
-\glnote{insist that we dont learn the simulator}
-
-As a first illustrative experiment, we evaluate adversarial variational
-optimization on a Poisson distribution with unknown scale parameter $\lambda$.
-We artificially consider the distribution as a parameterized black box
-simulator, from which we can only generate (discrete) data.
+As a first illustrative experiment, we evaluate inference for a discrete Poisson
+distribution with unknown scale parameter $\lambda$. We artificially consider
+the distribution as a parameterized simulator, from which we can only
+generate data.
 
@@
 Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
-and the following configuration. For the critic, we use a 3-layer MLP with 10
+and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
 hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
@@
 $\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
-each epoch in order to avoid building momentum in staled directions.  For
+each outer epoch in order to avoid building momentum in staled directions.  For
 estimating $\lambda^*$, we use a univariate Gaussian proposal distribution
 $q(\log \lambda|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
-each epoch, parameters $\psi$ are updated by making one Adam step, with
-$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. Penalty coefficients are set to
-$\lambda=0.025$ and $\gamma=5$.
+each epoch, parameters $\psi$ are updated by taking one Adam step, with
+$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. The gradient penalty coefficient is set to
+$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.
 
@@
     \centering
-    \vspace{-0.5cm}
-    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
-    \vspace{-1cm}
-    \caption{Toy problem on discrete data: Poisson with unknown scale.
-             This example illustrates adversarial variational optimization for a discrete non-differentiable generator.
-             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distribution $q(\lambda|\psi)$.
-             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic.
+    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
+    \caption{Discrete Poisson model with unknown scale.
+             (Top left) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization.
+For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
+                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
+             (Top right) Model distributions $p(\mathbf{x}|\psi)$ after training.
+             (Bottom) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
              }\label{fig:poisson}
@@
 
-The upper plot of Figure~\ref{fig:poisson} illustrates the resulting model
-after the optimization procedure. As theoretically expected from adversarial
-training, the model density $p(\mathbf{x}|\psi)$ closely matches the true
-density $p_r(\mathbf{x})$, as shown by the aligned histograms in blue and
-orange. Under the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we
-also observe that the proposal distribution has correctly concentrated its
-mass around $\log(7) = 1.94$. The bottom plot of
-Figure~\ref{fig:poisson} shows an empirical estimate of $-U_d$ with respect to
-the epoch. We observe that it quickly falls towards $0$, which indicates that
+The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
+distributions $p(\theta|\psi)$ after adversarial variational optimization.  For
+both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
+their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
+the effect of the positive entropy penalty $H(q(\theta|\psi))$,
+the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
+yielding in this case a more precise inference.  The top right plot compares the
+model distributions to the true distribution.  As theoretically expected from
+adversarial training, we see that the resulting distributions closely match
+the true distribution, with qualitatively  better results for the penalized
+model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
+of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
+the curves quickly fall towards $0$, which indicates that
 $\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
@@
 p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
-between true and model data. Despite the discreteness and the non-differentiability
-of the underlying generator, this confirms that inference with
-adversarial variational optimization works.
+between true and model data. Despite the discreteness and the
+non-differentiability of the underlying generator, this confirms that inference
+with adversarial variational optimization works.
 
@@
 
-\glnote{use 5d example from carl}
-\glnote{contour plots}
+\glnote{use 5d example from carl.}
+\glnote{contour plots for $\alpha$ and $\beta$.}
 

### 2017-06-29 code/proposals.py

@@ -6,8 +6,6 @@ from autograd.scipy.special import gammaln
 from scipy.stats import norm as sp_norm
 from scipy.stats import beta as sp_beta
 
-from sklearn.utils import check_random_state
-
 
 # Gaussian proposal (for unbounded support)
 
@@ -82,7 +80,7 @@ def beta_draw(params, n_samples, random_state=None):
     return thetas
 
 
-def beta_logpdf(params, theta):
+def beta_logpdf(params, theta, to_scalar=True):
     alpha = np.exp(params["log_alpha"])
     beta = np.exp(params["log_beta"])
 
@@ -90,7 +88,10 @@ def beta_logpdf(params, theta):
             (alpha - 1) * np.log(theta) +
             (beta - 1) * np.log(1 - theta))
 
-    return np.sum(logp)
+    if to_scalar:
+        return np.sum(logp)
+    else:
+        return logp
 
 
 def betaln(alpha, beta):

### 2017-06-29 code/toy-poisson.py

@@ -20,7 +20,6 @@ from sklearn.utils import check_random_state
 batch_size = 64
 n_epochs = 300+1
 lambda_gp = 0.025
-gamma = 5.0
 
 true_theta = np.array([np.log(7)])
 make_plots = True
@@ -38,12 +37,10 @@ n_params = len(true_theta)
 n_features = X_obs.shape[1]
 
 
-# Proposal distribution
-
-params_proposal = make_gaussian_proposal(n_params, mu=np.log(5.0))
-
-
 # Critic
+
+gammas = [0.0, 5.0]
+colors = ["C1", "C3"]
 
 def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
@@ -55,9 +52,6 @@ def make_critic(n_features, n_hidden, random_state=None):
                     np.zeros(1)]}
     return params
 
-params_critic = make_critic(1, 10)
-
-
 def predict(X, params):
     h = X
     h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
@@ -68,127 +62,167 @@ def predict(X, params):
 grad_predict_critic = ag.elementwise_grad(predict)
 
 
-def loss_critic(params_critic, i, lambda_gp=lambda_gp, batch_size=batch_size):
-    y_critic = np.zeros(batch_size)
-    # y_critic[:batch_size // 2] = 0.0  # 0 == fake
-    y_critic[batch_size // 2:] = 1.0
-
-    rng = check_random_state(i)
-
-    # WGAN loss
-    thetas = gaussian_draw(params_proposal, batch_size // 2)
-    _X_gen = np.zeros((batch_size // 2, n_features))
-    for j, theta in enumerate(thetas):
-        _X_gen[j, :] = simulator(theta, 1, random_state=rng).ravel()
-
-    indices = rng.permutation(len(X_obs))
-    _X_obs = X_obs[indices[:batch_size // 2]]
-    X = np.vstack([_X_gen, _X_obs])
-
-    y_pred = predict(X, params_critic)
-    l_wgan = np.mean(-y_critic * y_pred + (1. - y_critic) * y_pred)
-
-    # Gradient penalty
-    eps = rng.rand(batch_size // 2, 1)
-    _X_hat = eps * _X_obs + (1. - eps) * _X_gen
-    grad_Dx = grad_predict_critic(_X_hat, params_critic)
[…]

### 2017-06-29 code/weinberg.py

@@ -1,11 +1,19 @@
-import copy
 import autograd as ag
 import autograd.numpy as np
 import matplotlib.pyplot as plt
+import copy
 
 from nn import glorot_uniform
 from nn import relu
 from nn import AdamOptimizer
+
+# from proposals import make_gaussian_proposal
+# from proposals import gaussian_draw
+# from proposals import gaussian_logpdf
+# from proposals import grad_gaussian_logpdf
+# from proposals import grad_gaussian_entropy
+
+# XXX switch to a gaussian proposal
 
 from proposals import make_beta_proposal
 from proposals import beta_draw
@@ -18,12 +26,11 @@ from sklearn.utils import check_random_state
 # Global params
 
 batch_size = 64
-n_epochs = 200+1
-lambda_gp = 0.01
-gamma = 0.5
+n_epochs = 300+1
+lambda_gp = 0.025
+gamma = 0.0
 
-true_theta = np.array([(42 - 40) / (50 - 40),
-                       (0.9 - 0.5) / (1.5 - 0.5)])
+true_theta = np.array([(42 - 40) / (50 - 40)])
 make_plots = True
 
 
@@ -45,7 +52,7 @@ def diffxsec(costheta, sqrtshalf, gf):
 
 def rej_sample_costheta(n_samples, theta, rng):
     sqrtshalf = theta[0]
-    gf = theta[1]
+    gf = 0.9
 
     ntrials = 0
     samples = []
@@ -65,15 +72,14 @@ def rej_sample_costheta(n_samples, theta, rng):
 
 def simulator(theta, n_samples, random_state=None):
     theta = copy.copy(theta)
-    theta[0] = theta[0] * (50 - 40) + 40     # sqrtshalf
-    theta[1] = theta[1] * (1.5 - 0.5) + 0.5  # gf
+    theta[0] = theta[0] * (50 - 40) + 40
 
     rng = check_random_state(random_state)
     samples = rej_sample_costheta(n_samples, theta, rng)
 
     return samples.reshape(-1, 1)
 
-X_obs = simulator(true_theta, 10000, random_state=123)
+X_obs = simulator(true_theta, 20000, random_state=123)
 n_params = len(true_theta)
 n_features = X_obs.shape[1]
 
@@ -113,11 +119,11 @@ def predict(X, params):
 grad_predict_critic = ag.elementwise_grad(predict)
 
 
-y_critic = np.zeros(batch_size)
-y_critic[:batch_size // 2] = 0.0  # 0 == fake
[…]

### 2017-06-29 tex/paper.tex
save

@@
 inference algorithm for fitting a forward non-differentiable generative model to
-observed data. We adapt the adversarial training procedure of generative
+(continuous or discrete) observed data. We adapt the adversarial training procedure of generative
 adversarial networks by replacing the implicit generative network with a
@@
 specified as procedural implementations of forward stochastic processes that
-generate data. Because it is usually computationally intractable, most simulators
+generate discrete or continuous data. Because it is usually computationally intractable, most simulators
 do not provide a way to directly evaluate the likelihood function for a
@@
 joint collection of generated samples matches the observed data.
-\glnote{add note on discrete data}
 
@@
 of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
-More formally, the learned model $p(\mathbf{x}|\psi)$ therefore corresponds to the marginal distribution
-$\int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta$ of the generated data.
+More formally, the learned model  therefore corresponds to the marginal distribution
+of the generated data. That is,
+\begin{equation}
+    p(\mathbf{x}|\psi) = \int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta.
+\end{equation}
 
@@
              (Top right) Model distributions $p(\mathbf{x}|\psi)$ after training.
+This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution with, in this case, qualitatively
+                        better results for the penalized model.
              (Bottom) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.

### 2017-06-29 tex/paper.tex
save

@@
 each outer epoch in order to avoid building momentum in staled directions.  For
-estimating $\lambda^*$, we use a univariate Gaussian proposal distribution
-$q(\log \lambda|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
+estimating $\lambda^*$, we parameterize $\theta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
+$q(\theta|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
 each epoch, parameters $\psi$ are updated by taking one Adam step, with
@@
     \caption{Discrete Poisson model with unknown scale.
-             (Top left) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization.
+             ({\it Top left}) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization.
 For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                         the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
-             (Top right) Model distributions $p(\mathbf{x}|\psi)$ after training.
-This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution with, in this case, qualitatively
-                        better results for the penalized model.
-             (Bottom) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+             ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
+             ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
              }\label{fig:poisson}
@@
 adversarial training, we see that the resulting distributions closely match
-the true distribution, with qualitatively  better results for the penalized
+the true distribution, with in this case slightly  better results for the penalized
 model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
@@
 
-\glnote{use 5d example from carl.}
-\glnote{contour plots for $\alpha$ and $\beta$.}
+As a second example, we evaluate parameter inference for a generator producing
+5-dimensional continuous data, as originally specified in Section 4.2. of
+\citep{cranmer2015approximating}. More specifically, we consider the following
+generative process:
+\begin{itemize}
+    \item $\mathbf{z} = (z_0, z_1, z_2, z_3, z_4)$, such that
+        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
+        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
+        $z_2 \sim {\text{Mixture}}(\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
+        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
+        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
+    \item $\mathbf{x} = R  \mathbf{z}$, where $R$ is a fixed
+    semi-positive definite $5 \times 5$ matrix defining a fixed projection
+    of $\mathbf{z}$ into the observed space.
+\end{itemize}
+We consider observed data generated at the nominal values $\theta^* = (\alpha^*=1,\beta^*=-1)$.
+The simulator parameters are modeled with a factored Gaussian
+proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
+initialized with zero mean and unit variance. The entropy penalty is set
+to $\gamma=5$, while all other hyper-parameters are otherwise the same as in the previous example.
+
+Starting with a proposal distribution $q(\theta|\psi)$ largely spread over
+the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
+inference quickly converges towards a proposal distribution whose density
+concentrates around the nominal values $\theta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
+    \caption{Multidimensional continuous data.
+             ({\it Left}) Likelihood $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
+             ({\it Right}) Likelihood $q(\theta|\psi)$ after adverarial variational optimization. The proposal density correctly concentrates around $\theta^* = (1, -1)$.
+             }\label{fig:multi}
+\end{figure}
 

### 2017-06-30 tex/paper.tex
save

@@
 
-Complex computer simulators are increasingly used across fields of science to
+Complex computer simulators are used across fields of science to
 describe generative models tying parameters of an underlying theory to
@@
 inference algorithm for fitting a forward non-differentiable generative model to
-(continuous or discrete) observed data. We adapt the adversarial training procedure of generative
+continuous or discrete observed data. We adapt the adversarial training procedure of generative
 adversarial networks by replacing the implicit generative network with a
@@
 marginal distribution of the generated data matches the observations.
-\glnote{Mention experimental results.}
-\glnote{Add 'so what?' conclusion.}
+Preliminary results with non-differentiable simulators producing discrete/continuous univariate/multi-dimensional data
+validate the proposed method.
 
@@
 each epoch, parameters $\psi$ are updated by taking one Adam step, with
-$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. The gradient penalty coefficient is set to
+$\alpha=0.01$ and $\beta_1=\beta_2=0.5$. The gradient penalty coefficient is set to
 $\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.
@@
     \caption{Discrete Poisson model with unknown scale.
-             ({\it Top left}) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization.
+             ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization.
 For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
@@
 The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
-distributions $p(\theta|\psi)$ after adversarial variational optimization.  For
+distributions $q(\theta|\psi)$ after adversarial variational optimization.  For
 both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
@@
 proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
-initialized with zero mean and unit variance. The entropy penalty is set
-to $\gamma=5$, while all other hyper-parameters are otherwise the same as in the previous example.
+initialized with zero mean and unit variance. Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
+Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
 
@@
 concentrates around the nominal values $\theta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
+Overall, this example further illustrates and confirms the ability of adversarial
+variational optimization for inference with multiple parameters and multidimensional
+data, where reliable approximations of $p(\mathbf{x}|\theta)$ in a traditional
+MLE setting would otherwise be difficult to construct.
 
@@
 
-\subsection{Weinberg}
+\subsection{Electron--positron annihilation}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
+    \caption{Electron--positron annihilation.
+    ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization.
+The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\theta^*$.
+    ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
+    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+             }\label{fig:weinberg}
+\end{figure}
+
+As a more challenging example, we now consider a (simplified) simulator from
+particle physics for electron--positron collisions resulting in muon--antimuon
+pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
+distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
+polar angle of the outgoing muon with respect  to the originally incoming
+electron. This random variable is approximately distributed as
+\begin{equation}
+    p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
+\end{equation}
+where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
+function. Due to the linear term in the expression, the density $p(\mathbf{x} |
+E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
+size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
+and $G_f$ (the Fermi constant) through the coefficient function $c$. Given the
+known analytical form of the probability density function, a simulator for
+$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
+arbitrary number of random drawings may be required before passing the
+acceptance threshold, this illustrates how the latent variable
+$\mathbf{z}$ does not necessarily need to be fixed-sized vector, and how its
+density  $p(\mathbf{z}|\theta)$ may depend on $\theta$ in some complicated way
+(Section~\ref{sec:problem}).
+
+In this example, we consider observed data generated with $\theta^* =
+(E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
+factored Gaussian proposal distribution $q(\theta|\psi) = q(E_\text{beam}|\psi)
+q(G_f|\psi)$, where each component is respectively initialized with mean $45$
+and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
+$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
+$\beta_1=0.9$ and $\beta_2=0.999$. As for the first example, we compare
+entropy penalties $\gamma=0$ and $\gamma=5$.
+
+The top left plot in Figure~\ref{fig:poisson} illustates the resulting proposal
+distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
+variational optimization. We see that the distributions both arrive in the
+neighborhood of $\theta^*$, with a density highly more concentrated for
+$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
+distance with $\theta^*$, both models closely match the observed data, as shown
+in the top right plot of  Figure~\ref{fig:poisson}. This suggests either a
+relatively flat landscape around $\theta^*$ or that the observed data can in
+this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
+the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
+for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
+fall towards $0$, which indicates  convergence towards a distribution that the
+critic cannot distinguish from the true distribution.
 
@@
 \glnote{next version = try on an actual particle physics simulator.}
+\glnote{nes for better invariance and stability?}
 

### 2017-06-30 code/multi.py

@@ -1,0 +1,229 @@
+import autograd as ag
+import autograd.numpy as np
+import matplotlib.pyplot as plt
+
+from carl.distributions import Join
+from carl.distributions import Mixture
+from carl.distributions import Normal
+from carl.distributions import Exponential
+from carl.distributions import LinearTransform
+import theano
+
+from nn import glorot_uniform
+from nn import relu
+from nn import AdamOptimizer
+
+from proposals import make_gaussian_proposal
+from proposals import gaussian_draw
+from proposals import gaussian_logpdf
+from proposals import grad_gaussian_logpdf
+from proposals import grad_gaussian_entropy
+
+from sklearn.utils import check_random_state
+
+# Global params
+
+rng = check_random_state(777)
+
+batch_size = 64
+n_epochs = 300+1
+lambda_gp = 0.025
+gamma = 10.0
+
+true_theta = np.array([1.0, -1.0])
+make_plots = True
+
+
+# Simulator
+
+A = theano.shared(true_theta[0], name="A")
+B = theano.shared(true_theta[1], name="B")
+R = np.array([
+    [1.31229955,  0.10499961,  0.48310515, -0.3249938,  -0.26387927],
+    [0.10499961,  1.15833058, -0.55865473,  0.25275522, -0.39790775],
+    [0.48310515, -0.55865473,  2.25874579, -0.52087938, -0.39271231],
+    [0.3249938,   0.25275522, -0.52087938,  1.4034925,  -0.63521059],
+    [-0.26387927, -0.39790775, -0.39271231, -0.63521059,  1.]])
+
+p0 = LinearTransform(Join(components=[
+                    Normal(mu=A, sigma=1),
+                    Normal(mu=B, sigma=3),
+                    Mixture(components=[Normal(mu=-2, sigma=1),
+                                        Normal(mu=2, sigma=0.5)]),
+                    Exponential(inverse_scale=3.0),
+                    Exponential(inverse_scale=0.5)]), R)
+
+def simulator(theta, n_samples, random_state=None):
+    A.set_value(theta[0])
+    B.set_value(theta[1])
+    return p0.rvs(n_samples, random_state=random_state)
+
+X_obs = simulator(true_theta, 20000, random_state=rng)
+n_params = len(true_theta)
+n_features = X_obs.shape[1]
+
+
+# Proposal distribution
[…]

### 2017-06-30 code/nn.py

@@ -50,7 +50,7 @@ def orthogonal(shape, rng, scale=1.1):
 
 class AdamOptimizer:
     def __init__(self, grad, init_params, callback=None,
-                 step_size=0.01, b1=0.9, b2=0.99, eps=10**-8):
+                 step_size=0.01, b1=0.9, b2=0.999, eps=10**-8):
         self.grad = grad
         self.init_params = copy.copy(init_params)
         self.callback = callback

### 2017-06-30 code/toy-gaussian.py

@@ -48,7 +48,7 @@ params_proposal = make_gaussian_proposal(n_params, mu=0.0)
 
 def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
-    params = {"W": [glorot_uniform(n_hidden, 1, rng),
+    params = {"W": [glorot_uniform(n_hidden, n_features, rng),
                     glorot_uniform(n_hidden, n_hidden, rng),
                     glorot_uniform(n_hidden, 0, rng)],
               "b": [np.zeros(n_hidden),
@@ -56,7 +56,7 @@ def make_critic(n_features, n_hidden, random_state=None):
                     np.zeros(1)]}
     return params
 
-params_critic = make_critic(1, 10)
+params_critic = make_critic(n_features, 10)
 
 
 def predict(X, params):

### 2017-06-30 code/toy-poisson.py

@@ -17,6 +17,8 @@ from sklearn.utils import check_random_state
 
 # Global params
 
+rng = check_random_state(123)
+
 batch_size = 64
 n_epochs = 300+1
 lambda_gp = 0.025
@@ -32,7 +34,7 @@ def simulator(theta, n_samples, random_state=None):
                        size=n_samples,
                        random_state=random_state).reshape(-1, 1)
 
-X_obs = simulator(true_theta, 20000, random_state=123)
+X_obs = simulator(true_theta, 20000, random_state=rng)
 n_params = len(true_theta)
 n_features = X_obs.shape[1]
 
@@ -40,11 +42,11 @@ n_features = X_obs.shape[1]
 # Critic
 
 gammas = [0.0, 5.0]
-colors = ["C1", "C3"]
+colors = ["C1", "C2"]
 
 def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
-    params = {"W": [glorot_uniform(n_hidden, 1, rng),
+    params = {"W": [glorot_uniform(n_hidden, n_features, rng),
                     glorot_uniform(n_hidden, n_hidden, rng),
                     glorot_uniform(n_hidden, 0, rng)],
               "b": [np.zeros(n_hidden),
@@ -67,13 +69,14 @@ history = [{"gamma": gammas[i],
             "loss_d": [],
             "params_proposal": make_gaussian_proposal(n_params,
                                                       mu=np.log(5.0)),
-            "params_critic": make_critic(1, 10)}
+            "params_critic": make_critic(n_features, 10, random_state=rng)}
            for i in range(len(gammas))]
 
 
 # Global loop over gammas
 
 for state in history:
+    # WGAN + GP
     def loss_critic(params_critic, i, lambda_gp=lambda_gp, batch_size=batch_size):
         y_critic = np.zeros(batch_size)
         # y_critic[:batch_size // 2] = 0.0  # 0 == fake
@@ -82,7 +85,8 @@ for state in history:
         rng = check_random_state(i)
 
         # WGAN loss
-        thetas = gaussian_draw(state["params_proposal"], batch_size // 2)
+        thetas = gaussian_draw(state["params_proposal"],
+                               batch_size // 2, random_state=rng)
         _X_gen = np.zeros((batch_size // 2, n_features))
[…]

### 2017-06-30 code/weinberg.py

@@ -7,30 +7,25 @@ from nn import glorot_uniform
 from nn import relu
 from nn import AdamOptimizer
 
-# from proposals import make_gaussian_proposal
-# from proposals import gaussian_draw
-# from proposals import gaussian_logpdf
-# from proposals import grad_gaussian_logpdf
-# from proposals import grad_gaussian_entropy
+from proposals import make_gaussian_proposal
+from proposals import gaussian_draw
+from proposals import gaussian_logpdf
+from proposals import grad_gaussian_logpdf
+from proposals import grad_gaussian_entropy
 
-# XXX switch to a gaussian proposal
-
-from proposals import make_beta_proposal
-from proposals import beta_draw
-from proposals import beta_logpdf
-from proposals import grad_beta_logpdf
-from proposals import grad_beta_entropy
 
 from sklearn.utils import check_random_state
 
 # Global params
 
+rng = check_random_state(321)
+
 batch_size = 64
 n_epochs = 300+1
-lambda_gp = 0.025
-gamma = 0.0
+lambda_gp = 0.0025
 
-true_theta = np.array([(42 - 40) / (50 - 40)])
+true_theta = np.array([(42.0-40)/(50-40),
+                       (0.9 - 0.5) / (1.5-0.5)])
 make_plots = True
 
 
@@ -51,8 +46,8 @@ def diffxsec(costheta, sqrtshalf, gf):
     return ((1 + costheta ** 2) + a_fb(sqrtshalf, gf) * costheta) / norm
 
 def rej_sample_costheta(n_samples, theta, rng):
-    sqrtshalf = theta[0]
-    gf = 0.9
+    sqrtshalf = theta[0] * (50-40) + 40
+    gf = theta[1] * (1.5 - 0.5) + 0.5
 
     ntrials = 0
     samples = []
@@ -71,174 +66,205 @@ def rej_sample_costheta(n_samples, theta, rng):
     return np.array(samples)
 
 def simulator(theta, n_samples, random_state=None):
-    theta = copy.copy(theta)
-    theta[0] = theta[0] * (50 - 40) + 40
-
     rng = check_random_state(random_state)
     samples = rej_sample_costheta(n_samples, theta, rng)
 
     return samples.reshape(-1, 1)
 
-X_obs = simulator(true_theta, 20000, random_state=123)
+X_obs = simulator(true_theta, 50000, random_state=rng)
 n_params = len(true_theta)
 n_features = X_obs.shape[1]
 
 
[…]

### 2017-06-30 tex/paper.tex
save

@@
 That is,
-\begin{equation}
+\begin{equation}\label{eqn:grad-ug-approx}
     \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
@@
 distance with $\theta^*$, both models closely match the observed data, as shown
-in the top right plot of  Figure~\ref{fig:poisson}. This suggests either a
+in the top right plot of  Figure~\ref{fig:poisson}, with again
+slighlty better results for the entropy penalized model. This suggests either a
 relatively flat landscape around $\theta^*$ or that the observed data can in
@@
 
-\glnote{avo = extension of gans to non-differentiable generator}
-\glnote{nice for discrete data}
-\glnote{next version = try on an actual particle physics simulator.}
-\glnote{nes for better invariance and stability?}
+In this note, we develop a likelihood-free inference algorithm for fitting a
+forward non-differentiable simulator to continuous or discrete observed data.
+The algorithm combines adversarial training  with variational optimization to
+minimize variational upper bounds  on the otherwise non-differentiable
+adversarial objectives.
+Effectively, the procedure results in learning an abitrarily tight proposal
+distribution over the simulator parameters, such that corresponding the marginal
+distribution of the generated data matches the observations. Preliminary results
+with non-differentiable discrete or continuous simulators validate the proposed method.
 
-\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
-is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?}
+While the obtained results are encouraging, the complete validation of the
+method remains to be carried out in real conditions on a full fledged scientific
+simulator -- which we plan to achieve for a next version of this work.
+In terms of method, several components need to be further investigated.
+First, we need to better study the interplay between the entropy penalty and the adversarial objectives.
+Second, we should better understand the dynamics of the optimization
+procedure, in particular when combined with momentum-based optimizers like Adam.
+Third, we need to consider whether less noisy estimates of the gradients
+$\nabla_\psi U_g$ can be computed, in particular by exploiting the fact in the composition $d(g(\mathbf{z};\theta); \psi)$
+the derivatives of $d$ are known exactly.
 

### 2017-06-30 tex/paper.tex
save

@@
              ({\it Left}) Likelihood $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
-             ({\it Right}) Likelihood $q(\theta|\psi)$ after adverarial variational optimization. The proposal density correctly concentrates around $\theta^* = (1, -1)$.
+             ({\it Right}) Likelihood $q(\theta|\psi)$ after adversarial variational optimization. The proposal density correctly concentrates around $\theta^* = (1, -1)$.
              }\label{fig:multi}
@@
 and $G_f$ (the Fermi constant) through the coefficient function $c$. Given the
-known analytical form of the probability density function, a simulator for
+known analytic form of the probability density function, a simulator for
 $\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
@@
 
-The top left plot in Figure~\ref{fig:poisson} illustates the resulting proposal
+The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting proposal
 distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
@@
 distance with $\theta^*$, both models closely match the observed data, as shown
-in the top right plot of  Figure~\ref{fig:poisson}, with again
-slighlty better results for the entropy penalized model. This suggests either a
+in the top right plot of  Figure~\ref{fig:weinberg}, with again
+slightly better results for the entropy penalized model. This suggests either a
 relatively flat landscape around $\theta^*$ or that the observed data can in
@@
 adversarial objectives.
-Effectively, the procedure results in learning an abitrarily tight proposal
+Effectively, the procedure results in learning an arbitrarily tight proposal
 distribution over the simulator parameters, such that corresponding the marginal

### 2017-06-30 tex/paper.tex
save

@@
 proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
-initialized with zero mean and unit variance. Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
+initialized with zero mean and unit variance.
+Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
 Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
+The network architecture of the critic is the same as in the previous example.
 
@@
 
-The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting proposal
-distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
-variational optimization. We see that the distributions both arrive in the
-neighborhood of $\theta^*$, with a density highly more concentrated for
+The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
+proposal distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after
+adversarial variational optimization. We see that the distributions both arrive
+in the neighborhood of $\theta^*$, with a density highly more concentrated for
 $\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
 distance with $\theta^*$, both models closely match the observed data, as shown
-in the top right plot of  Figure~\ref{fig:weinberg}, with again
-slightly better results for the entropy penalized model. This suggests either a
-relatively flat landscape around $\theta^*$ or that the observed data can in
-this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
-the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
-for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
-fall towards $0$, which indicates  convergence towards a distribution that the
-critic cannot distinguish from the true distribution.
+in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
+results for the entropy penalized model. This suggests either a relatively flat
+landscape around $\theta^*$ or that the observed data can in this case also be
+reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of the simulator.
+Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
+$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
+towards $0$, which indicates  convergence towards a distribution that the critic
+cannot distinguish from the true distribution. In summary, despite the intricate
+relation betwen the observations $\mathbf{x}$, the latent variable $\mathbf{z}$
+and the parameters $\theta$, this example demonstrates once again the efficacy
+of adversarial variational optimization for parameter inference on black box
+simulator.
 
@@
 
+We would like to thank Lukas Heinrich for helpful comments regarding
+the electron--positron annihilation simulation.
+
 GL and KL are both supported through NSF ACI-1450310, additionally KC is

### 2017-06-30 tex/paper.tex
save

--- 2017-06-30 tex/paper.tex
@@
 
-Complex computer simulators are used across fields of science to
+Complex computer simulators are increasingly used across fields of science to
 describe generative models tying parameters of an underlying theory to
@@
 In consequence, the proposed inference algorithm does not necessarily guarantee that the
-proposal distribution $q(\theta|\psi)i$ will place its mass arbitrarily tight
+proposal distribution $q(\theta|\psi)$ will place its mass arbitrarily tight
 around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
@@
     \caption{Multidimensional continuous data.
-             ({\it Left}) Likelihood $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
-             ({\it Right}) Likelihood $q(\theta|\psi)$ after adversarial variational optimization. The proposal density correctly concentrates around $\theta^* = (1, -1)$.
+             ({\it Left}) Density $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
+             ({\it Right}) Density $q(\theta|\psi)$ after adversarial variational optimization.
+The proposal density correctly converges towards a distributon whose density concentrates around $\theta^* = (1, -1)$.
              }\label{fig:multi}
@@
 
-The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting proposal
-distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
-variational optimization. We see that the distributions both arrive in the
-neighborhood of $\theta^*$, with a density highly more concentrated for
+The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
+proposal distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after
+adversarial variational optimization. We see that the distributions both arrive
+in the neighborhood of $\theta^*$, with a density highly more concentrated for
 $\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
 distance with $\theta^*$, both models closely match the observed data, as shown
-in the top right plot of  Figure~\ref{fig:weinberg}, with again
-slightly better results for the entropy penalized model. This suggests either a
-relatively flat landscape around $\theta^*$ or that the observed data can in
-this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
-the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
-for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
-fall towards $0$, which indicates  convergence towards a distribution that the
-critic cannot distinguish from the true distribution.
+in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
+results for the entropy penalized model. This suggests either a relatively flat
+landscape around $\theta^*$ or that the observed data can in this case also be
+reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of the simulator.
+Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
+$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
+towards $0$, which indicates  convergence towards a distribution that the critic
+cannot distinguish from the true distribution. In summary, despite the intricate
+relation betwen the observations $\mathbf{x}$, the latent variable $\mathbf{z}$
+and the parameters $\theta$, this example demonstrates the efficacy
+of adversarial variational optimization for parameter inference on a black box
+simulator.
 
@@
 learning a differentiable implicit generative model. In the likelihood-free
-setup, \citep{cranmer2015approximating,2016arXiv161110242D} estimate likelihood
+setup, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
 ratios through supervised classification, which can in turn be used for
@@
 descent alternative to bayesian optimization based on density
-ratios~\citep{cranmer2015approximating} or to classifier
+ratios~\citep{cranmer2015approximating,cranmer2016experiments} or to classifier
 ABC~\citep{gutmann2017likelihood}.

### 2017-07-01 tex/paper.tex
save

@@
 
-Complex computer simulators are increasingly used across fields of science to
-describe generative models tying parameters of an underlying theory to
+Complex computer simulators are increasingly used across fields of science as
+generative models tying parameters of an underlying theory to
 experimental observations. Inference in this setup is often
 difficult, as simulators rarely provide a way to directly evaluate the likelihood
-function for a given observation. In this note, we develop a likelihood-free
-inference algorithm for fitting a forward non-differentiable generative model to
+function. In this note, we develop a likelihood-free
+inference algorithm for fitting a forward, non-differentiable generative model to
 continuous or discrete observed data. We adapt the adversarial training procedure of generative
-adversarial networks by replacing the implicit generative network with a
-domain-based scientific simulator, and solve the resulting non-differentiable
+adversarial networks by replacing the differentiable generative network with a
+domain-specific simulator and solve the resulting non-differentiable
 minimax problem by minimizing variational upper bounds of the adversarial objectives.
@@
 trained to optimality before each parameter update of the generator, it can
-be shown that the original adversarial learning procedure amounts to minimizing
+be shown that the original adversarial learning procedure~\cite{goodfellow2014generative} amounts to minimizing
 the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\theta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\theta)$.
@@
     \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] \nonumber \\
-    &= \nabla_\psi \int f(\theta)  q(\theta|\psi)  d\theta \nonumber \\
+    %&= \nabla_\psi \int f(\theta)  q(\theta|\psi)  d\theta \nonumber \\
     &= \int f(\theta) \nabla_\psi q(\theta|\psi)  d\theta \nonumber \\
@@
     \begin{algorithm}[H]
-    \caption{Adversarial variational optimization.}
+    \caption{Adversarial variational optimization (AVO).}
 
@@
 GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
-the setting where we are not interested in learning the implicit model itself but are
-rather interested in the inference of parameters of a fixed
-non-differentiable simulator (Section~\ref{sec:problem}),
+the setting where we are interested in the inference of the parameters of a
+fixed non-differentiable simulator (Section~\ref{sec:problem}),
+rather than in learning the generative model itself,
 gradients $\nabla_\theta g$ either do not exist or cannot be accessed. As a
@@
 For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
-optimization procedure, as built on top of WGAN-GP.
+optimization (AVO) procedure, as built on top of WGAN-GP.
 Obviously, the variational relaxation could similarly be coupled with
@@
 As a first illustrative experiment, we evaluate inference for a discrete Poisson
-distribution with unknown scale parameter $\lambda$. We artificially consider
+distribution with unknown mean $\lambda$. We artificially consider
 the distribution as a parameterized simulator, from which we can only
@@
 
-The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
+The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
 Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
@@
     \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
-    \caption{Discrete Poisson model with unknown scale.
+    \caption{Discrete Poisson model with unknown mean.
              ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization.
 For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
-                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
+                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
              ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
@@
 The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
-distributions $q(\theta|\psi)$ after adversarial variational optimization.  For
+distributions $q(\theta|\psi)$ after AVO.  For
 both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
@@
 adversarial training, we see that the resulting distributions closely match
-the true distribution, with in this case slightly  better results for the penalized
+the true distribution, with in this case visually slightly better results for the penalized
 model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
@@
 
-As a second example, we evaluate parameter inference for a generator producing
+As a second toy example, we evaluate parameter inference for a generator producing
 5-dimensional continuous data, as originally specified in Section 4.2. of
@@
 proposal distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after
-adversarial variational optimization. We see that the distributions both arrive
-in the neighborhood of $\theta^*$, with a density highly more concentrated for
+AVO. We see that the distributions both arrive
+in the neighborhood of $\theta^*$, with a density  more highly concentrated for
 $\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
@@
 and the parameters $\theta$, this example demonstrates the efficacy
-of adversarial variational optimization for parameter inference on a black box
+of AVO for parameter inference on a black box
 simulator.
@@
 non-differentiable simulators. It also constitutes an approximate gradient
-descent alternative to bayesian optimization based on density
+descent alternative to likelihood-free inference based on bayesian optimization and approximated density
 ratios~\citep{cranmer2015approximating,cranmer2016experiments} or to classifier

### 2017-07-01 code/multi.py
update code

@@ -23,7 +23,8 @@ from sklearn.utils import check_random_state
 
 # Global params
 
-rng = check_random_state(777)
+seed = 777
+rng = check_random_state(seed)
 
 batch_size = 64
 n_epochs = 300+1
@@ -134,7 +135,7 @@ def approx_grad_u(params_proposal, i, gamma=gamma):
     thetas = gaussian_draw(params_proposal, batch_size, random_state=rng)
 
     for theta in thetas:
-        x = simulator(theta, 1)
+        x = simulator(theta, 1, random_state=rng)
         dx = predict(x, params_critic).ravel()
 
         grad_q = grad_gaussian_logpdf(params_proposal, theta)
@@ -224,6 +225,6 @@ for i in range(n_epochs):
             plt.tight_layout()
 
             if i == n_epochs - 1:
-                plt.savefig("figs/multi-%d.pdf" % seed)
+                plt.savefig("figs/multi.pdf")
 
             plt.close()

### 2017-07-01 code/toy-poisson.py

@@ -17,7 +17,8 @@ from sklearn.utils import check_random_state
 
 # Global params
 
-rng = check_random_state(123)
+seed = 777
+rng = check_random_state(seed)
 
 batch_size = 64
 n_epochs = 300+1
@@ -119,7 +120,7 @@ for state in history:
         thetas = gaussian_draw(params_proposal, batch_size, random_state=rng)
 
         for theta in thetas:
-            x = simulator(theta, 1)
+            x = simulator(theta, 1, random_state=rng)
             dx = predict(x, state["params_critic"]).ravel()
 
             grad_q = grad_gaussian_logpdf(params_proposal, theta)
@@ -169,10 +170,10 @@ if make_plots:
 
     # proposal
     ax1 = plt.subplot2grid((2, 2), (0, 0))
-    plt.xlim(true_theta[0]-3, true_theta[0]+3)
+    plt.xlim(true_theta[0]-1, true_theta[0]+2)
 
     plt.axvline(x=true_theta[0], linestyle="--", label=r"$\theta^*$")
-    thetas = np.linspace(true_theta[0]-3, true_theta[0]+3, num=300)
+    thetas = np.linspace(true_theta[0]-1, true_theta[0]+2, num=300)
 
     for state in history:
         logp = np.array([gaussian_logpdf(state["params_proposal"],
@@ -180,7 +181,7 @@ if make_plots:
                          for theta in thetas])
         plt.plot(thetas,
                  np.exp([l[0] for l in logp]),
-                 label=r"$q(\theta|\psi) \quad\gamma=%d$" % state["gamma"],
+                 label=r"$q(\theta|\psi)\quad\gamma=%d$" % state["gamma"],
                  color=state["color"])
 
     plt.legend(loc="upper right")
@@ -201,7 +202,7 @@ if make_plots:
 
     plt.hist(Xs, histtype="bar",
              label=[r"$x \sim p_r(x)$"] +
-                   [r"$x \sim p(x|\psi) \quad\gamma=%d$" % state["gamma"] for state in history],
+                   [r"$x \sim p(x|\psi)\quad\gamma=%d$" % state["gamma"] for state in history],
              color=["C0"] + [state["color"] for state in history],
              range=(0, 15), bins=16, normed=1)
     plt.legend(loc="upper right")
@@ -213,7 +214,7 @@ if make_plots:
     for state in history:
         plt.plot(xs,
                  state["loss_d"],
[…]

### 2017-07-01 code/weinberg.py

@@ -18,7 +18,8 @@ from sklearn.utils import check_random_state
 
 # Global params
 
-rng = check_random_state(321)
+seed = 123
+rng = check_random_state(seed)
 
 batch_size = 64
 n_epochs = 300+1
@@ -156,7 +157,7 @@ for state in history:
         thetas = gaussian_draw(params_proposal, batch_size, random_state=rng)
 
         for theta in thetas:
-            x = simulator(theta, 1)
+            x = simulator(theta, 1, random_state=rng)
             dx = predict(x, state["params_critic"]).ravel()
 
             grad_q = grad_gaussian_logpdf(params_proposal, theta)
@@ -206,10 +207,11 @@ if make_plots:
 
     # proposal
     ax1 = plt.subplot2grid((2, 2), (0, 0))
+    offset = 0.15
 
     for state in history:
-        x = np.linspace(true_theta[0]-0.3, true_theta[0]+0.3, num=300)
-        y = np.linspace(true_theta[1]-0.3, true_theta[1]+0.3, num=300)
+        x = np.linspace(true_theta[0]-offset, true_theta[0]+offset, num=300)
+        y = np.linspace(true_theta[1]-offset, true_theta[1]+offset, num=300)
         X, Y = np.meshgrid(x, y)
         Z = [gaussian_logpdf(state["params_proposal"], theta)
              for theta in np.hstack([X.reshape(-1, 1), Y.reshape(-1, 1)])]
@@ -218,7 +220,7 @@ if make_plots:
         CS = plt.contour(X*(50-40)+40, Y*(1.5-0.5)+0.5, Z, colors=state["color"])
         plt.clabel(CS)
         plt.plot([-999], [-999],
-                 label=r"$q(\theta|\psi) \quad\gamma=%d$" % state["gamma"],
+                 label=r"$q(\theta|\psi)\quad\gamma=%d$" % state["gamma"],
                  color=state["color"])
 
     plt.scatter(true_theta[0]*(50-40)+40, true_theta[1]*(1.5-0.5)+0.5,
@@ -228,10 +230,10 @@ if make_plots:
 
     plt.xlabel(r"$E_{beam}$")
     plt.ylabel(r"$G_f$")
-    plt.xlim((true_theta[0]-0.3)*(50-40)+40, (true_theta[0]+0.3)*(50-40)+40)
[…]

### 2017-07-01 code/weinberg.py
save

@@ -18,7 +18,7 @@ from sklearn.utils import check_random_state
 
 # Global params
 
-seed = 123
+seed = 777
 rng = check_random_state(seed)
 
 batch_size = 64

### 2017-07-02 code/toy-poisson.py
save

@@ -181,7 +181,7 @@ if make_plots:
                          for theta in thetas])
         plt.plot(thetas,
                  np.exp([l[0] for l in logp]),
-                 label=r"$q(\theta|\psi)\quad\gamma=%d$" % state["gamma"],
+                 label=r"$q(\theta|\psi)\ \gamma=%d$" % state["gamma"],
                  color=state["color"])
 
     plt.legend(loc="upper right")
@@ -202,7 +202,7 @@ if make_plots:
 
     plt.hist(Xs, histtype="bar",
              label=[r"$x \sim p_r(x)$"] +
-                   [r"$x \sim p(x|\psi)\quad\gamma=%d$" % state["gamma"] for state in history],
+                   [r"$x \sim p(x|\psi)\ \gamma=%d$" % state["gamma"] for state in history],
              color=["C0"] + [state["color"] for state in history],
              range=(0, 15), bins=16, normed=1)
     plt.legend(loc="upper right")
@@ -214,13 +214,13 @@ if make_plots:
     for state in history:
         plt.plot(xs,
                  state["loss_d"],
-                 label=r"$-U_d\quad\gamma=%d$" % state["gamma"],
+                 label=r"$-U_d\ \gamma=%d$" % state["gamma"],
                  color=state["color"])
     plt.xlim(0, n_epochs)
     plt.ylim(0, 30)
     plt.legend(loc="upper right")
 
     plt.tight_layout()
-    plt.savefig("figs/poisson.pdf")
+    plt.savefig("figs/poisson-%d.pdf" % seed)
 
     plt.close()

### 2017-07-02 code/weinberg.py

@@ -220,7 +220,7 @@ if make_plots:
         CS = plt.contour(X*(50-40)+40, Y*(1.5-0.5)+0.5, Z, colors=state["color"])
         plt.clabel(CS)
         plt.plot([-999], [-999],
-                 label=r"$q(\theta|\psi)\quad\gamma=%d$" % state["gamma"],
+                 label=r"$q(\theta|\psi)\ \gamma=%d$" % state["gamma"],
                  color=state["color"])
 
     plt.scatter(true_theta[0]*(50-40)+40, true_theta[1]*(1.5-0.5)+0.5,
@@ -249,7 +249,7 @@ if make_plots:
 
     plt.hist(Xs, histtype="bar",
              label=[r"$x \sim p_r(x)$"] +
-                   [r"$x \sim p(x|\psi)\quad\gamma=%d$" % state["gamma"] for state in history],
+                   [r"$x \sim p(x|\psi)\ \gamma=%d$" % state["gamma"] for state in history],
              color=["C0"] + [state["color"] for state in history],
              range=(-1, 1), bins=15, normed=1)
     plt.legend(loc="upper right")
@@ -261,7 +261,7 @@ if make_plots:
     for state in history:
         plt.plot(xs,
                  state["loss_d"],
-                 label=r"$-U_d\quad\gamma=%d$" % state["gamma"],
+                 label=r"$-U_d\ \gamma=%d$" % state["gamma"],
                  color=state["color"])
     plt.xlim(0, n_epochs-1)
     plt.legend(loc="upper right")

### 2017-07-03 tex/paper.tex
save
ref for VO

@@
 non-differentiable minimax problem by minimizing variational upper
-bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
+bounds~\citep{2011arXiv1106.4487W,staines2013optimization} of the adversarial
 objectives. The procedure results in learning a proposal distribution over
@@
 
-Variational optimization~\cite{2012arXiv1212.4507S} and evolution strategies~\citep{2011arXiv1106.4487W} are general
+Variational optimization~\cite{staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
 optimization techniques that can be used to form a differentiable bound
@@
 
-Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
+Under mild restrictions outlined in  \citep{staines2013optimization}, the bound
 $U(\psi)$ is differentiable with respect to $\psi$, and using the log-likelihood
@@
              ({\it Left}) Density $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
-             ({\it Right}) Density $q(\theta|\psi)$ after adversarial variational optimization.
+             ({\it Right}) Density $q(\theta|\psi)$ after adversarial variational optimization ($\gamma=10$).
 The proposal density correctly converges towards a distributon whose density concentrates around $\theta^* = (1, -1)$.
@@
 \subsection{Electron--positron annihilation}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
-    \caption{Electron--positron annihilation.
-    ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization.
-The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\theta^*$.
-    ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
-    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
-             }\label{fig:weinberg}
-\end{figure}
 
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
+    \caption{Electron--positron annihilation.
+    ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization.
+The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\theta^*$.
+    ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
+    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+             }\label{fig:weinberg}
+\end{figure}
+
 % ==============================================================================
@@
 ABC~\citep{gutmann2017likelihood}.
+
+\glnote{Discuss 1702.08235 and references therein on adversarial variational bayes/inference.}
 

### 2017-07-04 code/weinberg.py
save

@@ -13,8 +13,8 @@ from proposals import gaussian_logpdf
 from proposals import grad_gaussian_logpdf
 from proposals import grad_gaussian_entropy
 
-
 from sklearn.utils import check_random_state
+from scipy.spatial.distance import mahalanobis
 
 # Global params
 
@@ -28,6 +28,7 @@ lambda_gp = 0.0025
 true_theta = np.array([(42.0-40)/(50-40),
                        (0.9 - 0.5) / (1.5-0.5)])
 make_plots = True
+plt.rcParams["figure.dpi"] = 200
 
 
 # Simulator
@@ -213,12 +214,20 @@ if make_plots:
         x = np.linspace(true_theta[0]-offset, true_theta[0]+offset, num=300)
         y = np.linspace(true_theta[1]-offset, true_theta[1]+offset, num=300)
         X, Y = np.meshgrid(x, y)
-        Z = [gaussian_logpdf(state["params_proposal"], theta)
-             for theta in np.hstack([X.reshape(-1, 1), Y.reshape(-1, 1)])]
-        Z = np.exp(np.array(Z).reshape(X.shape))
 
-        CS = plt.contour(X*(50-40)+40, Y*(1.5-0.5)+0.5, Z, colors=state["color"])
-        plt.clabel(CS)
+        mu = state["params_proposal"]["mu"]
+        sigma = np.diag(np.exp(state["params_proposal"]["log_sigma"])) ** 2.0
+        sigma_inv = np.linalg.inv(sigma)
+
+        Z = [mahalanobis(theta, mu, sigma_inv)
+             for theta in np.hstack([X.reshape(-1, 1), Y.reshape(-1, 1)])]
+        Z = np.array(Z).reshape(X.shape)
+
+        CS = plt.contour(X*(50-40)+40, Y*(1.5-0.5)+0.5, Z, [1.0, 2.0, 3.0], colors=state["color"])
+        fmt = {l:s for l, s in zip(CS.levels, [r"$1\sigma$", r"$2\sigma$", r"$3\sigma$"])}
+        plt.clabel(CS, fmt=fmt)
+
+        plt.scatter(mu[0]*(50-40)+40, mu[1]*(1.5-0.5)+0.5, c=state["color"], marker="+")
         plt.plot([-999], [-999],
                  label=r"$q(\theta|\psi)\ \gamma=%d$" % state["gamma"],
                  color=state["color"])

### 2017-07-06 tex/paper.tex
some notes on OPVI

--- 2017-07-01 tex/paper.tex
@@
 \begin{abstract}
-
 Complex computer simulators are increasingly used across fields of science as
@@
 experimental observations. Inference in this setup is often
-difficult, as simulators rarely provide a way to directly evaluate the likelihood
-function. In this note, we develop a likelihood-free
-inference algorithm for fitting a forward, non-differentiable generative model to
-continuous or discrete observed data. We adapt the adversarial training procedure of generative
+difficult, as simulators rarely admit a tractable likelihood
+function. In this note, we develop Adversarial Variational Optimization (AVO), a likelihood-free
+inference algorithm for fitting a non-differentiable generative model to
+continuous or discrete  data. We adapt the training procedure of generative
 adversarial networks by replacing the differentiable generative network with a
 domain-specific simulator and solve the resulting non-differentiable
-minimax problem by minimizing variational upper bounds of the adversarial objectives.
+minimax problem by minimizing variational upper bounds of the two adversarial objectives.
 Effectively, the procedure results in learning an arbitrarily tight
@@
 marginal distribution of the generated data matches the observations.
-Preliminary results with non-differentiable simulators producing discrete/continuous univariate/multi-dimensional data
+We present results of the method with with non-differentiable simulators producing discrete/continuous univariate/multi-dimensional data
 validate the proposed method.
@@
 replacing the adversarial classifier with a 1-Lipschitz adversarial critic
-$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
+$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. 
+\begin{align}
+   W( p_r(\mathbf{x} ), p(\mathbf{x} | \theta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\mathbf{x}|\theta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
+\end{align}
+Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
 for stabilizing the optimization procedure,
 training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize
+
 \begin{align}
-    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
-                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
-    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)]
+    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
+    {\cal L}_g =\,& - {\cal L}_W
 \end{align}
+%\begin{align}
+%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
+%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
+%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)]
+%\end{align}
 where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
@@
 
+\subsection{Operator Variational Optimization}
+
+Original OPVI
+\begin{eqnarray}
+\lambda^* = \inf_\lambda \sup_\phi \mathbb{E}_{ z \sim q( z | \lambda)} [ (O^{p,q} f_\lambda) ] 
+\end{eqnarray}
+with
+\begin{equation}
+(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
+\end{equation}
+
+%This is natural here, but not quite OPVI formalism:
+%\begin{equation}
+%\psi^* = \inf_\psi \sup_\phi \mathbb{E}_{ \theta \sim q(\theta | \psi)} [ (O^{p,q} f_\phi) ]
+%\end{equation}
+%
+%\begin{equation}
+%(O^{p,q} f) = W(p,q) = \sup_f \mathbb{E}_{x\sim p}[f(x)] - \mathbb{E}_{x \sim q}[f(x)]
+%\end{equation}
+%The expectation should be $x \sim p(x|\psi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$
+
+To get into OPVI formalism
+
+\begin{eqnarray}
+\psi^* &=& \inf_\psi \sup_\phi\mathbb{E}_{x\sim p_r(x)}[f(x)] - \mathbb{E}_{x \sim p(x|\psi)}[f(x)] \\
+&=&  \inf_\psi \sup_\phi \mathbb{E}_{ x \sim p( x | \psi)} [ (O^{p_r,p_\psi} f_\phi) ] 
+\end{eqnarray}
+with
+\begin{equation}
+(O^{p,q} f) = \left( \frac{p_r(x)}{p(x|\psi)}  -1 \right) f(x)
+\end{equation}
+
+We can think of $p(x|\psi)$ as a \textit{variational program} as described in ~\cite{opvi}, though more complicated than the a simple reparametrization of normally distributed noise $\epsilon$ through a differentiable function $R(\epsilon, \psi)$.
+In our case, the variational program is a marginalized non-differentiable simulator simulator.
+Nevertheless, it can generate samples for $x$, is differentiable with respect to $\phi$, and its density is be intractable.
+
 % ==============================================================================

### 2017-07-06 tex/paper.tex
some non-final notes on other works

@@
 \begin{align}
-   W( p_r(\mathbf{x} ), p(\mathbf{x} | \theta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\mathbf{x}|\theta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
+   W( p_r(\mathbf{x} ), p(\mathbf{x} | \theta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\tilde{\mathbf{x}}|\theta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
 \end{align}
@@
 
-Variational optimization~\cite{2012arXiv1212.4507S} and evolution strategies~\citep{2011arXiv1106.4487W} are general
+Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
 optimization techniques that can be used to form a differentiable bound
@@
 
-Original OPVI
+Original OPVI  \citep{2016arXiv161009033R}
+
 \begin{eqnarray}
@@
 
-\glnote{Discuss 1702.08235 and references therein on adversarial variational bayes/inference.}
+This work has many connections to variational inference, in which the goal is to optimize the recognition model $q(\theta|\psi)$ so that it is close to the true posterior $p(\theta|x)$.
+This is slighty different from variational optimization (or variational learning), where the toal is to find $\theta^*$ such that $p(x|\theta^*) \approx p_r(x)$.
+Typically in variational inference the model $p(x|\theta)$ admits a density, so that the likelihood is known.
+Thus, assuming the prior also admits a density, it is tractable to evaluate the posterior is (up to a constant factor).
+In Adversarial Variational Bayes \citep{DBLP:journals/corr/MeschederNG17} and , equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H}), the authors use an adversarial setup, but assume a known density $p(x|\theta)$ that is differentiable with respect to $\theta$.
+
+In \citep{tran2017variational}, the authors consider Variational Bayes with an intractable likelihood.
+In that approach "The only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of most simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(x|\theta)$.
+
+In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.
+The  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\theta$''.
 
@@
 
+\textbf{KC: could also use this to tune non-differentialble fast simulator to slower full simulator. Then fixed data size not an issue}
+
 % ==============================================================================

### 2017-07-10 tex/paper.tex
minor notation changes

@@
 \usepackage{bbold}
+\usepackage{bm}
 \usepackage{slashed}
@@
 \newcommand{\kcnote}[1]{\textcolor{red}{[KC: #1]}}
+
+\newcommand{\qxpsi}{q(\mathbf{x}|\bfpsi)}
+\newcommand{\bftheta}{{\bm \theta}}
+\newcommand{\bfpsi}{{\bm \psi}}
+\newcommand{\bfx}{\mathbf{x}}
+\newcommand{\bfz}{\mathbf{z}}
 
@@
 
-We consider a family of parameterized densities $p(\mathbf{x}|\theta)$
+We consider a family of parameterized densities $p(\mathbf{x}|\bftheta)$
 defined implicitly through the simulation of a stochastic generative process,
-where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
+where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
 parameters of interest. The simulation may involve some complicated latent
@@
 \begin{equation}\label{eqn:p_x}
-    p(\mathbf{x}|\theta) = \int p(\mathbf{x}|\mathbf{z},\theta) p(\mathbf{z}|\theta) d\mathbf{z}
+    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
 \end{equation}
-where $\mathbf{z} \in {\cal Z}$ is a latent variable providing an external
-source of randomness. In particular, $\mathbf{z}$ is not necessarily assumed to
+where $\bfz \in {\cal Z}$ is a latent variable providing an external
+source of randomness. In particular, $\bfz$ is not necessarily assumed to
 be a fixed-size vector (e.g., it can be a sequence of variable length) and its
-distribution $p(\mathbf{z}|\theta)$ may itself depend on $\theta$ in some intricate way.
+distribution $p(\bfz|\bftheta)$ may itself depend on $\bftheta$ in some intricate way.
 
 We assume that we already have an accurate simulation of the stochastic
-generative process that defines $p(\mathbf{x}|\mathbf{z},\theta)$, as
-specified through a highly regularized deterministic function $g(\cdot; \theta) : {\cal Z} \to
+generative process that defines $p(\mathbf{x}|\bfz,\bftheta)$, as
+specified through a highly regularized deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
 \mathbb{R}^d$ with usually few parameters. That is, we consider
 \begin{equation}\label{eqn:p_theta}
-    \mathbf{x} \sim p(\mathbf{x}|\theta) \equiv \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta)
+    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
 \end{equation}
-such that the likelihood $p(\mathbf{x}|\theta)$ can be rewritten as
-\begin{equation}\label{eqn:p_x_sim}
-    p(\mathbf{x}|\theta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p(\mathbf{z}|\theta) \mu(d\mathbf{z}),
-\end{equation}
-where $\mu$ is a probability measure.
+such that the likelihood $p(\mathbf{x}|\bftheta)$ can be 
+considered as a non-differentiable reparameterization of the random variable $\bfz$.
+%rewritten as
+%\begin{equation}\label{eqn:p_x_sim}
+%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
+%\end{equation}
+%where $\mu$ is a probability measure.
 Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
@@
 in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
-in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$,
+in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
 e.g. as when $g$ is specified as a computer program.
@@
 (unknown) true distribution $p_r(\mathbf{x})$, our goal is the inference of the parameters
-of interest $\theta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
-the modeled data distribution $p(\mathbf{x}|\theta)$ induced by $g(\cdot;
-\theta)$ over $\mathbf{z}$. That is,
+of interest $\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
+the modeled data distribution $p(\mathbf{x}|\bftheta)$ induced by $g(\cdot;
+\bftheta)$ over $\bfz$. That is,
 \begin{equation}
-    \theta^* = \arg \min_{\theta \in \Theta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\theta)),
+    \bftheta^* = \arg \min_{\bftheta \in \bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
 \end{equation}
@@
 \cite{goodfellow2014generative} as a way to build an implicit generative model
-capable of producing samples from random noise $\mathbf{z}$. More specifically,
-a generative model $g(\cdot; \theta)$ is pit against an adversarial
+capable of producing samples from random noise $\bfz$. More specifically,
+a generative model $g(\cdot; \bftheta)$ is pit against an adversarial
 classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
-from generated data $\tilde{\mathbf{x}} = g(\mathbf{z}; \theta)$. Both models $g$ and $d$
+from generated data $\tilde{\mathbf{x}} = g(\bfz; \bftheta)$. Both models $g$ and $d$
 are trained simultaneously, in such a way that $g$ learns to fool
@@
 be shown that the original adversarial learning procedure~\cite{goodfellow2014generative} amounts to minimizing
-the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\theta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\theta)$.
+the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\bftheta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\bftheta)$.
 
@@
 Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
-setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\theta))$ by
+setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta))$ by
 replacing the adversarial classifier with a 1-Lipschitz adversarial critic
@@
 \begin{align}
-   W( p_r(\mathbf{x} ), p(\mathbf{x} | \theta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\tilde{\mathbf{x}}|\theta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
+   W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
 \end{align}
@@
 for stabilizing the optimization procedure,
-training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize
+training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\bftheta$ in order to respectively minimize
 
@@
 %\begin{align}
-%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
+%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
 %                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
-%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)]
+%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\phi)]
 %\end{align}
@@
 (1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
-p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)$.
+p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.
 
@@
 \begin{equation}
-    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] = U(\psi),
+    \min_{\bftheta \in \bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
 \end{equation}
-where $q(\theta|\psi)$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
+where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
 That is, the minimum of a set of function values is always less than or equal
-to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
-can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.
+to any of their average. Provided that the proposal is flexible enough, the parameters $\bfpsi$
+can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \bftheta} f(\bftheta)$.
 
 Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
-$U(\psi)$ is differentiable with respect to $\psi$, and using the log-likelihood
+$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
 trick it comes:
 \begin{align}\label{eqn:approx-grad}
-    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] \nonumber \\
-    %&= \nabla_\psi \int f(\theta)  q(\theta|\psi)  d\theta \nonumber \\
-    &= \int f(\theta) \nabla_\psi q(\theta|\psi)  d\theta \nonumber \\
-    &= \int \left[ f(\theta) \nabla_\psi \log q(\theta|\psi) \right]  q(\theta|\psi) d\theta \nonumber \\
-    &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta) \nabla_\psi \log q(\theta|\psi)]
+    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
+    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
+    &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
+    &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
+    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
 \end{align}
-Effectively, this means that provided that the score function $\nabla_\psi \log
-q(\theta|\psi)$ of the proposal is known and that one can evaluate
-$f(\mathbf{\theta})$ for any $\theta$, then one can construct empirical
+Effectively, this means that provided that the score function $\nabla_\bfpsi \log
+q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
+$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
 estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
-$U(\psi)$ with stochastic gradient descent (or a variant thereof, like
+$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, like
 Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
@@
 % \begin{equation}\label{eqn:approx-grad-nat}
-%     \tilde{\nabla}_\psi U(\psi) = \mathbf{F}(\psi)^{-1} \nabla_\psi U(\psi),
+%     \tilde{\nabla}_\bfpsi U(\bfpsi) = \mathbf{F}(\bfpsi)^{-1} \nabla_\bfpsi U(\bfpsi),
 % \end{equation}
-% where $\mathbf{F}(\psi)$ is the Fisher information matrix
+% where $\mathbf{F}(\bfpsi)$ is the Fisher information matrix
 % \begin{equation}
-%     \mathbf{F}(\psi) = \mathbb{E}_{\theta \sim q_\phi}[ \nabla_\psi \log q_\psi(\theta) \nabla_\psi \log q_\psi(\theta)^{\top} ]
+%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\phi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
 % \end{equation}
-% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\psi \log q_\psi(\theta)$ already
-% evaluated to form $\nabla_\psi U(\psi)$.
+% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\bfpsi \log q_\bfpsi(\bftheta)$ already
+% evaluated to form $\nabla_\bfpsi U(\bfpsi)$.
 
@@
         {\it Inputs:} observed data $\{ \mathbf{x}_i \sim p_r(\mathbf{x}) \}_{i=1}^N$, simulator $g$.\\
-        {\it Outputs:} proposal distribution $q(\theta|\psi)$, such that $p_r(\mathbf{x}) \approx p(\mathbf{x}|\psi)$.\\
+        {\it Outputs:} proposal distribution $q(\bftheta|\bfpsi)$, such that $\qxpsi \approx p_r(\mathbf{x})$.\\
         {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
@@
     \begin{algorithmic}[1]
-        \State{$q(\theta|\psi) \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
-        \While{$\psi$ has not converged}
+        \State{$q(\bftheta|\bfpsi) \leftarrow \text{prior on $\bftheta$ (with differentiable and known density)}$}
+        \While{$\bfpsi$ has not converged}
             \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
-                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \theta_m \sim q(\theta|\psi), \mathbf{z}_m \sim p(\mathbf{z}|\theta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
+                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \bftheta_m \sim q(\bftheta|\bfpsi), \bfz_m \sim p(\bfz|\bftheta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                 \For{$m=1$ to $M$}
-                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
+                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\bfz_m; \bftheta_m)$}
                     \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
@@
             \EndFor
-            \State{Sample a mini-batch $\{ \theta_m \sim q(\theta|\psi),  \mathbf{z}_m \sim p(\mathbf{z}|\theta_m) \}_{m=1}^M$.}  \Comment{Update $q(\theta|\psi)$}
-            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
-            % \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
-            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$}
-            \State{$\psi \leftarrow \text{Adam}(\nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi))$}
+            \State{Sample a mini-batch $\{ \bftheta_m \sim q(\bftheta|\bfpsi),  \bfz_m \sim p(\bfz|\bftheta_m) \}_{m=1}^M$.}  \Comment{Update $q(\bftheta|\bfpsi)$}
+            \State{$\nabla_\bfpsi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m)) \nabla_\bfpsi \log q_\bfpsi(\bftheta_m)$}
+            % \State{$\mathbf{F}(\bfpsi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\bfpsi \log q_\bfpsi(\bftheta_m) \nabla_\bfpsi \log q_\bfpsi(\bftheta_m)^{\top}$}
+            \State{$\nabla_\bfpsi H(q_\bfpsi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\bfpsi q_\bfpsi(\bftheta_m) \log q_\bfpsi(\bftheta_m)$}
+            \State{$\bfpsi \leftarrow \text{Adam}(\nabla_\bfpsi U_g + \gamma \nabla_\bfpsi H(q_\bfpsi))$}
         \EndWhile
@@
 rather than in learning the generative model itself,
-gradients $\nabla_\theta g$ either do not exist or cannot be accessed. As a
-result, gradients $\nabla_\theta {\cal L}_g$ cannot be constructed and the
+gradients $\nabla_\bftheta g$ either do not exist or cannot be accessed. As a
+result, gradients $\nabla_\bftheta {\cal L}_g$ cannot be constructed and the
 optimization procedure cannot be carried out.
@@
 L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
-specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
-parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
+specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
+parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and minimize in alternation the variational upper bounds
 \begin{align}
-    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
-    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
-\end{align} respectively over $\phi$ and $\psi$.
+    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
+    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
+\end{align} respectively over $\phi$ and $\bfpsi$.
 When updating
@@
 generated data, as ordinarily done in stochastic gradient descent. When updating
-$\psi$, estimates of $\nabla_\psi U_g$ can be derived with forward
+$\bfpsi$, estimates of $\nabla_\bfpsi U_g$ can be derived with forward
 simulations, as described in the previous section.
@@
 \begin{equation}\label{eqn:grad-ug-approx}
-    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
+    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\phi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
 \end{equation} which we can approximate with mini-batches of
@@
 \begin{equation}
-    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
+    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \phi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
 \end{equation}
-for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$.
+for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
 For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
@@
 have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
-a distribution parameterized in terms of $\psi$:
+a distribution parameterized in terms of $\bfpsi$:
 \begin{equation}\label{eqn:p_psi}
-    \mathbf{x} \sim p(\mathbf{x}|\psi) \equiv \theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta).
+    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
 \end{equation}
 Intuitively, this corresponds to a family of simulators, each configured
-with randomly sampled parameters $\theta \sim q(\theta|\psi)$, whose joint collection
+with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
 of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
@@
 \begin{equation}
-    p(\mathbf{x}|\psi) = \int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta.
+    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
 \end{equation}
@@
 In consequence, the proposed inference algorithm does not necessarily guarantee that the
-proposal distribution $q(\theta|\psi)$ will place its mass arbitrarily tight
-around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
+proposal distribution $q(\bftheta|\bfpsi)$ will place its mass arbitrarily tight
+around the parameters of interest, which might be an issue when one is rather interested in point estimates $\bftheta^*$.
 For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
@@
 \begin{equation}
-    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
+    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
 \end{equation}
@@
 distribution with low entropy, which has the effect of concentrating its density
-tightly around one or a few $\theta$ values. A too large penalty may however
-eventually make the optimization unstable, as the variance of $\nabla_\psi \log q(\theta_m|\psi)$
+tightly around one or a few $\bftheta$ values. A too large penalty may however
+eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
 typically increases as the entropy of the proposal decreases.
@@
 %\begin{equation}
-%\psi^* = \inf_\psi \sup_\phi \mathbb{E}_{ \theta \sim q(\theta | \psi)} [ (O^{p,q} f_\phi) ]
+%\bfpsi^* = \inf_\bfpsi \sup_\phi \mathbb{E}_{ \bftheta \sim q(\bftheta | \bfpsi)} [ (O^{p,q} f_\phi) ]
 %\end{equation}
@@
 %\end{equation}
-%The expectation should be $x \sim p(x|\psi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$
+%The expectation should be $x \sim p(x|\bfpsi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$
 
@@
 \begin{eqnarray}
-\psi^* &=& \inf_\psi \sup_\phi\mathbb{E}_{x\sim p_r(x)}[f(x)] - \mathbb{E}_{x \sim p(x|\psi)}[f(x)] \\
-&=&  \inf_\psi \sup_\phi \mathbb{E}_{ x \sim p( x | \psi)} [ (O^{p_r,p_\psi} f_\phi) ] 
+\bfpsi^* &=& \inf_\bfpsi \sup_\phi\mathbb{E}_{x\sim p_r(x)}[f(x)] - \mathbb{E}_{x \sim p(x|\bfpsi)}[f(x)] \\
+&=&  \inf_\bfpsi \sup_\phi \mathbb{E}_{ x \sim p( x | \bfpsi)} [ (O^{p_r,p_\bfpsi} f_\phi) ] 
 \end{eqnarray}
@@
 \begin{equation}
-(O^{p,q} f) = \left( \frac{p_r(x)}{p(x|\psi)}  -1 \right) f(x)
+(O^{p,q} f) = \left( \frac{p_r(x)}{p(x|\bfpsi)}  -1 \right) f(x)
 \end{equation}
 
-We can think of $p(x|\psi)$ as a \textit{variational program} as described in ~\cite{opvi}, though more complicated than the a simple reparametrization of normally distributed noise $\epsilon$ through a differentiable function $R(\epsilon, \psi)$.
+We can think of $p(x|\bfpsi)$ as a \textit{variational program} as described in ~\citep{2016arXiv161009033R}, though more complicated than the a simple reparametrization of normally distributed noise $\epsilon$ through a differentiable function $R(\epsilon, \bfpsi)$.
 In our case, the variational program is a marginalized non-differentiable simulator simulator.
@@
 each outer epoch in order to avoid building momentum in staled directions.  For
-estimating $\lambda^*$, we parameterize $\theta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
-$q(\theta|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
-each epoch, parameters $\psi$ are updated by taking one Adam step, with
+estimating $\lambda^*$, we parameterize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
+$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
+each epoch, parameters $\bfpsi$ are updated by taking one Adam step, with
 $\alpha=0.01$ and $\beta_1=\beta_2=0.5$. The gradient penalty coefficient is set to
@@
     \caption{Discrete Poisson model with unknown mean.
-             ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization.
+             ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
 For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                         the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
-             ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
+             ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
              ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
@@
 The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
-distributions $q(\theta|\psi)$ after AVO.  For
+distributions $q(\bftheta|\bfpsi)$ after AVO.  For
 both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
 their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
-the effect of the positive entropy penalty $H(q(\theta|\psi))$,
+the effect of the positive entropy penalty $H(q(\bftheta|\bfpsi))$,
 the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
@@
 the curves quickly fall towards $0$, which indicates that
-$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
+$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
 [d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
@@
 \begin{itemize}
-    \item $\mathbf{z} = (z_0, z_1, z_2, z_3, z_4)$, such that
+    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that
         $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
@@
         $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
-    \item $\mathbf{x} = R  \mathbf{z}$, where $R$ is a fixed
+    \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
     semi-positive definite $5 \times 5$ matrix defining a fixed projection
-    of $\mathbf{z}$ into the observed space.
+    of $\bfz$ into the observed space.
 \end{itemize}
-We consider observed data generated at the nominal values $\theta^* = (\alpha^*=1,\beta^*=-1)$.
+We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
 The simulator parameters are modeled with a factored Gaussian
-proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
+proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component is
 initialized with zero mean and unit variance.
@@
 
-Starting with a proposal distribution $q(\theta|\psi)$ largely spread over
+Starting with a proposal distribution $q(\bftheta|\bfpsi)$ largely spread over
 the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
 inference quickly converges towards a proposal distribution whose density
-concentrates around the nominal values $\theta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
+concentrates around the nominal values $\bftheta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
 Overall, this example further illustrates and confirms the ability of adversarial
 variational optimization for inference with multiple parameters and multidimensional
-data, where reliable approximations of $p(\mathbf{x}|\theta)$ in a traditional
+data, where reliable approximations of $p(\mathbf{x}|\bftheta)$ in a traditional
 MLE setting would otherwise be difficult to construct.
@@
     \caption{Multidimensional continuous data.
-             ({\it Left}) Density $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
-             ({\it Right}) Density $q(\theta|\psi)$ after adversarial variational optimization ($\gamma=10$).
-The proposal density correctly converges towards a distributon whose density concentrates around $\theta^* = (1, -1)$.
+             ({\it Left}) Density $q(\bftheta|\bfpsi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
+             ({\it Right}) Density $q(\bftheta|\bfpsi)$ after adversarial variational optimization ($\gamma=10$).
+The proposal density correctly converges towards a distributon whose density concentrates around $\bftheta^* = (1, -1)$.
              }\label{fig:multi}
@@
 acceptance threshold, this illustrates how the latent variable
-$\mathbf{z}$ does not necessarily need to be fixed-sized vector, and how its
-density  $p(\mathbf{z}|\theta)$ may depend on $\theta$ in some complicated way
+$\bfz$ does not necessarily need to be fixed-sized vector, and how its
+density  $p(\bfz|\bftheta)$ may depend on $\bftheta$ in some complicated way
 (Section~\ref{sec:problem}).
 
-In this example, we consider observed data generated with $\theta^* =
+In this example, we consider observed data generated with $\bftheta^* =
 (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
-factored Gaussian proposal distribution $q(\theta|\psi) = q(E_\text{beam}|\psi)
-q(G_f|\psi)$, where each component is respectively initialized with mean $45$
+factored Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
+q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
 and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
@@
 The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
-proposal distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after
+proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
 AVO. We see that the distributions both arrive
-in the neighborhood of $\theta^*$, with a density  more highly concentrated for
+in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
 $\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
-distance with $\theta^*$, both models closely match the observed data, as shown
+distance with $\bftheta^*$, both models closely match the observed data, as shown
 in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
 results for the entropy penalized model. This suggests either a relatively flat
-landscape around $\theta^*$ or that the observed data can in this case also be
-reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of the simulator.
+landscape around $\bftheta^*$ or that the observed data can in this case also be
+reproduced with a smeared variation $\qxpsi$ of the simulator.
 Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
@@
 cannot distinguish from the true distribution. In summary, despite the intricate
-relation betwen the observations $\mathbf{x}$, the latent variable $\mathbf{z}$
-and the parameters $\theta$, this example demonstrates the efficacy
+relation betwen the observations $\mathbf{x}$, the latent variable $\bfz$
+and the parameters $\bftheta$, this example demonstrates the efficacy
 of AVO for parameter inference on a black box
@@
     \caption{Electron--positron annihilation.
-    ({\it Top left}) Proposal distributions $q(\theta|\psi)$ after adversarial variational optimization.
-The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\theta^*$.
-    ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
+    ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
+The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\bftheta^*$.
+    ({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
     ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
@@
 
-This work has many connections to variational inference, in which the goal is to optimize the recognition model $q(\theta|\psi)$ so that it is close to the true posterior $p(\theta|x)$.
-This is slighty different from variational optimization (or variational learning), where the toal is to find $\theta^*$ such that $p(x|\theta^*) \approx p_r(x)$.
-Typically in variational inference the model $p(x|\theta)$ admits a density, so that the likelihood is known.
+This work has many connections to variational inference, in which the goal is to optimize the recognition model $q(\bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bftheta|x)$.
+This is slighty different from variational optimization (or variational learning), where the toal is to find $\bftheta^*$ such that $p(x|\bftheta^*) \approx p_r(x)$.
+Typically in variational inference the model $p(x|\bftheta)$ admits a density, so that the likelihood is known.
 Thus, assuming the prior also admits a density, it is tractable to evaluate the posterior is (up to a constant factor).
-In Adversarial Variational Bayes \citep{DBLP:journals/corr/MeschederNG17} and , equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H}), the authors use an adversarial setup, but assume a known density $p(x|\theta)$ that is differentiable with respect to $\theta$.
+In Adversarial Variational Bayes \citep{DBLP:journals/corr/MeschederNG17} and , equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H}), the authors use an adversarial setup, but assume a known density $p(x|\bftheta)$ that is differentiable with respect to $\bftheta$.
 
 In \citep{tran2017variational}, the authors consider Variational Bayes with an intractable likelihood.
-In that approach "The only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of most simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(x|\theta)$.
+In that approach "The only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of most simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(x|\bftheta)$.
 
 In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.
-The  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\theta$''.
+The  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''.
 
@@
 Third, we need to consider whether less noisy estimates of the gradients
-$\nabla_\psi U_g$ can be computed, in particular by exploiting the fact in the composition $d(g(\mathbf{z};\theta); \psi)$
+$\nabla_\bfpsi U_g$ can be computed, in particular by exploiting the fact in the composition $d(g(\bfz;\bftheta); \bfpsi)$
 the derivatives of $d$ are known exactly.

### 2017-07-11 tex/paper.tex
modified intro

@@
 experimental observations. Inference in this setup is often
-difficult, as simulators rarely admit a tractable likelihood
+difficult, as simulators rarely admit a tractable density or likelihood
 function. In this note, we develop Adversarial Variational Optimization (AVO), a likelihood-free
@@
 adversarial networks by replacing the differentiable generative network with a
-domain-specific simulator and solve the resulting non-differentiable
+domain-specific simulator. We solve the resulting non-differentiable
 minimax problem by minimizing variational upper bounds of the two adversarial objectives.
@@
 marginal distribution of the generated data matches the observations.
-We present results of the method with with non-differentiable simulators producing discrete/continuous univariate/multi-dimensional data
-validate the proposed method.
+We present results of the method with simulators producing both discrete and continuous data.
 
@@
 
-In many fields of science such as particle physics, climatology or population
-genetics, computer simulators are used to describe complex processes that tie
-parameters of an underlying theory to high dimensional observations. In most
-cases, these implicit generative models~\citep{2016arXiv161003483M} are
-specified as procedural implementations of forward stochastic processes that
-generate discrete or continuous data. Because it is usually computationally intractable, most simulators
-do not provide a way to directly evaluate the likelihood function for a
-given observation, thereby making inference difficult.
-In addition, most scientific simulators are written using opaque low-level programming
-languages, hence raising the bar for likelihood-free inference
-algorithms relying e.g. on the simulator being a function with computable derivatives~\citep{2016arXiv160507826G}
-or being a controllable probabilistic program~\citep{2016arXiv161009900L}.
+%Computer simulators are used to describe complex data generation processes in a diverse set of fields of science including particle physics, climatology, epidemiology,  and population
+%genetics,
+
+In many fields of science such as particle physics, epidemiology,  and population
+genetics, computer simulators are used to describe complex data generation processes. These simulators relate 
+observations $\bfx$ to the parameters $\bftheta$ of an underlying theory or mechanistic model. 
+In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
+% and may generate either discrete or continuous data. 
+Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx | \bftheta)$.
+The prevalence and significance of this problem has motivated an active research effort in so-called \textit{likelihood-free inference} algorithms such as Approximate Bayesian Computation (ABC)~\cite{abc, cranmer2015approximating}.
+Often the simulation involves control-flow that leads to models such that even $p(\bfx, \bfz | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.
+
+In parallel, with the introduction of variational autoencoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial networks~\cite{goodfellow2014generative}, there has been a vibrant research program around implicit generative models based on neural networks~\citep{2016arXiv161003483M}.
+While these neural generative models also do not admit a tractable density, they are differentiable.
+
+%%In contrast, scientific simulators involve control-flow that leads to models such that $p(\bfx, \bfz | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.
+%Scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters  endowed with some level of interpretation.
+In contrast, generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation.
+In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.
+
+Generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation.
+In contrast, scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters that are endowed with some level of interpretation.
+In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.
+
+%While generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation, scientific simulators typically have relatively few parameters that endowed with some level of interpretation.
+Thus simulators can be thought of as highly regularized, non-differentiable generative models.
+In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latents $\bfz$.
+
+%Generative models based on neural networks are differentiable with respect to their high dimensional model parameters $\bftheta$, which  with no particular , differentiable with respect to $\bfz$ and $\bftheta$, and the distribution of the latent $\bfz$
+%
+%In the context of variational inference, one assumes the the likelihood $p(\bfx | \bftheta)$ is available and posterior $p(\bftheta | \bfx)$ is tractable up to a constant (ie.
+($p(\bfx, \bfz | \bftheta)$ is tractable).
+Moreover, one is typically interested in estimating the posterior distribution of the latent variables instead of the parameters of the simulator $\bftheta$.
+
+%In the context of variational inference, these technihave also been used as the recognition model ~\citep{2015arXiv150505770J, DBLP:journals/corr/KingmaSW16, DBLP:conf/icml/RezendeMW14}; however, in those cases the likelihood $p(\bfx | \bftheta)$ is available and posterior $p(\bftheta | \bfx)$ is tractable up to a constant.
+
+%Because it is usually computationally intractable, most simulators
+%
+%implicit generative models~\citep{2016arXiv161003483M} 
+%
+%In most
+%cases, these implicit generative models~\citep{2016arXiv161003483M} are
+%specified as procedural implementations of forward stochastic processes that
+%generates discrete or continuous data, and rarely do these simulators admit a tractable density. Because it is usually computationally intractable, most simulators
+%do not provide a way to directly evaluate the likelihood function for a
+%given observation, thereby making inference difficult.
+%In addition, most scientific simulators are written using opaque low-level programming
+%languages, hence raising the bar for likelihood-free inference
+%algorithms relying e.g. on the simulator being a function with computable derivatives~\citep{2016arXiv160507826G}
+%or being a controllable probabilistic program~\citep{2016arXiv161009900L}.
 
 In this note, we develop a likelihood-free inference algorithm for the point
-estimation of the parameters of a forward non-differentiable
+estimation of the parameters $\bftheta$ of a non-differentiable, implicit
 generative model. We adapt the adversarial
@@
 simulator parameters, hence producing an arbitrarily tight family of models whose
-joint collection of generated samples matches the observed data.
+marginal distribution of generated data matches the observed data.
 

### 2017-07-11 tex/paper.tex
rewritten problem statement

@@
 The prevalence and significance of this problem has motivated an active research effort in so-called \textit{likelihood-free inference} algorithms such as Approximate Bayesian Computation (ABC)~\cite{abc, cranmer2015approximating}.
-Often the simulation involves control-flow that leads to models such that even $p(\bfx, \bfz | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.
+Often the simulation code involves control-flow that implies the dependence of $\bfx$ on $\bfz$ is non-differentiable, making inference even more difficult and precludes approaches such as Ref.~\citep{2016arXiv160507826G}.
+
+%Often the simulation code involves control-flow that leads to models such that even $p(\bfx | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.
 
 In parallel, with the introduction of variational autoencoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial networks~\cite{goodfellow2014generative}, there has been a vibrant research program around implicit generative models based on neural networks~\citep{2016arXiv161003483M}.
-While these neural generative models also do not admit a tractable density, they are differentiable.
+While these implicit generative networks also do not admit a tractable density, they are differentiable.
 
@@
 In contrast, generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation.
-In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.
-
-Generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation.
-In contrast, scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters that are endowed with some level of interpretation.
 In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.
@@
 
-In this note, we develop a likelihood-free inference algorithm for the point
+Generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation.
+In contrast, scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters that are endowed with some level of interpretation.
+In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.
+
+In this note, we develop an unsupervised learning algorithm for the point
 estimation of the parameters $\bftheta$ of a non-differentiable, implicit
@@
 bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
-objectives. The procedure results in learning a proposal distribution over
-simulator parameters, hence producing an arbitrarily tight family of models whose
-marginal distribution of generated data matches the observed data.
+objectives. The procedure results in learning an arbitrarily tight
+proposal distribution over simulator parameters, such that the corresponding
+marginal distribution of the generated data matches the observations.
+
+%In this note, we develop a likelihood-free inference algorithm for the point
+%estimation of the parameters $\bftheta$ of a non-differentiable, implicit
+%generative model. We adapt the adversarial
+%training procedure of generative adversarial
+%networks~\cite{goodfellow2014generative} by replacing the implicit generative
+%network with a domain-based scientific simulator, and solve the resulting
+%non-differentiable minimax problem by minimizing variational upper
+%bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
+%objectives. The procedure results in learning a proposal distribution over
+%simulator parameters, hence producing an arbitrarily tight family of models whose
+%marginal distribution of generated data matches the observed data.
 
@@
 parameters of interest. The simulation may involve some complicated latent
-process, such that
-\begin{equation}\label{eqn:p_x}
-    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
-\end{equation}
+process
+%, such that
+%\begin{equation}\label{eqn:p_x}
+%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
+%\end{equation}
+%\begin{equation}\label{eqn:p_x_sim}
+%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
+%\end{equation}
 where $\bfz \in {\cal Z}$ is a latent variable providing an external
-source of randomness. In particular, $\bfz$ is not necessarily assumed to
-be a fixed-size vector (e.g., it can be a sequence of variable length) and its
-distribution $p(\bfz|\bftheta)$ may itself depend on $\bftheta$ in some intricate way.
+source of randomness. 
+Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
+be a fixed-size vector with a simple density.
+Instead, the dimensionality of $z$ and the nature of it's components (uniform, normal, discrete, continuous, etc.) are inherited from the control flow of the simulation code and may depend on $\bftheta$ in some intricate way.
+Moreover, the dimensionality of $\bfz$ may be much larger than the dimensionality of $\bfx$.
 
-We assume that we already have an accurate simulation of the stochastic
-generative process that defines $p(\mathbf{x}|\bfz,\bftheta)$, as
-specified through a highly regularized deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
+%Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
+%be a fixed-size vector (e.g., it can be a sequence of variable length) and its
+%distribution $p(\bfz|\bftheta)$ may itself depend on $\bftheta$ in some intricate way
+%inherited from the control flow of the simulation code.
+
+We assume that the stochastic generative process that defines $p(\mathbf{x}|\bfz,\bftheta)$ is
+specified through a deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
 \mathbb{R}^d$ with usually few parameters. That is, we consider
@@
 such that the likelihood $p(\mathbf{x}|\bftheta)$ can be 
-considered as a non-differentiable reparameterization of the random variable $\bfz$.
-%rewritten as
-%\begin{equation}\label{eqn:p_x_sim}
-%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
-%\end{equation}
-%where $\mu$ is a probability measure.
-Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
-used to generate data in forward mode. For this reason, evaluating the integral
-in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
-in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
-e.g. as when $g$ is specified as a computer program.
+considered as a non-differentiable reparameterization of the random variable $\bfz$
+\begin{equation}\label{eqn:p_x_sim}
+    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
+\end{equation}
+where $\mu$ is a probability measure.
+%Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
+%used to generate data in forward mode. For this reason, evaluating the integral
+%in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
+%in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
+%e.g. as when $g$ is specified as a computer program.
 

### 2017-07-11 tex/paper.tex
reworded problem statement

@@
 
-We assume that the stochastic generative process that defines $p(\mathbf{x}|\bfz,\bftheta)$ is
-specified through a deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
-\mathbb{R}^d$ with usually few parameters. That is, we consider
+We assume that the stochastic generative process that defines $p(\mathbf{x}|\bftheta)$ is
+specified through a non-differentiable deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
+\mathbb{R}^d$. Operationally, %That is, we consider
 \begin{equation}\label{eqn:p_theta}
@@
 \end{equation}
-such that the likelihood $p(\mathbf{x}|\bftheta)$ can be 
-considered as a non-differentiable reparameterization of the random variable $\bfz$
+such that the density $p(\mathbf{x}|\bftheta)$ can be 
+rewritten as
+%considered as a non-differentiable reparameterization of the random variable $\bfz$
+%\begin{equation}\label{eqn:p_x_sim}
+%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g_i(\bfz;\bftheta) \leq x_i\}} p(\bfz|\bftheta) \mu(d\bfz),
+%\end{equation}
 \begin{equation}\label{eqn:p_x_sim}
-    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
+    p(\mathbf{x}|\bftheta) = \int_{\{\bfz:g(\bfz;\bftheta) = \bfx \}} p(\bfz|\bftheta) \mu(d\bfz),
 \end{equation}
@@
 Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
-(unknown) true distribution $p_r(\mathbf{x})$, our goal is the inference of the parameters
-of interest $\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
-the modeled data distribution $p(\mathbf{x}|\bftheta)$ induced by $g(\cdot;
-\bftheta)$ over $\bfz$. That is,
+(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters 
+$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
+the implicit model $p(\mathbf{x}|\bftheta)$. That is,
 \begin{equation}

### 2017-07-11 tex/paper.tex
ABC and HABC notes
ABC citations

@@
 Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx | \bftheta)$.
-The prevalence and significance of this problem has motivated an active research effort in so-called \textit{likelihood-free inference} algorithms such as Approximate Bayesian Computation (ABC)~\cite{abc, cranmer2015approximating}.
+The prevalence and significance of this problem has motivated an active research effort in so-called \textit{likelihood-free inference} algorithms such as Approximate Bayesian Computation (ABC)~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential, sisson2011likelihood, marin2012approximate, cranmer2015approximating}.
 Often the simulation code involves control-flow that implies the dependence of $\bfx$ on $\bfz$ is non-differentiable, making inference even more difficult and precludes approaches such as Ref.~\citep{2016arXiv160507826G}.
+
+%\kcnote{HMC https://arxiv.org/pdf/1503.01916.pdf}
+
+%\kcnote{add ABC refs mentioned in tex}
+%(Beaumont et al., 2002; Marjoram et al., 2003; Sisson et al., 2007; Sisson \& Fan, 2010; Marin et al., 2012; Fan et al., 2013) }
 
@@
 
+\citep{meeds2015hamiltonian}
+
 In this context, the proposed method can be considered as a direct adaptation of
@@
 
-\bibliographystyle{acm}
+%\bibliographystyle{hacm}
+\bibliographystyle{hunsrt}
 \bibliography{bibliography.bib}

### 2017-07-11 tex/paper.tex
connection to variational inference via posterior predictive distilled

@@
 \newcommand{\bfpsi}{{\bm \psi}}
+\newcommand{\bfphi}{{\bm \phi}}
 \newcommand{\bfx}{\mathbf{x}}
@@
 \end{equation}
-where $\rho$ is some distance or divergence.
+where $\rho$ is some distance or divergence. For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate.
 
@@
 a generative model $g(\cdot; \bftheta)$ is pit against an adversarial
-classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
+classifier $d(\cdot; \bfphi):\mathbb{R}^d \to [0,1]$ with parameters $\bfphi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
 from generated data $\tilde{\mathbf{x}} = g(\bfz; \bftheta)$. Both models $g$ and $d$
@@
 replacing the adversarial classifier with a 1-Lipschitz adversarial critic
-$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. 
+$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$. 
 \begin{align}
-   W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
+W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) = \sup_{d} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [d(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [d(\mathbf{x})]
 \end{align}
@@
 for stabilizing the optimization procedure,
-training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\bftheta$ in order to respectively minimize
-
+training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
 \begin{align}
-    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
-    {\cal L}_g =\,& - {\cal L}_W
+    {\cal L}_d =\,&  W(p_r, p_\theta) + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
+    {\cal L}_g =\,& - W(p_r, p_\theta)
 \end{align}
+%
 %\begin{align}
-%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
-%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
-%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\phi)]
+%    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
+%    {\cal L}_g =\,& - {\cal L}_W
+%\end{align}
+%\begin{align}
+%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]  \nonumber \\
+%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
+%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)]
 %\end{align}
@@
 \begin{equation}
-    \min_{\bftheta \in \bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
+    \min_{\bftheta \in {\Theta}} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
 \end{equation}
@@
 to any of their average. Provided that the proposal is flexible enough, the parameters $\bfpsi$
-can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \bftheta} f(\bftheta)$.
+can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.
 
@@
 $U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
-trick it comes:
+trick it can be rewritten as:
 \begin{align}\label{eqn:approx-grad}
@@
     %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
-    &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
-    &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
+ %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
+   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
     &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
@@
 % \begin{equation}
-%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\phi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
+%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\bfphi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
 % \end{equation}
@@
                     \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
-                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
+                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \bfphi) - d(\mathbf{x}_m; \bfphi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \bfphi) ||_2 - 1)^2 $}
                 \EndFor
-                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
+                \State{$\bfphi \leftarrow \text{Adam}(\nabla_\bfphi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
             \EndFor
@@
 GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
-the setting where we are interested in the inference of the parameters of a
+the setting where we are interested in estimating the parameters of a
 fixed non-differentiable simulator (Section~\ref{sec:problem}),
 rather than in learning the generative model itself,
-gradients $\nabla_\bftheta g$ either do not exist or cannot be accessed. As a
+gradients $\nabla_\bftheta g$ either do not exist or are inaccessible. As a
 result, gradients $\nabla_\bftheta {\cal L}_g$ cannot be constructed and the
@@
 specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
-parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and minimize in alternation the variational upper bounds
+parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and alternately minimize the variational upper bounds
 \begin{align}
@@
     U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
-\end{align} respectively over $\phi$ and $\bfpsi$.
+\end{align} respectively over $\bfphi$ and $\bfpsi$.
 When updating
-$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
-evaluating the exact and known gradient of $U_d$ over mini-batches of true and
-generated data, as ordinarily done in stochastic gradient descent. When updating
-$\bfpsi$, estimates of $\nabla_\bfpsi U_g$ can be derived with forward
-simulations, as described in the previous section.
+$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
+directly evaluating the gradient of $U_d$ over mini-batches of real and
+generated data, as is ordinarily done in stochastic gradient descent. When updating
+$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
 That is,
+%\begin{equation}\label{eqn:grad-ug-approx}
+%   \nabla_\bfpsi U_g = \mathbb{E}_{\subarray{l}{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta) }}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
+%\end{equation} 
 \begin{equation}\label{eqn:grad-ug-approx}
-    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\phi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
-\end{equation} which we can approximate with mini-batches of
+   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
+\end{equation} 
+%\begin{equation}\label{eqn:grad-ug-approx}
+%    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
+%\end{equation} 
+which we can approximate with mini-batches of
 generated data
 \begin{equation}
-    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \phi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
+    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
 \end{equation}
@@
 
-Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
+%Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
+%have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
+%a distribution parameterized in terms of $\bfpsi$:
+%\begin{equation}\label{eqn:p_psi}
+%    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
+%\end{equation}
+%Intuitively, this corresponds to a family of simulators, each configured
+%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
+%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
+%More formally, the learned model  therefore corresponds to the marginal distribution
+%of the generated data. That is,
+%\begin{equation}
+%    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
+%\end{equation}
+%Intuitively, this corresponds to a family of simulators, each configured
+%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
+%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
+
+The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
 have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
-a distribution parameterized in terms of $\bfpsi$:
+the marginal distribution parameterized by $\bfpsi$
+\begin{equation}
+    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
+\end{equation}
+Operationally, we sample from this marginal model via
 \begin{equation}\label{eqn:p_psi}
@@
 \end{equation}
-Intuitively, this corresponds to a family of simulators, each configured
-with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
-of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
-More formally, the learned model  therefore corresponds to the marginal distribution
-of the generated data. That is,
-\begin{equation}
-    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
-\end{equation}
+Effectively, we are optimizing a variational posterior $q(\bftheta | \psi)$ by minimizing the 
+divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
+%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
+Consequently, the proposed algorithm is a form of variational Bayesian inference through model criticism~\citep{box1980sampling}.
 
-In consequence, the proposed inference algorithm does not necessarily guarantee that the
-proposal distribution $q(\bftheta|\bfpsi)$ will place its mass arbitrarily tight
-around the parameters of interest, which might be an issue when one is rather interested in point estimates $\bftheta^*$.
-For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
+%Consequently, the proposed inference algorithm does not necessarily guarantee that the
+%proposal distribution $q(\bftheta|\bfpsi)$ will place its mass arbitrarily tight
+%around the parameters of interest, which might be an issue when one is rather interested in point estimates $\bftheta^*$.
+%For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
+
+If instead, one is interested in point estimates $\bftheta^*$, then we augment Eqn.~\ref{eqn:vo-ug}
 with a regularization term corresponding to the differential entropy $H$ of
@@
 For small values of $\gamma$,
-proposal distributions with large entropy are not penalized, which may result
-in learning a smeared variation of the original simulator. On the other hand,
-for large values of $\gamma$, the procedure is constrained to fit a proposal
+proposal distributions with large entropy are not penalized, which results
+in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
+for large values of $\gamma$, the procedure is encouraged to fit a proposal
 distribution with low entropy, which has the effect of concentrating its density
-tightly around one or a few $\bftheta$ values. A too large penalty may however
+tightly around one or a few $\bftheta$ values. Very large penalties may 
 eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
-typically increases as the entropy of the proposal decreases.
-
-\subsection{Operator Variational Optimization}
-
-Original OPVI  \citep{2016arXiv161009033R}
-
-\begin{eqnarray}
-\lambda^* = \inf_\lambda \sup_\phi \mathbb{E}_{ z \sim q( z | \lambda)} [ (O^{p,q} f_\lambda) ] 
-\end{eqnarray}
-with
-\begin{equation}
-(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
-\end{equation}
-
-%This is natural here, but not quite OPVI formalism:
-%\begin{equation}
-%\bfpsi^* = \inf_\bfpsi \sup_\phi \mathbb{E}_{ \bftheta \sim q(\bftheta | \bfpsi)} [ (O^{p,q} f_\phi) ]
-%\end{equation}
-%
-%\begin{equation}
-%(O^{p,q} f) = W(p,q) = \sup_f \mathbb{E}_{x\sim p}[f(x)] - \mathbb{E}_{x \sim q}[f(x)]
-%\end{equation}
-%The expectation should be $x \sim p(x|\bfpsi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$
-
-To get into OPVI formalism
-
-\begin{eqnarray}
-\bfpsi^* &=& \inf_\bfpsi \sup_\phi\mathbb{E}_{x\sim p_r(x)}[f(x)] - \mathbb{E}_{x \sim p(x|\bfpsi)}[f(x)] \\
-&=&  \inf_\bfpsi \sup_\phi \mathbb{E}_{ x \sim p( x | \bfpsi)} [ (O^{p_r,p_\bfpsi} f_\phi) ] 
-\end{eqnarray}
-with
-\begin{equation}
-(O^{p,q} f) = \left( \frac{p_r(x)}{p(x|\bfpsi)}  -1 \right) f(x)
-\end{equation}
-
-We can think of $p(x|\bfpsi)$ as a \textit{variational program} as described in ~\citep{2016arXiv161009033R}, though more complicated than the a simple reparametrization of normally distributed noise $\epsilon$ through a differentiable function $R(\epsilon, \bfpsi)$.
-In our case, the variational program is a marginalized non-differentiable simulator simulator.
-Nevertheless, it can generate samples for $x$, is differentiable with respect to $\phi$, and its density is be intractable.
+typically increases as the entropy of the proposal decreases. \kcnote{maybe, but not obvious why}
 
@@
 $\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
-[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
-p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
+[d(\tilde{\mathbf{x}};\bfphi)] \approx \mathbb{E}_{\mathbf{x} \sim
+p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]$ and that the critic cannot distinguish
 between true and model data. Despite the discreteness and the
@@
 This work has many connections to variational inference, in which the goal is to optimize the recognition model $q(\bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bftheta|x)$.
-This is slighty different from variational optimization (or variational learning), where the toal is to find $\bftheta^*$ such that $p(x|\bftheta^*) \approx p_r(x)$.
-Typically in variational inference the model $p(x|\bftheta)$ admits a density, so that the likelihood is known.
+This is slighty different from variational optimization (or variational learning), where the toal is to find $\bftheta^*$ such that $p(\bfx|\bftheta^*) \approx p_r(\bfx)$.
+Typically in variational inference the model $p(\bfx|\bftheta)$ admits a density, so that the likelihood is known.
 Thus, assuming the prior also admits a density, it is tractable to evaluate the posterior is (up to a constant factor).
-In Adversarial Variational Bayes \citep{DBLP:journals/corr/MeschederNG17} and , equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H}), the authors use an adversarial setup, but assume a known density $p(x|\bftheta)$ that is differentiable with respect to $\bftheta$.
+In Adversarial Variational Bayes \citep{DBLP:journals/corr/MeschederNG17} and , equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H}), the authors use an adversarial setup, but assume a known density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$.
 
 In \citep{tran2017variational}, the authors consider Variational Bayes with an intractable likelihood.
-In that approach "The only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of most simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(x|\bftheta)$.
+In that approach "The only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of most simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$.
 
@@
 The  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''.
+
+\subsection{Operator Variational Optimization}
+
+Original OPVI  \citep{2016arXiv161009033R}
+
+\begin{eqnarray}
+\lambda^* = \inf_\lambda \sup_\bfphi \mathbb{E}_{ z \sim q( z | \lambda)} [ (O^{p,q} f_\lambda) ] 
+\end{eqnarray}
+with
+\begin{equation}
+(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
+\end{equation}
+
+%This is natural here, but not quite OPVI formalism:
+%\begin{equation}
+%\bfpsi^* = \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bftheta \sim q(\bftheta | \bfpsi)} [ (O^{p,q} f_\bfphi) ]
+%\end{equation}
+%
+%\begin{equation}
+%(O^{p,q} f) = W(p,q) = \sup_f \mathbb{E}_{x\sim p}[f(x)] - \mathbb{E}_{x \sim q}[f(x)]
+%\end{equation}
+%The expectation should be $x \sim p(x|\bfpsi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$
+
+To get into OPVI formalism
+
+\begin{eqnarray}
+\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi\mathbb{E}_{x\sim p_r(x)}[f(x)] - \mathbb{E}_{x \sim p(x|\bfpsi)}[f(x)] \\
+&=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ x \sim p( x | \bfpsi)} [ (O^{p_r,p_\bfpsi} f_\bfphi) ] 
+\end{eqnarray}
+with
+\begin{equation}
+(O^{p,q} f) = \left( \frac{p_r(x)}{p(x|\bfpsi)}  -1 \right) f(x)
+\end{equation}
+
+We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in ~\citep{2016arXiv161009033R}, though more complicated than the a simple reparametrization of normally distributed noise $\epsilon$ through a differentiable function $R(\epsilon, \bfpsi)$.
+In our case, the variational program is a marginalized non-differentiable simulator simulator.
+Nevertheless, it can generate samples for $x$, is differentiable with respect to $\bfphi$, and its density is be intractable.
 

### 2017-07-11 tex/paper.tex
more on AVO and AVI-MC

@@
 \begin{equation}\label{eqn:grad-ug-approx}
-   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
+   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \tilde{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \tilde{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
 \end{equation} 
@@
 optimization (AVO) procedure, as built on top of WGAN-GP.
-Obviously, the variational relaxation could similarly be coupled with
-other variants of GANs and/or of evolution strategies.
+%
+%Obviously, the variational relaxation could similarly be coupled with
+%other variants of GANs and/or of evolution strategies.
 
@@
 
+Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems.
+Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.
+
+\subsection{Adversarial variational inference through model criticism}
+
 The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
@@
 \end{equation}
+We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in \citep{2016arXiv161009033R}, though more complicated than a simple reparameterization of normally distributed noise $\bfz$ through a differentiable function $g(\bfz, \bfpsi)$.
+In our case, the variational program is a marginalized, non-differentiable  simulator.
+Its density is  intractable; nevertheless, it can generate samples for $\bfx$ and is differentiable with respect to $\bfpsi$ and $\bfphi$.
 Operationally, we sample from this marginal model via
@@
 \end{equation}
-Effectively, we are optimizing a variational posterior $q(\bftheta | \psi)$ by minimizing the 
+
+%Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by directly minimizing the 
+%divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the unknown joint density $p(\bfx,\bftheta)$.
+
+Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by minimizing the 
 divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
 %Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
-Consequently, the proposed algorithm is a form of variational Bayesian inference through model criticism~\citep{box1980sampling}.
+Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.
 
@@
 
+\subsection{Adversarial variational learning}
+
 If instead, one is interested in point estimates $\bftheta^*$, then we augment Eqn.~\ref{eqn:vo-ug}
 with a regularization term corresponding to the differential entropy $H$ of
-the proposal distribution. That is,
+the proposal distribution $q(\bftheta | \bfpsi)$. That is,
 \begin{equation}
@@
 
+%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems.
+Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here analogous to their use in black box variational inference~\citep{ranganath2014black}.
+
+\kcnote{Discuss hierarchical models and iid data x. Factor of $N/M$ for AVI-MC?}
+
+\newpage
 % ==============================================================================
@@
 
-\textbf{KC: could also use this to tune non-differentialble fast simulator to slower full simulator. Then fixed data size not an issue}
+\kcnote{could also use this to tune non-differentialble fast simulator to slower full simulator. Then fixed data size not an issue}
 

### 2017-07-11 tex/paper.tex
Updated experiments section

@@
 
-%Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by directly minimizing the 
-%divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the unknown joint density $p(\bfx,\bftheta)$.
+Typically in variational inference one would has access to the posterior $p(\bfx | \bftheta)$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
+Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by directly minimizing the 
+divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the unknown posterior $p(\bfx | \bftheta)$ or joint density $p(\bfx,\bftheta)$.
+Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.
 
-Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by minimizing the 
-divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
-%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
-Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.
+%Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by minimizing the 
+%divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
+%%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
+%Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.
 
@@
 
-If instead, one is interested in point estimates $\bftheta^*$, then we augment Eqn.~\ref{eqn:vo-ug}
+In order to target point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug}
 with a regularization term corresponding to the differential entropy $H$ of
@@
 distribution with low entropy, which has the effect of concentrating its density
-tightly around one or a few $\bftheta$ values. Very large penalties may 
+tightly around one or a few $\bftheta$ values.
+
+Very large penalties may 
 eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
@@
 p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]$ and that the critic cannot distinguish
-between true and model data. Despite the discreteness and the
-non-differentiability of the underlying generator, this confirms that inference
-with adversarial variational optimization works.
+between true and model data. This confirms that adversarial variational optimization 
+works despite the discreteness of the data and lack of access to the 
+density $p(\bfx | \bftheta)$ or its gradient.
 
@@
 
-As a second toy example, we evaluate parameter inference for a generator producing
+As a second toy example, we consider a generator producing
 5-dimensional continuous data, as originally specified in Section 4.2. of
@@
 \begin{itemize}
-    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that
+    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
         $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
         $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
-        $z_2 \sim {\text{Mixture}}(\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
+        $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
         $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
@@
 \end{itemize}
+Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model
 We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
-The simulator parameters are modeled with a factored Gaussian
-proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component is
+The simulator parameters are modeled with a factored  (mean field) Gaussian
+variational distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
 initialized with zero mean and unit variance.
@@
 the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
-inference quickly converges towards a proposal distribution whose density
+AVO quickly converges towards a variational distribution whose density
 concentrates around the nominal values $\bftheta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
@@
 
-As a more challenging example, we now consider a (simplified) simulator from
+As a more realistic example, we now consider a (simplified) simulator from
 particle physics for electron--positron collisions resulting in muon--antimuon
@@
 polar angle of the outgoing muon with respect  to the originally incoming
-electron. This random variable is approximately distributed as
-\begin{equation}
+electron. Neglecting measurement uncertainty induced from the particle detectors, 
+this random variable is approximately distributed as
+\begin{equation}\label{eq:simple_weinberg}
     p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
@@
 size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
-and $G_f$ (the Fermi constant) through the coefficient function $c$. Given the
-known analytic form of the probability density function, a simulator for
-$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
-arbitrary number of random drawings may be required before passing the
-acceptance threshold, this illustrates how the latent variable
-$\bfz$ does not necessarily need to be fixed-sized vector, and how its
-density  $p(\bfz|\bftheta)$ may depend on $\bftheta$ in some complicated way
-(Section~\ref{sec:problem}).
+and $G_f$ (the Fermi constant) through the coefficient function $c$.
 
-In this example, we consider observed data generated with $\bftheta^* =
-(E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
-factored Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
+A typical physics simulator for this process includes a more precise treatment of the 
+quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering 
+using \texttt{MadGraph}~\citep{Alwall:2011uj},  ionization of matter in the 
+detector due to the passage of the out-going $\mu^+\mu^-$ particles using 
+\texttt{GEANT4}~\citep{Agostinelli:2002hh}, electronic noise and other details of the sensors 
+that measure the ionization signal, and the deterministic algorithms that 
+estimate the polar angle $A$ based on the sensor readouts. The simulation of 
+this process is highly non-trivial as is the space of latent variables $\mathcal{Z}$.
+
+%Given the
+%known analytic form of the probability density function, a simulator for
+%$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
+%arbitrary number of random drawings may be required before passing the
+%acceptance threshold, this illustrates how the latent variable
+%$\bfz$ does not necessarily need to be fixed-sized vector, and how its
+%density  $p(\bfz|\bftheta)$ may depend on $\bftheta$ in some complicated way
+%(Section~\ref{sec:problem}).
+
+In this example, we consider observed data generated with the simplified generator of Eq.~\ref{eq:simple_weinberg}
+using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
+factored (mean field) Gaussian variational distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
 q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
@@
 $n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
-$\beta_1=0.9$ and $\beta_2=0.999$. As for the first example, we compare
+$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
 entropy penalties $\gamma=0$ and $\gamma=5$.
@@
 landscape around $\bftheta^*$ or that the observed data can in this case also be
-reproduced with a smeared variation $\qxpsi$ of the simulator.
+reproduced with the posterior predictive distribution $\qxpsi$.
 Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
@@
 towards $0$, which indicates  convergence towards a distribution that the critic
-cannot distinguish from the true distribution. In summary, despite the intricate
-relation betwen the observations $\mathbf{x}$, the latent variable $\bfz$
-and the parameters $\bftheta$, this example demonstrates the efficacy
-of AVO for parameter inference on a black box
-simulator.
+cannot distinguish from the true distribution. 
+%In summary, despite the intricate
+%relation between the observations $\mathbf{x}$, the latent variable $\bfz$
+%and the parameters $\bftheta$, this example demonstrates the efficacy
+%of AVO for parameter inference on a black box
+%simulator.
 
@@
 
-\section{Related works}
+\section{Related work}
 
@@
 
+\kcnote{proposal distribution $\to$ variational distribution}
+
 % ==============================================================================

### 2017-07-12 tex/paper.tex
code to test various M values. Text updates

@@
 \begin{align}
-W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) = \sup_{d} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [d(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [d(\mathbf{x})]
+W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) =& \sup_{\bfphi} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [d(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [d(\mathbf{x})] \\
+\equiv& \sup_{\bfphi} \mathcal{L}_W
 \end{align}
@@
 \begin{align}
-    {\cal L}_d =\,&  W(p_r, p_\theta) + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
-    {\cal L}_g =\,& - W(p_r, p_\theta)
+    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
+    {\cal L}_g =\,& - \mathcal{L}_W
 \end{align}
@@
 estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
-$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, like
-Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
-algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
-robustness to noisy gradients).
+$U(\bfpsi)$ with stochastic gradient descent. We will use 
+Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to incorporate the Natural Evolution Strategy
+algorithm~\citep{2011arXiv1106.4487W} and variance reduction techniques~\citep{ranganath2014black}.
+%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems.
+Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.
 
@@
 
-Typically in variational inference one would has access to the posterior $p(\bfx | \bftheta)$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
+%Typically in variational inference one would has access to the posterior $p(\bftheta | \bfx )$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
+Typically in variational inference one starts with a tractable prior $p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the posterior $p(\bftheta | \bfx )$.
+After optimizing $q(\bftheta | \bfpsi)$ to approximate the posterior, one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
 Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by directly minimizing the 
-divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the unknown posterior $p(\bfx | \bftheta)$ or joint density $p(\bfx,\bftheta)$.
+divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
 Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.
+While the likelihood is intractable, it is fixed, thus one can view the optimization of $\bfpsi$ through the lens of emperical Bayes, though the prior family $p(\bftheta | \bfpsi)$ is not explicitly specified.
+
+%\kcnote{where is the prior? comment}
 
@@
 eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
-typically increases as the entropy of the proposal decreases. \kcnote{maybe, but not obvious why}
+typically increases as the entropy of the proposal decreases. 
+%\kcnote{maybe, but not obvious why}
 
@@
 
-\newpage
+%\newpage
 % ==============================================================================
@@
 
+This work sits at the intersection of several lines of work related to 
+likelihood-free inference, approximate Bayesian computation (ABC),
+implicit generative networks, and variational inference.  Our focus
+is on extending existing work to non-differentiable generative models
+that do not admit a tractable density.
+
+Viewed from the literature around implicit generative models based on neural networks, 
+the proposed method can be considered as a direct adaptation of
+generative adversarial networks~\citep{goodfellow2014generative,2017arXiv170104862A, 2017arXiv170107875A} to
+non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization}.
+
+From the point of view of likelihood-free inference, the closest to our work are procedures that rely on a classifier to estimate the
+discrepancy between the observed data and the model distributions. For example,
+\citep{gutmann2012noise} uses non linear logistic regression for fitting
+unnormalized differentiable statistical models. In the likelihood-free
+setup, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
+ratios through supervised classification, which can in turn be used for
+parameter inference in combination with a gradient-free optimization algorithm.
+Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
+a summary statistics for approximate Bayesian computation.
+
 As reviewed in \cite{2016arXiv161003483M}, likelihood-free inference is
@@
 
-Closest to our work are procedures that rely on a classifier to estimate the
-discrepancy between the true and the model distributions. For example,
-\citep{gutmann2012noise} uses non linear logistic regression for fitting
-unnormalized differentiable statistical models, while
-\citep{goodfellow2014generative} exploits an adversarial neural network for
-learning a differentiable implicit generative model. In the likelihood-free
-setup, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
-ratios through supervised classification, which can in turn be used for
-parameter inference in combination with a gradient-free optimization algorithm.
-Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
-a summary statistics for approximate Bayesian computation.
-
-\citep{meeds2015hamiltonian}
-
-In this context, the proposed method can be considered as a direct adaptation of
-generative adversarial networks~\citep{goodfellow2014generative} to
-non-differentiable simulators. It also constitutes an approximate gradient
+%In this context, the proposed method can be considered as a direct adaptation of
+%generative adversarial networks~\citep{goodfellow2014generative} to
+%non-differentiable simulators. 
+It also constitutes an approximate gradient
 descent alternative to likelihood-free inference based on bayesian optimization and approximated density
@@
 
-This work has many connections to variational inference, in which the goal is to optimize the recognition model $q(\bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bftheta|x)$.
-This is slighty different from variational optimization (or variational learning), where the toal is to find $\bftheta^*$ such that $p(\bfx|\bftheta^*) \approx p_r(\bfx)$.
+This work has many connections to work on variational inference, in which the goal is to optimize the recognition model $q(\bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bftheta|x)$.
+This is slightly different from variational optimization (or variational learning), where the goal is to find $\bftheta^*$ such that $p(\bfx|\bftheta^*) \approx p_r(\bfx)$.
 Typically in variational inference the model $p(\bfx|\bftheta)$ admits a density, so that the likelihood is known.
 Thus, assuming the prior also admits a density, it is tractable to evaluate the posterior is (up to a constant factor).
-In Adversarial Variational Bayes \citep{DBLP:journals/corr/MeschederNG17} and , equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H}), the authors use an adversarial setup, but assume a known density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$.
+In Adversarial Variational Bayes \citep{DBLP:journals/corr/MeschederNG17} and, equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H}), the authors use an adversarial setup, but assume a known density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$.
 
+There have been efforts to extend variational inference to intractable likelihoods; however, they require restrictive assumptions. 
 In \citep{tran2017variational}, the authors consider Variational Bayes with an intractable likelihood.
-In that approach "The only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of most simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$.
+In that approach "the only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of many simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$.
+In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.
+Furthermore, the  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''.
+Lastly, in \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients
 
-In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.
-The  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''.
+\kcnote{F.H. Implicit model is differentiable. Is what we are doing JC-Adv with our variational program}
 

### 2017-07-12 code/toy-poisson-Mtest.py

@@ -1,0 +1,237 @@
+import autograd as ag
+import autograd.numpy as np
+import matplotlib.pyplot as plt
+from scipy.stats import poisson
+
+from nn import glorot_uniform
+from nn import relu
+from nn import AdamOptimizer
+
+from proposals import make_gaussian_proposal
+from proposals import gaussian_draw
+from proposals import gaussian_logpdf
+from proposals import grad_gaussian_logpdf
+from proposals import grad_gaussian_entropy
+
+from sklearn.utils import check_random_state
+
+# Global params
+
+seed = 777
+rng = check_random_state(seed)
+
+def_batch_size = 256
+n_epochs = 300+1
+n_epochs = 100+1
+lambda_gp = 0.025
+
+true_theta = np.array([np.log(7)])
+make_plots = True
+
+
+# Simulator
+
+def simulator(theta, n_samples, random_state=None):
+    return poisson.rvs(np.exp(theta[0]),
+                       size=n_samples,
+                       random_state=random_state).reshape(-1, 1)
+
+X_obs = simulator(true_theta, 20000, random_state=rng)
+n_params = len(true_theta)
+n_features = X_obs.shape[1]
+
+
+# Critic
+
+#gammas = [0.0, 5.0]
+gammas = [0.0, 0.0] #, 0.0, 0.0]
+colors = ["C1", "C2", "C3","C4"]
+batch_sizes = [16, 64, 256, 1024]
+
+def make_critic(n_features, n_hidden, random_state=None):
+    rng = check_random_state(random_state)
+    params = {"W": [glorot_uniform(n_hidden, n_features, rng),
+                    glorot_uniform(n_hidden, n_hidden, rng),
+                    glorot_uniform(n_hidden, 0, rng)],
+              "b": [np.zeros(n_hidden),
+                    np.zeros(n_hidden),
+                    np.zeros(1)]}
+    return params
+
+def predict(X, params):
+    h = X
+    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
+    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.1)
+    h = np.dot(params["W"][2], h.T).T + params["b"][2]
+    return h
+
+grad_predict_critic = ag.elementwise_grad(predict)
+
+
+history = [{"gamma": gammas[i],
+            "color": colors[i],
+            "loss_d": [],
+            "batch_size" : batch_sizes[i],
[…]

### 2017-07-12 tex/paper.tex
stopping for dinner.

@@
 Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems.
-Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.
+Variance reduction techniques such as Rao-Blackwellization, Control Variates, and quasi-Monte Carlo can be applied here as they are in black box variational inference~\citep{ranganath2014black}.
 
@@
 
-This work sits at the intersection of several lines of work related to 
+This work sits at the intersection of several lines of research related to 
 likelihood-free inference, approximate Bayesian computation (ABC),
-implicit generative networks, and variational inference.  Our focus
-is on extending existing work to non-differentiable generative models
-that do not admit a tractable density.
-
+implicit generative networks, and variational inference.  
+%Our focus
+%is on extending existing work to non-differentiable generative models
+%that do not admit a tractable density.
 Viewed from the literature around implicit generative models based on neural networks, 
@@
 
-From the point of view of likelihood-free inference, the closest to our work are procedures that rely on a classifier to estimate the
+ALI~\citep{dumoulin2016adversarially}, BiGANs ~\citep{donahue2016adversarial}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and, equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.
+The BiGAN, AVB, and PC-Adv setups are quite similar, but those models assume a tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$, which are not satisfied in the likelihood-free setting.
+Our lifting of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational program $q(\bfx | \bftheta)$ provides the ability to differentiate expectations with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density $q(\bfx | \bftheta)$ is still intractable.
+Moreover, the latent space $\mathcal{Z}$ of many real-world simulators is complicated and not amenable to neural a recognition model.
+One similarity to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) ?
+\log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.
+
+From the point of view of likelihood-free inference, where intractable non-differentiable simulators are the norm, our contributions are four fold.
+First is the process of lifting expectations with respect to the non-differentiable simulator $\mathbb{E}_{\tilde{\bfx} \sim p(\bfx | \bftheta)}$ to a differentiable expectations with respect to the variational program $\mathbb{E}_{\tilde{\bfx} \sim q(\bfx | \bfpsi)}$.
+Secondly, is the introduction of a form of variational inference that works in a likelihood-free setting.
+Thirdly, the AVI-MC algorithm can be viewed as a form of empirical Bayes where the prior is optimized based on the data.
+Fourthly, the AVL algorithm is not a Bayesian inference algorithm, but instead provides point estimates $\bftheta^*$.
+
+Perhaps the closest to our work is~\cite{bernton2017inference}, which uses Wasserstein distance for a rejection sampler in an ABC-like setup (as opposed to variational inference).
+They emphasize that this approach allows one to eliminate the summary statistics typically used in ABC and calculate the Wasserstein distance explicitly without use Kantorovich-Rubinstein duality and the critic $d$.
+For high-dimensional data, they note that this is computationally expensive and introduce an approximation based on projection onto Hilbert space filling curves.
+Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an approximation of the gradient of an entropy-regularized Wasserstein distance, which uses a similar duality.
+They note that ``Unfortunately, it is not applicable in the setting of purely generative models, as it involves point-wise evaluations of the derivative of the log-likelihood.'' \kcnote{Can we evaluate their Eq.
+6 with our Eq.
+9} Thus, our contribution is to provide gradients of an approximate MEWE by taking expectations with the variational program $q(\bfx | \bf\theta)$.
+This paired with the critic Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the critic $d$ allows us to work in high dimensions without summary statistics and optimize $\bfpsi$ with stochastic gradient descent.
+
+\kcnote{Could we actually replace the optimization of the critic with a direct Wasserstein distance calculation in fewer dimensions.
+If the data is drawn from the variational program, it would still be differentiable with respect to $\bfpsi$}
+
+Within the literature on ABC and likelihood-free inference, 
+the closest to our work are procedures that rely on a classifier to estimate the
 discrepancy between the observed data and the model distributions. For example,
 \citep{gutmann2012noise} uses non linear logistic regression for fitting
-unnormalized differentiable statistical models. In the likelihood-free
-setup, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
+unnormalized differentiable statistical models. Similarly, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
 ratios through supervised classification, which can in turn be used for
-parameter inference in combination with a gradient-free optimization algorithm.
-Similarly, \citep{gutmann2017likelihood} makes use of classification accuracy as
+ Bayesian or frequentist inference. Similarly, \citep{gutmann2017likelihood} 
+ makes use of classification accuracy as
 a summary statistics for approximate Bayesian computation.
@@
 In that approach "the only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of many simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$.
+\kcnote{They suggest to use ABC likelihood... is this only unbiased as $\epsilon \to 0$?}
+
  In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.

### 2017-07-12 code/toy-poisson-Mtest.py

@@ -22,7 +22,7 @@ rng = check_random_state(seed)
 
 def_batch_size = 256
 n_epochs = 300+1
-n_epochs = 100+1
+#n_epochs = 100+1
 lambda_gp = 0.025
 
 true_theta = np.array([np.log(7)])
@@ -44,7 +44,7 @@ n_features = X_obs.shape[1]
 # Critic
 
 #gammas = [0.0, 5.0]
-gammas = [0.0, 0.0] #, 0.0, 0.0]
+gammas = [0.0, 0.0, 0.0, 0.0]
 colors = ["C1", "C2", "C3","C4"]
 batch_sizes = [16, 64, 256, 1024]
 

### 2017-07-13 tex/paper.tex
updated related work, various notes.

@@
 
+\kcnote{connect to MWE and MEWE here?}
+
 In order to target point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug}
@@
 
-ALI~\citep{dumoulin2016adversarially}, BiGANs ~\citep{donahue2016adversarial}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and, equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.
-The BiGAN, AVB, and PC-Adv setups are quite similar, but those models assume a tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$, which are not satisfied in the likelihood-free setting.
-Our lifting of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational program $q(\bfx | \bftheta)$ provides the ability to differentiate expectations with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density $q(\bfx | \bftheta)$ is still intractable.
-Moreover, the latent space $\mathcal{Z}$ of many real-world simulators is complicated and not amenable to neural a recognition model.
-One similarity to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) ?
+Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially}, Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.
+Each of these assume a tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$, which is  not satisfied in the likelihood-free setting.
+Our lifting of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density $q(\bfx | \bfpsi)$ is still intractable.
+Moreover, we do not attempt to define a recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$ of many real-world simulators is complicated and not amenable to neural a recognition model.
+One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) ?
 \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.
 
-From the point of view of likelihood-free inference, where intractable non-differentiable simulators are the norm, our contributions are four fold.
+\kcnote{does \citep{rosca2017variational} go in the above paragraph?}
+
+From the point of view of likelihood-free inference, where  non-differentiable simulators are the norm, our contributions are threefold.
 First is the process of lifting expectations with respect to the non-differentiable simulator $\mathbb{E}_{\tilde{\bfx} \sim p(\bfx | \bftheta)}$ to a differentiable expectations with respect to the variational program $\mathbb{E}_{\tilde{\bfx} \sim q(\bfx | \bfpsi)}$.
-Secondly, is the introduction of a form of variational inference that works in a likelihood-free setting.
+Secondly, is the introduction of a novel form of variational inference that works in a likelihood-free setting.
 Thirdly, the AVI-MC algorithm can be viewed as a form of empirical Bayes where the prior is optimized based on the data.
-Fourthly, the AVL algorithm is not a Bayesian inference algorithm, but instead provides point estimates $\bftheta^*$.
 
-Perhaps the closest to our work is~\cite{bernton2017inference}, which uses Wasserstein distance for a rejection sampler in an ABC-like setup (as opposed to variational inference).
+%Fourthly, the AVL algorithm is not a Bayesian inference algorithm, but instead provides point estimates $\bftheta^*$.
+
+Perhaps the closest to our work is~\cite{bernton2017inference}, which uses Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of a rejection sampler in an ABC-like setup (as opposed to variational inference).
 They emphasize that this approach allows one to eliminate the summary statistics typically used in ABC and calculate the Wasserstein distance explicitly without use Kantorovich-Rubinstein duality and the critic $d$.
-For high-dimensional data, they note that this is computationally expensive and introduce an approximation based on projection onto Hilbert space filling curves.
+For high-dimensional data, they note that this is computationally expensive and introduce an approximation based on projection of the data onto Hilbert space-filling curves.
 Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an approximation of the gradient of an entropy-regularized Wasserstein distance, which uses a similar duality.
@@
 
-\kcnote{Could we actually replace the optimization of the critic with a direct Wasserstein distance calculation in fewer dimensions.
-If the data is drawn from the variational program, it would still be differentiable with respect to $\bfpsi$}
+%\kcnote{Could we actually replace the optimization of the critic with a direct Wasserstein distance calculation in fewer dimensions.
+If the data is drawn from the variational program, it would still be differentiable with respect to $\bfpsi$.
+No, because it still requires $\nabla_\psi \log p(\bfx | \bfpsi)$}
 
-Within the literature on ABC and likelihood-free inference, 
-the closest to our work are procedures that rely on a classifier to estimate the
-discrepancy between the observed data and the model distributions. For example,
-\citep{gutmann2012noise} uses non linear logistic regression for fitting
-unnormalized differentiable statistical models. Similarly, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
-ratios through supervised classification, which can in turn be used for
- Bayesian or frequentist inference. Similarly, \citep{gutmann2017likelihood} 
- makes use of classification accuracy as
-a summary statistics for approximate Bayesian computation.
+%Within the literature on ABC and likelihood-free inference, 
+%the closest to our work are procedures that rely on a classifier to estimate the
+%discrepancy between the observed data and the model distributions. For example,
+%\citep{gutmann2012noise} uses non linear logistic regression for fitting
+%unnormalized differentiable statistical models. Similarly, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
+%ratios through supervised classification, which can in turn be used for
+% Bayesian or frequentist inference. Similarly, \citep{gutmann2017likelihood} 
+% makes use of classification accuracy as
+%a summary statistics for approximate Bayesian computation.
 
@@
 first compared to the true data distribution and then updated to make it more
-comparable to the latter.
+comparable to the latter. Relevant work in this direction includes \citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood}.
 
-%In this context, the proposed method can be considered as a direct adaptation of
-%generative adversarial networks~\citep{goodfellow2014generative} to
-%non-differentiable simulators. 
-It also constitutes an approximate gradient
-descent alternative to likelihood-free inference based on bayesian optimization and approximated density
-ratios~\citep{cranmer2015approximating,cranmer2016experiments} or to classifier
-ABC~\citep{gutmann2017likelihood}.
+%%In this context, the proposed method can be considered as a direct adaptation of
+%%generative adversarial networks~\citep{goodfellow2014generative} to
+%%non-differentiable simulators. 
+%It also constitutes an approximate gradient
+%descent alternative to likelihood-free inference based on bayesian optimization and approximated density
+%ratios~\citep{cranmer2015approximating,cranmer2016experiments} or to classifier
+%ABC~\citep{gutmann2017likelihood}.
 
-This work has many connections to work on variational inference, in which the goal is to optimize the recognition model $q(\bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bftheta|x)$.
-This is slightly different from variational optimization (or variational learning), where the goal is to find $\bftheta^*$ such that $p(\bfx|\bftheta^*) \approx p_r(\bfx)$.
+%This work has many connections to work on variational inference, in which the goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bfz, \bftheta |x)$.
+%This is slightly different from variational optimization (or variational learning), where the goal is to find $\bftheta^*$ such that $p(\bfx|\bftheta^*) \approx p_r(\bfx)$.
 Typically in variational inference the model $p(\bfx|\bftheta)$ admits a density, so that the likelihood is known.
@@
 
+This work has many connections to work on variational inference, in which the goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bfz, \bftheta |x)$.
 There have been efforts to extend variational inference to intractable likelihoods; however, they require restrictive assumptions.
-In \citep{tran2017variational}, the authors consider Variational Bayes with an intractable likelihood.
-In that approach "the only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of many simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$.
-\kcnote{They suggest to use ABC likelihood... is this only unbiased as $\epsilon \to 0$?}
+In \citep{tran2017variational}, the authors consider Variational Bayes with an Intractable Likelihood (VBIL).
+In that approach "the only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of simulators, they propose to use the ABC-likelihood with an $\epsilon$ kernel.
+\kcnote{Is this an unbiased estimator if $\epsilon > 0$?} This approach
+This approach 
+n the case of many simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$.
+\kcnote{They suggest to use ABC likelihood...
+is this only unbiased as $\epsilon \to 0$?} Thus this method inherits the drawbacks of the ABC-likelihood including the choice of summary statistics and the inefficiency in evaluating the ABC likelihood for high-dimensional data and small $\epsilon$.
 
+\kcnote{do we keep JC-Adv  notes? Should say something about Hamiltonian ABC}
 In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.
 Furthermore, the  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''.
-Lastly, in \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients
+Lastly, in \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients  \kcnote{todo}
 
-\kcnote{F.H. Implicit model is differentiable. Is what we are doing JC-Adv with our variational program}
+%\kcnote{F.H. Implicit model is differentiable. Is what we are doing JC-Adv with our variational program}
 
 \subsection{Operator Variational Optimization}
+
+\kcnote{to do}
 
@@
 \end{eqnarray}
-with
-\begin{equation}
-(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
-\end{equation}
+and variational inference with the KL distance corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
+
+%\begin{equation}
+%(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
+%\end{equation}
 
@@
 
-We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in ~\citep{2016arXiv161009033R}, though more complicated than the a simple reparametrization of normally distributed noise $\epsilon$ through a differentiable function $R(\epsilon, \bfpsi)$.
+%We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in ~\citep{2016arXiv161009033R}, though more complicated than the a simple reparametrization of normally distributed noise $\epsilon$ through a differentiable function $R(\epsilon, \bfpsi)$.
 In our case, the variational program is a marginalized non-differentiable simulator simulator.

### 2017-07-13 tex/paper.tex
OPVI section

@@
 \newcommand{\bfphi}{{\bm \phi}}
+\newcommand{\bflambda}{{\bm \lambda}}
 \newcommand{\bfx}{\mathbf{x}}
@@
 Moreover, we do not attempt to define a recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$ of many real-world simulators is complicated and not amenable to neural a recognition model.
-One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) ?
-\log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.
+%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.
 
@@
 This paired with the critic Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the critic $d$ allows us to work in high dimensions without summary statistics and optimize $\bfpsi$ with stochastic gradient descent.
+\kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL.
+We also don't specify a prior b/c it's like implicit empirical bayes.
+They also don't specify a recognition model to latents.
+}
 
@@
 In that approach "the only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of simulators, they propose to use the ABC-likelihood with an $\epsilon$ kernel.
-\kcnote{Is this an unbiased estimator if $\epsilon > 0$?} This approach
-This approach 
-n the case of many simulators it is not clear how one would obtain the unbiased estimator $\hat{p}(\bfx|\bftheta)$.
-\kcnote{They suggest to use ABC likelihood...
-is this only unbiased as $\epsilon \to 0$?} Thus this method inherits the drawbacks of the ABC-likelihood including the choice of summary statistics and the inefficiency in evaluating the ABC likelihood for high-dimensional data and small $\epsilon$.
+The ABC likelihood is only unbiased as $\epsilon > 0$,
+thus this method inherits the drawbacks of the ABC-likelihood including the choice of summary statistics and the inefficiency in evaluating the ABC likelihood for high-dimensional data and small $\epsilon$.
 
-\kcnote{do we keep JC-Adv  notes? Should say something about Hamiltonian ABC}
-In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.
+%\kcnote{do we keep JC-Adv  notes? Should say something about Hamiltonian ABC}
+%In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.
 Furthermore, the  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''.
-Lastly, in \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients  \kcnote{todo}
+%``Unfortunately, none of the algorithms presented here allow for fully variational learning
+%of model parameters $\bftheta$ when $p_\bftheta(\bfx|\bfz)$ is implicit.''
+%%JC-Adv doesn't let you optimize theta, theta is fixed and you get variational inference on z. Need a recognition model for latents
+
+In \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients ... \kcnote{todo}
 
@@
 
-\subsection{Operator Variational Optimization}
+%\subsection{Operator Variational Optimization}
 
-\kcnote{to do}
-
-Original OPVI  \citep{2016arXiv161009033R}
-
-\begin{eqnarray}
-\lambda^* = \inf_\lambda \sup_\bfphi \mathbb{E}_{ z \sim q( z | \lambda)} [ (O^{p,q} f_\lambda) ] 
-\end{eqnarray}
-and variational inference with the KL distance corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
-
+%Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R}, 
+%which is a generalization of variational inference formulated as the following optimization problem
+%\begin{eqnarray}
+%\bflambda^* = \inf_\bfpsi \sup_\bflambda \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
+%\end{eqnarray}
+%In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
+%AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
+%\begin{eqnarray} \label{eq:opvi_avo}
+%\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \\
+%\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \;.
+%\end{eqnarray}
+%Rewriting Eq.~\ref{eq:opvi_avo} as Eq.~\ref{eq:opvi_avo2} is possible through importance sampling
+%and reinforces the link to density ratio estimation as this corresponds to the operator
 %\begin{equation}
-%(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
+%(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, .
 %\end{equation}
 
-%This is natural here, but not quite OPVI formalism:
+Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R}, 
+which is a generalization of variational inference formulated as the following optimization problem
+\begin{eqnarray}
+\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
+\end{eqnarray}
+In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
+AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
+\begin{eqnarray} \label{eq:opvi_avo}
+\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \\\
+\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
+\end{eqnarray}
+Rewriting Eq.~\ref{eq:opvi_avo} as Eq.~\ref{eq:opvi_avo2} is possible through importance sampling, 
+corresponding to an implicit form of the operator
+\begin{equation}
+(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
+\end{equation}
+which reinforces the link to density ratio estimation and inference with implicit models.
+
+%
+%%\begin{equation}
+%%(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
+%%\end{equation}
+%
+%
+%%This is natural here, but not quite OPVI formalism:
+%%\begin{equation}
+%%\bfpsi^* = \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bftheta \sim q(\bftheta | \bfpsi)} [ (O^{p,q} f_\bfphi) ]
+%%\end{equation}
+%%
+%%\begin{equation}
+%%(O^{p,q} f) = W(p,q) = \sup_f \mathbb{E}_{x\sim p}[f(x)] - \mathbb{E}_{x \sim q}[f(x)]
+%%\end{equation}
+%%The expectation should be $x \sim p(x|\bfpsi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$
+%
+%To get into OPVI formalism
+%
+%\begin{eqnarray}
+%\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi\mathbb{E}_{x\sim p_r(x)}[f(x)] - \mathbb{E}_{x \sim p(x|\bfpsi)}[f(x)] \\
+%&=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ x \sim p( x | \bfpsi)} [ (O^{p_r,p_\bfpsi} f_\bfphi) ] 
+%\end{eqnarray}
+%with
 %\begin{equation}
-%\bfpsi^* = \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bftheta \sim q(\bftheta | \bfpsi)} [ (O^{p,q} f_\bfphi) ]
+%(O^{p,q} f) = \left( \frac{p_r(x)}{p(x|\bfpsi)}  -1 \right) f(x)
 %\end{equation}
-%
-%\begin{equation}
-%(O^{p,q} f) = W(p,q) = \sup_f \mathbb{E}_{x\sim p}[f(x)] - \mathbb{E}_{x \sim q}[f(x)]
-%\end{equation}
-%The expectation should be $x \sim p(x|\bfpsi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$
-
-To get into OPVI formalism
-
-\begin{eqnarray}
-\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi\mathbb{E}_{x\sim p_r(x)}[f(x)] - \mathbb{E}_{x \sim p(x|\bfpsi)}[f(x)] \\
-&=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ x \sim p( x | \bfpsi)} [ (O^{p_r,p_\bfpsi} f_\bfphi) ] 
-\end{eqnarray}
-with
-\begin{equation}
-(O^{p,q} f) = \left( \frac{p_r(x)}{p(x|\bfpsi)}  -1 \right) f(x)
-\end{equation}
 

### 2017-07-13 tex/paper.tex
sentence for Hamiltonian ABC.. lacking context.

@@
 difficult, as simulators rarely admit a tractable density or likelihood
-function. In this note, we develop Adversarial Variational Optimization (AVO), a likelihood-free
+function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
 inference algorithm for fitting a non-differentiable generative model to
-continuous or discrete  data. We adapt the training procedure of generative
+continuous or discrete  data. \kcnote{emperical variational bayes}. We adapt the training procedure of generative
 adversarial networks by replacing the differentiable generative network with a
@@
 Generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation.
-In contrast, scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters that are endowed with some level of interpretation.
+In contrast, scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters and they are endowed with some level of interpretation.
 In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.
@@
 be a fixed-size vector with a simple density.
-Instead, the dimensionality of $z$ and the nature of it's components (uniform, normal, discrete, continuous, etc.) are inherited from the control flow of the simulation code and may depend on $\bftheta$ in some intricate way.
+Instead, the dimensionality of $\bfz$ and the nature of its components (uniform, normal, discrete, continuous, etc.) are inherited from the control flow of the simulation code and may depend on $\bftheta$ in some intricate way.
 Moreover, the dimensionality of $\bfz$ may be much larger than the dimensionality of $\bfx$.
+Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.
 
@@
 such that the density $p(\mathbf{x}|\bftheta)$ can be 
-rewritten as
+written as
 %considered as a non-differentiable reparameterization of the random variable $\bfz$
@@
 \end{equation}
-where $\rho$ is some distance or divergence. For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate.
+where $\rho$ is some distance or divergence.
+%~\citep{bernton2017inference}. 
+%For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate if $ \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta^*)) \to 0$.
+
+%By plugging Wp in place of D in Eqs.
+(1) and (2), we obtain the minimum Wasserstein estimator (MWE) and minimum expected Wasserstein estimator (MEWE) of order p, denoted ??n and ??n,m respectively.
+Some properties of the MWE have been studied in Bassetti et al.
+(2006), for well-specified models and i.i.d.
+data; we propose new results in Section 4.
+Intuitively, under some conditions we can expect ??n to converge to ??, in the sense that Wp(??n,??) ?
+0.
+Consequently, the minimum of ?
+?
+Wp(??n,??) might converge to the minimum of ?
+?
+Wp(??,??), denoted by ??, assuming its existence and unicity.
+In the well-specified case, ??
+coincides with the data-generating parameter.
+In the misspecified case, ??
+is typically different from the limit of the maximum likelihood estimator (MLE), which is the minimizer of KL(??|??).
 
@@
 
-\subsection{Adversarial variational learning}
+\kcnote{Discuss hierarchical models and iid data x. Factor of $N/M$ for AVI-MC?}
 
-\kcnote{connect to MWE and MEWE here?}
+\subsection{Point estimates}
 
-In order to target point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug}
+%\kcnote{connect to MWE and MEWE here?}
+
+When $\rho$ is the Wasserstein distance, $\bftheta^*$ is refered to as the minimum Wasserstein estimator.
+When the model is well specified, $\bftheta^*$ coincides with the true data-generating parameter; however, if the model is misspecified, $\bftheta^*$ is typically different from the maximum likelihood estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx | \bftheta)$~\citep{bernton2017inference}.
+In order to more effectively target point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug}
 with a regularization term corresponding to the differential entropy $H$ of
@@
 Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here analogous to their use in black box variational inference~\citep{ranganath2014black}.
-
-\kcnote{Discuss hierarchical models and iid data x. Factor of $N/M$ for AVI-MC?}
 
@@
 
-Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially}, Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.
+Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially}, Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial}, $\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.
 Each of these assume a tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$, which is  not satisfied in the likelihood-free setting.
@@
 Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an approximation of the gradient of an entropy-regularized Wasserstein distance, which uses a similar duality.
-They note that ``Unfortunately, it is not applicable in the setting of purely generative models, as it involves point-wise evaluations of the derivative of the log-likelihood.'' \kcnote{Can we evaluate their Eq.
-6 with our Eq.
-9} Thus, our contribution is to provide gradients of an approximate MEWE by taking expectations with the variational program $q(\bfx | \bf\theta)$.
+They note that ``Unfortunately, it is not applicable in the setting of purely generative models, as it involves point-wise evaluations of the derivative of the log-likelihood.'' Thus, our contribution is to provide gradients of an approximate MEWE by taking expectations with the variational program $q(\bfx | \bf\theta)$.
 This paired with the critic Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the critic $d$ allows us to work in high dimensions without summary statistics and optimize $\bfpsi$ with stochastic gradient descent.
@@
 first compared to the true data distribution and then updated to make it more
-comparable to the latter. Relevant work in this direction includes \citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood}.
+comparable to the latter.
+Relevant work in this direction includes \citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
 
@@
 
-In \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients ... \kcnote{todo}
+In \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients with respect to $\bftheta$ through finite differences from multiple forward passes of the simulator with variance reduction strategies based on controlling the source of randomness used for the latent variable $\bfz$.
 

### 2017-07-13 code/multi.py
sync with master

@@ -20,6 +20,7 @@ from proposals import grad_gaussian_logpdf
 from proposals import grad_gaussian_entropy
 
 from sklearn.utils import check_random_state
+from scipy.spatial.distance import mahalanobis
 
 # Global params
 
@@ -184,16 +185,22 @@ for i in range(n_epochs):
             x = np.arange(0.0, 2.0, delta)
             y = np.arange(-2.0, 0.0, delta)
             X, Y = np.meshgrid(x, y)
-            Z = [gaussian_logpdf(params_proposal, theta)
-                 for theta in np.hstack([X.reshape(-1, 1), Y.reshape(-1, 1)])]
-            Z = np.exp(np.array(Z).reshape(X.shape))
 
-            CS = plt.contour(X, Y, Z, label=r"$q(\theta|\psi)$")
-            plt.clabel(CS)
+            mu = params_proposal["mu"]
+            sigma = np.diag(np.exp(params_proposal["log_sigma"])) ** 2.0
+            sigma_inv = np.linalg.inv(sigma)
+
+            Z = [mahalanobis(theta, mu, sigma_inv)
+                 for theta in np.hstack([X.reshape(-1, 1), Y.reshape(-1, 1)])]
+            Z = np.array(Z).reshape(X.shape)
+
+            CS = plt.contour(X, Y, Z, [1.0, 2.0, 3.0], colors="C1")
+            fmt = {l:s for l, s in zip(CS.levels, [r"$1\sigma$", r"$2\sigma$", r"$3\sigma$"])}
+            plt.clabel(CS, fmt=fmt)
+
+            plt.scatter(mu[0], mu[1], c="C1", marker="+")
             plt.scatter(true_theta[0], true_theta[1],
-                        c="C0",
-                        label=r"$\theta^* = (%d, %d)$" % (true_theta[0],
-                                                          true_theta[1]))
+                        c="C0")
             plt.xlabel(r"$\alpha$")
             plt.ylabel(r"$\beta$")
             plt.yticks([-1.5, -1.0, -0.5])
@@ -206,12 +213,20 @@ for i in range(n_epochs):
             x = np.arange(0.0, 2.0, delta)
             y = np.arange(-2.0, 0.0, delta)
             X, Y = np.meshgrid(x, y)
-            Z = [gaussian_logpdf(params_proposal, theta)
[…]

### 2017-07-13 tex/paper.tex
indentation

@@
 % and may generate either discrete or continuous data. 
-Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx | \bftheta)$.
-The prevalence and significance of this problem has motivated an active research effort in so-called \textit{likelihood-free inference} algorithms such as Approximate Bayesian Computation (ABC)~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential, sisson2011likelihood, marin2012approximate, cranmer2015approximating}.
-Often the simulation code involves control-flow that implies the dependence of $\bfx$ on $\bfz$ is non-differentiable, making inference even more difficult and precludes approaches such as Ref.~\citep{2016arXiv160507826G}.
+Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx |
+\bftheta)$. The prevalence and significance of this problem has motivated an
+active research effort in so-called \textit{likelihood-free inference}
+algorithms such as Approximate Bayesian Computation
+(ABC)~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential,
+sisson2011likelihood, marin2012approximate, cranmer2015approximating}.  Often
+the simulation code involves control-flow that implies the dependence of $\bfx$
+on $\bfz$ is non-differentiable, making inference even more difficult and
+precludes approaches such as Ref.~\citep{2016arXiv160507826G}.
 
@@
 
-In parallel, with the introduction of variational autoencoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial networks~\cite{goodfellow2014generative}, there has been a vibrant research program around implicit generative models based on neural networks~\citep{2016arXiv161003483M}.
-While these implicit generative networks also do not admit a tractable density, they are differentiable.
+In parallel, with the introduction of variational
+autoencoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
+networks~\cite{goodfellow2014generative}, there has been a vibrant research
+program around implicit generative models based on neural
+networks~\citep{2016arXiv161003483M}.  While these implicit generative networks
+also do not admit a tractable density, they are differentiable.
 
@@
 
-Generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation.
-In contrast, scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters and they are endowed with some level of interpretation.
-In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.
+Generative models based on neural networks are highly parametrized and the model
+parameters have no obvious interpretation. In contrast, scientific simulators
+can be thought of as highly regularized generative models as they typically have
+relatively few parameters and they are endowed with some level of
+interpretation. In this setting, inference on the model parameters $\bftheta$ is
+often of more interest than the latent variables $\bfz$.
 
@@
 source of randomness.
-Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
-be a fixed-size vector with a simple density.
-Instead, the dimensionality of $\bfz$ and the nature of its components (uniform, normal, discrete, continuous, etc.) are inherited from the control flow of the simulation code and may depend on $\bftheta$ in some intricate way.
-Moreover, the dimensionality of $\bfz$ may be much larger than the dimensionality of $\bfx$.
-Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.
+Unlike implicit generative models defined by neural networks, we do not assume
+$\bfz$ to be a fixed-size vector with a simple density. Instead, the
+dimensionality of $\bfz$ and the nature of its components (uniform, normal,
+discrete, continuous, etc.) are inherited from the control flow of the
+simulation code and may depend on $\bftheta$ in some intricate way. Moreover,
+the dimensionality of $\bfz$ may be much larger than the dimensionality of
+$\bfx$. Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.
 
@@
 
-Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems.
-Variance reduction techniques such as Rao-Blackwellization, Control Variates, and quasi-Monte Carlo can be applied here as they are in black box variational inference~\citep{ranganath2014black}.
+Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
+variance of the noisy estimator of the gradients may be too large to be useful
+in many problems. Variance reduction techniques such as Rao-Blackwellization,
+Control Variates, and quasi-Monte Carlo can be applied here as they are in black
+box variational inference~\citep{ranganath2014black}.
 
@@
 %Typically in variational inference one would has access to the posterior $p(\bftheta | \bfx )$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
-Typically in variational inference one starts with a tractable prior $p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the posterior $p(\bftheta | \bfx )$.
-After optimizing $q(\bftheta | \bfpsi)$ to approximate the posterior, one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
-Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by directly minimizing the 
-divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
-Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.
-While the likelihood is intractable, it is fixed, thus one can view the optimization of $\bfpsi$ through the lens of emperical Bayes, though the prior family $p(\bftheta | \bfpsi)$ is not explicitly specified.
+
+Typically in variational inference one starts with a tractable prior
+$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
+posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
+approximate the posterior, one would criticize the model through the the
+posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
+Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by
+directly minimizing the  divergence of the corresponding posterior predictive
+distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating
+the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
+Consequently, the proposed algorithm is a form of variational inference through
+model criticism~\citep{box1980sampling}, which we denote AVI-MC. While the
+likelihood is intractable, it is fixed, thus one can view the optimization of
+$\bfpsi$ through the lens of emperical Bayes, though the prior family
+$p(\bftheta | \bfpsi)$ is not explicitly specified.
 
@@
 
-When $\rho$ is the Wasserstein distance, $\bftheta^*$ is refered to as the minimum Wasserstein estimator.
-When the model is well specified, $\bftheta^*$ coincides with the true data-generating parameter; however, if the model is misspecified, $\bftheta^*$ is typically different from the maximum likelihood estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx | \bftheta)$~\citep{bernton2017inference}.
-In order to more effectively target point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug}
-with a regularization term corresponding to the differential entropy $H$ of
-the proposal distribution $q(\bftheta | \bfpsi)$. That is,
+When $\rho$ is the Wasserstein distance, $\bftheta^*$ is refered to as the
+minimum Wasserstein estimator. When the model is well specified, $\bftheta^*$
+coincides with the true data-generating parameter; however, if the model is
+misspecified, $\bftheta^*$ is typically different from the maximum likelihood
+estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx |
+\bftheta)$~\citep{bernton2017inference}. In order to more effectively target
+point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
+regularization term corresponding to the differential entropy $H$ of the
+proposal distribution $q(\bftheta | \bfpsi)$. That is,
 \begin{equation}
@@
 
-Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially}, Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial}, $\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that add an inference network to the generative model.
-Each of these assume a tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$, which is  not satisfied in the likelihood-free setting.
-Our lifting of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density $q(\bfx | \bfpsi)$ is still intractable.
-Moreover, we do not attempt to define a recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$ of many real-world simulators is complicated and not amenable to neural a recognition model.
+Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
+Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
+$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
+(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
+algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that
+add an inference network to the generative model.  Each of these assume a
+tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
+$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
+of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
+program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
+with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
+$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
+recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$
+of many real-world simulators is complicated and not amenable to neural a
+recognition model.
 %One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.
@@
 
-From the point of view of likelihood-free inference, where  non-differentiable simulators are the norm, our contributions are threefold.
-First is the process of lifting expectations with respect to the non-differentiable simulator $\mathbb{E}_{\tilde{\bfx} \sim p(\bfx | \bftheta)}$ to a differentiable expectations with respect to the variational program $\mathbb{E}_{\tilde{\bfx} \sim q(\bfx | \bfpsi)}$.
-Secondly, is the introduction of a novel form of variational inference that works in a likelihood-free setting.
-Thirdly, the AVI-MC algorithm can be viewed as a form of empirical Bayes where the prior is optimized based on the data.
+From the point of view of likelihood-free inference, where  non-differentiable
+simulators are the norm, our contributions are threefold. First is the process
+of lifting expectations with respect to the non-differentiable simulator
+$\mathbb{E}_{\tilde{\bfx} \sim p(\bfx | \bftheta)}$ to a differentiable
+expectations with respect to the variational program $\mathbb{E}_{\tilde{\bfx}
+\sim q(\bfx | \bfpsi)}$. Secondly, is the introduction of a novel form of
+variational inference that works in a likelihood-free setting. Thirdly, the
+AVI-MC algorithm can be viewed as a form of empirical Bayes where the prior is
+optimized based on the data.
 
@@
 
-Perhaps the closest to our work is~\cite{bernton2017inference}, which uses Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of a rejection sampler in an ABC-like setup (as opposed to variational inference).
-They emphasize that this approach allows one to eliminate the summary statistics typically used in ABC and calculate the Wasserstein distance explicitly without use Kantorovich-Rubinstein duality and the critic $d$.
-For high-dimensional data, they note that this is computationally expensive and introduce an approximation based on projection of the data onto Hilbert space-filling curves.
-Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an approximation of the gradient of an entropy-regularized Wasserstein distance, which uses a similar duality.
-They note that ``Unfortunately, it is not applicable in the setting of purely generative models, as it involves point-wise evaluations of the derivative of the log-likelihood.'' Thus, our contribution is to provide gradients of an approximate MEWE by taking expectations with the variational program $q(\bfx | \bf\theta)$.
-This paired with the critic Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the critic $d$ allows us to work in high dimensions without summary statistics and optimize $\bfpsi$ with stochastic gradient descent.
+Perhaps the closest to our work is~\cite{bernton2017inference}, which uses
+Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
+a rejection sampler in an ABC-like setup (as opposed to variational inference).
+They emphasize that this approach allows one to eliminate the summary statistics
+typically used in ABC and calculate the Wasserstein distance explicitly without
+use Kantorovich-Rubinstein duality and the critic $d$. For high-dimensional
+data, they note that this is computationally expensive and introduce an
+approximation based on projection of the data onto Hilbert space-filling curves.
+Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
+approximation of the gradient of an entropy-regularized Wasserstein distance,
+which uses a similar duality. They note that ``Unfortunately, it is not
+applicable in the setting of purely generative models, as it involves point-wise
+evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
+to provide gradients of an approximate MEWE by taking expectations with the
+variational program $q(\bfx | \bf\theta)$. This paired with the critic
+Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the
+critic $d$ allows us to work in high dimensions without summary statistics and
+optimize $\bfpsi$ with stochastic gradient descent.
 \kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL.
@@
 first compared to the true data distribution and then updated to make it more
-comparable to the latter.
-Relevant work in this direction includes \citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
+comparable to the latter. Relevant work in this direction includes
+\citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
 
@@
 
-This work has many connections to work on variational inference, in which the goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bfz, \bftheta |x)$.
-There have been efforts to extend variational inference to intractable likelihoods; however, they require restrictive assumptions.
-In \citep{tran2017variational}, the authors consider Variational Bayes with an Intractable Likelihood (VBIL).
-In that approach "the only requirement is that the intractable likelihood can be estimated unbiasedly.'' In the case of simulators, they propose to use the ABC-likelihood with an $\epsilon$ kernel.
-The ABC likelihood is only unbiased as $\epsilon > 0$,
-thus this method inherits the drawbacks of the ABC-likelihood including the choice of summary statistics and the inefficiency in evaluating the ABC likelihood for high-dimensional data and small $\epsilon$.
+This work has many connections to work on variational inference, in which the
+goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
+is close to the true posterior $p(\bfz, \bftheta |x)$. There have been efforts
+to extend variational inference to intractable likelihoods; however, they
+require restrictive assumptions.  In \citep{tran2017variational}, the authors
+consider Variational Bayes with an Intractable Likelihood (VBIL). In that
+approach "the only requirement is that the intractable likelihood can be
+estimated unbiasedly.'' In the case of simulators, they propose to use the
+ABC-likelihood with an $\epsilon$ kernel. The ABC likelihood is only unbiased as
+$\epsilon > 0$,   thus this method inherits the drawbacks of the ABC-likelihood
+including the choice of summary statistics and the inefficiency in evaluating
+the ABC likelihood for high-dimensional data and small $\epsilon$.
 
@@
 
-In \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which applies in the likelihood-free setting, and they estimate gradients with respect to $\bftheta$ through finite differences from multiple forward passes of the simulator with variance reduction strategies based on controlling the source of randomness used for the latent variable $\bfz$.
+In \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which
+applies in the likelihood-free setting, and they estimate gradients with respect
+to $\bftheta$ through finite differences from multiple forward passes of the
+simulator with variance reduction strategies based on controlling the source of
+randomness used for the latent variable $\bfz$.
 

### 2017-07-13 tex/paper.tex
spell check

@@
 inference algorithm for fitting a non-differentiable generative model to
-continuous or discrete  data. \kcnote{emperical variational bayes}. We adapt the training procedure of generative
+continuous or discrete  data. \kcnote{empirical variational bayes}. We adapt the training procedure of generative
 adversarial networks by replacing the differentiable generative network with a
@@
 In parallel, with the introduction of variational
-autoencoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
+auto-encoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
 networks~\cite{goodfellow2014generative}, there has been a vibrant research
@@
 
-We consider a family of parameterized densities $p(\mathbf{x}|\bftheta)$
+We consider a family of parametrized densities $p(\mathbf{x}|\bftheta)$
 defined implicitly through the simulation of a stochastic generative process,
@@
 $\bfz$ to be a fixed-size vector with a simple density. Instead, the
-dimensionality of $\bfz$ and the nature of its components (uniform, normal,
+dimension of $\bfz$ and the nature of its components (uniform, normal,
 discrete, continuous, etc.) are inherited from the control flow of the
 simulation code and may depend on $\bftheta$ in some intricate way. Moreover,
-the dimensionality of $\bfz$ may be much larger than the dimensionality of
+the dimension of $\bfz$ may be much larger than the dimension of
 $\bfx$. Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.
@@
 have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
-the marginal distribution parameterized by $\bfpsi$
+the marginal distribution parametrized by $\bfpsi$
 \begin{equation}
@@
 likelihood is intractable, it is fixed, thus one can view the optimization of
-$\bfpsi$ through the lens of emperical Bayes, though the prior family
+$\bfpsi$ through the lens of empirical Bayes, though the prior family
 $p(\bftheta | \bfpsi)$ is not explicitly specified.
@@
 
-When $\rho$ is the Wasserstein distance, $\bftheta^*$ is refered to as the
+When $\rho$ is the Wasserstein distance, $\bftheta^*$ is referred to as the
 minimum Wasserstein estimator. When the model is well specified, $\bftheta^*$
@@
 distribution with unknown mean $\lambda$. We artificially consider
-the distribution as a parameterized simulator, from which we can only
+the distribution as a parametrized simulator, from which we can only
 generate data.
@@
              ({\it Right}) Density $q(\bftheta|\bfpsi)$ after adversarial variational optimization ($\gamma=10$).
-The proposal density correctly converges towards a distributon whose density concentrates around $\bftheta^* = (1, -1)$.
+The proposal density correctly converges towards a distribution whose density concentrates around $\bftheta^* = (1, -1)$.
              }\label{fig:multi}
@@
 
-In this example, we consider observed data generated with the simplified generator of Eq.~\ref{eq:simple_weinberg}
+In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
 using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
@@
 %\end{eqnarray}
-%Rewriting Eq.~\ref{eq:opvi_avo} as Eq.~\ref{eq:opvi_avo2} is possible through importance sampling
+%Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling
 %and reinforces the link to density ratio estimation as this corresponds to the operator
@@
 \end{eqnarray}
-Rewriting Eq.~\ref{eq:opvi_avo} as Eq.~\ref{eq:opvi_avo2} is possible through importance sampling,
+Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling,
 corresponding to an implicit form of the operator

### 2017-07-13 tex/paper.tex
layout

@@
 Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
-setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta))$ by
-replacing the adversarial classifier with a 1-Lipschitz adversarial critic
-$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$.
+setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
+by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
+$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, and where
 \begin{align}
-W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) =& \sup_{\bfphi} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [d(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [d(\mathbf{x})] \\
-\equiv& \sup_{\bfphi} \mathcal{L}_W
+& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
+&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
+&\equiv \sup_{\bfphi} \mathcal{L}_W.
 \end{align}
@@
 
-\subsection{Adversarial variational inference through model criticism}
+\subsection{Adversarial variational inference\\
+            through model criticism}
 
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
+    \caption{Electron--positron annihilation.
+    ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
+The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\bftheta^*$.
+    ({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
+    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+             }\label{fig:weinberg}
+\end{figure}
+
 As a more realistic example, we now consider a (simplified) simulator from
@@
 %simulator.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
-    \caption{Electron--positron annihilation.
-    ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
-The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\bftheta^*$.
-    ({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
-    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
-             }\label{fig:weinberg}
-\end{figure}
 

### 2017-07-13 tex/paper.tex
minor edits

@@
 (AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
-algorithm of  \citep{2017arXiv170208235H})  are a recent extension to GANs that
+algorithm of  \citep{2017arXiv170208235H}  are a recent extension to GANs that
 add an inference network to the generative model.  Each of these assume a
@@
 They emphasize that this approach allows one to eliminate the summary statistics
-typically used in ABC and calculate the Wasserstein distance explicitly without
-use Kantorovich-Rubinstein duality and the critic $d$. For high-dimensional
+typically used in ABC and calculate the Wasserstein distance explicitly, without
+making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
 data, they note that this is computationally expensive and introduce an
@@
 to provide gradients of an approximate MEWE by taking expectations with the
-variational program $q(\bfx | \bf\theta)$. This paired with the critic
-Kantorovich-Rubinstein dual formulation of the Wasserstein distance using the
-critic $d$ allows us to work in high dimensions without summary statistics and
+variational program $q(\bfx | \bfpsi)$. This paired with the critic
+Kantorovich-Rubinstein dual formulation of the Wasserstein distance
+allows us to work in high dimensions without summary statistics and
 optimize $\bfpsi$ with stochastic gradient descent.
@@
 
-This work has many connections to work on variational inference, in which the
+This work has also many connections to work on variational inference, in which the
 goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
-is close to the true posterior $p(\bfz, \bftheta |x)$. There have been efforts
+is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have been efforts
 to extend variational inference to intractable likelihoods; however, they
@@
 consider Variational Bayes with an Intractable Likelihood (VBIL). In that
-approach "the only requirement is that the intractable likelihood can be
+approach ``the only requirement is that the intractable likelihood can be
 estimated unbiasedly.'' In the case of simulators, they propose to use the
@@
 \end{eqnarray}
-In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
+In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
 AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and

### 2017-07-13 tex/paper.tex
AVO-EB

@@
 function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
-inference algorithm for fitting a non-differentiable generative model to
-continuous or discrete  data. \kcnote{empirical variational bayes}. We adapt the training procedure of generative
+inference algorithm for fitting a non-differentiable generative model or to perform empirical Bayes through variational inference. 
+We adapt the training procedure of generative
 adversarial networks by replacing the differentiable generative network with a
@@
 
-\subsection{Adversarial variational inference\\
-            through model criticism}
+\subsection{Empirical Bayes through Adversarial Variational Inference}
 
@@
 Consequently, the proposed algorithm is a form of variational inference through
-model criticism~\citep{box1980sampling}, which we denote AVI-MC. While the
+model criticism~\citep{box1980sampling}, which we denote AVO-EB. While the
 likelihood is intractable, it is fixed, thus one can view the optimization of
@@
 
-\kcnote{Discuss hierarchical models and iid data x. Factor of $N/M$ for AVI-MC?}
+\kcnote{Discuss hierarchical models and iid data x. Factor of $N/M$ for AVO-EB?}
 
@@
 between the generator objective and the tightness of the proposal distribution.
+We refer to the algorithm with $\gamma>0$ as AVO-MWE.
 For small values of $\gamma$,

### 2017-07-13 tex/paper.tex
mention fast exact Wasserstein

@@
 in many problems. Variance reduction techniques such as Rao-Blackwellization,
-Control Variates, and quasi-Monte Carlo can be applied here as they are in black
-box variational inference~\citep{ranganath2014black}.
+Control Variates, and quasi-Monte Carlo can be applied here~\citep{ranganath2014black, tran2017variational}. 
+Similarly, various fast 
+algorithms for calculating the exact (entropically regularized) Wasserstein distance on
+empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
 
@@
 When $\rho$ is the Wasserstein distance, $\bftheta^*$ is referred to as the
-minimum Wasserstein estimator. When the model is well specified, $\bftheta^*$
+minimum Wasserstein estimator (MWE). When the model is well specified, $\bftheta^*$
 coincides with the true data-generating parameter; however, if the model is
@@
 estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx |
-\bftheta)$~\citep{bernton2017inference}. In order to more effectively target
+\bftheta))$~\citep{bernton2017inference}. In order to more effectively target
 point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
@@
 
-\kcnote{does \citep{rosca2017variational} go in the above paragraph?}
+%\kcnote{does \citep{rosca2017variational} go in the above paragraph?}
 
@@
 evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
-to provide gradients of an approximate MEWE by taking expectations with the
+to provide gradients of an approximate MWE by taking expectations with the
 variational program $q(\bfx | \bfpsi)$. This paired with the critic
@@
 allows us to work in high dimensions without summary statistics and
-optimize $\bfpsi$ with stochastic gradient descent.
-\kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL.
+optimize $\bfpsi$ with stochastic gradient descent. Additionally,
+our our AVO-EB procedure avoids the inefficiencies of 
+their Wasserstein rejection sampler incurred from using the ABC-likelihood.
+
+%\kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL.
 We also don't specify a prior b/c it's like implicit empirical bayes.
@@
 
-\kcnote{proposal distribution $\to$ variational distribution}
+%\kcnote{proposal distribution $\to$ variational distribution}
 

### 2017-07-13 tex/paper.tex
version to Rajesh

@@
 
-\kcnote{Discuss hierarchical models and iid data x. Factor of $N/M$ for AVO-EB?}
+\kcnote{Clarify hierarchical models and iid data x. Factor of $N/M$ for AVO-EB?}
 
@@
 
-In this note, we develop a likelihood-free inference algorithm for fitting a
-forward non-differentiable simulator to continuous or discrete observed data.
-The algorithm combines adversarial training  with variational optimization to
+In this note, we develop likelihood-free inference algorithms for non-differentiable, implicit generative models.
+The algorithms combine adversarial training  with variational optimization to
 minimize variational upper bounds  on the otherwise non-differentiable
 adversarial objectives.
-Effectively, the procedure results in learning an arbitrarily tight proposal
-distribution over the simulator parameters, such that corresponding the marginal
-distribution of the generated data matches the observations. Preliminary results
-with non-differentiable discrete or continuous simulators validate the proposed method.
+The AVO-EB algorithm enables empirical Bayesian through variational inference in the likelihood free setting.
+This approach does not incur the inefficiencies of an ABC-like rejection sampler. The AVO-MWE algorithm
+provides point estimates for the generative model, which asymptotically correspond to the data generating parameter when the model is well-specified.
+Both algorithms work on continuous or discrete observed data.
 
+Preliminary results on toy problems with discrete and continuous data validate the proposed method.
 While the obtained results are encouraging, the complete validation of the
@@
 Third, we need to consider whether less noisy estimates of the gradients
-$\nabla_\bfpsi U_g$ can be computed, in particular by exploiting the fact in the composition $d(g(\bfz;\bftheta); \bfpsi)$
-the derivatives of $d$ are known exactly.
+$\nabla_\bfpsi U_g$ can be computed.
+%, in particular by exploiting the fact in the composition $d(g(\bfz;\bftheta); \bfpsi)$
+%the derivatives of $d$ are known exactly.
 
-\kcnote{could also use this to tune non-differentialble fast simulator to slower full simulator. Then fixed data size not an issue}
+%We note that the technique can be used either to fit a generative model to an observed data distribution 
+%or to fit a fast simulator to a slower, higher-fidelity simulator.
+
+%\kcnote{could also use this to tune non-differentialble fast simulator to slower full simulator. Then fixed data size not an issue}
 

### 2017-07-14 tex/paper.tex
small things

@@
 \end{equation}
-We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in \citep{2016arXiv161009033R}, though more complicated than a simple reparameterization of normally distributed noise $\bfz$ through a differentiable function $g(\bfz, \bfpsi)$.
-In our case, the variational program is a marginalized, non-differentiable  simulator.
-Its density is  intractable; nevertheless, it can generate samples for $\bfx$ and is differentiable with respect to $\bfpsi$ and $\bfphi$.
+We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
+in \citep{2016arXiv161009033R}, though more complicated than a simple
+reparameterization of normally distributed noise $\bfz$ through a differentiable
+function $g(\bfz, \bfpsi)$. In our case, the variational program is a
+marginalized, non-differentiable  simulator.  Its density is  intractable;
+nevertheless, it can generate samples for $\bfx$ and is differentiable with
+respect to $\bfpsi$.
 Operationally, we sample from this marginal model via
@@
 posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
-Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by
+Effectively, we are optimizing variational posterior $q(\bftheta | \bfpsi)$ by
 directly minimizing the  divergence of the corresponding posterior predictive

### 2017-07-14 tex/paper.tex
save

@@
 have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
-the marginal distribution parametrized by $\bfpsi$
+the parameterized marginal distribution of the generated data
 \begin{equation}
@@
 reparameterization of normally distributed noise $\bfz$ through a differentiable
-function $g(\bfz, \bfpsi)$. In our case, the variational program is a
+function. In our case, the variational program is a
 marginalized, non-differentiable  simulator.  Its density is  intractable;
@@
 posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
-Effectively, we are optimizing variational posterior $q(\bftheta | \bfpsi)$ by
+Effectively, we are optimizing the variational posterior $q(\bftheta | \bfpsi)$ by
 directly minimizing the  divergence of the corresponding posterior predictive

### 2017-07-14 tex/paper.tex
save

@@
 Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
-variance of the noisy estimator of the gradients may be too large to be useful
-in many problems. Variance reduction techniques such as Rao-Blackwellization,
-Control Variates, and quasi-Monte Carlo can be applied here~\citep{ranganath2014black, tran2017variational}.
-Similarly, various fast
-algorithms for calculating the exact (entropically regularized) Wasserstein distance on
+variance of the noisy estimator of the gradients may be too large to be useful in many problems. Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
+Control Variates, and quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}. 
+Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
 empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}. 
+Similarly, the algorithm could be adopted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
+These variants may be particularly useful for small $N$.
 
@@
 Consequently, the proposed algorithm is a form of variational inference through
-model criticism~\citep{box1980sampling}, which we denote AVO-EB. While the
-likelihood is intractable, it is fixed, thus one can view the optimization of
-$\bfpsi$ through the lens of empirical Bayes, though the prior family
-$p(\bftheta | \bfpsi)$ is not explicitly specified.
+model criticism~\citep{box1980sampling}.
+
+We can also view the optimization  through the lens of empirical Bayes (EB),
+where one specifies a family of priors and $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
+%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the 
+%posterior predictive distribution. 
+Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
+ $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi^*) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi^*) d\bftheta $
+as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly 
+specified a prior family. However, the procedure corresponds to empirical Bayes if we identify 
+$q(\bftheta | \bfpsi^*) = p(\bftheta | \bfx , \bfpsi)$. Thus we call our method AVO-EB and note that it is applicable even outside of the likelihood-free context.
+
+Algorithm~\ref{alg:avo}, can be easily modified to have different mini-batch $M_r$ for the observed data and $M_g$ for the simulator. 
+When $M_g \gg M_r$, then the variance of $\bftheta^*$ goes like $M_r$~\citep{bernton2017inference}.
+A final $\sqrt{N/M}$ scaling to the log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of the data.
+
+%While the
+%likelihood is intractable, it is fixed, thus one can view the optimization of
+%$\bfpsi$ through the lens of empirical Bayes, though the prior family
+%$p(\bftheta | \bfpsi)$ is not explicitly specified, 
+%ie. $q(\bftheta | \bfpsi^*) = p(\bfx | \bftheta) p(\bftheta | \bfpsi^*) / \int d\bftheta p(\bfx | \bftheta) p(\theta | \bfpsi^*)$.
 
@@
 %For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
-
-\kcnote{Clarify hierarchical models and iid data x. Factor of $N/M$ for AVO-EB?}
 

### 2017-07-14 tex/paper.tex
modified section title and fix typos

@@
 empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}. 
-Similarly, the algorithm could be adopted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
+Moreover, the algorithm could be adopted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
 These variants may be particularly useful for small $N$.
 
-\subsection{Empirical Bayes through Adversarial Variational Inference}
+\subsection{Empirical Bayes through  Variational Inference}
 
@@
 marginalized, non-differentiable  simulator.  Its density is  intractable;
-nevertheless, it can generate samples for $\bfx$ and is differentiable with
+nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
 respect to $\bfpsi$.
@@
 Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
- $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi^*) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi^*) d\bftheta $
+ $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi) d\bftheta $
 as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly 
@@
 Algorithm~\ref{alg:avo}, can be easily modified to have different mini-batch $M_r$ for the observed data and $M_g$ for the simulator. 
-When $M_g \gg M_r$, then the variance of $\bftheta^*$ goes like $M_r$~\citep{bernton2017inference}.
+When $M_g \gg M_r$, then the variance of $\bftheta^*$ scales ad $\sqrt{M_r}$~\citep{bernton2017inference}.
 A final $\sqrt{N/M}$ scaling to the log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of the data.

### 2017-07-15 tex/paper.tex
save

@@
 written as
-%considered as a non-differentiable reparameterization of the random variable $\bfz$
+%considered as a non-differentiable reparametrization of the random variable $\bfz$
 %\begin{equation}\label{eqn:p_x_sim}
@@
 % Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
-% of scale invariance. To avoid the dependence on the parameterization of the
+% of scale invariance. To avoid the dependence on the parametrization of the
 % proposal distribution, natural gradients~\citep{amari1998natural} can be used
@@
 %have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
-%a distribution parameterized in terms of $\bfpsi$:
+%a distribution parametrized in terms of $\bfpsi$:
 %\begin{equation}\label{eqn:p_psi}
@@
 have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
-the parameterized marginal distribution of the generated data
+the parametrized marginal distribution of the generated data
 \begin{equation}
@@
 in \citep{2016arXiv161009033R}, though more complicated than a simple
-reparameterization of normally distributed noise $\bfz$ through a differentiable
+reparametrization of normally distributed noise $\bfz$ through a differentiable
 function. In our case, the variational program is a
@@
 each outer epoch in order to avoid building momentum in staled directions.  For
-estimating $\lambda^*$, we parameterize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
+estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
 $q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
@@
                         the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
-             ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
+             ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
              ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
@@
 \begin{eqnarray} \label{eq:opvi_avo}
-\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \\\
+\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \nonumber \\
 \label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .

### 2017-07-15 tex/paper.tex
small edits

@@
 \end{itemize}
-Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model
+Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model.
 We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
@@
 Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
-The network architecture of the critic is the same as in the previous example.
+The architecture of the critic is the same as in the previous example.
 
@@
 $q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
-recognition model $q(\bfz | \bftheta, \bfx)$ as the latent space $\mathcal{Z}$
+recognition model  $q(\bfz, \bftheta|\bfpsi)$ as the latent space $\mathcal{Z}$
 of many real-world simulators is complicated and not amenable to neural a
@@
 simulators are the norm, our contributions are threefold. First is the process
-of lifting expectations with respect to the non-differentiable simulator
+of lifting the expectation with respect to the non-differentiable simulator
 $\mathbb{E}_{\tilde{\bfx} \sim p(\bfx | \bftheta)}$ to a differentiable
-expectations with respect to the variational program $\mathbb{E}_{\tilde{\bfx}
+expectation with respect to the variational program $\mathbb{E}_{\tilde{\bfx}
 \sim q(\bfx | \bfpsi)}$. Secondly, is the introduction of a novel form of
 variational inference that works in a likelihood-free setting. Thirdly, the
-AVI-MC algorithm can be viewed as a form of empirical Bayes where the prior is
+AVO-EB algorithm can be viewed as a form of empirical Bayes where the prior is
 optimized based on the data.
@@
 approximation of the gradient of an entropy-regularized Wasserstein distance,
-which uses a similar duality. They note that ``Unfortunately, it is not
+which uses a similar duality. They note that ``unfortunately, it is not
 applicable in the setting of purely generative models, as it involves point-wise
@@
 optimize $\bfpsi$ with stochastic gradient descent. Additionally,
-our our AVO-EB procedure avoids the inefficiencies of
+our AVO-EB procedure avoids the inefficiencies of
 their Wasserstein rejection sampler incurred from using the ABC-likelihood.
@@
 ABC-likelihood with an $\epsilon$ kernel. The ABC likelihood is only unbiased as
-$\epsilon > 0$,   thus this method inherits the drawbacks of the ABC-likelihood
+$\epsilon \to 0$,   thus this method inherits the drawbacks of the ABC-likelihood
 including the choice of summary statistics and the inefficiency in evaluating
@@
 Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
-which is a generalization of variational inference formulated as the following optimization problem
+which is a generalization of variational inference formulated as the following optimization problem:
 \begin{eqnarray}
@@
 \end{eqnarray}
-In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
+In traditional VI with the KL distance, we have $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
 AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
 \begin{eqnarray} \label{eq:opvi_avo}
-\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \nonumber \\
-\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
+\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,q_\bfpsi} d_\bfphi) ] \\
+\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx; \bfphi)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx; \bfphi)] \nonumber \, .
 \end{eqnarray}
-Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling,
+Rewriting Eqn.~\ref{eq:opvi_avo} as above is possible through importance sampling,
 corresponding to an implicit form of the operator
 \begin{equation}
-(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
+(O^{p_r,q_\bfpsi} d_\bfphi)(\bfx) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx; \bfphi) \, ,
 \end{equation}
@@
 adversarial objectives.
-The AVO-EB algorithm enables empirical Bayesian through variational inference in the likelihood free setting.
+The AVO-EB algorithm enables empirical Bayes through variational inference in the likelihood-free setting.
 This approach does not incur the inefficiencies of an ABC-like rejection sampler. The AVO-MWE algorithm
@@
 
-\section*{Acknowledgments}
+\vspace{0.25cm}
+\subsection*{Acknowledgments}
 
@@
 the electron--positron annihilation simulation.
-
 GL and KL are both supported through NSF ACI-1450310, additionally KC is

### 2017-07-15 tex/paper.tex
save

@@
 \end{eqnarray}
-In traditional VI with the KL distance, we have $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
+In traditional VI with the KL distance, this corresponds to $(O^{p,q} f)(\bfz) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
 AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
@@
 
-\vspace{0.25cm}
-\subsection*{Acknowledgments}
+\vspace{0.4cm}
+\section*{Acknowledgments}
 
@@
 
+\vspace{2cm}
+
 %\bibliographystyle{hacm}

### 2017-07-16 tex/paper.tex
save

@@
 the dimension of $\bfz$ may be much larger than the dimension of
-$\bfx$. Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.
+$\bfx$.
+%Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.
 
@@
 specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
-parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and alternately minimize the variational upper bounds
+parameters of $g$ and alternately minimize the variational upper bounds
 \begin{align}
@@
 empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
-Moreover, the algorithm could be adopted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
+Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
 These variants may be particularly useful for small $N$.
@@
 We can also view the optimization  through the lens of empirical Bayes (EB),
-where one specifies a family of priors and $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
+where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
 %In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
@@
 
-Algorithm~\ref{alg:avo}, can be easily modified to have different mini-batch $M_r$ for the observed data and $M_g$ for the simulator.
-When $M_g \gg M_r$, then the variance of $\bftheta^*$ scales ad $\sqrt{M_r}$~\citep{bernton2017inference}.
-A final $\sqrt{N/M}$ scaling to the log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of the data.
+Algorithm~\ref{alg:avo}, can be easily modified to have different sizes of mini-batches,
+$M_r$ for the observed data and $M_g$ for the simulator. When $M_g \gg M_r$,
+then the variance of $\bftheta^*$ scales as
+$\sqrt{M_r}$~\citep{bernton2017inference}. A final $\sqrt{N/M}$ scaling to the
+log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of
+the data.
 

### 2017-07-17 tex/paper.tex
Rajesh and Dustin
reworded last paragraph of intro

@@
 
-In this note, we develop an unsupervised learning algorithm for the point
-estimation of the parameters $\bftheta$ of a non-differentiable, implicit
-generative model. We adapt the adversarial
+In this note, we develop a likelihood-free inference algorithms 
+non-differentiable, implicit generative mode. The first allows us 
+to perform empirical Bayes through variational inference, and the 
+second provides a point estimator of the parameters $\bftheta$. 
+We adapt the adversarial
 training procedure of generative adversarial
@@
 bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
-objectives. The procedure results in learning an arbitrarily tight
-proposal distribution over simulator parameters, such that the corresponding
-marginal distribution of the generated data matches the observations.
+objectives. 
+The objective of both algorithms is to match marginal distribution of 
+the generated data to the empirical distribution of the observations.
+
+%The procedure results in learning an arbitrarily tight
+%proposal distribution over simulator parameters, such that the corresponding
+%marginal distribution of the generated data matches the observations.
+
+%In this note, we develop an unsupervised learning algorithm for the point
+%estimation of the parameters $\bftheta$ of a non-differentiable, implicit
+%generative model. We adapt the adversarial
+%training procedure of generative adversarial
+%networks~\cite{goodfellow2014generative} by replacing the implicit generative
+%network with a domain-based scientific simulator, and solve the resulting
+%non-differentiable minimax problem by minimizing variational upper
+%bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
+%objectives. 
+%The procedure results in learning an arbitrarily tight
+%proposal distribution over simulator parameters, such that the corresponding
+%marginal distribution of the generated data matches the observations.
 
@@
 We would like to thank Lukas Heinrich for helpful comments regarding
-the electron--positron annihilation simulation.
+the electron--positron annihilation simulation. We would also like to thank
+Rajesh Ranganath and Dustin Tran for enlightening discussions and feedback.
 GL and KL are both supported through NSF ACI-1450310, additionally KC is

### 2017-07-18 tex/paper.tex
some fixes, moved a few things around

@@
 algorithms such as Approximate Bayesian Computation
-(ABC)~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential,
-sisson2011likelihood, marin2012approximate, cranmer2015approximating}.  Often
-the simulation code involves control-flow that implies the dependence of $\bfx$
-on $\bfz$ is non-differentiable, making inference even more difficult and
-precludes approaches such as Ref.~\citep{2016arXiv160507826G}.
+(ABC) and density estimation-by-comparison algorithms~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential,
+sisson2011likelihood, marin2012approximate, cranmer2015approximating}.
+% Often
+% the simulation code involves control-flow that implies the dependence of $\bfx$
+% on $\bfz$ is non-differentiable, making inference even more difficult and
+% precludes approaches such as Ref.~\citep{2016arXiv160507826G}.
 
@@
 also do not admit a tractable density, they are differentiable.
+Generative models based on neural networks are highly parametrized and the model
+parameters have no obvious interpretation. In contrast, scientific simulators
+can be thought of as highly regularized generative models as they typically have
+relatively few parameters and they are endowed with some level of
+interpretation. In this setting, inference on the model parameters $\bftheta$ is
+often of more interest than the latent variables $\bfz$.
 
@@
 
-Generative models based on neural networks are highly parametrized and the model
-parameters have no obvious interpretation. In contrast, scientific simulators
-can be thought of as highly regularized generative models as they typically have
-relatively few parameters and they are endowed with some level of
-interpretation. In this setting, inference on the model parameters $\bftheta$ is
-often of more interest than the latent variables $\bfz$.
-
-In this note, we develop a likelihood-free inference algorithms 
-non-differentiable, implicit generative mode. The first allows us 
+In this note, we develop two likelihood-free inference algorithms
+for non-differentiable, implicit generative models. The first allows us
 to perform empirical Bayes through variational inference, and the
@@
 \begin{equation}
-    \bftheta^* = \arg \min_{\bftheta \in \bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
+    \bftheta^* = \arg \min_{\bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
 \end{equation}
@@
 by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
-$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, and where
+$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, with
 \begin{align}
@@
 \begin{equation}
-    \min_{\bftheta \in {\Theta}} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
+    \min_{\bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
 \end{equation}
@@
 That is, the minimum of a set of function values is always less than or equal
-to any of their average. Provided that the proposal is flexible enough, the parameters $\bfpsi$
+to any of their average. Provided that the proposal distribution is flexible enough, the parameters $\bfpsi$
 can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.
@@
 $U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
-trick it can be rewritten as:
+trick its gradient can be rewritten as:
 \begin{align}\label{eqn:approx-grad}
@@
 $U(\bfpsi)$ with stochastic gradient descent. We will use
-Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to incorporate the Natural Evolution Strategy
-algorithm~\citep{2011arXiv1106.4487W} and variance reduction techniques~\citep{ranganath2014black}.
+Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to make  use instead of the Natural Evolution Strategy
+algorithm~\citep{2011arXiv1106.4487W} or of variance reduction techniques~\citep{ranganath2014black}.
 %Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems.
@@
 variance of the noisy estimator of the gradients may be too large to be useful in many problems. Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
-Control Variates, and quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
+Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
 Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
 empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
-Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
+%Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
 These variants may be particularly useful for small $N$.
@@
 posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
-approximate the posterior, one would criticize the model through the the
+approximate the posterior, one would criticize the model through the
 posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
@@
 directly minimizing the  divergence of the corresponding posterior predictive
-distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating
+distribution $\qxpsi$ and the data distribution $p_r(\bfx)$, without evaluating
 the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
@@
 
-Very large penalties may
+Finally, we note that very large penalties may
 eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
@@
 The simulator parameters are modeled with a factored  (mean field) Gaussian
-variational distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
+proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
 initialized with zero mean and unit variance.
@@
 using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
-factored (mean field) Gaussian variational distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
+factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
 q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
@@
 
-Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
-Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
-$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
-(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
-algorithm of  \citep{2017arXiv170208235H}  are a recent extension to GANs that
-add an inference network to the generative model.  Each of these assume a
-tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
-$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
-of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
-program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
-with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
-$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
-recognition model  $q(\bfz, \bftheta|\bfpsi)$ as the latent space $\mathcal{Z}$
-of many real-world simulators is complicated and not amenable to neural a
-recognition model.
-%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.
-
-%\kcnote{does \citep{rosca2017variational} go in the above paragraph?}
-
 From the point of view of likelihood-free inference, where  non-differentiable
@@
 
-Perhaps the closest to our work is~\cite{bernton2017inference}, which uses
+Perhaps the closest to our work is~\cite{bernton2017inference}, which uses the
 Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
@@
 Kantorovich-Rubinstein dual formulation of the Wasserstein distance
-allows us to work in high dimensions without summary statistics and
+allows us to work in high dimensions without summary statistics and to
 optimize $\bfpsi$ with stochastic gradient descent. Additionally,
@@
 
-As reviewed in \cite{2016arXiv161003483M}, likelihood-free inference is
-intimately tied to a class of algorithms that can be framed as density
-estimation-by-comparison. In most cases, these inference algorithms are
-formulated as an iterative two-step process where the model distribution is
-first compared to the true data distribution and then updated to make it more
-comparable to the latter. Relevant work in this direction includes
+More generally, likelihood-free
+inference is intimately tied to a class of algorithms that can be framed as
+density estimation-by-comparison, as reviewed in \cite{2016arXiv161003483M}. In most cases, these
+inference algorithms are formulated as an iterative two-step process where the
+model distribution is first compared to the true data distribution and then
+updated to make it more comparable to the latter. Relevant work in this
+direction includes those that rely on a classifier to estimate the discrepancy betwen the observed data and the model distributions
 \citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
+Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\citep{meeds2015hamiltonian}
+estimate gradients with respect
+to $\bftheta$ through finite differences from multiple forward passes of the
+simulator with variance reduction strategies based on controlling the source of
+randomness used for the latent variable $\bfz$.
+
+Likewise, AVO closely relates to recent extensions of GANs, such as
+Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
+Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
+$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
+(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
+algorithm of  \citep{2017arXiv170208235H}, which
+add an inference network to the generative model.  Each of these assume a
+tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
+$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
+of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
+program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
+with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
+$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
+recognition model  $q(\bfz, \bftheta|\bfpsi)$ as the latent space $\mathcal{Z}$
+of many real-world simulators is complicated and not amenable to neural a
+recognition model.
+%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.
 
@@
 estimated unbiasedly.'' In the case of simulators, they propose to use the
-ABC-likelihood with an $\epsilon$ kernel. The ABC likelihood is only unbiased as
+ABC-likelihood with an $\epsilon$-kernel. The ABC likelihood is only unbiased as
 $\epsilon \to 0$,   thus this method inherits the drawbacks of the ABC-likelihood
@@
 %%JC-Adv doesn't let you optimize theta, theta is fixed and you get variational inference on z. Need a recognition model for latents
-
-In \citep{meeds2015hamiltonian}, the authors introduce Hamiltonian ABC which
-applies in the likelihood-free setting, and they estimate gradients with respect
-to $\bftheta$ through finite differences from multiple forward passes of the
-simulator with variance reduction strategies based on controlling the source of
-randomness used for the latent variable $\bfz$.
 
@@
 
-In this note, we develop likelihood-free inference algorithms for non-differentiable, implicit generative models.
-The algorithms combine adversarial training  with variational optimization to
-minimize variational upper bounds  on the otherwise non-differentiable
-adversarial objectives.
-The AVO-EB algorithm enables empirical Bayes through variational inference in the likelihood-free setting.
-This approach does not incur the inefficiencies of an ABC-like rejection sampler. The AVO-MWE algorithm
-provides point estimates for the generative model, which asymptotically correspond to the data generating parameter when the model is well-specified.
-Both algorithms work on continuous or discrete observed data.
-
-Preliminary results on toy problems with discrete and continuous data validate the proposed method.
-While the obtained results are encouraging, the complete validation of the
-method remains to be carried out in real conditions on a full fledged scientific
-simulator -- which we plan to achieve for a next version of this work.
-In terms of method, several components need to be further investigated.
-First, we need to better study the interplay between the entropy penalty and the adversarial objectives.
-Second, we should better understand the dynamics of the optimization
-procedure, in particular when combined with momentum-based optimizers like Adam.
-Third, we need to consider whether less noisy estimates of the gradients
-$\nabla_\bfpsi U_g$ can be computed.
+In this note, we develop likelihood-free inference algorithms for
+non-differentiable, implicit generative models. The algorithms combine
+adversarial training  with variational optimization to minimize variational
+upper bounds  on the otherwise non-differentiable adversarial objectives. The
+AVO-EB algorithm enables empirical Bayes through variational inference in the
+likelihood-free setting. This approach does not incur the inefficiencies of an
+ABC-like rejection sampler. The AVO-MWE algorithm provides point estimates for
+the generative model, which asymptotically corresponds to the data generating
+parameters when the model is well-specified. Both algorithms work on continuous
+or discrete observed data.
+
+Preliminary results on toy problems with discrete and continuous data validate
+the proposed method. While the obtained results are encouraging, the complete
+validation of the method remains to be carried out in real conditions on a full
+fledged scientific simulator -- which we plan to achieve for a next version of
+this work. In terms of method, several components need further investigation.
+First, we need to better study the interplay between the entropy penalty and the
+adversarial objectives. Second, we should better understand the dynamics of the
+optimization procedure, in particular when combined with momentum-based
+optimizers like Adam. Third, we need to consider whether less noisy estimates of
+the gradients $\nabla_\bfpsi U_g$ can be computed.
 %, in particular by exploiting the fact in the composition $d(g(\bfz;\bftheta); \bfpsi)$
 %the derivatives of $d$ are known exactly.
-
-%We note that the technique can be used either to fit a generative model to an observed data distribution
-%or to fit a fast simulator to a slower, higher-fidelity simulator.
-
-%\kcnote{could also use this to tune non-differentialble fast simulator to slower full simulator. Then fixed data size not an issue}
-
-%\kcnote{proposal distribution $\to$ variational distribution}
 

### 2017-07-18 tex/paper.tex
save

@@
 posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
-Effectively, we are optimizing the variational posterior $q(\bftheta | \bfpsi)$ by
+In our case, we can identify the proposal distribution with the variational posterior
+$q(\bftheta | \bfpsi)$, which we effectively optimize by
 directly minimizing the  divergence of the corresponding posterior predictive
@@
 updated to make it more comparable to the latter. Relevant work in this
-direction includes those that rely on a classifier to estimate the discrepancy betwen the observed data and the model distributions
+direction includes those that rely on a classifier to estimate the discrepancy between the observed data and the model distributions
 \citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.

### 2017-07-19 tex/paper.tex
save

@@
 estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
-$U(\bfpsi)$ with stochastic gradient descent. We will use
-Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to make  use instead of the Natural Evolution Strategy
-algorithm~\citep{2011arXiv1106.4487W} or of variance reduction techniques~\citep{ranganath2014black}.
-%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems.
+$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, robust to noise
+and parameter scaling).
+% We will use
+% Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to make  use instead of the Natural Evolution Strategy
+% algorithm~\citep{2011arXiv1106.4487W} or of variance reduction techniques
+% such as Rao-Blackwellization,
+% Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
+%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the
+%gradients may be too large to be useful in many problems.
 Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.
@@
 
+% Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
+% variance of the noisy estimator of the gradients may be too large to be useful in many problems.
+% Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
+% Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
+% Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
+% empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
+% %Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
+% These variants may be particularly useful for small $N$.
+
 Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
-variance of the noisy estimator of the gradients may be too large to be useful in many problems. Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
+variance of the noisy estimator of the gradients may be too large to be useful in many problems.
+We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity use instead the Natural Evolution Strategy
+algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques
+such as Rao-Blackwellization,
 Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
-Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
-empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
-%Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
-These variants may be particularly useful for small $N$.
 
@@
 misspecified, $\bftheta^*$ is typically different from the maximum likelihood
-estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx |
+estimator (MLE), which is the minimizer of $\text{KL}(p_r(\bfx) \parallel p(\bfx |
 \bftheta))$~\citep{bernton2017inference}. In order to more effectively target
@@
 We refer to the algorithm with $\gamma>0$ as AVO-MWE.
+
 For small values of $\gamma$,
@@
 their Wasserstein rejection sampler incurred from using the ABC-likelihood.
+In the case of small $N$, we note however that fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
+empirical distributions~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}
+are alternatives to adversarial learning that are worth considering.
+
+% Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
+% empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
+% %Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
+% These variants may be particularly useful for small $N$.
 

### 2017-07-21 code/toy-gaussian.py
clean code

@@ -1,196 +1,0 @@
-import autograd as ag
-import autograd.numpy as np
-import matplotlib.pyplot as plt
-from scipy.stats import norm
-
-from nn import glorot_uniform
-from nn import relu
-from nn import AdamOptimizer
-
-from proposals import make_gaussian_proposal
-from proposals import gaussian_draw
-from proposals import gaussian_logpdf
-from proposals import grad_gaussian_logpdf
-from proposals import grad_gaussian_entropy
-
-from sklearn.utils import check_random_state
-
-# Global params
-
-batch_size = 64
-n_epochs = 300+1
-lambda_gp = 0.025
-gamma = 5.0
-
-true_theta = np.array([0.5, np.log(0.5)])
-make_plots = True
-
-
-# Simulator
-
-def simulator(theta, n_samples, random_state=None):
-    return norm.rvs(size=n_samples,
-                    loc=theta[0],
-                    scale=np.exp(theta[1]),
-                    random_state=random_state).reshape(-1, 1)
-
-X_obs = simulator(true_theta, 20000, random_state=123)
-n_params = len(true_theta)
-n_features = X_obs.shape[1]
-
-
-# Proposal distribution
-
-params_proposal = make_gaussian_proposal(n_params, mu=0.0)
-
-
-# Critic
-
-def make_critic(n_features, n_hidden, random_state=None):
-    rng = check_random_state(random_state)
-    params = {"W": [glorot_uniform(n_hidden, n_features, rng),
-                    glorot_uniform(n_hidden, n_hidden, rng),
-                    glorot_uniform(n_hidden, 0, rng)],
-              "b": [np.zeros(n_hidden),
-                    np.zeros(n_hidden),
-                    np.zeros(1)]}
-    return params
-
-params_critic = make_critic(n_features, 10)
-
-
-def predict(X, params):
-    h = X
-    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
-    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.1)
-    h = np.dot(params["W"][2], h.T).T + params["b"][2]
-    return h
-
-grad_predict_critic = ag.elementwise_grad(predict)
-
-
[…]

### 2017-07-22 tex/paper.tex
contour levels
fix rajesh small comments

@@
 program around implicit generative models based on neural
-networks~\citep{2016arXiv161003483M}.  While these implicit generative networks
-also do not admit a tractable density, they are differentiable.
-Generative models based on neural networks are highly parametrized and the model
+networks~\citep{2016arXiv161003483M}.  While some of these models
+also do not admit a tractable density, they are all differentiable by construction.
+In addition, generative models based on neural networks are highly parametrized and the model
 parameters have no obvious interpretation. In contrast, scientific simulators
@@
 approximate the posterior, one would criticize the model through the
-posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
+posterior predictive distribution~\citep{box1980sampling,rubin1984bayesianly}.
 In our case, we can identify the proposal distribution with the variational posterior
@@
 Consequently, the proposed algorithm is a form of variational inference through
-model criticism~\citep{box1980sampling}.
+model criticism.
 
@@
              ({\it Left}) Density $q(\bftheta|\bfpsi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
+                          Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
              ({\it Right}) Density $q(\bftheta|\bfpsi)$ after adversarial variational optimization ($\gamma=10$).
@@
     ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
+                     Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
                      The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\bftheta^*$.

### 2017-07-22 tex/paper.tex
Fixed Empirical Bayes confusion

--- 2017-07-19 tex/paper.tex
@@
 
-\subsection{Empirical Bayes through  Variational Inference}
+\subsection{Point Estimates through Marginal Likelihood}
+
+%\subsection{Empirical Bayes through  Variational Inference}
 
@@
 the parametrized marginal distribution of the generated data
-\begin{equation}
+\begin{equation}\label{eq:marginal_likelihood}
     \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
@@
 
-%Typically in variational inference one would has access to the posterior $p(\bftheta | \bfx )$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
-
-Typically in variational inference one starts with a tractable prior
-$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
-posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
-approximate the posterior, one would criticize the model through the
-posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
-In our case, we can identify the proposal distribution with the variational posterior
-$q(\bftheta | \bfpsi)$, which we effectively optimize by
-directly minimizing the  divergence of the corresponding posterior predictive
-distribution $\qxpsi$ and the data distribution $p_r(\bfx)$, without evaluating
-the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
-Consequently, the proposed algorithm is a form of variational inference through
-model criticism~\citep{box1980sampling}.
-
-We can also view the optimization  through the lens of empirical Bayes (EB),
-where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
-%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
-%posterior predictive distribution.
-Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
- $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi) d\bftheta $
-as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly
-specified a prior family. However, the procedure corresponds to empirical Bayes if we identify
-$q(\bftheta | \bfpsi^*) = p(\bftheta | \bfx , \bfpsi)$. Thus we call our method AVO-EB and note that it is applicable even outside of the likelihood-free context.
-
-Algorithm~\ref{alg:avo}, can be easily modified to have different sizes of mini-batches,
-$M_r$ for the observed data and $M_g$ for the simulator. When $M_g \gg M_r$,
-then the variance of $\bftheta^*$ scales as
-$\sqrt{M_r}$~\citep{bernton2017inference}. A final $\sqrt{N/M}$ scaling to the
-log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of
-the data.
+%where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
+%%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
+%%posterior predictive distribution.
+
+%%Typically in variational inference one would has access to the posterior $p(\bftheta | \bfx )$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
+%
+%Typically in variational inference one starts with a tractable prior
+%$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
+%posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
+%approximate the posterior, one would criticize the model through the
+%posterior predictive distribution as advocated by Rubin~\citep{rubin1984, box1980sampling}.
+%In our case, we can identify the proposal distribution with the variational posterior
+%$q(\bftheta | \bfpsi)$, which we effectively optimize by
+%directly minimizing the  divergence of the corresponding posterior predictive
+%distribution $\qxpsi$ and the data distribution $p_r(\bfx)$, without evaluating
+%the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
+%Consequently, the proposed algorithm is a form of variational inference through
+%model criticism~\citep{box1980sampling}.
+%
+%%% KC: Confusion here about $q(\bftheta | \bfpsi)$ as posterior or prior.
+%%% need differentiate xnew vs. xold and q_new and q_old
+%%% Seems like there is still a route to get at likelihood or posterior
+%%% Could be that issue is we loose normalizing constant between q_old(theta|psi) and likleihood
+%
+%We can also view the optimization  through the lens of empirical Bayes (EB),
+%where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
+%%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
+%%posterior predictive distribution.
+%Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
+% $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi) d\bftheta $
+%as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly
+%specified a prior family. However, the procedure corresponds to empirical Bayes if we identify
+%$q(\bftheta | \bfpsi) = p(\bftheta | \bfx , \bfpsi)$. Thus we call our method AVO-EB and note that it is applicable even outside of the likelihood-free context.
+
+%%KC: Not necessary if we aren't interpreting as posterior
+%Algorithm~\ref{alg:avo}, can be easily modified to have different sizes of mini-batches,
+%$M_r$ for the observed data and $M_g$ for the simulator. When $M_g \gg M_r$,
+%then the variance of $\bftheta^*$ scales as
+%$\sqrt{M_r}$~\citep{bernton2017inference}. A final $\sqrt{N/M}$ scaling to the
+%log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of
+%the data.
 
@@
 
-\subsection{Point estimates}
-
-%\kcnote{connect to MWE and MEWE here?}
-
-When $\rho$ is the Wasserstein distance, $\bftheta^*$ is referred to as the
-minimum Wasserstein estimator (MWE). When the model is well specified, $\bftheta^*$
+We can view the optimization of the marginal likelihood with respect to $\bfpsi$ through the lens of empirical Bayes (EB), where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
+If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\citep{rubin1984}.
+When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
+minimum Wasserstein estimator (MWE). 
+When the model is well specified, the MWE
 coincides with the true data-generating parameter; however, if the model is
-misspecified, $\bftheta^*$ is typically different from the maximum likelihood
-estimator (MLE), which is the minimizer of $\text{KL}(p_r(\bfx) \parallel p(\bfx |
-\bftheta))$~\citep{bernton2017inference}. In order to more effectively target
+misspecified, the MWE is typically different from the maximum likelihood
+estimator (MLE).
+Moreover, if the simulator $p(\bfx |
+\bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal
+model $q(\bfx | \bfpsi)$ is closer to $p_r(\bfx)$. However, if the simulator is well specified, 
+then $q(\bftheta | \bfpsi)$ will concentrate its mass around the true data-generating parameter.
+
+%When the model is well specified, $\bfpsi^*$
+%coincides with the true data-generating parameter; however, if the model is
+%misspecified, $\bftheta^*$ is typically different from the maximum likelihood
+%estimator (MLE), which is the minimizer of $\text{KL}(p_r(\bfx) \parallel p(\bfx |
+%\bftheta))$~\citep{bernton2017inference}.
+
+In order to more effectively target
 point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
@@
 between the generator objective and the tightness of the proposal distribution.
-We refer to the algorithm with $\gamma>0$ as AVO-MWE.
+%We refer to the algorithm with $\gamma>0$ as AVO-MWE.
 
@@
 proposal distributions with large entropy are not penalized, which results
-in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
+in learning a smeared variation of the original simulator corresponding to the marginal or prior predictive model. On the other hand,
 for large values of $\gamma$, the procedure is encouraged to fit a proposal
@@
     ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
-The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\bftheta^*$.
-    ({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
+                     Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
+                     The density of the penalized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
+    ({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
     ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
@@
 landscape around $\bftheta^*$ or that the observed data can in this case also be
-reproduced with the posterior predictive distribution $\qxpsi$.
+reproduced with the  predictive distribution $\qxpsi$.
 Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both

### 2017-07-22 tex/paper.tex

@@
 
-We can view the optimization of the marginal likelihood with respect to $\bfpsi$ through the lens of empirical Bayes (EB), where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
+We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes (EB), where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
 If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\citep{rubin1984}.
@@
 estimator (MLE).
-Moreover, if the simulator $p(\bfx |
+Thus, if the simulator $p(\bfx |
 \bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal

### 2017-07-22 tex/paper.tex
Tweaks to summary

@@
 variational inference that works in a likelihood-free setting. Thirdly, the
-AVO-EB algorithm can be viewed as a form of empirical Bayes where the prior is
+AVO algorithm can be viewed as a form of empirical Bayes where the prior is
 optimized based on the data.
@@
 upper bounds  on the otherwise non-differentiable adversarial objectives. The
-AVO-EB algorithm enables empirical Bayes through variational inference in the
+AVO algorithm enables empirical Bayes through variational inference in the
 likelihood-free setting. This approach does not incur the inefficiencies of an
-ABC-like rejection sampler. The AVO-MWE algorithm provides point estimates for
+ABC-like rejection sampler. When the model is well-specified, the AVO algorithm 
+provides point estimates for
 the generative model, which asymptotically corresponds to the data generating
-parameters when the model is well-specified. Both algorithms work on continuous
-or discrete observed data.
+parameters. The algorithm works on continuous or discrete data.
 

### 2017-07-22 tex/paper.tex
tweak abstract

@@
 function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
-inference algorithm for fitting a non-differentiable generative model or to perform empirical Bayes through variational inference.
+inference algorithm for fitting a non-differentiable generative model incorporating ideas from empirical Bayes and variational inference.
 We adapt the training procedure of generative
@@
 minimax problem by minimizing variational upper bounds of the two adversarial objectives.
-Effectively, the procedure results in learning an arbitrarily tight
+Effectively, the procedure results in learning a
 proposal distribution over simulator parameters, such that the corresponding

### 2017-07-22 tex/paper.tex
from two to one algorithm

@@
 
-In this note, we develop two likelihood-free inference algorithms
-for non-differentiable, implicit generative models. The first allows us
-to perform empirical Bayes through variational inference, and the
-second provides a point estimator of the parameters $\bftheta$.
+In this note, we develop a likelihood-free inference algorithm
+for non-differentiable, implicit generative models.
+% The first allows us
+% to perform empirical Bayes through variational inference, and the
+% second provides a point estimator of the parameters $\bftheta$.
 We adapt the adversarial
@@
 optimize $\bfpsi$ with stochastic gradient descent. Additionally,
-our AVO-EB procedure avoids the inefficiencies of
+our procedure avoids the inefficiencies of
 their Wasserstein rejection sampler incurred from using the ABC-likelihood.
@@
 
-In this note, we develop likelihood-free inference algorithms for
-non-differentiable, implicit generative models. The algorithms combine
+In this note, we develop a likelihood-free inference algorithm for
+non-differentiable, implicit generative models. The algorithm combines
 adversarial training  with variational optimization to minimize variational

### 2017-07-22 tex/paper.tex
small tweaks

@@
 \section{Adversarial variational optimization}
+
+\subsection{Algorithm}
 
@@
 
-We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes (EB), where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
+We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
 If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\citep{rubin1984}.

### 2017-07-28 tex/paper.tex
Some cleanup

@@
 \begin{abstract}
+
 Complex computer simulators are increasingly used across fields of science as
@@
 
-%Computer simulators are used to describe complex data generation processes in a diverse set of fields of science including particle physics, climatology, epidemiology,  and population
-%genetics,
-
 In many fields of science such as particle physics, epidemiology,  and population
@@
 In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
-% and may generate either discrete or continuous data.
 Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx |
@@
 sisson2011likelihood, marin2012approximate, cranmer2015approximating}.
-% Often
-% the simulation code involves control-flow that implies the dependence of $\bfx$
-% on $\bfz$ is non-differentiable, making inference even more difficult and
-% precludes approaches such as Ref.~\citep{2016arXiv160507826G}.
-
-%\kcnote{HMC https://arxiv.org/pdf/1503.01916.pdf}
-
-%\kcnote{add ABC refs mentioned in tex}
-%(Beaumont et al., 2002; Marjoram et al., 2003; Sisson et al., 2007; Sisson \& Fan, 2010; Marin et al., 2012; Fan et al., 2013) }
-
-%Often the simulation code involves control-flow that leads to models such that even $p(\bfx | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.
 
@@
 
-%%In contrast, scientific simulators involve control-flow that leads to models such that $p(\bfx, \bfz | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.
-%Scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters  endowed with some level of interpretation.
-In contrast, generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation.
-In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.
-
-%While generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation, scientific simulators typically have relatively few parameters that endowed with some level of interpretation.
-Thus simulators can be thought of as highly regularized, non-differentiable generative models.
-In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latents $\bfz$.
-
-%Generative models based on neural networks are differentiable with respect to their high dimensional model parameters $\bftheta$, which  with no particular , differentiable with respect to $\bfz$ and $\bftheta$, and the distribution of the latent $\bfz$
-%
-%In the context of variational inference, one assumes the the likelihood $p(\bfx | \bftheta)$ is available and posterior $p(\bftheta | \bfx)$ is tractable up to a constant (ie.
-($p(\bfx, \bfz | \bftheta)$ is tractable).
-Moreover, one is typically interested in estimating the posterior distribution of the latent variables instead of the parameters of the simulator $\bftheta$.
-
-%In the context of variational inference, these technihave also been used as the recognition model ~\citep{2015arXiv150505770J, DBLP:journals/corr/KingmaSW16, DBLP:conf/icml/RezendeMW14}; however, in those cases the likelihood $p(\bfx | \bftheta)$ is available and posterior $p(\bftheta | \bfx)$ is tractable up to a constant.
-
-%Because it is usually computationally intractable, most simulators
-%
-%implicit generative models~\citep{2016arXiv161003483M}
-%
-%In most
-%cases, these implicit generative models~\citep{2016arXiv161003483M} are
-%specified as procedural implementations of forward stochastic processes that
-%generates discrete or continuous data, and rarely do these simulators admit a tractable density. Because it is usually computationally intractable, most simulators
-%do not provide a way to directly evaluate the likelihood function for a
-%given observation, thereby making inference difficult.
-%In addition, most scientific simulators are written using opaque low-level programming
-%languages, hence raising the bar for likelihood-free inference
-%algorithms relying e.g. on the simulator being a function with computable derivatives~\citep{2016arXiv160507826G}
-%or being a controllable probabilistic program~\citep{2016arXiv161009900L}.
-
 In this note, we develop a likelihood-free inference algorithm
 for non-differentiable, implicit generative models.
-% The first allows us
-% to perform empirical Bayes through variational inference, and the
-% second provides a point estimator of the parameters $\bftheta$.
 We adapt the adversarial
@@
 the generated data to the empirical distribution of the observations.
-
-%The procedure results in learning an arbitrarily tight
-%proposal distribution over simulator parameters, such that the corresponding
-%marginal distribution of the generated data matches the observations.
-
-%In this note, we develop an unsupervised learning algorithm for the point
-%estimation of the parameters $\bftheta$ of a non-differentiable, implicit
-%generative model. We adapt the adversarial
-%training procedure of generative adversarial
-%networks~\cite{goodfellow2014generative} by replacing the implicit generative
-%network with a domain-based scientific simulator, and solve the resulting
-%non-differentiable minimax problem by minimizing variational upper
-%bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
-%objectives.
-%The procedure results in learning an arbitrarily tight
-%proposal distribution over simulator parameters, such that the corresponding
-%marginal distribution of the generated data matches the observations.
-
-%In this note, we develop a likelihood-free inference algorithm for the point
-%estimation of the parameters $\bftheta$ of a non-differentiable, implicit
-%generative model. We adapt the adversarial
-%training procedure of generative adversarial
-%networks~\cite{goodfellow2014generative} by replacing the implicit generative
-%network with a domain-based scientific simulator, and solve the resulting
-%non-differentiable minimax problem by minimizing variational upper
-%bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
-%objectives. The procedure results in learning a proposal distribution over
-%simulator parameters, hence producing an arbitrarily tight family of models whose
-%marginal distribution of generated data matches the observed data.
 
@@
 process
-%, such that
-%\begin{equation}\label{eqn:p_x}
-%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
-%\end{equation}
-%\begin{equation}\label{eqn:p_x_sim}
-%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
-%\end{equation}
 where $\bfz \in {\cal Z}$ is a latent variable providing an external
@@
 $\bfx$.
-%Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.
-
-%Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
-%be a fixed-size vector (e.g., it can be a sequence of variable length) and its
-%distribution $p(\bfz|\bftheta)$ may itself depend on $\bftheta$ in some intricate way
-%inherited from the control flow of the simulation code.
 
@@
 written as
-%considered as a non-differentiable reparametrization of the random variable $\bfz$
-%\begin{equation}\label{eqn:p_x_sim}
-%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g_i(\bfz;\bftheta) \leq x_i\}} p(\bfz|\bftheta) \mu(d\bfz),
-%\end{equation}
 \begin{equation}\label{eqn:p_x_sim}
@@
 where $\mu$ is a probability measure.
-%Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
-%used to generate data in forward mode. For this reason, evaluating the integral
-%in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
-%in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
-%e.g. as when $g$ is specified as a computer program.
 
@@
 where $\rho$ is some distance or divergence.
-%~\citep{bernton2017inference}.
-%For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate if $ \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta^*)) \to 0$.
-
-%By plugging Wp in place of D in Eqs.
-(1) and (2), we obtain the minimum Wasserstein estimator (MWE) and minimum expected Wasserstein estimator (MEWE) of order p, denoted ??n and ??n,m respectively.
-Some properties of the MWE have been studied in Bassetti et al.
-(2006), for well-specified models and i.i.d.
-data; we propose new results in Section 4.
-Intuitively, under some conditions we can expect ??n to converge to ??, in the sense that Wp(??n,??) ?
-0.
-Consequently, the minimum of ?
-?
-Wp(??n,??) might converge to the minimum of ?
-?
-Wp(??,??), denoted by ??, assuming its existence and unicity.
-In the well-specified case, ??
-coincides with the data-generating parameter.
-In the misspecified case, ??
-is typically different from the limit of the maximum likelihood estimator (MLE), which is the minimizer of KL(??|??).
 
@@
 by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
-$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, with
+$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
 \begin{align}
@@
 \end{align}
-%
-%\begin{align}
-%    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
-%    {\cal L}_g =\,& - {\cal L}_W
-%\end{align}
-%\begin{align}
-%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]  \nonumber \\
-%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
-%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)]
-%\end{align}
 where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
@@
 and parameter scaling).
-% We will use
-% Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to make  use instead of the Natural Evolution Strategy
-% algorithm~\citep{2011arXiv1106.4487W} or of variance reduction techniques
-% such as Rao-Blackwellization,
-% Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
-%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the
-%gradients may be too large to be useful in many problems.
-Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.
-
-% In this simple form, stochastic gradient descent based on
-% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
-% of scale invariance. To avoid the dependence on the parametrization of the
-% proposal distribution, natural gradients~\citep{amari1998natural} can be used
-% instead to point towards the steepest descent in the space of realizable
-% proposal distributions, rather than in the space of their parameters. Effectively,
-% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
-% corrects the gradient of $U$ according to the local curvature of the KL-divergence
-% surface of proposal distributions, which results in forming
-% \begin{equation}\label{eqn:approx-grad-nat}
-%     \tilde{\nabla}_\bfpsi U(\bfpsi) = \mathbf{F}(\bfpsi)^{-1} \nabla_\bfpsi U(\bfpsi),
-% \end{equation}
-% where $\mathbf{F}(\bfpsi)$ is the Fisher information matrix
-% \begin{equation}
-%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\bfphi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
-% \end{equation}
-% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\bfpsi \log q_\bfpsi(\bftheta)$ already
-% evaluated to form $\nabla_\bfpsi U(\bfpsi)$.
 
@@
 That is,
-%\begin{equation}\label{eqn:grad-ug-approx}
-%   \nabla_\bfpsi U_g = \mathbb{E}_{\subarray{l}{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta) }}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
-%\end{equation}
 \begin{equation}\label{eqn:grad-ug-approx}
@@
 \end{equation}
-%\begin{equation}\label{eqn:grad-ug-approx}
-%    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
-%\end{equation}
 which we can approximate with mini-batches of
@@
 optimization (AVO) procedure, as built on top of WGAN-GP.
-%
-%Obviously, the variational relaxation could similarly be coupled with
-%other variants of GANs and/or of evolution strategies.
-
-%Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
-%have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
-%a distribution parametrized in terms of $\bfpsi$:
-%\begin{equation}\label{eqn:p_psi}
-%    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
-%\end{equation}
-%Intuitively, this corresponds to a family of simulators, each configured
-%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
-%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
-%More formally, the learned model  therefore corresponds to the marginal distribution
-%of the generated data. That is,
-%\begin{equation}
-%    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
-%\end{equation}
-%Intuitively, this corresponds to a family of simulators, each configured
-%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
-%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
-
-% Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
-% variance of the noisy estimator of the gradients may be too large to be useful in many problems.
-% Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
-% Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
-% Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
-% empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
-% %Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
-% These variants may be particularly useful for small $N$.
 
@@
 variance of the noisy estimator of the gradients may be too large to be useful in many problems.
-We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity use instead the Natural Evolution Strategy
+We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
 algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques
@@
 \subsection{Point Estimates through Marginal Likelihood}
-
 %\subsection{Empirical Bayes through  Variational Inference}
@@
 
-%where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
-%%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
-%%posterior predictive distribution.
-
-%%Typically in variational inference one would has access to the posterior $p(\bftheta | \bfx )$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
-%
-%Typically in variational inference one starts with a tractable prior
-%$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
-%posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
-%approximate the posterior, one would criticize the model through the
-%posterior predictive distribution as advocated by Rubin~\citep{rubin1984, box1980sampling}.
-%In our case, we can identify the proposal distribution with the variational posterior
-%$q(\bftheta | \bfpsi)$, which we effectively optimize by
-%directly minimizing the  divergence of the corresponding posterior predictive
-%distribution $\qxpsi$ and the data distribution $p_r(\bfx)$, without evaluating
-%the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
-%Consequently, the proposed algorithm is a form of variational inference through
-%model criticism~\citep{box1980sampling}.
-%
-%%% KC: Confusion here about $q(\bftheta | \bfpsi)$ as posterior or prior.
-%%% need differentiate xnew vs. xold and q_new and q_old
-%%% Seems like there is still a route to get at likelihood or posterior
-%%% Could be that issue is we loose normalizing constant between q_old(theta|psi) and likleihood
-%
-%We can also view the optimization  through the lens of empirical Bayes (EB),
-%where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
-%%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
-%%posterior predictive distribution.
-%Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
-% $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi) d\bftheta $
-%as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly
-%specified a prior family. However, the procedure corresponds to empirical Bayes if we identify
-%$q(\bftheta | \bfpsi) = p(\bftheta | \bfx , \bfpsi)$. Thus we call our method AVO-EB and note that it is applicable even outside of the likelihood-free context.
-
-%%KC: Not necessary if we aren't interpreting as posterior
-%Algorithm~\ref{alg:avo}, can be easily modified to have different sizes of mini-batches,
-%$M_r$ for the observed data and $M_g$ for the simulator. When $M_g \gg M_r$,
-%then the variance of $\bftheta^*$ scales as
-%$\sqrt{M_r}$~\citep{bernton2017inference}. A final $\sqrt{N/M}$ scaling to the
-%log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of
-%the data.
-
-%While the
-%likelihood is intractable, it is fixed, thus one can view the optimization of
-%$\bfpsi$ through the lens of empirical Bayes, though the prior family
-%$p(\bftheta | \bfpsi)$ is not explicitly specified,
-%ie. $q(\bftheta | \bfpsi^*) = p(\bfx | \bftheta) p(\bftheta | \bfpsi^*) / \int d\bftheta p(\bfx | \bftheta) p(\theta | \bfpsi^*)$.
-
-%\kcnote{where is the prior? comment}
-
-%Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by minimizing the
-%divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$.
-%%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
-%Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.
-
-%Consequently, the proposed inference algorithm does not necessarily guarantee that the
-%proposal distribution $q(\bftheta|\bfpsi)$ will place its mass arbitrarily tight
-%around the parameters of interest, which might be an issue when one is rather interested in point estimates $\bftheta^*$.
-%For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
-
 We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
@@
 
-%When the model is well specified, $\bfpsi^*$
-%coincides with the true data-generating parameter; however, if the model is
-%misspecified, $\bftheta^*$ is typically different from the maximum likelihood
-%estimator (MLE), which is the minimizer of $\text{KL}(p_r(\bfx) \parallel p(\bfx |
-%\bftheta))$~\citep{bernton2017inference}.
-
 In order to more effectively target
@@
 between the generator objective and the tightness of the proposal distribution.
-%We refer to the algorithm with $\gamma>0$ as AVO-MWE.
 
@@
 typically increases as the entropy of the proposal decreases.
-%\kcnote{maybe, but not obvious why}
 
-%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems.
-Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here analogous to their use in black box variational inference~\citep{ranganath2014black}.
-
-%\newpage
 % ==============================================================================
@@
 
-%Given the
-%known analytic form of the probability density function, a simulator for
-%$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
-%arbitrary number of random drawings may be required before passing the
-%acceptance threshold, this illustrates how the latent variable
-%$\bfz$ does not necessarily need to be fixed-sized vector, and how its
-%density  $p(\bfz|\bftheta)$ may depend on $\bftheta$ in some complicated way
-%(Section~\ref{sec:problem}).
-
 In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
@@
 cannot distinguish from the true distribution.
-%In summary, despite the intricate
-%relation between the observations $\mathbf{x}$, the latent variable $\bfz$
-%and the parameters $\bftheta$, this example demonstrates the efficacy
-%of AVO for parameter inference on a black box
-%simulator.
 
@@
 implicit generative networks, and variational inference.
-%Our focus
-%is on extending existing work to non-differentiable generative models
-%that do not admit a tractable density.
 Viewed from the literature around implicit generative models based on neural networks,
@@
 optimized based on the data.
-
-%Fourthly, the AVL algorithm is not a Bayesian inference algorithm, but instead provides point estimates $\bftheta^*$.
 
@@
 are alternatives to adversarial learning that are worth considering.
-
-% Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
-% empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
-% %Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
-% These variants may be particularly useful for small $N$.
-
-%\kcnote{We don't use ABC-likelihood (more efficient?) used here and in VBIL.
-We also don't specify a prior b/c it's like implicit empirical bayes.
-They also don't specify a recognition model to latents.
-}
-
-%\kcnote{Could we actually replace the optimization of the critic with a direct Wasserstein distance calculation in fewer dimensions.
-If the data is drawn from the variational program, it would still be differentiable with respect to $\bfpsi$.
-No, because it still requires $\nabla_\psi \log p(\bfx | \bfpsi)$}
-
-%Within the literature on ABC and likelihood-free inference,
-%the closest to our work are procedures that rely on a classifier to estimate the
-%discrepancy between the observed data and the model distributions. For example,
-%\citep{gutmann2012noise} uses non linear logistic regression for fitting
-%unnormalized differentiable statistical models. Similarly, \citep{cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D} estimate likelihood
-%ratios through supervised classification, which can in turn be used for
-% Bayesian or frequentist inference. Similarly, \citep{gutmann2017likelihood}
-% makes use of classification accuracy as
-%a summary statistics for approximate Bayesian computation.
 
@@
 recognition model.
-%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.
-
-%%In this context, the proposed method can be considered as a direct adaptation of
-%%generative adversarial networks~\citep{goodfellow2014generative} to
-%%non-differentiable simulators.
-%It also constitutes an approximate gradient
-%descent alternative to likelihood-free inference based on bayesian optimization and approximated density
-%ratios~\citep{cranmer2015approximating,cranmer2016experiments} or to classifier
-%ABC~\citep{gutmann2017likelihood}.
-
-%This work has many connections to work on variational inference, in which the goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it is close to the true posterior $p(\bfz, \bftheta |x)$.
-%This is slightly different from variational optimization (or variational learning), where the goal is to find $\bftheta^*$ such that $p(\bfx|\bftheta^*) \approx p_r(\bfx)$.
-Typically in variational inference the model $p(\bfx|\bftheta)$ admits a density, so that the likelihood is known.
-Thus, assuming the prior also admits a density, it is tractable to evaluate the posterior is (up to a constant factor).
-In Adversarial Variational Bayes \citep{DBLP:journals/corr/MeschederNG17} and, equivalently, the \texttt{PC-Adv} algorithm of  \citep{2017arXiv170208235H}), the authors use an adversarial setup, but assume a known density $p(\bfx|\bftheta)$ that is differentiable with respect to $\bftheta$.
 
@@
 the ABC likelihood for high-dimensional data and small $\epsilon$.
-
-%\kcnote{do we keep JC-Adv  notes? Should say something about Hamiltonian ABC}
-%In \citep{2017arXiv170208235H} the author defines implicit distributions to be ``probability models whose probability density function may be intractable, but there is a way to 1) sample from them exactly and/or calculate and approximate expectations under them, and 2) calculate or estimate gradients of such expectations with respect to model parameters.'' This notion of implicit model does not include most scientific simulators as 2) is not satisfied.
-Furthermore, the  \texttt{JC-Adv} and \texttt{JC-Den} algorithm defined there applies to implicit models, but, as the author points out, is aimed at variational inference and ``does not provide a direct way to learn model parameters $\bftheta$''.
-%``Unfortunately, none of the algorithms presented here allow for fully variational learning
-%of model parameters $\bftheta$ when $p_\bftheta(\bfx|\bfz)$ is implicit.''
-%%JC-Adv doesn't let you optimize theta, theta is fixed and you get variational inference on z. Need a recognition model for latents
-
-%\kcnote{F.H. Implicit model is differentiable. Is what we are doing JC-Adv with our variational program}
-
-%\subsection{Operator Variational Optimization}
-
-%Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
-%which is a generalization of variational inference formulated as the following optimization problem
-%\begin{eqnarray}
-%\bflambda^* = \inf_\bfpsi \sup_\bflambda \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
-%\end{eqnarray}
-%In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
-%AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
-%\begin{eqnarray} \label{eq:opvi_avo}
-%\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \\
-%\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \;.
-%\end{eqnarray}
-%Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling
-%and reinforces the link to density ratio estimation as this corresponds to the operator
-%\begin{equation}
-%(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, .
-%\end{equation}
 
@@
 which reinforces the link to density ratio estimation and inference with implicit models.
-
-%
-%%\begin{equation}
-%%(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
-%%\end{equation}
-%
-%
-%%This is natural here, but not quite OPVI formalism:
-%%\begin{equation}
-%%\bfpsi^* = \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bftheta \sim q(\bftheta | \bfpsi)} [ (O^{p,q} f_\bfphi) ]
-%%\end{equation}
-%%
-%%\begin{equation}
-%%(O^{p,q} f) = W(p,q) = \sup_f \mathbb{E}_{x\sim p}[f(x)] - \mathbb{E}_{x \sim q}[f(x)]
-%%\end{equation}
-%%The expectation should be $x \sim p(x|\bfpsi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$
-%
-%To get into OPVI formalism
-%
-%\begin{eqnarray}
-%\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi\mathbb{E}_{x\sim p_r(x)}[f(x)] - \mathbb{E}_{x \sim p(x|\bfpsi)}[f(x)] \\
-%&=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ x \sim p( x | \bfpsi)} [ (O^{p_r,p_\bfpsi} f_\bfphi) ]
-%\end{eqnarray}
-%with
-%\begin{equation}
-%(O^{p,q} f) = \left( \frac{p_r(x)}{p(x|\bfpsi)}  -1 \right) f(x)
-%\end{equation}
-
-%We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in ~\citep{2016arXiv161009033R}, though more complicated than the a simple reparametrization of normally distributed noise $\epsilon$ through a differentiable function $R(\epsilon, \bfpsi)$.
-In our case, the variational program is a marginalized non-differentiable simulator simulator.
-Nevertheless, it can generate samples for $x$, is differentiable with respect to $\bfphi$, and its density is be intractable.
 
@@
 the gradients $\nabla_\bfpsi U_g$ can be computed.
-%, in particular by exploiting the fact in the composition $d(g(\bfz;\bftheta); \bfpsi)$
-%the derivatives of $d$ are known exactly.
 

### 2017-07-28 code/nn.py

@@ -91,45 +91,3 @@ class AdamOptimizer:
         for k, v in self.unflatten(self.x).items():
             params[k] = v
         return params
-
-
-class RmsPropOptimizer:
-    def __init__(self, grad, init_params, callback=None,
-                 step_size=0.01, gamma=0.9, eps=10**-8):
-        self.grad = grad
-        self.init_params = copy.copy(init_params)
-        self.callback = callback
-        self.step_size = step_size
-        self.gamma = gamma
-        self.eps = eps
-
-        self.flattened_grad, self.unflatten, self.x = flatten_func(
-            self.grad, self.init_params)
-        self.reset()
-
-    def reset(self):
-        self.avg_sq_grad = np.ones(len(self.x))
-        self.counter = 0
-
-    def step(self,  num_iters=1):
-        for i in range(num_iters):
-            g = self.flattened_grad(self.x, self.counter)
-
-            if self.callback:
-                self.callback(self.unflatten(self.x),
-                              self.counter,
-                              self.unflatten(g))
-
-            self.avg_sq_grad = (self.avg_sq_grad *
-                                self.gamma + g**2 * (1 - self.gamma))
-            self.x = (self.x -
-                      self.step_size * g / (np.sqrt(self.avg_sq_grad) +
-                                            self.eps))
-            self.counter += 1
-
-        return self.unflatten(self.x)
-
-    def move_to(self, params):
-        for k, v in self.unflatten(self.x).items():
-            params[k] = v
-        return params

### 2017-07-28 code/toy-poisson-Mtest.py

@@ -1,237 +1,0 @@
-import autograd as ag
-import autograd.numpy as np
-import matplotlib.pyplot as plt
-from scipy.stats import poisson
-
-from nn import glorot_uniform
-from nn import relu
-from nn import AdamOptimizer
-
-from proposals import make_gaussian_proposal
-from proposals import gaussian_draw
-from proposals import gaussian_logpdf
-from proposals import grad_gaussian_logpdf
-from proposals import grad_gaussian_entropy
-
-from sklearn.utils import check_random_state
-
-# Global params
-
-seed = 777
-rng = check_random_state(seed)
-
-def_batch_size = 256
-n_epochs = 300+1
-#n_epochs = 100+1
-lambda_gp = 0.025
-
-true_theta = np.array([np.log(7)])
-make_plots = True
-
-
-# Simulator
-
-def simulator(theta, n_samples, random_state=None):
-    return poisson.rvs(np.exp(theta[0]),
-                       size=n_samples,
-                       random_state=random_state).reshape(-1, 1)
-
-X_obs = simulator(true_theta, 20000, random_state=rng)
-n_params = len(true_theta)
-n_features = X_obs.shape[1]
-
-
-# Critic
-
-#gammas = [0.0, 5.0]
-gammas = [0.0, 0.0, 0.0, 0.0]
-colors = ["C1", "C2", "C3","C4"]
-batch_sizes = [16, 64, 256, 1024]
-
-def make_critic(n_features, n_hidden, random_state=None):
-    rng = check_random_state(random_state)
-    params = {"W": [glorot_uniform(n_hidden, n_features, rng),
-                    glorot_uniform(n_hidden, n_hidden, rng),
-                    glorot_uniform(n_hidden, 0, rng)],
-              "b": [np.zeros(n_hidden),
-                    np.zeros(n_hidden),
-                    np.zeros(1)]}
-    return params
-
-def predict(X, params):
-    h = X
-    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
-    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.1)
-    h = np.dot(params["W"][2], h.T).T + params["b"][2]
-    return h
-
-grad_predict_critic = ag.elementwise_grad(predict)
-
-
-history = [{"gamma": gammas[i],
-            "color": colors[i],
-            "loss_d": [],
-            "batch_size" : batch_sizes[i],
[…]

### 2017-07-28 code/weinberg.py

@@ -9,7 +9,6 @@ from nn import AdamOptimizer
 
 from proposals import make_gaussian_proposal
 from proposals import gaussian_draw
-from proposals import gaussian_logpdf
 from proposals import grad_gaussian_logpdf
 from proposals import grad_gaussian_entropy
 
@@ -47,6 +46,7 @@ def diffxsec(costheta, sqrtshalf, gf):
     norm = 2. * (1. + 1. / 3.)
     return ((1 + costheta ** 2) + a_fb(sqrtshalf, gf) * costheta) / norm
 
+
 def rej_sample_costheta(n_samples, theta, rng):
     sqrtshalf = theta[0] * (50-40) + 40
     gf = theta[1] * (1.5 - 0.5) + 0.5
@@ -67,6 +67,7 @@ def rej_sample_costheta(n_samples, theta, rng):
 
     return np.array(samples)
 
+
 def simulator(theta, n_samples, random_state=None):
     rng = check_random_state(random_state)
     samples = rej_sample_costheta(n_samples, theta, rng)
@@ -83,6 +84,7 @@ n_features = X_obs.shape[1]
 gammas = [0.0, 5.0]
 colors = ["C1", "C2"]
 
+
 def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
     params = {"W": [glorot_uniform(n_hidden, n_features, rng),
@@ -92,6 +94,7 @@ def make_critic(n_features, n_hidden, random_state=None):
                     np.zeros(n_hidden),
                     np.zeros(1)]}
     return params
+
 
 def predict(X, params):
     h = X
@@ -117,7 +120,8 @@ history = [{"gamma": gammas[i],
 
 for state in history:
     # WGAN + GP
-    def loss_critic(params_critic, i, lambda_gp=lambda_gp, batch_size=batch_size):
+    def loss_critic(params_critic, i, lambda_gp=lambda_gp,
+                    batch_size=batch_size):
         y_critic = np.zeros(batch_size)
         # y_critic[:batch_size // 2] = 0.0  # 0 == fake
         y_critic[batch_size // 2:] = 1.0
@@ -125,7 +129,8 @@ for state in history:
         rng = check_random_state(i)
 
         # WGAN loss
-        thetas = gaussian_draw(state["params_proposal"], batch_size // 2, random_state=rng)
+        thetas = gaussian_draw(state["params_proposal"],
+                               batch_size // 2, random_state=rng)
[…]

### 2017-10-07 aistats/sample_paper.tex
preparing submission for aistats

--- 2017-07-28 tex/paper.tex
@@
-\documentclass[twocolumn,superscriptaddress,aps]{revtex4-1}
-
-\usepackage[utf8]{inputenc}
+\documentclass[twoside]{article}
+\usepackage{aistats2018}
+ % If your paper is accepted, change the options for the package
+% aistats2018 as follows:
+%
+%\usepackage[accepted]{aistats2018}
+%
+% This option will print headings for the title of your paper and
+% headings for the authors names, plus a copyright note at the end of
+% the first column of the first page.
 
@@
 \usepackage{color}
-\usepackage{hyperref}
 \usepackage{algorithm}
@@
 
-% ==============================================================================
+% If your paper is accepted and the title of your paper is very long,
+% the style will print as headings an error message. Use the following
+% command to supply a shorter title of your paper so that it can be
+% used as headings.
+%
+%\runningtitle{I use this title instead because the last one was very long}
 
-\title{\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}}
-\vspace{1cm}
-\author{\small{\bf Gilles Louppe}\thanks{\texttt{g.louppe@nyu.edu}}}
-\affiliation{New York University}
-\author{\small{\bf Kyle Cranmer}\thanks{\texttt{kyle.cranmer@nyu.edu}}}
-\affiliation{New York University}
+% If your paper is accepted and the number of authors is large, the
+% style will print as headings an error message. Use the following
+% command to supply a shorter version of the authors names so that
+% they can be used as headings (for example, use only the surnames)
+%
+%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
+
+\twocolumn[
+
+\aistatstitle{Adversarial Variational Optimization of\\
+              Non-Differentiable Simulators}
+
+\aistatsauthor{Author 1 \And Author 2}
+\aistatsaddress{Institution 1 \And Institution 2}
+
+% \aistatsauthor{Gilles Louppe \And Kyle Cranmer}
+%
+% \aistatsaddress{New York University\\
+%                 \& University of Li{\`e}ge
+%                 \And New York University}
+                ]
 
 \begin{abstract}
-
     Complex computer simulators are increasingly used across fields of science as
@@
     We present results of the method with simulators producing both discrete and continuous data.
-
 \end{abstract}
-
-\maketitle
 
@@
 In parallel, with the introduction of variational
-auto-encoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
+auto-encoders~\cite{DBLP:journals/corr/KingmaW13} and generative adversarial
 networks~\cite{goodfellow2014generative}, there has been a vibrant research
 program around implicit generative models based on neural
-networks~\citep{2016arXiv161003483M}.  While some of these models
+networks~\cite{2016arXiv161003483M}.  While some of these models
 also do not admit a tractable density, they are all differentiable by construction.
@@
 non-differentiable minimax problem by minimizing variational upper
-bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
+bounds~\cite{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
 objectives.
@@
 
-As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
+As thoroughly explored in \cite{2017arXiv170104862A}, GANs remain remarkably
 difficult to train because of vanishing gradients as $d$ saturates, or because of
 unreliable updates when the training procedure is relaxed. As a remedy,
-Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
+Wasserstein GANs~\cite{2017arXiv170107875A} reformulate the adversarial
 setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
@@
 
-Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
+Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\cite{2011arXiv1106.4487W} are general
 optimization techniques that can be used to form a differentiable bound
@@
 
-Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
+Under mild restrictions outlined in  \cite{2012arXiv1212.4507S}, the bound
 $U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
@@
 We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
-algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques
+algorithm~\cite{2011arXiv1106.4487W} or variance reduction techniques
 such as Rao-Blackwellization,
-Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
+Control Variates, or quasi-Monte Carlo~\cite{ranganath2014black, tran2017variational}.
 
@@
 We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
-in \citep{2016arXiv161009033R}, though more complicated than a simple
+in \cite{2016arXiv161009033R}, though more complicated than a simple
 reparametrization of normally distributed noise $\bfz$ through a differentiable
@@
 We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
-If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\citep{rubin1984}.
+If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\cite{rubin1984}.
 When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
-    \caption{Discrete Poisson model with unknown mean.
-             ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
-For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
-                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
-             ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
-             ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
-             }\label{fig:poisson}
-\end{figure}
-
 The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
+\caption{Discrete Poisson model with unknown mean.
+ ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
+            the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
+ ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
+ ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+ }\label{fig:poisson}
+\end{figure}
+
 \subsection{Multidimensional continuous data}
@@
 5-dimensional continuous data, as originally specified in Section 4.2. of
-\citep{cranmer2015approximating}. More specifically, we consider the following
+\cite{cranmer2015approximating}. More specifically, we consider the following
 generative process:
@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering
-using \texttt{MadGraph}~\citep{Alwall:2011uj},  ionization of matter in the
+using \texttt{MadGraph}~\cite{Alwall:2011uj},  ionization of matter in the
 detector due to the passage of the out-going $\mu^+\mu^-$ particles using
-\texttt{GEANT4}~\citep{Agostinelli:2002hh}, electronic noise and other details of the sensors
+\texttt{GEANT4}~\cite{Agostinelli:2002hh}, electronic noise and other details of the sensors
 that measure the ionization signal, and the deterministic algorithms that
@@
 the proposed method can be considered as a direct adaptation of
-generative adversarial networks~\citep{goodfellow2014generative,2017arXiv170104862A, 2017arXiv170107875A} to
+generative adversarial networks~\cite{goodfellow2014generative,2017arXiv170104862A, 2017arXiv170107875A} to
 non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization}.
@@
 approximation based on projection of the data onto Hilbert space-filling curves.
-Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
+Their Remark 5.1 points out that~\cite{montavon2016wasserstein} proposed an
 approximation of the gradient of an entropy-regularized Wasserstein distance,
@@
 In the case of small $N$, we note however that fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
-empirical distributions~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}
+empirical distributions~\cite{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}
 are alternatives to adversarial learning that are worth considering.
@@
 direction includes those that rely on a classifier to estimate the discrepancy between the observed data and the model distributions
-\citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
-Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\citep{meeds2015hamiltonian}
+\cite{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
+Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
 estimate gradients with respect
@@
 Likewise, AVO closely relates to recent extensions of GANs, such as
-Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
-Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
-$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
-(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
-algorithm of  \citep{2017arXiv170208235H}, which
+Adversarial Learned Inference (ALI)~\cite{dumoulin2016adversarially},
+Bidirectional GANs (BiGANs)~\cite{donahue2016adversarial},
+$\alpha$-GAN~\cite{rosca2017variational}, Adversarial Variational Bayes
+(AVB)~\cite{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
+algorithm of  \cite{2017arXiv170208235H}, which
 add an inference network to the generative model.  Each of these assume a
@@
 to extend variational inference to intractable likelihoods; however, they
-require restrictive assumptions.  In \citep{tran2017variational}, the authors
+require restrictive assumptions.  In \cite{tran2017variational}, the authors
 consider Variational Bayes with an Intractable Likelihood (VBIL). In that
@@
 
-Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
+Lastly, we make a connection to Operator Variational Inference (OPVI)~\cite{2016arXiv161009033R},
 which is a generalization of variational inference formulated as the following optimization problem:
@@
 
-\vspace{0.4cm}
-\section*{Acknowledgments}
+% \section*{Acknowledgements}
 
-We would like to thank Lukas Heinrich for helpful comments regarding
-the electron--positron annihilation simulation. We would also like to thank
-Rajesh Ranganath and Dustin Tran for enlightening discussions and feedback.
-GL and KL are both supported through NSF ACI-1450310, additionally KC is
-supported through PHY-1505463 and PHY-1205376.
+% We would like to thank Lukas Heinrich for helpful comments regarding
+% the electron--positron annihilation simulation. We would also like to thank
+% Rajesh Ranganath and Dustin Tran for enlightening discussions and feedback.
+% GL and KL are both supported through NSF ACI-1450310, additionally KC is
+% supported through PHY-1505463 and PHY-1205376.
 
-% ==============================================================================
-
-\vspace{2cm}
-
-%\bibliographystyle{hacm}
 \bibliographystyle{hunsrt}

### 2018-01-25 code/nn.py
save code

@@ -1,5 +1,5 @@
 import autograd.numpy as np
-from autograd.util import flatten_func
+from autograd.misc.flatten import flatten_func
 import copy
 
 

### 2018-01-25 code/toy-poisson.py

@@ -42,8 +42,11 @@ n_features = X_obs.shape[1]
 
 # Critic
 
-gammas = [0.0, 5.0]
-colors = ["C1", "C2"]
+# gammas = [0.0, 5.0]
+# colors = ["C1", "C2"]
+
+gammas = [5.0]
+colors = ["C2"]
 
 def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)

### 2018-01-25 icml/avo.tex
init icml version

--- 2017-10-07 aistats/sample_paper.tex
@@
-\documentclass[twoside]{article}
-\usepackage{aistats2018}
+\documentclass{article}
+\usepackage[accepted]{icml2018}
  % If your paper is accepted, change the options for the package
@@
 \usepackage{tikz}
+\usepackage{hyperref}
+
+% Attempt to make hyperref and algorithmic work together better:
+\newcommand{\theHalgorithm}{\arabic{algorithm}}
 
@@
 \newtheorem{proposition}[theorem]{Proposition}
+
+\icmltitlerunning{Adversarial Variational Optimization of Non-Differentiable Simulators}
 
@@
 
-\aistatstitle{Adversarial Variational Optimization of\\
+\icmltitle{Adversarial Variational Optimization of\\
            Non-Differentiable Simulators}
 
-\aistatsauthor{Author 1 \And Author 2}
-\aistatsaddress{Institution 1 \And Institution 2}
+\begin{icmlauthorlist}
+\icmlauthor{Glles Louppe}{uliege}
+\icmlauthor{Kyle Cranmer}{nyu}
+\end{icmlauthorlist}
 
-% \aistatsauthor{Gilles Louppe \And Kyle Cranmer}
-%
-% \aistatsaddress{New York University\\
-%                 \& University of Li{\`e}ge
-%                 \And New York University}
+\icmlaffiliation{uliege}{University of Li{\`e}ge, Belgium}
+\icmlaffiliation{nyu}{New York University, USA}
+
+\icmlcorrespondingauthor{Gilles Louppe}{g.louppe@uliege.be}
+\icmlcorrespondingauthor{Kyle Cranmer}{kyle.cranmer@nyu.edu}
+
+% You may provide any keywords that you
+% find helpful for describing your paper; these are used to populate
+% the "keywords" metadata in the PDF but will not be shown in the document
+\icmlkeywords{XXXXX}
+
+\vskip 0.3in
 ]
+
+\printAffiliationsAndNotice{\icmlEqualContribution}
 
@@
 objectives.
-The objective of both algorithms is to match marginal distribution of
+The objective of the algorithm is to match the marginal distribution of
 the generated data to the empirical distribution of the observations.
@@
 (unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters
-$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
+$\bftheta^*$ that minimize the divergence or the distance $\rho$ between $p_r(\mathbf{x})$ and
 the implicit model $p(\mathbf{x}|\bftheta)$. That is,
 \begin{equation}
-    \bftheta^* = \arg \min_{\bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
+    \bftheta^* = \arg \min_{\bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)).
 \end{equation}
-where $\rho$ is some distance or divergence.
 
@@
 unreliable updates when the training procedure is relaxed. As a remedy,
-Wasserstein GANs~\cite{2017arXiv170107875A} reformulate the adversarial
+Wasserstein GANs~\cite{2017arXiv170107875A} (WGANs) reformulate the adversarial
 setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
@@
 &= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
-&\equiv \sup_{\bfphi} \mathcal{L}_W.
+&\triangleq \sup_{\bfphi} \mathcal{L}_W.
 \end{align}
@@
 on the optima of a non-differentiable function. Given a function $f$ to minimize,
-these techniques are based on the simple fact that
+these techniques are based on the observation that
 \begin{equation}
@@
 The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
-GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
+WGANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
 the setting where we are interested in estimating the parameters of a
@@
 specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
-parameters of $g$ and alternately minimize the variational upper bounds
+parameters of the simulator $g$ and alternately minimize the variational upper bounds
 \begin{align}
@@
 directly evaluating the gradient of $U_d$ over mini-batches of real and
-generated data, as is ordinarily done in stochastic gradient descent. When updating
+generated data. When updating
 $\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
@@
 such as Rao-Blackwellization,
-Control Variates, or quasi-Monte Carlo~\cite{ranganath2014black, tran2017variational}.
+control variates, or quasi-Monte Carlo~\cite{ranganath2014black, tran2017variational}.
 
-\subsection{Point Estimates through Marginal Likelihood}
+\subsection{Parameter Point Estimates}
 %\subsection{Empirical Bayes through  Variational Inference}
@@
 
-We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
+We can view the optimization of $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
 If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\cite{rubin1984}.
@@
 In order to more effectively target
-point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
-regularization term corresponding to the differential entropy $H$ of the
-proposal distribution $q(\bftheta | \bfpsi)$. That is,
+point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with
+an entropic regularization term $H(q(\bftheta|\bfpsi))$, that is
 \begin{equation}
@@
 between the generator objective and the tightness of the proposal distribution.
-
 For small values of $\gamma$,
@@
 \end{figure}
-
-\subsection{Electron--positron annihilation}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
-\caption{Electron--positron annihilation.
-({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
-         Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
-         The density of the penalized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
-({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
-({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
- }\label{fig:weinberg}
-\end{figure}
-
-As a more realistic example, we now consider a (simplified) simulator from
-particle physics for electron--positron collisions resulting in muon--antimuon
-pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
-distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
-polar angle of the outgoing muon with respect  to the originally incoming
-electron. Neglecting measurement uncertainty induced from the particle detectors,
-this random variable is approximately distributed as
-\begin{equation}\label{eq:simple_weinberg}
-p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
-\end{equation}
-where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
-function. Due to the linear term in the expression, the density $p(\mathbf{x} |
-E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
-size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
-and $G_f$ (the Fermi constant) through the coefficient function $c$.
-
-A typical physics simulator for this process includes a more precise treatment of the
-quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering
-using \texttt{MadGraph}~\cite{Alwall:2011uj},  ionization of matter in the
-detector due to the passage of the out-going $\mu^+\mu^-$ particles using
-\texttt{GEANT4}~\cite{Agostinelli:2002hh}, electronic noise and other details of the sensors
-that measure the ionization signal, and the deterministic algorithms that
-estimate the polar angle $A$ based on the sensor readouts. The simulation of
-this process is highly non-trivial as is the space of latent variables $\mathcal{Z}$.
-
-In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
-using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
-factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
-q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
-and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
-$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
-$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
-entropy penalties $\gamma=0$ and $\gamma=5$.
-
-The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
-proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
-AVO. We see that the distributions both arrive
-in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
-$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
-distance with $\bftheta^*$, both models closely match the observed data, as shown
-in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
-results for the entropy penalized model. This suggests either a relatively flat
-landscape around $\bftheta^*$ or that the observed data can in this case also be
-reproduced with the  predictive distribution $\qxpsi$.
-Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
-$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
-towards $0$, which indicates  convergence towards a distribution that the critic
-cannot distinguish from the true distribution.
 
@@
 
-\bibliographystyle{hunsrt}
 \bibliography{bibliography.bib}
+\bibliographystyle{icml2018}
+
+% ==============================================================================
+
+\appendix
+
+\section{Further experiments}
+
+\subsection{Electron--positron annihilation}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
+\caption{Electron--positron annihilation.
+({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
+         Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
+         The density of the penalized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
+({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
+({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+ }\label{fig:weinberg}
+\end{figure}
+
+As a more realistic example, we now consider a (simplified) simulator from
+particle physics for electron--positron collisions resulting in muon--antimuon
+pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
+distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
+polar angle of the outgoing muon with respect  to the originally incoming
+electron. Neglecting measurement uncertainty induced from the particle detectors,
+this random variable is approximately distributed as
+\begin{equation}\label{eq:simple_weinberg}
+p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
+\end{equation}
+where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
+function. Due to the linear term in the expression, the density $p(\mathbf{x} |
+E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
+size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
+and $G_f$ (the Fermi constant) through the coefficient function $c$.
+
+A typical physics simulator for this process includes a more precise treatment of the
+quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering
+using \texttt{MadGraph}~\cite{Alwall:2011uj},  ionization of matter in the
+detector due to the passage of the out-going $\mu^+\mu^-$ particles using
+\texttt{GEANT4}~\cite{Agostinelli:2002hh}, electronic noise and other details of the sensors
+that measure the ionization signal, and the deterministic algorithms that
+estimate the polar angle $A$ based on the sensor readouts. The simulation of
+this process is highly non-trivial as is the space of latent variables $\mathcal{Z}$.
+
+In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
+using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
+factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
+q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
+and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
+$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
+$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
+entropy penalties $\gamma=0$ and $\gamma=5$.
+
+The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
+proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
+AVO. We see that the distributions both arrive
+in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
+$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
+distance with $\bftheta^*$, both models closely match the observed data, as shown
+in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
+results for the entropy penalized model. This suggests either a relatively flat
+landscape around $\bftheta^*$ or that the observed data can in this case also be
+reproduced with the  predictive distribution $\qxpsi$.
+Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
+$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
+towards $0$, which indicates  convergence towards a distribution that the critic
+cannot distinguish from the true distribution.
 

### 2018-01-25 icml/icml2018_style/example_paper.tex
init icml version

%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2018}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2018}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Submission and Formatting Instructions for ICML 2018}

\begin{document}

\twocolumn[
\icmltitle{Submission and Formatting Instructions for \\
           International Conference on Machine Learning (ICML 2018)}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2018
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Aeiau Zzzz}{equal,to}
\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
\icmlauthor{Cieua Vvvvv}{goo}
\icmlauthor{Iaesut Saoeu}{ed}
\icmlauthor{Fiuea Rrrr}{to}
\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
\icmlauthor{Aaoeu Iasoh}{goo}
\icmlauthor{Buiui Eueu}{ed}
\icmlauthor{Aeuia Zzzz}{ed}
\icmlauthor{Bieea C.~Yyyy}{to,goo}
\icmlauthor{Teoau Xxxx}{ed}
\icmlauthor{Eee Pppp}{ed}
\end{icmlauthorlist}

\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}

\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
This document provides a basic paper template and submission guidelines.
Abstracts must be a single paragraph, ideally between 4--6 sentences long.
Gross violations will trigger corrections at the camera-ready phase.
\end{abstract}

\section{Electronic Submission}
\label{submission}

Submission to ICML 2018 will be entirely electronic, via a web site
(not email). Information about the submission process and \LaTeX\ templates
are available on the conference web site at:
\begin{center}
\textbf{\texttt{http://icml.cc/2018/}}
\end{center}

The guidelines below will be enforced for initial submissions and
camera-ready copies. Here is a brief summary:
\begin{itemize}
\item Submissions must be in PDF\@.
\item The maximum paper length is \textbf{8 pages excluding references and
    acknowledgements, and 10 pages including references and acknowledgements}
    (pages 9 and 10 must contain only references and acknowledgements).
\item \textbf{Do not include author information or acknowledgements} in your
    initial submission.
\item Your paper should be in \textbf{10 point Times font}.
\item Make sure your PDF file only uses Type-1 fonts.
\item Place figure captions \emph{under} the figure (and omit titles from inside
    the graphic file itself). Place table captions \emph{over} the table.
\item References must include page numbers whenever possible and be as complete
    as possible. Place multiple citations in chronological order.
\item Do not alter the style template; in particular, do not compress the paper
    format by reducing the vertical spaces.
\item Keep your abstract brief and self-contained, one paragraph and roughly
    4--6 sentences. Gross violations will require correction at the
    camera-ready phase. The title should have content words capitalized.
\end{itemize}

\subsection{Submitting Papers}

\textbf{Paper Deadline:} The deadline for paper submission that is
advertised on the conference website is strict. If your full,
anonymized, submission does not reach us on time, it will not be
considered for publication. There is no separate abstract submission.

\textbf{Anonymous Submission:} ICML uses double-blind review: no identifying
author information may appear on the title page or in the paper
itself. Section~\ref{author info} gives further details.

\textbf{Simultaneous Submission:} ICML will not accept any paper which,
at the time of submission, is under review for another conference or
has already been published. This policy also applies to papers that
overlap substantially in technical content with conference papers
under review or previously published. ICML submissions must not be
submitted to other conferences during ICML's review period. Authors
may submit to ICML substantially different versions of journal papers
that are currently under review by the journal, but not yet accepted
at the time of submission. Informal publications, such as technical
reports or papers in workshop proceedings which do not appear in
print, do not fall under these restrictions.

\medskip

Authors must provide their manuscripts in \textbf{PDF} format.
Furthermore, please make sure that files contain only embedded Type-1 fonts
(e.g.,~using the program \texttt{pdffonts} in linux or using
File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
might come from graphics files imported into the document.

Authors using \textbf{Word} must convert their document to PDF\@. Most
of the latest versions of Word have the facility to do this
automatically. Submissions will not be accepted in Word format or any
format other than PDF\@. Really. We're not joking. Don't send Word.

Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
Those using \texttt{latex} and \texttt{dvips} may need the following
two commands:

{\footnotesize
\begin{verbatim}
dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
ps2pdf paper.ps
\end{verbatim}}
It is a zero following the ``-G'', which tells dvips to use
the config.pdf file. Newer \TeX\ distributions don't always need this
option.

Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
results. This program avoids the Type-3 font problem, and supports more
advanced features in the \texttt{microtype} package.

\textbf{Graphics files} should be a reasonable size, and included from
an appropriate format. Use vector formats (.eps/.pdf) for plots,
lossless bitmap formats (.png) for raster graphics with sharp lines, and
jpeg for photo-like images.

The style file uses the \texttt{hyperref} package to make clickable
links in documents. If this causes problems for you, add
\texttt{nohyperref} as one of the options to the \texttt{icml2018}
usepackage statement.

\subsection{Submitting Final Camera-Ready Copy}

The final versions of papers accepted for publication should follow the
same format and naming convention as initial submissions, except that
author information (names and affiliations) should be given. See
Section~\ref{final author} for formatting instructions.

The footnote, ``Preliminary work. Under review by the International
Conference on Machine Learning (ICML). Do not distribute.'' must be
modified to ``\textit{Proceedings of the
$\mathit{35}^{th}$ International Conference on Machine Learning},
Stockholm, Sweden, PMLR 80, 2018.
Copyright 2018 by the author(s).''

For those using the \textbf{\LaTeX} style file, this change (and others) is
handled automatically by simply changing
$\mathtt{\backslash usepackage\{icml2018\}}$ to
$$\mathtt{\backslash usepackage[accepted]\{icml2018\}}$$
Authors using \textbf{Word} must edit the
footnote on the first page of the document themselves.

Camera-ready copies should have the title of the paper as running head
on each page except the first one. The running title consists of a
single line centered above a horizontal rule which is $1$~point thick.
The running head should be centered, bold and in $9$~point type. The
rule should be $10$~points above the main text. For those using the
\textbf{\LaTeX} style file, the original title is automatically set as running
head using the \texttt{fancyhdr} package which is included in the ICML
2018 style file package. In case that the original title exceeds the
size restrictions, a shorter form can be supplied by using

\verb|\icmltitlerunning{...}|

just before $\mathtt{\backslash begin\{document\}}$.
Authors using \textbf{Word} must edit the header of the document themselves.

\section{Format of the Paper}

All submissions must follow the specified format.

\subsection{Length and Dimensions}

Papers must not exceed eight (8) pages, including all figures, tables,
and appendices, but excluding references and acknowledgements. When references and acknowledgements are included,
the paper must not exceed ten (10) pages.
Acknowledgements should be limited to grants and people who contributed to the paper.
Any submission that exceeds
this page limit, or that diverges significantly from the specified format,
will be rejected without review.

The text of the paper should be formatted in two columns, with an
overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
between the columns. The left margin should be 0.75~inches and the top
margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
whether you print on US letter or A4 paper, but all final versions
must be produced for US letter size.

The paper body should be set in 10~point type with a vertical spacing
of 11~points. Please use Times typeface throughout the text.

\subsection{Title}

The paper title should be set in 14~point bold type and centered
between two horizontal rules that are 1~point thick, with 1.0~inch
between the top rule and the top edge of the page. Capitalize the
first letter of content words and put the rest of the title in lower
case.

\subsection{Author Information for Submission}
\label{author info}

ICML uses double-blind review, so author information must not appear. If
you are using \LaTeX\/ and the \texttt{icml2018.sty} file, use
\verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
will not be printed unless \texttt{accepted} is passed as an argument to the
style file.
Submissions that include the author information will not
be reviewed.

\subsubsection{Self-Citations}

If you are citing published papers for which you are an author, refer
to yourself in the third person. In particular, do not use phrases
that reveal your identity (e.g., ``in previous work \cite{langley00}, we
have shown \ldots'').

Do not anonymize citations in the reference section. The only exception are manuscripts that are
not yet published (e.g., under submission). If you choose to refer to
such unpublished manuscripts \cite{anonymous}, anonymized copies have
to be submitted
as Supplementary Material via CMT\@. However, keep in mind that an ICML
paper should be self contained and should contain sufficient detail
for the reviewers to evaluate the work. In particular, reviewers are
not required to look at the Supplementary Material when writing their
review.

\subsubsection{Camera-Ready Author Information}
\label{final author}

If a paper is accepted, a final camera-ready copy must be prepared.
%
For camera-ready papers, author information should start 0.3~inches below the
bottom rule surrounding the title. The authors' names should appear in 10~point
bold type, in a row, separated by white space, and centered. Author names should
not be broken across lines. Unbolded superscripted numbers, starting 1, should
be used to refer to affiliations.

Affiliations should be numbered in the order of appearance. A single footnote
block of text should be used to list all the affiliations. (Academic
affiliations should list Department, University, City, State/Region, Country.
Similarly for industrial affiliations.)

Each distinct affiliations should be listed once. If an author has multiple
affiliations, multiple superscripts should be placed after the name, separated
by thin spaces. If the authors would like to highlight equal contribution by
multiple first authors, those authors should have an asterisk placed after their
name in superscript, and the term ``\textsuperscript{*}Equal contribution"
should be placed in the footnote block ahead of the list of affiliations. A
list of corresponding authors and their emails (in the format Full Name
\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
Ideally only one or two names should be listed.

A sample file with author names is included in the ICML2018 style file
package. Turn on the \texttt{[accepted]} option to the stylefile to
see the names rendered. All of the guidelines above are implemented
by the \LaTeX\ style file.

\subsection{Abstract}

The paper abstract should begin in the left column, 0.4~inches below the final
address. The heading `Abstract' should be centered, bold, and in 11~point type.
The abstract body should use 10~point type, with a vertical spacing of
11~points, and should be indented 0.25~inches more than normal on left-hand and
right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
sentences. Gross violations will require correction at the camera-ready phase.

\subsection{Partitioning the Text}

You should organize your paper into sections and paragraphs to help
readers place a structure on the material and understand its
contributions.

\subsubsection{Sections and Subsections}

Section headings should be numbered, flush left, and set in 11~pt bold
type with the content words capitalized. Leave 0.25~inches of space
before the heading and 0.15~inches after the heading.

Similarly, subsection headings should be numbered, flush left, and set
in 10~pt bold type with the content words capitalized. Leave
0.2~inches of space before the heading and 0.13~inches afterward.

Finally, subsubsection headings should be numbered, flush left, and
set in 10~pt small caps with the content words capitalized. Leave
0.18~inches of space before the heading and 0.1~inches after the
heading.

Please use no more than three levels of headings.

\subsubsection{Paragraphs and Footnotes}

Within each section or subsection, you should further partition the
paper into paragraphs. Do not indent the first line of a given
paragraph, but insert a blank line between succeeding ones.

You can use footnotes\footnote{Footnotes
should be complete sentences.} to provide readers with additional
information about a topic without interrupting the flow of the paper.
Indicate footnotes with a number in the text where the point is most
relevant. Place the footnote in 9~point type at the bottom of the
column in which it appears. Precede the first footnote in a column
with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
appear in each column, in the same order as they appear in the text,
but spread them across columns and pages if possible.}

\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
\caption{Historical locations and number of accepted papers for International
Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
produced, the number of accepted papers for ICML 2008 was unknown and instead
estimated.}
\label{icml-historical}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Figures}

You may want to include figures in the paper to illustrate
your approach and results. Such artwork should be centered,
legible, and separated from the text. Lines should be dark and at
least 0.5~points thick for purposes of reproduction, and text should
not appear on a gray background.

Label all distinct components of each figure. If the figure takes the
form of a graph, then give a name for each axis and include a legend
that briefly describes each curve. Do not include a title inside the
figure; instead, the caption should serve this function.

Number figures sequentially, placing the figure number and caption
\emph{after} the graphics, with at least 0.1~inches of space before
the caption and 0.1~inches after it, as in
Figure~\ref{icml-historical}. The figure caption should be set in
9~point type and centered unless it runs two or more lines, in which
case it should be flush left. You may float figures to the top or
bottom of a column, and you may set wide figures across both columns
(use the environment \texttt{figure*} in \LaTeX). Always place
two-column figures at the top or bottom of the page.

\subsection{Algorithms}

If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
environments to format pseudocode. These require
the corresponding stylefiles, algorithm.sty and
algorithmic.sty, which are supplied with this package.
Algorithm~\ref{alg:example} shows an example.

\begin{algorithm}[tb]
   \caption{Bubble Sort}
   \label{alg:example}
\begin{algorithmic}
   \STATE {\bfseries Input:} data $x_i$, size $m$
   \REPEAT
   \STATE Initialize $noChange = true$.
   \FOR{$i=1$ {\bfseries to} $m-1$}
   \IF{$x_i > x_{i+1}$}
   \STATE Swap $x_i$ and $x_{i+1}$
   \STATE $noChange = false$
   \ENDIF
   \ENDFOR
   \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

\subsection{Tables}

You may also want to include tables that summarize material. Like
figures, these should be centered, legible, and numbered consecutively.
However, place the title \emph{above} the table with at least
0.1~inches of space before the title and the same after it, as in
Table~\ref{sample-table}. The table title should be set in 9~point
type and centered unless it runs two or more lines, in which case it
should be flush left.

% Note use of \abovespace and \belowspace to get reasonable spacing
% above and below tabular lines.

\begin{table}[t]
\caption{Classification accuracies for naive Bayes and flexible
Bayes on various data sets.}
\label{sample-table}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
Data set & Naive & Flexible & Better? \\
\midrule
Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

Tables contain textual material, whereas figures contain graphical material.
Specify the contents of each row and column in the table's topmost
row. Again, you may float tables to a column's top or bottom, and set
wide tables across both columns. Place two-column tables at the
top or bottom of the page.

\subsection{Citations and References}

Please use APA reference format regardless of your formatter
or word processor. If you rely on the \LaTeX\/ bibliographic
facility, use \texttt{natbib.sty} and \texttt{icml2018.bst}
included in the style-file package to obtain this format.

Citations within the text should include the authors' last names and
year. If the authors' names are included in the sentence, place only
the year in parentheses, for example when referencing Arthur Samuel's
pioneering work \yrcite{Samuel59}. Otherwise place the entire
reference in parentheses with the authors and year separated by a
comma \cite{Samuel59}. List multiple references separated by
semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
construct only for citations with three or more authors or after
listing all authors to a publication in an earlier reference \cite{MachineLearningI}.

Authors should cite their own work in the third person
in the initial version of their paper submitted for blind review.
Please refer to Section~\ref{author info} for detailed instructions on how to
cite your own papers.

Use an unnumbered first-level section heading for the references, and use a
hanging indent style, with the first line of the reference flush against the
left margin and subsequent lines indented by 10 points. The references at the
end of this document give examples for journal articles \cite{Samuel59},
conference publications \cite{langley00}, book chapters \cite{Newell81}, books
\cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
\cite{mitchell80}, and dissertations \cite{kearns89}.

Alphabetize references by the surnames of the first authors, with
single author entries preceding multiple author entries. Order
references for the same authors by year of publication, with the
earliest first. Make sure that each reference includes all relevant
information (e.g., page numbers).

Please put some effort into making references complete, presentable, and
consistent. If using bibtex, please protect capital letters of names and
abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
in your .bib file.

\subsection{Software and Data}

We strongly encourage the publication of software and data with the
camera-ready version of the paper whenever appropriate. This can be
done by including a URL in the camera-ready copy. However, do not
include URLs that reveal your institution or identity in your
submission for review. Instead, provide an anonymous URL or upload
the material as ``Supplementary Material'' into the CMT reviewing
system. Note that reviewers are not required to look a this material
when writing their review.

% Acknowledgements should only appear in the accepted version.
\section*{Acknowledgements}

\textbf{Do not} include acknowledgements in the initial version of
the paper submitted for blind review.

If a paper is accepted, the final camera-ready version can (and
probably should) include acknowledgements. In this case, please
place such acknowledgements in an unnumbered section at the
end of the paper. Typically, this will include thanks to reviewers
who gave useful comments, to colleagues who contributed to the ideas,
and to funding agencies and corporate sponsors that provided financial
support.

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\nocite{langley00}

\bibliography{example_paper}
\bibliographystyle{icml2018}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DELETE THIS PART. DO NOT PLACE CONTENT AFTER THE REFERENCES!
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Do \emph{not} have an appendix here}

\textbf{\emph{Do not put content after the references.}}
%
Put anything that you might normally include after the references in a separate
supplementary file.

We recommend that you build supplementary material in a separate document.
If you must create one PDF and cut it up, please be careful to use a tool that
doesn't alter the margins, and that doesn't aggressively rewrite the PDF file.
pdftk usually works fine.

\textbf{Please do not use Apple's preview to cut off supplementary material.} In
previous years it has altered margins, and created headaches at the camera-ready
stage. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018. It was modified from a version from Dan Roy in
% 2017, which was based on a version from Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2018-01-30 icml/avo.tex
save

@@
 \begin{abstract}
-    Complex computer simulators are increasingly used across fields of science as
-    generative models tying parameters of an underlying theory to
-    experimental observations. Inference in this setup is often
-    difficult, as simulators rarely admit a tractable density or likelihood
-    function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
-    inference algorithm for fitting a non-differentiable generative model incorporating ideas from empirical Bayes and variational inference.
-    We adapt the training procedure of generative
-    adversarial networks by replacing the differentiable generative network with a
-    domain-specific simulator. We solve the resulting non-differentiable
-    minimax problem by minimizing variational upper bounds of the two adversarial objectives.
-    Effectively, the procedure results in learning a
-    proposal distribution over simulator parameters, such that the corresponding
-    marginal distribution of the generated data matches the observations.
-    We present results of the method with simulators producing both discrete and continuous data.
+    Complex computer simulators are increasingly used across fields
+of science as generative models tying parameters of an underlying theory to
+experimental observations. Inference in this setup is often difficult, as
+simulators rarely admit a tractable density or likelihood function. We introduce
+Adversarial Variational Optimization (AVO), a likelihood-free inference
+algorithm for fitting a non-differentiable generative model incorporating ideas
+from generative adversarial networks (GANs) and variational inference. We adapt the
+training procedure of (Wasserstein) GANs by replacing the
+differentiable generative network with a domain-specific simulator. We solve the
+resulting non-differentiable minimax problem by minimizing variational upper
+bounds of the two adversarial objectives. Effectively, the procedure results in
+learning a proposal distribution over simulator parameters, such that the
+corresponding marginal distribution of the generated data matches the
+observations. We present results of the method with simulators producing both
+discrete and continuous data.
 \end{abstract}
@@
 
-In many fields of science such as particle physics, epidemiology,  and population
-genetics, computer simulators are used to describe complex data generation processes. These simulators relate
-observations $\bfx$ to the parameters $\bftheta$ of an underlying theory or mechanistic model.
-In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
-Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx |
-\bftheta)$. The prevalence and significance of this problem has motivated an
-active research effort in so-called \textit{likelihood-free inference}
-algorithms such as Approximate Bayesian Computation
-(ABC) and density estimation-by-comparison algorithms~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential,
-sisson2011likelihood, marin2012approximate, cranmer2015approximating}.
+In many fields of science such as particle physics, epidemiology,  and
+population genetics, computer simulators are used to describe complex data
+generation processes. These simulators relate observations $\bfx$ to the
+parameters $\bftheta$ of an underlying theory or mechanistic model. In most
+cases, these simulators are specified as procedural implementations of forward,
+stochastic processes involving latent variables $\bfz$. Rarely do these
+simulators admit a tractable density (or likelihood) $p(\bfx | \bftheta)$. The
+prevalence and significance of this problem has motivated an active research
+effort in so-called \textit{likelihood-free inference} algorithms such as
+Approximate Bayesian Computation (ABC) and density estimation-by-comparison
+algorithms~\cite{beaumont2002approximate, marjoram2003markov,
+sisson2007sequential, sisson2011likelihood, marin2012approximate,
+cranmer2015approximating}.
 
@@
 \begin{equation}\label{eqn:p_theta}
-    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
+    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \triangleq \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
 \end{equation}
@@
 We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
-algorithm~\cite{2011arXiv1106.4487W} or variance reduction techniques
-such as Rao-Blackwellization,
-control variates, or quasi-Monte Carlo~\cite{ranganath2014black, tran2017variational}.
+algorithm~\cite{2011arXiv1106.4487W} or variance reduction techniques as
+control variates~\cite{2017arXiv171100123G} or finite differences~\cite{buesingstochastic}.
 
@@
 \begin{equation}\label{eqn:p_psi}
-\mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
+\mathbf{x} \sim \qxpsi \triangleq \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
 \end{equation}
@@
 proposal distributions with large entropy are not penalized, which results
-in learning a smeared variation of the original simulator corresponding to the marginal or prior predictive model. On the other hand,
+in learning a smeared variation of the original simulator. On the other hand,
 for large values of $\gamma$, the procedure is encouraged to fit a proposal
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
+\caption{Discrete Poisson model with unknown mean.
+ ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=1$, the distributions correctly concentrate their density around
+            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=1$) results in a tighter density.
+ ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
+ ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+ }\label{fig:poisson}
+\end{figure}
+
 As a first illustrative experiment, we evaluate inference for a discrete Poisson
@@
 The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
-Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
+Algorithm~\ref{alg:avo} is run for 500 epochs with mini-batches of size $M=64$
 and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
 hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
-$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
-$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
+$n_\text{critic}=5$ iterations with a step size $\alpha=0.01$ and otherwise default parameters.
+Its inner first and second moment vectors are reset at
 each outer epoch in order to avoid building momentum in staled directions.  For
@@
 $q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
-each epoch, parameters $\bfpsi$ are updated by taking one Adam step, with
-$\alpha=0.01$ and $\beta_1=\beta_2=0.5$. The gradient penalty coefficient is set to
-$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.
+each epoch, parameters $\bfpsi$ are updated by taking one Adam step, with step size
+$\alpha=0.01$. The gradient penalty coefficient is set to
+$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=1$.
 
@@
 distributions $q(\bftheta|\bfpsi)$ after AVO.  For
-both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
+both $\gamma=0$ and $\gamma=1$, the proposal distributions correctly concentrate
 their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
-the effect of the positive entropy penalty $H(q(\bftheta|\bfpsi))$,
-the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
-yielding in this case a more precise inference.  The top right plot compares the
+the effect of entropic regularization,
+the proposal distribution for $\gamma=1$ concentrates its mass tightly,
+yielding in this case precise inference.  The top right plot compares the
 model distributions to the true distribution.  As theoretically expected from
-adversarial training, we see that the resulting distributions closely match
+adversarial training, we see that the resulting distributions align with
 the true distribution, with in this case visually slightly better results for the penalized
 model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
-of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
+of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=1$,
 the curves quickly fall towards $0$, which indicates that
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
-\caption{Discrete Poisson model with unknown mean.
- ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
-            the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
- ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
- ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
- }\label{fig:poisson}
-\end{figure}
-
 \subsection{Multidimensional continuous data}
@@
 As a second toy example, we consider a generator producing
-5-dimensional continuous data, as originally specified in Section 4.2. of
-\cite{cranmer2015approximating}. More specifically, we consider the following
-generative process:
-\begin{itemize}
-\item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
-$z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
-$z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
-$z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
-$z_3 \sim {\text{Exponential}(\lambda=3)}$, and
-$z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
-\item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
-semi-positive definite $5 \times 5$ matrix defining a fixed projection
-of $\bfz$ into the observed space.
-\end{itemize}
-Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model.
+5-dimensional continuous data, as originally specified in Section 4.2 of
+\cite{cranmer2015approximating}.
+% More specifically, we consider the following
+% generative process:
+% \begin{itemize}
+% \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
+% $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
+% $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
+% $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
+% $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
+% $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
+% \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
+% semi-positive definite $5 \times 5$ matrix defining a fixed projection
+% of $\bfz$ into the observed space.
+% \end{itemize}
+Again, AVO does not have access to the density or its gradient, only samples from the generative model.
 We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
@@
 Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
-Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
+Adam configured with a step size $\alpha=0.01$ and otherwise default parameters.
 The architecture of the critic is the same as in the previous example.
@@
 the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
-AVO quickly converges towards a variational distribution whose density
+AVO quickly converges towards a proposal distribution whose density
 concentrates around the nominal values $\bftheta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
@@
 
+\subsection{Benchmarks}
+
+\glnote{Compare against ABC and others.}
+\glnote{Sections 6.1 and 6.4 of Bernton.}
+
 % ==============================================================================
@@
 likelihood-free inference, approximate Bayesian computation (ABC),
-implicit generative networks, and variational inference.
+implicit generative models, and variational inference.
 Viewed from the literature around implicit generative models based on neural networks,
@@
 recognition model  $q(\bfz, \bftheta|\bfpsi)$ as the latent space $\mathcal{Z}$
-of many real-world simulators is complicated and not amenable to neural a
+of many real-world simulators is complicated and not amenable to a neural
 recognition model.
@@
 this work. In terms of method, several components need further investigation.
-First, we need to better study the interplay between the entropy penalty and the
+First, we need to better study the interplay between the entropic regularization and the
 adversarial objectives. Second, we should better understand the dynamics of the
@@
 
-As a more realistic example, we now consider a (simplified) simulator from
+As a more realistic example, we here consider a (simplified) simulator from
 particle physics for electron--positron collisions resulting in muon--antimuon

### 2018-01-30 code/multi.py

@@ -240,6 +240,6 @@ for i in range(n_epochs):
             plt.tight_layout()
 
             if i == n_epochs - 1:
-                plt.savefig("figs/multi-%d.pdf" % seed)
+                plt.savefig("multi-%d.pdf" % seed)
 
             plt.close()

### 2018-01-30 code/toy-poisson.py

@@ -21,7 +21,7 @@ seed = 777
 rng = check_random_state(seed)
 
 batch_size = 64
-n_epochs = 300+1
+n_epochs = 500+1
 lambda_gp = 0.025
 
 true_theta = np.array([np.log(7)])
@@ -42,11 +42,11 @@ n_features = X_obs.shape[1]
 
 # Critic
 
-# gammas = [0.0, 5.0]
-# colors = ["C1", "C2"]
+gammas = [0.0, 1.0]
+colors = ["C1", "C2"]
 
-gammas = [5.0]
-colors = ["C2"]
+# gammas = [5.0]
+# colors = ["C2"]
 
 def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
@@ -142,24 +142,23 @@ for state in history:
         return grad_u
 
     # Training loop
-    opt_critic = AdamOptimizer(grad_loss_critic, state["params_critic"],
-                               step_size=0.01, b1=0.5, b2=0.5)
-    opt_proposal = AdamOptimizer(approx_grad_u, state["params_proposal"],
-                                 step_size=0.01, b1=0.5, b2=0.5)
+    opt_critic = AdamOptimizer(grad_loss_critic, state["params_critic"]) #, step_size=0.01, b1=0.5, b2=0.5)
+    opt_proposal = AdamOptimizer(approx_grad_u, state["params_proposal"]) #, step_size=0.01, b1=0.5, b2=0.5)
 
     opt_critic.step(100)
     opt_critic.move_to(state["params_critic"])
 
     for i in range(n_epochs):
-        print(i, state["gamma"], state["params_proposal"])
+        print(i, state["gamma"], state["params_proposal"], np.mean((true_theta - state["params_proposal"]["mu"]) ** 2))
 
         # fit simulator
+        opt_proposal.reset()
         opt_proposal.step(1)
         opt_proposal.move_to(state["params_proposal"])
 
         # fit critic
         opt_critic.reset()   # reset moments
-        opt_critic.step(100)
+        opt_critic.step(5)
         opt_critic.move_to(state["params_critic"])
 
         state["loss_d"].append(-loss_critic(state["params_critic"], i,
@@ -188,6 +187,7 @@ if make_plots:
                  color=state["color"])
 
     plt.legend(loc="upper right")
+    plt.ylim(0, 15)
 
     # histograms
[…]

### 2018-01-30 icml/avo.tex
add lfvi

@@
 the proposed method can be considered as a direct adaptation of
-generative adversarial networks~\cite{goodfellow2014generative,2017arXiv170104862A, 2017arXiv170107875A} to
-non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization}.
+generative adversarial networks~\cite{goodfellow2014generative,2017arXiv170104862A} to
+non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S}.
 
@@
 \cite{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
-Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
-estimate gradients with respect
-to $\bftheta$ through finite differences from multiple forward passes of the
-simulator with variance reduction strategies based on controlling the source of
-randomness used for the latent variable $\bfz$.
+% Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
+% estimate gradients with respect
+% to $\bftheta$ through finite differences from multiple forward passes of the
+% simulator with variance reduction strategies based on controlling the source of
+% randomness used for the latent variable $\bfz$.
 
@@
 
-This work has also many connections to work on variational inference, in which the
-goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
-is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have been efforts
-to extend variational inference to intractable likelihoods; however, they
-require restrictive assumptions.  In \cite{tran2017variational}, the authors
-consider Variational Bayes with an Intractable Likelihood (VBIL). In that
-approach ``the only requirement is that the intractable likelihood can be
+This work has also many connections to work on variational inference, in which
+the goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that
+it is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have
+been efforts to extend variational inference to intractable likelihoods;
+however, many require restrictive assumptions.  In \cite{tran2017variational},
+the authors consider Variational Bayes with an Intractable Likelihood (VBIL). In
+that approach ``the only requirement is that the intractable likelihood can be
 estimated unbiasedly.'' In the case of simulators, they propose to use the
 ABC-likelihood with an $\epsilon$-kernel. The ABC likelihood is only unbiased as
-$\epsilon \to 0$,   thus this method inherits the drawbacks of the ABC-likelihood
-including the choice of summary statistics and the inefficiency in evaluating
-the ABC likelihood for high-dimensional data and small $\epsilon$.
+$\epsilon \to 0$,   thus this method inherits the drawbacks of the
+ABC-likelihood including the choice of summary statistics and the inefficiency
+in evaluating the ABC likelihood for high-dimensional data and small $\epsilon$.
+More recently, \cite{2017arXiv170208896T} adapted variational inference to
+hierarchical implicit models defined on simulators. In this work, the authors
+step around the intractable likelihoods by reformulating the optimization of the
+ELBO in terms of a neural and differentiable approximation $r$ of the log-likelihood
+ratio $\log \tfrac{p}{q}$.
 

### 2018-01-30 icml/avo.tex
save

@@
 fledged scientific simulator -- which we plan to achieve for a next version of
-this work. In terms of method, several components need further investigation.
+this work \glnote{to update}. In terms of method, several components need further investigation.
 First, we need to better study the interplay between the entropic regularization and the

### 2018-01-31 icml/avo.tex
save

@@
 \caption{Discrete Poisson model with unknown mean.
- ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=1$, the distributions correctly concentrate their density around
-            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=1$) results in a tighter density.
+ ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
+            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=5$) results in a tighter density.
  ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
@@
 hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
-$n_\text{critic}=5$ iterations with a step size $\alpha=0.01$ and otherwise default parameters.
-Its inner first and second moment vectors are reset at
-each outer epoch in order to avoid building momentum in staled directions.  For
+$n_\text{critic}=10$ iterations with a step size $\alpha=0.005$ and $\beta_1=\beta_2=0.05$.
+% Its inner first and second moment vectors are reset at
+% each outer epoch in order to avoid building momentum in staled directions.
+For
 estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
-$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
-each epoch, parameters $\bfpsi$ are updated by taking one Adam step, with step size
-$\alpha=0.01$. The gradient penalty coefficient is set to
-$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=1$.
+$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(10)$ and unit variance. At
+each epoch, parameters $\bfpsi$ are updated by taking one Adam step and the same hyper-parameters as for the critic.
+The gradient penalty coefficient is set to
+$\lambda=0.1$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.
 
@@
 distributions $q(\bftheta|\bfpsi)$ after AVO.  For
-both $\gamma=0$ and $\gamma=1$, the proposal distributions correctly concentrate
+both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
 their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
 the effect of entropic regularization,
-the proposal distribution for $\gamma=1$ concentrates its mass tightly,
+the proposal distribution for $\gamma=5$ concentrates its mass tightly,
 yielding in this case precise inference.  The top right plot compares the
@@
 model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
-of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=1$,
-the curves quickly fall towards $0$, which indicates that
+of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
+the curves fall towards $0$, which indicates that
 $\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
@@
 initialized with zero mean and unit variance.
-Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
-Adam configured with a step size $\alpha=0.01$ and otherwise default parameters.
+Hyper-parameters are set to $M=64$, $n_\text{critic}=10$, $\lambda=0.1$, $\gamma=1$ and
+Adam configured with a step size $\alpha=0.005$ and $\beta_1=\beta_2=0.05$.
 The architecture of the critic is the same as in the previous example.
@@
 the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
-AVO quickly converges towards a proposal distribution whose density
+AVO converges towards a proposal distribution whose density
 concentrates around the nominal values $\bftheta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
@@
 \cite{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
-% Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
-% estimate gradients with respect
-% to $\bftheta$ through finite differences from multiple forward passes of the
-% simulator with variance reduction strategies based on controlling the source of
-% randomness used for the latent variable $\bfz$.
+Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
+estimate gradients with respect
+to $\bftheta$ through finite differences from multiple forward passes of the
+simulator with variance reduction strategies based on controlling the source of
+randomness used for the latent variable $\bfz$.
 
@@
 ELBO in terms of a neural and differentiable approximation $r$ of the log-likelihood
-ratio $\log \tfrac{p}{q}$.
+ratio $\log \tfrac{p}{q}$, thereby effectively using the same core principle as used
+in GANs~\cite{2016arXiv161003483M}.
 
@@
 in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
-results for the entropy penalized model. This suggests either a relatively flat
+results when using entropic regularization. This suggests either a relatively flat
 landscape around $\bftheta^*$ or that the observed data can in this case also be

### 2018-01-31 code/multi.py

@@ -22,15 +22,21 @@ from proposals import grad_gaussian_entropy
 from sklearn.utils import check_random_state
 from scipy.spatial.distance import mahalanobis
 
+# What works:
+# remove the reset!!
+# effect of lambda_gp?  lambda_gp=0.1 works
+# effect of gamma? works ok for gamma=0, tighter bands with gamma=1.0
+# refit critic periodically? some zigzag with 1:10 but fine; refitting not necessary
+
 # Global params
 
 seed = 777
 rng = check_random_state(seed)
 
 batch_size = 64
-n_epochs = 300+1
-lambda_gp = 0.025
-gamma = 10.0
+n_epochs = 1000+1
+lambda_gp = 0.1
+gamma = 1.0
 
 true_theta = np.array([1.0, -1.0])
 make_plots = True
@@ -157,22 +163,25 @@ def approx_grad_u(params_proposal, i, gamma=gamma):
 
 # Training loop
 
-opt_critic = AdamOptimizer(grad_loss_critic, params_critic, step_size=0.01)
-opt_proposal = AdamOptimizer(approx_grad_u, params_proposal, step_size=0.01)
+opt_critic = AdamOptimizer(grad_loss_critic, params_critic, step_size=0.005, b1=0.05, b2=0.05)
+opt_proposal = AdamOptimizer(approx_grad_u, params_proposal, step_size=0.005, b1=0.05, b2=0.05)
 
 opt_critic.step(100)
 opt_critic.move_to(params_critic)
+opt_critic.reset()
 
 for i in range(n_epochs):
     print(i, params_proposal)
 
     # fit simulator
+    #opt_proposal.reset()
     opt_proposal.step(1)
     opt_proposal.move_to(params_proposal)
 
     # fit critic
-    opt_critic.reset()   # reset moments
-    opt_critic.step(100)
+    #opt_critic.reset()   # reset moments
+    #opt_critic.step(1000 if i % 50 == 0 else 10)
+    opt_critic.step(10)
     opt_critic.move_to(params_critic)
 
     # plot
@@ -195,7 +204,7 @@ for i in range(n_epochs):
             Z = np.array(Z).reshape(X.shape)
 
             CS = plt.contour(X, Y, Z, [1.0, 2.0, 3.0], colors="C1")
[…]

### 2018-01-31 code/proposals.py

@@ -46,7 +46,10 @@ def gaussian_logpdf(params, theta, to_scalar=True):
 
 def gaussian_entropy(params):
     sigma = np.exp(params["log_sigma"])
-    return np.sum(np.log(sigma * (2. * np.pi * np.e) ** 0.5))
+    if np.sum(sigma) > 0.05:
+        return np.sum(np.log(sigma * (2. * np.pi * np.e) ** 0.5))
+    else:
+        return np.log(0.05 * (2. * np.pi * np.e) ** 0.5)
 
 
 grad_gaussian_logpdf = ag.grad(gaussian_logpdf)

### 2018-01-31 code/toy-poisson.py

@@ -20,9 +20,11 @@ from sklearn.utils import check_random_state
 seed = 777
 rng = check_random_state(seed)
 
-batch_size = 64
+batch_size = 64   # try increasing, default=64
 n_epochs = 500+1
-lambda_gp = 0.025
+lambda_gp = 0.1 # ok~0.1; try reducing (=> worse), try increasing (0.4, not sure what to conclude)
+
+# see zigzag near the end... hence the reset? but works better without in multi.py
 
 true_theta = np.array([np.log(7)])
 make_plots = True
@@ -42,7 +44,7 @@ n_features = X_obs.shape[1]
 
 # Critic
 
-gammas = [0.0, 1.0]
+gammas = [0.0, 5.0]
 colors = ["C1", "C2"]
 
 # gammas = [5.0]
@@ -72,7 +74,7 @@ history = [{"gamma": gammas[i],
             "color": colors[i],
             "loss_d": [],
             "params_proposal": make_gaussian_proposal(n_params,
-                                                      mu=np.log(5.0)),
+                                                      mu=np.log(10.0)),
             "params_critic": make_critic(n_features, 10, random_state=rng)}
            for i in range(len(gammas))]
 
@@ -142,23 +144,24 @@ for state in history:
         return grad_u
 
     # Training loop
-    opt_critic = AdamOptimizer(grad_loss_critic, state["params_critic"]) #, step_size=0.01, b1=0.5, b2=0.5)
-    opt_proposal = AdamOptimizer(approx_grad_u, state["params_proposal"]) #, step_size=0.01, b1=0.5, b2=0.5)
+    opt_critic = AdamOptimizer(grad_loss_critic, state["params_critic"], step_size=0.005, b1=0.05, b2=0.05) #, step_size=0.01, b1=0.5, b2=0.5)
+    opt_proposal = AdamOptimizer(approx_grad_u, state["params_proposal"], step_size=0.005, b1=0.05, b2=0.05) #, step_size=0.01, b1=0.5, b2=0.5)
 
     opt_critic.step(100)
     opt_critic.move_to(state["params_critic"])
+    opt_critic.reset()
 
     for i in range(n_epochs):
         print(i, state["gamma"], state["params_proposal"], np.mean((true_theta - state["params_proposal"]["mu"]) ** 2))
 
         # fit simulator
-        opt_proposal.reset()
+        #opt_proposal.reset()
         opt_proposal.step(1)
[…]

### 2018-02-01 icml/avo.tex
save

@@
 \begin{icmlauthorlist}
-\icmlauthor{Glles Louppe}{uliege}
+\icmlauthor{Gilles Louppe}{uliege}
 \icmlauthor{Kyle Cranmer}{nyu}
@@
 algorithm for fitting a non-differentiable generative model incorporating ideas
-from generative adversarial networks (GANs) and variational inference. We adapt the
-training procedure of (Wasserstein) GANs by replacing the
+from generative adversarial networks (GANs) and variational optimization. We adapt the
+training procedure of Wasserstein GANs by replacing the
 differentiable generative network with a domain-specific simulator. We solve the
 resulting non-differentiable minimax problem by minimizing variational upper
-bounds of the two adversarial objectives. Effectively, the procedure results in
-learning a proposal distribution over simulator parameters, such that the
-corresponding marginal distribution of the generated data matches the
-observations. We present results of the method with simulators producing both
+bounds of the two adversarial objectives.
+Effectively, the procedure results in
+learning a proposal distribution over simulator parameters, such that
+the Wasserstein distance between the marginal distribution of the synthetic data and
+the empirical distribution of observed data is minimized.
+We present results of the method with simulators producing both
 discrete and continuous data.
@@
 
-In many fields of science such as particle physics, epidemiology,  and
+In many fields of science such as particle physics, epidemiology  or
 population genetics, computer simulators are used to describe complex data
@@
 
-In this note, we develop a likelihood-free inference algorithm
+In this work, we develop a likelihood-free inference algorithm
 for non-differentiable, implicit generative models.
@@
 
-As thoroughly explored in \cite{2017arXiv170104862A}, GANs remain remarkably
+As explored in \cite{2017arXiv170104862A}, GANs remain remarkably
 difficult to train because of vanishing gradients as $d$ saturates, or because of
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
-\caption{Discrete Poisson model with unknown mean.
- ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
-            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=5$) results in a tighter density.
- ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
- ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
- }\label{fig:poisson}
-\end{figure}
-
 As a first illustrative experiment, we evaluate inference for a discrete Poisson
@@
 density $p(\bfx | \bftheta)$ or its gradient.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
+\caption{Discrete Poisson model with unknown mean.
+ ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
+            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=5$) results in a tighter density.
+ ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
+ ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+ }\label{fig:poisson}
+\end{figure}
 
@@
 
-\subsection{Benchmarks}
+% \subsection{Benchmarks}
+%
+% \glnote{Compare against ABC and others.}
+% \glnote{Sections 6.1 and 6.4 of Bernton.}
 
-\glnote{Compare against ABC and others.}
-\glnote{Sections 6.1 and 6.4 of Bernton.}
+\subsection{Electron--positron annihilation}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
+\caption{Electron--positron annihilation.
+({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
+         Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
+         The density of the entropically regularized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
+({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
+({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+ }\label{fig:weinberg}
+\end{figure}
+
+As a more realistic example, we here consider a (simplified) simulator from
+particle physics for electron--positron collisions resulting in muon--antimuon
+pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
+distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
+polar angle of the outgoing muon with respect  to the originally incoming
+electron. Neglecting measurement uncertainty induced from the particle detectors,
+this random variable is approximately distributed as
+\begin{equation}\label{eq:simple_weinberg}
+p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
+\end{equation}
+where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
+function. Due to the linear term in the expression, the density $p(\mathbf{x} |
+E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
+size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
+and $G_f$ (the Fermi constant) through the coefficient function $c$.
+
+A typical physics simulator for this process includes a more precise treatment of the
+quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering
+using \texttt{MadGraph}~\cite{Alwall:2011uj},  ionization of matter in the
+detector due to the passage of the out-going $\mu^+\mu^-$ particles using
+\texttt{GEANT4}~\cite{Agostinelli:2002hh}, electronic noise and other details of the sensors
+that measure the ionization signal, and the deterministic algorithms that
+estimate the polar angle $A$ based on the sensor readouts. The simulation of
+this process is highly non-trivial as is the space of latent variables $\mathcal{Z}$.
+
+In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
+using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
+factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
+q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
+and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
+$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
+$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
+entropy penalties $\gamma=0$ and $\gamma=5$. \glnote{Double check}
+
+The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
+proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
+AVO. We see that the distributions both arrive
+in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
+$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
+distance with $\bftheta^*$, both models closely match the observed data, as shown
+in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
+results when using entropic regularization. This suggests either a relatively flat
+landscape around $\bftheta^*$ or that the observed data can in this case also be
+reproduced with the  predictive distribution $\qxpsi$.
+Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
+$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ fall
+towards $0$, which indicates  convergence towards a distribution that the critic
+cannot distinguish from the true distribution.
 
@@
 
-\appendix
-
-\section{Further experiments}
-
-\subsection{Electron--positron annihilation}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
-\caption{Electron--positron annihilation.
-({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
-         Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
-         The density of the penalized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
-({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
-({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
- }\label{fig:weinberg}
-\end{figure}
-
-As a more realistic example, we here consider a (simplified) simulator from
-particle physics for electron--positron collisions resulting in muon--antimuon
-pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
-distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
-polar angle of the outgoing muon with respect  to the originally incoming
-electron. Neglecting measurement uncertainty induced from the particle detectors,
-this random variable is approximately distributed as
-\begin{equation}\label{eq:simple_weinberg}
-p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
-\end{equation}
-where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
-function. Due to the linear term in the expression, the density $p(\mathbf{x} |
-E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
-size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
-and $G_f$ (the Fermi constant) through the coefficient function $c$.
-
-A typical physics simulator for this process includes a more precise treatment of the
-quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering
-using \texttt{MadGraph}~\cite{Alwall:2011uj},  ionization of matter in the
-detector due to the passage of the out-going $\mu^+\mu^-$ particles using
-\texttt{GEANT4}~\cite{Agostinelli:2002hh}, electronic noise and other details of the sensors
-that measure the ionization signal, and the deterministic algorithms that
-estimate the polar angle $A$ based on the sensor readouts. The simulation of
-this process is highly non-trivial as is the space of latent variables $\mathcal{Z}$.
-
-In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
-using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
-factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
-q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
-and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
-$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
-$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
-entropy penalties $\gamma=0$ and $\gamma=5$.
-
-The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
-proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
-AVO. We see that the distributions both arrive
-in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
-$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
-distance with $\bftheta^*$, both models closely match the observed data, as shown
-in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
-results when using entropic regularization. This suggests either a relatively flat
-landscape around $\bftheta^*$ or that the observed data can in this case also be
-reproduced with the  predictive distribution $\qxpsi$.
-Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
-$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
-towards $0$, which indicates  convergence towards a distribution that the critic
-cannot distinguish from the true distribution.
+% \appendix
+%
+% \section{Further experiments}
 

### 2018-02-01 code/toy-poisson.py

@@ -54,7 +54,7 @@ def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
     params = {"W": [glorot_uniform(n_hidden, n_features, rng),
                     glorot_uniform(n_hidden, n_hidden, rng),
-                    glorot_uniform(n_hidden, 0, rng)],
+                    glorot_uniform(n_hidden, 0, rng, scale=0.01)],
               "b": [np.zeros(n_hidden),
                     np.zeros(n_hidden),
                     np.zeros(1)]}

### 2018-02-01 code/weinberg.py

@@ -21,7 +21,7 @@ seed = 777
 rng = check_random_state(seed)
 
 batch_size = 64
-n_epochs = 300+1
+n_epochs = 1000+1
 lambda_gp = 0.0025
 
 true_theta = np.array([(42.0-40)/(50-40),
@@ -89,7 +89,7 @@ def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
     params = {"W": [glorot_uniform(n_hidden, n_features, rng),
                     glorot_uniform(n_hidden, n_hidden, rng),
-                    glorot_uniform(n_hidden, 0, rng)],
+                    glorot_uniform(n_hidden, 0, rng, scale=0.01)],
               "b": [np.zeros(n_hidden),
                     np.zeros(n_hidden),
                     np.zeros(1)]}
@@ -184,12 +184,13 @@ for state in history:
     # Training loop
     init_critic = copy.copy(state["params_critic"])
     opt_critic = AdamOptimizer(grad_loss_critic, state["params_critic"],
-                               step_size=0.01)
+                               step_size=0.005, b1=0.05, b2=0.05)
     opt_proposal = AdamOptimizer(approx_grad_u, state["params_proposal"],
-                                 step_size=0.01)
+                                 step_size=0.005, b1=0.05, b2=0.05)
 
-    opt_critic.step(1000)
+    opt_critic.step(100)
     opt_critic.move_to(state["params_critic"])
+    opt_critic.reset()
 
     for i in range(n_epochs):
         # fit simulator
@@ -197,14 +198,14 @@ for state in history:
         opt_proposal.move_to(state["params_proposal"])
 
         # fit critic
-        opt_critic.reset()   # reset moments
-        opt_critic.step(100)
+        #opt_critic.reset()   # reset moments
+        opt_critic.step(10)
         opt_critic.move_to(state["params_critic"])
 
         state["loss_d"].append(-loss_critic(state["params_critic"], i,
                                             batch_size=10000))
 
-        print(i, state["gamma"], state["params_proposal"], state["loss_d"][-1])
+        print(i, state["gamma"], state["params_proposal"], true_theta - state["params_proposal"]["mu"], state["loss_d"][-1])
 
 
 # Plot
@@ -261,6 +262,9 @@ if make_plots:
             X_[j, :] = simulator(theta, 1).ravel()
[…]

### 2018-02-01 icml/avo.tex
save

@@
 
-% \subsection{Benchmarks}
-%
-% \glnote{Compare against ABC and others.}
-% \glnote{Sections 6.1 and 6.4 of Bernton.}
+\subsection{Benchmarks}
+
+\glnote{Compare against ABC and others.}
+\glnote{Sections 6.1 and 6.4 of Bernton.}
 
@@
 
-In this note, we develop a likelihood-free inference algorithm for
+In this work, we develop a likelihood-free inference algorithm for
 non-differentiable, implicit generative models. The algorithm combines
@@
 
+\glnote{to update}
 Preliminary results on toy problems with discrete and continuous data validate
@@
 fledged scientific simulator -- which we plan to achieve for a next version of
-this work \glnote{to update}. In terms of method, several components need further investigation.
+this work. In terms of method, several components need further investigation.
 First, we need to better study the interplay between the entropic regularization and the

### 2018-02-06 icml/avo.tex
save

@@
 % the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{XXXXX}
+\icmlkeywords{likelihood-free inference, implicit models, generative adversarial networks}
 
@@
 algorithm for fitting a non-differentiable generative model incorporating ideas
-from generative adversarial networks (GANs) and variational optimization. We adapt the
+from generative adversarial networks, variational optimization and empirical Bayes. We adapt the
 training procedure of Wasserstein GANs by replacing the
@@
 The objective of the algorithm is to match the marginal distribution of
-the generated data to the empirical distribution of the observations.
+the synthetic data to the empirical distribution of observations.
 
@@
 ratio $\log \tfrac{p}{q}$, thereby effectively using the same core principle as used
-in GANs~\cite{2016arXiv161003483M}.
+in GANs~\cite{2016arXiv161003483M}. With a similar objective, \cite{2017arXiv171203353M}
+adapt variational inference to a non-differentiable cardiac simulator by maximizing
+the ELBO using Bayesian optimization, hence bypassing altogether the need for gradient estimates.
 

### 2018-02-06 code/weinberg.py

@@ -21,8 +21,8 @@ seed = 777
 rng = check_random_state(seed)
 
 batch_size = 64
-n_epochs = 1000+1
-lambda_gp = 0.0025
+n_epochs = 5000+1
+lambda_gp = 0.001
 
 true_theta = np.array([(42.0-40)/(50-40),
                        (0.9 - 0.5) / (1.5-0.5)])
@@ -184,11 +184,11 @@ for state in history:
     # Training loop
     init_critic = copy.copy(state["params_critic"])
     opt_critic = AdamOptimizer(grad_loss_critic, state["params_critic"],
-                               step_size=0.005, b1=0.05, b2=0.05)
+                               step_size=0.0025, b1=0.025, b2=0.025)
     opt_proposal = AdamOptimizer(approx_grad_u, state["params_proposal"],
-                                 step_size=0.005, b1=0.05, b2=0.05)
+                                 step_size=0.0025, b1=0.025, b2=0.025)
 
-    opt_critic.step(100)
+    opt_critic.step(10000)
     opt_critic.move_to(state["params_critic"])
     opt_critic.reset()
 

### 2018-02-09 icml/avo.tex
save

@@
 
-\subsection{Benchmarks}
-
-\glnote{Compare against ABC and others.}
-\glnote{Sections 6.1 and 6.4 of Bernton.}
+% \subsection{Benchmarks}
+%
+% \glnote{Compare against ABC and others.}
+% \glnote{Sections 6.1 and 6.4 of Bernton.}
 
@@
 
-\glnote{to update}
+% \glnote{to update}
 Preliminary results on toy problems with discrete and continuous data validate
@@
 
-% \section*{Acknowledgements}
+\section*{Acknowledgements}
 
-% We would like to thank Lukas Heinrich for helpful comments regarding
-% the electron--positron annihilation simulation. We would also like to thank
-% Rajesh Ranganath and Dustin Tran for enlightening discussions and feedback.
-% GL and KL are both supported through NSF ACI-1450310, additionally KC is
-% supported through PHY-1505463 and PHY-1205376.
+We would like to thank Lukas Heinrich for helpful comments regarding
+the electron--positron annihilation simulation. We would also like to thank
+Rajesh Ranganath and Dustin Tran for enlightening discussions and feedback.
+GL and KL are both supported through NSF ACI-1450310, additionally KC is
+supported through PHY-1505463 and PHY-1205376.
 

### 2018-02-09 code/multi.py

@@ -15,27 +15,21 @@ from nn import AdamOptimizer
 
 from proposals import make_gaussian_proposal
 from proposals import gaussian_draw
-from proposals import gaussian_logpdf
 from proposals import grad_gaussian_logpdf
 from proposals import grad_gaussian_entropy
 
 from sklearn.utils import check_random_state
 from scipy.spatial.distance import mahalanobis
 
-# What works:
-# remove the reset!!
-# effect of lambda_gp?  lambda_gp=0.1 works
-# effect of gamma? works ok for gamma=0, tighter bands with gamma=1.0
-# refit critic periodically? some zigzag with 1:10 but fine; refitting not necessary
 
 # Global params
 
 seed = 777
 rng = check_random_state(seed)
 
-batch_size = 64
-n_epochs = 1000+1
-lambda_gp = 0.1
+batch_size = 50
+n_epochs = 3000+1
+lambda_gp = 0.01
 gamma = 1.0
 
 true_theta = np.array([1.0, -1.0])
@@ -82,7 +76,7 @@ def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
     params = {"W": [glorot_uniform(n_hidden, n_features, rng),
                     glorot_uniform(n_hidden, n_hidden, rng),
-                    glorot_uniform(n_hidden, 0, rng)],
+                    glorot_uniform(n_hidden, 0, rng, scale=0.01)],
               "b": [np.zeros(n_hidden),
                     np.zeros(n_hidden),
                     np.zeros(1)]}
@@ -93,8 +87,8 @@ params_critic = make_critic(n_features, 10, random_state=rng)
 
 def predict(X, params):
     h = X
-    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
-    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.1)
+    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.2)
+    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.2)
     h = np.dot(params["W"][2], h.T).T + params["b"][2]
     return h
 
@@ -163,26 +157,44 @@ def approx_grad_u(params_proposal, i, gamma=gamma):
 
 # Training loop
 
-opt_critic = AdamOptimizer(grad_loss_critic, params_critic, step_size=0.005, b1=0.05, b2=0.05)
[…]

### 2018-02-09 code/nn.py

@@ -27,6 +27,12 @@ def softmax(X):
     return np.exp(X - logsumexp(X))
 
 
+def batch_normalize(activations):
+    mbmean = np.mean(activations, axis=0, keepdims=True)
+    return (activations - mbmean) / (np.std(activations, axis=0,
+                                            keepdims=True) + 1)
+
+
 # Initializations
 
 def glorot_uniform(fan_in, fan_out, rng, scale=0.1):
@@ -83,6 +89,48 @@ class AdamOptimizer:
             mhat = self.m / (1 - self.b1 ** (self.counter + 1))
             vhat = self.v / (1 - self.b2 ** (self.counter + 1))
             self.x = self.x - self.step_size*mhat/(np.sqrt(vhat) + self.eps)
+            self.counter += 1
+
+        return self.unflatten(self.x)
+
+    def move_to(self, params):
+        for k, v in self.unflatten(self.x).items():
+            params[k] = v
+        return params
+
+
+class RmsPropOptimizer:
+    def __init__(self, grad, init_params, callback=None,
+                 step_size=0.01, gamma=0.9, eps=10**-8):
+        self.grad = grad
+        self.init_params = copy.copy(init_params)
+        self.callback = callback
+        self.step_size = step_size
+        self.gamma = gamma
+        self.eps = eps
+
+        self.flattened_grad, self.unflatten, self.x = flatten_func(
+            self.grad, self.init_params)
+        self.reset()
+
+    def reset(self):
+        self.avg_sq_grad = np.ones(len(self.x))
+        self.counter = 0
+
+    def step(self,  num_iters=1):
+        for i in range(num_iters):
+            g = self.flattened_grad(self.x, self.counter)
+
+            if self.callback:
+                self.callback(self.unflatten(self.x),
+                              self.counter,
+                              self.unflatten(g))
+
+            self.avg_sq_grad = (self.avg_sq_grad *
+                                self.gamma + g**2 * (1 - self.gamma))
+            self.x = (self.x -
+                      self.step_size * g / (np.sqrt(self.avg_sq_grad) +
+                                            self.eps))
             self.counter += 1
 
         return self.unflatten(self.x)

### 2018-02-09 code/proposals.py

@@ -46,10 +46,7 @@ def gaussian_logpdf(params, theta, to_scalar=True):
 
 def gaussian_entropy(params):
     sigma = np.exp(params["log_sigma"])
-    if np.sum(sigma) > 0.05:
-        return np.sum(np.log(sigma * (2. * np.pi * np.e) ** 0.5))
-    else:
-        return np.log(0.05 * (2. * np.pi * np.e) ** 0.5)
+    return np.sum(np.log(sigma * (2. * np.pi * np.e) ** 0.5))
 
 
 grad_gaussian_logpdf = ag.grad(gaussian_logpdf)

### 2018-02-09 code/toy-poisson.py

@@ -15,16 +15,15 @@ from proposals import grad_gaussian_entropy
 
 from sklearn.utils import check_random_state
 
+
 # Global params
 
 seed = 777
 rng = check_random_state(seed)
 
-batch_size = 64   # try increasing, default=64
-n_epochs = 500+1
-lambda_gp = 0.1 # ok~0.1; try reducing (=> worse), try increasing (0.4, not sure what to conclude)
-
-# see zigzag near the end... hence the reset? but works better without in multi.py
+batch_size = 50
+n_epochs = 3000+1
+lambda_gp = 0.001
 
 true_theta = np.array([np.log(7)])
 make_plots = True
@@ -44,11 +43,9 @@ n_features = X_obs.shape[1]
 
 # Critic
 
-gammas = [0.0, 5.0]
+gammas = [0.0, 3.0]
 colors = ["C1", "C2"]
 
-# gammas = [5.0]
-# colors = ["C2"]
 
 def make_critic(n_features, n_hidden, random_state=None):
     rng = check_random_state(random_state)
@@ -60,10 +57,11 @@ def make_critic(n_features, n_hidden, random_state=None):
                     np.zeros(1)]}
     return params
 
+
 def predict(X, params):
     h = X
-    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
-    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.1)
+    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.2)
+    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.2)
     h = np.dot(params["W"][2], h.T).T + params["b"][2]
     return h
 
@@ -74,7 +72,8 @@ history = [{"gamma": gammas[i],
             "color": colors[i],
             "loss_d": [],
             "params_proposal": make_gaussian_proposal(n_params,
-                                                      mu=np.log(10.0)),
+                                                      mu=np.log(5),
+                                                      log_sigma=0.0),
             "params_critic": make_critic(n_features, 10, random_state=rng)}
            for i in range(len(gammas))]
 
@@ -138,34 +137,62 @@ for state in history:
 
         M = len(thetas)
 
-        for k in grad_u:
-            grad_u[k] = 1. / M * grad_u[k] + state["gamma"] * grad_ent[k]
+        if state["gamma"] == 0.0:
[…]

### 2018-02-09 code/weinberg.py

@@ -20,12 +20,13 @@ from scipy.spatial.distance import mahalanobis
 seed = 777
 rng = check_random_state(seed)
 
-batch_size = 64
-n_epochs = 5000+1
-lambda_gp = 0.001
+batch_size = 50
+n_epochs = 3000+1
+lambda_gp = 0.0
 
-true_theta = np.array([(42.0-40)/(50-40),
-                       (0.9 - 0.5) / (1.5-0.5)])
+true_theta = np.array([(43.0-40)/(50-40),
+                       (1.1 - 0.5) / (1.5-0.5)])
+
 make_plots = True
 plt.rcParams["figure.dpi"] = 200
 
@@ -74,14 +75,14 @@ def simulator(theta, n_samples, random_state=None):
 
     return samples.reshape(-1, 1)
 
-X_obs = simulator(true_theta, 50000, random_state=rng)
+X_obs = simulator(true_theta, 20000, random_state=rng)
 n_params = len(true_theta)
 n_features = X_obs.shape[1]
 
 
 # Critic
 
-gammas = [0.0, 5.0]
+gammas = [0.0, 3.0]
 colors = ["C1", "C2"]
 
 
@@ -95,11 +96,13 @@ def make_critic(n_features, n_hidden, random_state=None):
                     np.zeros(1)]}
     return params
 
+params_critic = make_critic(n_features, 50, random_state=rng)
+
 
 def predict(X, params):
     h = X
-    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.1)
-    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.1)
+    h = relu(np.dot(params["W"][0], h.T).T + params["b"][0], alpha=0.2)
+    h = relu(np.dot(params["W"][1], h.T).T + params["b"][1], alpha=0.2)
     h = np.dot(params["W"][2], h.T).T + params["b"][2]
     return h
 
@@ -111,8 +114,8 @@ history = [{"gamma": gammas[i],
             "loss_d": [],
             "params_proposal": make_gaussian_proposal(n_params,
                                                       mu=0.5,
-                                                      log_sigma=np.log(0.1)),
-            "params_critic": make_critic(n_features, 10, random_state=rng)}
+                                                      log_sigma=np.log(0.25)),
+            "params_critic": make_critic(n_features, 20, random_state=rng)}
            for i in range(len(gammas))]
 
 
[…]

### 2018-02-10 colt/avo.tex
colt submission

--- 2018-02-09 icml/avo.tex
@@
-\documentclass{article}
-\usepackage[accepted]{icml2018}
- % If your paper is accepted, change the options for the package
-% aistats2018 as follows:
-%
-%\usepackage[accepted]{aistats2018}
-%
-% This option will print headings for the title of your paper and
-% headings for the authors names, plus a copyright note at the end of
-% the first column of the first page.
+\documentclass[12pt]{colt2018} % Anonymized submission
+% \documentclass{colt2017} % Include author names
 
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsthm}
+% The following packages will be automatically loaded:
+% amsmath, amssymb, natbib, graphicx, url, algorithm2e
+
+\title[Adversarial Variational Optimization]{Adversarial Variational Optimization\\
+                                             of Non-Differentiable Simulators}
+\usepackage{times}
+ % Use \Name{Author Name} to specify the name.
+ % If the surname contains spaces, enclose the surname
+ % in braces, e.g. \Name{John {Smith Jones}} similarly
+ % if the name has a "von" part, e.g \Name{Jane {de Winter}}.
+ % If the first letter in the forenames is a diacritic
+ % enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}
+
+ % Two authors with the same address
+  % \coltauthor{\Name{Author Name1} \Email{abc@sample.com}\and
+  %  \Name{Author Name2} \Email{xyz@sample.com}\\
+  %  \addr Address}
+
+ % Three or more authors with the same address:
+ % \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
+ %  \Name{Author Name2} \Email{an2@sample.com}\\
+ %  \Name{Author Name3} \Email{an3@sample.com}\\
+ %  \addr Address}
+
+ % Authors with different addresses:
+ \coltauthor{\Name{Gilles Louppe} \Email{g.louppe@uliege.be}\\
+ \addr University of Li\`{e}ge, Belgium
+ \AND
+ \Name{Kyle Cranmer} \Email{kyle.cranmer@nyu.edu}\\
+ \addr New York University, USA
+ }
+
+ % \usepackage{amsfonts}
+ % \usepackage{amssymb}
+ % \usepackage{amsmath}
+ % \usepackage{amsthm}
 
@@
  \usepackage{slashed}
-\usepackage{graphicx}
+ % \usepackage{graphicx}
  \usepackage{color}
@@
  % Attempt to make hyperref and algorithmic work together better:
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
+ % \newcommand{\theHalgorithm}{\arabic{algorithm}}
 
@@
 
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\icmltitlerunning{Adversarial Variational Optimization of Non-Differentiable Simulators}
+ % \theoremstyle{plain}
+ % \newtheorem{theorem}{Theorem}
+ % \newtheorem{proposition}[theorem]{Proposition}
 
@@
 
-% If your paper is accepted and the title of your paper is very long,
-% the style will print as headings an error message. Use the following
-% command to supply a shorter title of your paper so that it can be
-% used as headings.
-%
-%\runningtitle{I use this title instead because the last one was very long}
-
-% If your paper is accepted and the number of authors is large, the
-% style will print as headings an error message. Use the following
-% command to supply a shorter version of the authors names so that
-% they can be used as headings (for example, use only the surnames)
-%
-%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
-
-\twocolumn[
-
-\icmltitle{Adversarial Variational Optimization of\\
-           Non-Differentiable Simulators}
-
-\begin{icmlauthorlist}
-\icmlauthor{Gilles Louppe}{uliege}
-\icmlauthor{Kyle Cranmer}{nyu}
-\end{icmlauthorlist}
-
-\icmlaffiliation{uliege}{University of Li{\`e}ge, Belgium}
-\icmlaffiliation{nyu}{New York University, USA}
-
-\icmlcorrespondingauthor{Gilles Louppe}{g.louppe@uliege.be}
-\icmlcorrespondingauthor{Kyle Cranmer}{kyle.cranmer@nyu.edu}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{likelihood-free inference, implicit models, generative adversarial networks}
-
-\vskip 0.3in
-]
-
-\printAffiliationsAndNotice{\icmlEqualContribution}
+\maketitle
 
@@
 
+\begin{keywords}
+List of keywords
+\end{keywords}
+
 % ==============================================================================
@@
 Approximate Bayesian Computation (ABC) and density estimation-by-comparison
-algorithms~\cite{beaumont2002approximate, marjoram2003markov,
+algorithms~\citep{beaumont2002approximate, marjoram2003markov,
 sisson2007sequential, sisson2011likelihood, marin2012approximate,
@@
 In parallel, with the introduction of variational
-auto-encoders~\cite{DBLP:journals/corr/KingmaW13} and generative adversarial
-networks~\cite{goodfellow2014generative}, there has been a vibrant research
+auto-encoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
+networks~\citep{goodfellow2014generative}, there has been a vibrant research
 program around implicit generative models based on neural
-networks~\cite{2016arXiv161003483M}.  While some of these models
+networks~\citep{2016arXiv161003483M}.  While some of these models
 also do not admit a tractable density, they are all differentiable by construction.
@@
 training procedure of generative adversarial
-networks~\cite{goodfellow2014generative} by replacing the implicit generative
+networks \citep{goodfellow2014generative} by replacing the implicit generative
 network with a domain-based scientific simulator, and solve the resulting
 non-differentiable minimax problem by minimizing variational upper
-bounds~\cite{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
+bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
 objectives.
@@
 Generative adversarial networks (GANs) were first proposed by
-\cite{goodfellow2014generative} as a way to build an implicit generative model
+\citep{goodfellow2014generative} as a way to build an implicit generative model
 capable of producing samples from random noise $\bfz$. More specifically,
@@
 trained to optimality before each parameter update of the generator, it can
-be shown that the original adversarial learning procedure~\cite{goodfellow2014generative} amounts to minimizing
+be shown that the original adversarial learning procedure~\citep{goodfellow2014generative} amounts to minimizing
 the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\bftheta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\bftheta)$.
 
-As explored in \cite{2017arXiv170104862A}, GANs remain remarkably
+As explored in \citep{2017arXiv170104862A}, GANs remain remarkably
 difficult to train because of vanishing gradients as $d$ saturates, or because of
 unreliable updates when the training procedure is relaxed. As a remedy,
-Wasserstein GANs~\cite{2017arXiv170107875A} (WGANs) reformulate the adversarial
+Wasserstein GANs~\citep{2017arXiv170107875A} (WGANs) reformulate the adversarial
 setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
@@
 \end{align}
-Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
+Under the WGAN-GP formulation of \citep{2017arXiv170400028G}
 for stabilizing the optimization procedure,
@@
 
-Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\cite{2011arXiv1106.4487W} are general
+Variational optimization~\citep{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
 optimization techniques that can be used to form a differentiable bound
@@
 
-Under mild restrictions outlined in  \cite{2012arXiv1212.4507S}, the bound
+Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
 $U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
@@
     \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
-    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
- %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
-   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
+    &= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
+   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
+   &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
     &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
@@
 variance of the noisy estimator of the gradients may be too large to be useful in many problems.
-We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
-algorithm~\cite{2011arXiv1106.4487W} or variance reduction techniques as
-control variates~\cite{2017arXiv171100123G} or finite differences~\cite{buesingstochastic}.
+We use Adam~\citep{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
+algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques such as
+control variates~\citep{2017arXiv171100123G} or finite differences~\citep{buesingstochastic}.
 
@@
 We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
-in \cite{2016arXiv161009033R}, though more complicated than a simple
+in \citep{2016arXiv161009033R}, though more complicated than a simple
 reparametrization of normally distributed noise $\bfz$ through a differentiable
@@
 We can view the optimization of $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$.
-If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\cite{rubin1984}.
+If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\citep{rubin1984}.
 When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{figures/poisson.pdf}
+\caption{Discrete Poisson model with unknown mean.
+ ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=3$, the distributions correctly concentrate their density around
+            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=3$) results in a tighter density.
+ ({\it Top right}) Model distributions $\qxpsi$ after training.
+This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution, with better results when enabling entropic regularization.
+ ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
+ }\label{fig:poisson}
+\end{figure}
+
 As a first illustrative experiment, we evaluate inference for a discrete Poisson
@@
 The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
-Algorithm~\ref{alg:avo} is run for 500 epochs with mini-batches of size $M=64$
+Algorithm~\ref{alg:avo} is run for 3000 iterations with mini-batches of size $M=50$
 and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
 hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
-$n_\text{critic}=10$ iterations with a step size $\alpha=0.005$ and $\beta_1=\beta_2=0.05$.
+$n_\text{critic}=100$ iterations with a step size $\alpha=0.001$, with $\beta_1=0.5$ and $\beta_2=0.9$.
 % Its inner first and second moment vectors are reset at
@@
 estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
-$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(10)$ and unit variance. At
-each epoch, parameters $\bfpsi$ are updated by taking one Adam step and the same hyper-parameters as for the critic.
+$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
+each iteration, parameters $\bfpsi$ are updated by taking one Adam step and the same hyper-parameters as for the critic.
 The gradient penalty coefficient is set to
-$\lambda=0.1$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.
+$\lambda=0.001$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=3$.
 
@@
 distributions $q(\bftheta|\bfpsi)$ after AVO.  For
-both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
+both $\gamma=0$ and $\gamma=3$, the proposal distributions correctly concentrate
 their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
 the effect of entropic regularization,
-the proposal distribution for $\gamma=5$ concentrates its mass tightly,
+the proposal distribution for $\gamma=3$ concentrates its mass tightly,
 yielding in this case precise inference.  The top right plot compares the
@@
 model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
-of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
+of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=3$,
 the curves fall towards $0$, which indicates that
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
-\caption{Discrete Poisson model with unknown mean.
- ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
-            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=5$) results in a tighter density.
- ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
- ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
- }\label{fig:poisson}
-\end{figure}
-
 \subsection{Multidimensional continuous data}
 
-As a second toy example, we consider a generator producing
+As a second illustrative example, we consider a generator producing
 5-dimensional continuous data, as originally specified in Section 4.2 of
-\cite{cranmer2015approximating}.
-% More specifically, we consider the following
-% generative process:
-% \begin{itemize}
-% \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
-% $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
-% $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
-% $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
-% $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
-% $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
-% \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
-% semi-positive definite $5 \times 5$ matrix defining a fixed projection
-% of $\bfz$ into the observed space.
-% \end{itemize}
+\citep{cranmer2015approximating}.
+More specifically, we consider the following
+generative process:
+\begin{itemize}
+\item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
+$z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
+$z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
+$z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
+$z_3 \sim {\text{Exponential}(\lambda=3)}$, and
+$z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
+\item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
+semi-positive definite $5 \times 5$ matrix defining a fixed projection
+of $\bfz$ into the observed space.
+\end{itemize}
 Again, AVO does not have access to the density or its gradient, only samples from the generative model.
@@
 initialized with zero mean and unit variance.
-Hyper-parameters are set to $M=64$, $n_\text{critic}=10$, $\lambda=0.1$, $\gamma=1$ and
-Adam configured with a step size $\alpha=0.005$ and $\beta_1=\beta_2=0.05$.
+Hyper-parameters are set to $M=50$, $n_\text{critic}=100$, $\lambda=0.01$, $\gamma=1$ and
+Adam configured with a step size $\alpha=0.001$ with $\beta_1=0.5$ and $\beta_2=0.9$.
 The architecture of the critic is the same as in the previous example.
@@
 \centering
-\includegraphics[width=0.48\textwidth]{figures/multi.pdf}
+\includegraphics[width=0.6\textwidth]{figures/multi.pdf}
 \caption{Multidimensional continuous data.
@@
               Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
- ({\it Right}) Density $q(\bftheta|\bfpsi)$ after adversarial variational optimization ($\gamma=10$).
+ ({\it Right}) Density $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
                The proposal density correctly converges towards a distribution whose density concentrates around $\bftheta^* = (1, -1)$.
@@
 
-\begin{figure}[h]
+\begin{figure}
 \centering
-\includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
+\includegraphics[width=0.6\textwidth]{figures/weinberg.pdf}
 \caption{Electron--positron annihilation.
@@
          Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
-         The density of the entropically regularized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
-({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
+         The density of the proposal distribution concentrates in the neighborhood of $\bftheta^*$.
+({\it Top right}) Model distribution $\qxpsi$ after training. Despite the width of the proposal distribution with respect to $\bftheta^*$, the synthetic data distribution
+matches the observed data.
 ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
@@
 quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering
-using \texttt{MadGraph}~\cite{Alwall:2011uj},  ionization of matter in the
+using \texttt{MadGraph}~\citep{Alwall:2011uj},  ionization of matter in the
 detector due to the passage of the out-going $\mu^+\mu^-$ particles using
-\texttt{GEANT4}~\cite{Agostinelli:2002hh}, electronic noise and other details of the sensors
+\texttt{GEANT4}~\citep{Agostinelli:2002hh}, electronic noise and other details of the sensors
 that measure the ionization signal, and the deterministic algorithms that
@@
 In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
-using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
+using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=1.1)$. The simulator parameters are modeled with a
 factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
 q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
-and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
-$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
-$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
-entropy penalties $\gamma=0$ and $\gamma=5$. \glnote{Double check}
+and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=50$,
+$n_\text{critic}=100$, $\lambda=0.0001$, $\gamma=5$ and Adam configured with $\alpha=0.001$,
+$\beta_1=0.5$ and $\beta_2=0.9$.
+% As with the first example, we compare
+% entropy penalties $\gamma=0$ and $\gamma=5$. \glnote{Double check}
 
 The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
-proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
-AVO. We see that the distributions both arrive
-in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
-$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
-distance with $\bftheta^*$, both models closely match the observed data, as shown
-in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
-results when using entropic regularization. This suggests either a relatively flat
+proposal distributions $q(\bftheta|\bfpsi)$ after
+AVO. We see that the proposal distribution concentrates
+in the neighborhood of $\bftheta^*$.  Despite the width of the resulting proposal distribution with respect to $\bftheta^*$
+in the top right plot of  Figure~\ref{fig:weinberg}. This suggests either a relatively flat
 landscape around $\bftheta^*$ or that the observed data can in this case also be
@@
 Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
-$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ fall
+the variational upper bound $-U_d$ fall
 towards $0$, which indicates  convergence towards a distribution that the critic
@@
 the proposed method can be considered as a direct adaptation of
-generative adversarial networks~\cite{goodfellow2014generative,2017arXiv170104862A} to
-non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S}.
+generative adversarial networks~\citep{goodfellow2014generative,2017arXiv170104862A} to
+non-differentiable simulators using variational optimization~\citep{2012arXiv1212.4507S}.
 
@@
 
-Perhaps the closest to our work is~\cite{bernton2017inference}, which uses the
+Perhaps the closest to our work is~\citep{bernton2017inference}, which uses the
 Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
@@
 approximation based on projection of the data onto Hilbert space-filling curves.
-Their Remark 5.1 points out that~\cite{montavon2016wasserstein} proposed an
+Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
 approximation of the gradient of an entropy-regularized Wasserstein distance,
@@
 In the case of small $N$, we note however that fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
-empirical distributions~\cite{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}
+empirical distributions~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}
 are alternatives to adversarial learning that are worth considering.
@@
 inference is intimately tied to a class of algorithms that can be framed as
-density estimation-by-comparison, as reviewed in \cite{2016arXiv161003483M}. In most cases, these
+density estimation-by-comparison, as reviewed in \citep{2016arXiv161003483M}. In most cases, these
 inference algorithms are formulated as an iterative two-step process where the
@@
 direction includes those that rely on a classifier to estimate the discrepancy between the observed data and the model distributions
-\cite{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
-Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
+\citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
+Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\citep{meeds2015hamiltonian}
 estimate gradients with respect
@@
 Likewise, AVO closely relates to recent extensions of GANs, such as
-Adversarial Learned Inference (ALI)~\cite{dumoulin2016adversarially},
-Bidirectional GANs (BiGANs)~\cite{donahue2016adversarial},
-$\alpha$-GAN~\cite{rosca2017variational}, Adversarial Variational Bayes
-(AVB)~\cite{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
-algorithm of  \cite{2017arXiv170208235H}, which
+Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
+Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
+$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
+(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
+algorithm of  \citep{2017arXiv170208235H}, which
 add an inference network to the generative model.  Each of these assume a
@@
 been efforts to extend variational inference to intractable likelihoods;
-however, many require restrictive assumptions.  In \cite{tran2017variational},
+however, many require restrictive assumptions.  In \citep{tran2017variational},
 the authors consider Variational Bayes with an Intractable Likelihood (VBIL). In
@@
 in evaluating the ABC likelihood for high-dimensional data and small $\epsilon$.
-More recently, \cite{2017arXiv170208896T} adapted variational inference to
+More recently, \citep{2017arXiv170208896T} adapted variational inference to
 hierarchical implicit models defined on simulators. In this work, the authors
@@
 ratio $\log \tfrac{p}{q}$, thereby effectively using the same core principle as used
-in GANs~\cite{2016arXiv161003483M}. With a similar objective, \cite{2017arXiv171203353M}
+in GANs~\citep{2016arXiv161003483M}. With a similar objective, \citep{2017arXiv171203353M}
 adapt variational inference to a non-differentiable cardiac simulator by maximizing
@@
 
-Lastly, we make a connection to Operator Variational Inference (OPVI)~\cite{2016arXiv161009033R},
+Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
 which is a generalization of variational inference formulated as the following optimization problem:
@@
 
-% \glnote{to update}
-Preliminary results on toy problems with discrete and continuous data validate
+Results on illustrative problems with discrete and continuous data validate
 the proposed method. While the obtained results are encouraging, the complete
 validation of the method remains to be carried out in real conditions on a full
-fledged scientific simulator -- which we plan to achieve for a next version of
-this work. In terms of method, several components need further investigation.
+fledged scientific simulator -- which is planned for future works.
+In terms of method, several components need further investigation.
 First, we need to better study the interplay between the entropic regularization and the
@@
 
-% ==============================================================================
-
-\section*{Acknowledgements}
-
-We would like to thank Lukas Heinrich for helpful comments regarding
+% Acknowledgments---Will not appear in anonymized version
+\acks{We would like to thank Lukas Heinrich for helpful comments regarding
 the electron--positron annihilation simulation. We would also like to thank
@@
 GL and KL are both supported through NSF ACI-1450310, additionally KC is
-supported through PHY-1505463 and PHY-1205376.
+supported through PHY-1505463 and PHY-1205376.}
 
-\bibliography{bibliography.bib}
-\bibliographystyle{icml2018}
-
-% ==============================================================================
-
-% \appendix
-%
-% \section{Further experiments}
+\bibliography{bibliography}
 

### 2018-02-10 code/weinberg.py

@@ -17,15 +17,15 @@ from scipy.spatial.distance import mahalanobis
 
 # Global params
 
-seed = 777
+seed = 123
 rng = check_random_state(seed)
 
 batch_size = 50
-n_epochs = 3000+1
-lambda_gp = 0.0
+n_epochs = 1000+1
+lambda_gp = 0.0001  # works well for 0.0; 0.001
 
-true_theta = np.array([(43.0-40)/(50-40),
-                       (1.1 - 0.5) / (1.5-0.5)])
+true_theta = np.array([(42.0-40) / (50-40),
+                       (1.1-0.5) / (1.5-0.5)])
 
 make_plots = True
 plt.rcParams["figure.dpi"] = 200
@@ -82,8 +82,8 @@ n_features = X_obs.shape[1]
 
 # Critic
 
-gammas = [0.0, 3.0]
-colors = ["C1", "C2"]
+gammas = [5.0]
+colors = ["C1"]
 
 
 def make_critic(n_features, n_hidden, random_state=None):
@@ -114,8 +114,8 @@ history = [{"gamma": gammas[i],
             "loss_d": [],
             "params_proposal": make_gaussian_proposal(n_params,
                                                       mu=0.5,
-                                                      log_sigma=np.log(0.25)),
-            "params_critic": make_critic(n_features, 20, random_state=rng)}
+                                                      log_sigma=np.log(0.1)), 
+            "params_critic": make_critic(n_features, 50, random_state=rng)}
            for i in range(len(gammas))]
 
 
@@ -195,7 +195,7 @@ for state in history:
                                  step_size=10e-4, b1=0.5, b2=0.9)
 
     print(predict(X_obs, state["params_critic"]).mean())
-    opt_critic.step(1500)
+    opt_critic.step(1000)
     opt_critic.move_to(state["params_critic"])
     print(predict(X_obs, state["params_critic"]).mean())
 
@@ -203,10 +203,8 @@ for state in history:
         # fit simulator
         opt_proposal.step(1)
         opt_proposal.move_to(state["params_proposal"])
-        #opt_proposal.reset()
 
         # fit critic
-        #opt_critic.reset()   # reset moments
         opt_critic.step(10)
         opt_critic.move_to(state["params_critic"])
 
@@ -215,7 +213,7 @@ for state in history:
[…]

### 2018-02-11 colt/avo.tex
keywords
save

@@
 \begin{keywords}
-List of keywords
+Likelihood-free inference, implicit models, adversarial training, variational inference.
 \end{keywords}
@@
 \begin{align}
-    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
+    {\cal L}_d =\,&  \mathcal{L}_W + \lambda_{GP} \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
     {\cal L}_g =\,& - \mathcal{L}_W
@@
 The gradient penalty coefficient is set to
-$\lambda=0.001$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=3$.
+$\lambda_{GP}=0.001$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=3$.
 
@@
 initialized with zero mean and unit variance.
-Hyper-parameters are set to $M=50$, $n_\text{critic}=100$, $\lambda=0.01$, $\gamma=1$ and
+Hyper-parameters are set to $M=50$, $n_\text{critic}=100$, $\lambda_{GP}=0.01$, $\gamma=1$ and
 Adam configured with a step size $\alpha=0.001$ with $\beta_1=0.5$ and $\beta_2=0.9$.
@@
 $n_\text{critic}=100$, $\lambda=0.0001$, $\gamma=5$ and Adam configured with $\alpha=0.001$,
-$\beta_1=0.5$ and $\beta_2=0.9$.
+$\beta_1=0.5$ and $\beta_2=0.9$. For the critic $d$, we use a 3-layer MLP with 50
+hidden nodes per layer and ReLU activations.
 % As with the first example, we compare
@@
 AVO. We see that the proposal distribution concentrates
-in the neighborhood of $\bftheta^*$.  Despite the width of the resulting proposal distribution with respect to $\bftheta^*$
+in the neighborhood of $\bftheta^*$.
+Despite the width of the resulting proposal distribution with respect to $\bftheta^*$,
+the synthetic data closely match with the observed data, as shown
 in the top right plot of  Figure~\ref{fig:weinberg}. This suggests either a relatively flat

### 2018-02-11 colt/avo.tex
save

@@
 landscape around $\bftheta^*$ or that the observed data can in this case also be
-reproduced with the  predictive distribution $\qxpsi$.
+reproduced through a different predictive distribution $\qxpsi$.
 Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both

### 2018-07-20 colt/avo.tex
update code

@@
-\documentclass[12pt]{colt2018} % Anonymized submission
+\documentclass[anon,12pt]{colt2018} % Anonymized submission
 % \documentclass{colt2017} % Include author names
@@
         {\it Outputs:} proposal distribution $q(\bftheta|\bfpsi)$, such that $\qxpsi \approx p_r(\mathbf{x})$.\\
-        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
+        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda_{GP}$; the entropy penalty coefficient $\gamma$.
     \end{flushleft}
@@
                     \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
-                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \bfphi) - d(\mathbf{x}_m; \bfphi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \bfphi) ||_2 - 1)^2 $}
+                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \bfphi) - d(\mathbf{x}_m; \bfphi) + \lambda_{GP} (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \bfphi) ||_2 - 1)^2 $}
                 \EndFor

### 2018-07-20 code/multi.py

@@ -11,7 +11,7 @@ import theano
 
 from nn import glorot_uniform
 from nn import relu
-from nn import AdamOptimizer
+from nn import RmsPropOptimizer
 
 from proposals import make_gaussian_proposal
 from proposals import gaussian_draw
@@ -27,10 +27,10 @@ from scipy.spatial.distance import mahalanobis
 seed = 777
 rng = check_random_state(seed)
 
-batch_size = 50
-n_epochs = 3000+1
-lambda_gp = 0.01
-gamma = 1.0
+batch_size = 64
+n_epochs = 500+1
+lambda_reg = 20.
+gamma = 0.005
 
 true_theta = np.array([1.0, -1.0])
 make_plots = True
@@ -54,6 +54,7 @@ p0 = LinearTransform(Join(components=[
                                         Normal(mu=2, sigma=0.5)]),
                     Exponential(inverse_scale=3.0),
                     Exponential(inverse_scale=0.5)]), R)
+
 
 def simulator(theta, n_samples, random_state=None):
     A.set_value(theta[0])
@@ -95,7 +96,7 @@ def predict(X, params):
 grad_predict_critic = ag.elementwise_grad(predict)
 
 
-def loss_critic(params_critic, i, lambda_gp=lambda_gp, batch_size=batch_size):
+def loss_critic(params_critic, i, lambda_reg=lambda_reg, batch_size=batch_size):
     y_critic = np.zeros(batch_size)
     # y_critic[:batch_size // 2] = 0.0  # 0 == fake
     y_critic[batch_size // 2:] = 1.0
@@ -115,14 +116,19 @@ def loss_critic(params_critic, i, lambda_gp=lambda_gp, batch_size=batch_size):
     y_pred = predict(X, params_critic)
     l_wgan = np.mean(-y_critic * y_pred + (1. - y_critic) * y_pred)
 
-    # Gradient penalty
-    eps = rng.rand(batch_size // 2, 1)
-    _X_hat = eps * _X_obs + (1. - eps) * _X_gen
-    grad_Dx = grad_predict_critic(_X_hat, params_critic)
-    norms = np.sum(grad_Dx ** 2, axis=1) ** 0.5
-    l_gp = np.mean((norms - 1.0) ** 2.0)
+    # # Gradient penalty
+    # eps = rng.rand(batch_size // 2, 1)
+    # _X_hat = eps * _X_obs + (1. - eps) * _X_gen
[…]

### 2018-07-20 code/nn.py

@@ -117,7 +117,7 @@ class RmsPropOptimizer:
         self.avg_sq_grad = np.ones(len(self.x))
         self.counter = 0
 
-    def step(self,  num_iters=1):
+    def step(self, num_iters=1):
         for i in range(num_iters):
             g = self.flattened_grad(self.x, self.counter)
 

### 2018-07-20 code/toy-poisson.py

@@ -6,6 +6,7 @@ from scipy.stats import poisson
 from nn import glorot_uniform
 from nn import relu
 from nn import AdamOptimizer
+from nn import RmsPropOptimizer
 
 from proposals import make_gaussian_proposal
 from proposals import gaussian_draw
@@ -18,15 +19,24 @@ from sklearn.utils import check_random_state
 
 # Global params
 
-seed = 777
+seed = 666
 rng = check_random_state(seed)
 
-batch_size = 50
-n_epochs = 3000+1
-lambda_gp = 0.001
+batch_size = 64  # was 32
+n_epochs = 500+1  # was 500
+lambda_reg = 20.
 
 true_theta = np.array([np.log(7)])
 make_plots = True
+
+# dirac-gan:
+# add r1 reg                                    OK
+# switch to rmsprop, alpha=0.9, lr=10-4         OK, but adjust lr and batch_size
+# 1:1 d and g updates                           OK
+# no-pretraining                                OK
+#
+# add history (Apple paper)
+# or use historical averaging (Salimans)
 
 
 # Simulator
@@ -43,7 +53,7 @@ n_features = X_obs.shape[1]
 
 # Critic
 
-gammas = [0.0, 3.0]
+gammas = [0.0, 0.005]
 colors = ["C1", "C2"]
 
 
@@ -71,10 +81,12 @@ grad_predict_critic = ag.elementwise_grad(predict)
 history = [{"gamma": gammas[i],
             "color": colors[i],
             "loss_d": [],
+            "mse": [],
             "params_proposal": make_gaussian_proposal(n_params,
                                                       mu=np.log(5),
                                                       log_sigma=0.0),
-            "params_critic": make_critic(n_features, 10, random_state=rng)}
+            "params_critic": make_critic(n_features, 10, random_state=rng),
+            "seed": rng.randint(10000)}
            for i in range(len(gammas))]
 
 
@@ -82,12 +94,12 @@ history = [{"gamma": gammas[i],
 
 for state in history:
     # WGAN + GP
-    def loss_critic(params_critic, i, lambda_gp=lambda_gp, batch_size=batch_size):
+    def loss_critic(params_critic, i, lambda_reg=lambda_reg, batch_size=batch_size):
         y_critic = np.zeros(batch_size)
         # y_critic[:batch_size // 2] = 0.0  # 0 == fake
         y_critic[batch_size // 2:] = 1.0
 
-        rng = check_random_state(i)
[…]

### 2018-07-20 code/weinberg.py

@@ -5,7 +5,7 @@ import copy
 
 from nn import glorot_uniform
 from nn import relu
-from nn import AdamOptimizer
+from nn import RmsPropOptimizer
 
 from proposals import make_gaussian_proposal
 from proposals import gaussian_draw
@@ -20,9 +20,9 @@ from scipy.spatial.distance import mahalanobis
 seed = 123
 rng = check_random_state(seed)
 
-batch_size = 50
-n_epochs = 1000+1
-lambda_gp = 0.0001  # works well for 0.0; 0.001
+batch_size = 64
+n_epochs = 5000+1
+lambda_reg = 20.
 
 true_theta = np.array([(42.0-40) / (50-40),
                        (1.1-0.5) / (1.5-0.5)])
@@ -82,7 +82,7 @@ n_features = X_obs.shape[1]
 
 # Critic
 
-gammas = [5.0]
+gammas = [0.005]
 colors = ["C1"]
 
 
@@ -112,9 +112,10 @@ grad_predict_critic = ag.elementwise_grad(predict)
 history = [{"gamma": gammas[i],
             "color": colors[i],
             "loss_d": [],
+            "mse": [],
             "params_proposal": make_gaussian_proposal(n_params,
                                                       mu=0.5,
-                                                      log_sigma=np.log(0.1)), 
+                                                      log_sigma=0),  # was np.log(0.1)
             "params_critic": make_critic(n_features, 50, random_state=rng)}
            for i in range(len(gammas))]
 
@@ -123,7 +124,7 @@ history = [{"gamma": gammas[i],
 
 for state in history:
     # WGAN + GP
-    def loss_critic(params_critic, i, lambda_gp=lambda_gp,
+    def loss_critic(params_critic, i, lambda_reg=lambda_reg,
                     batch_size=batch_size):
         y_critic = np.zeros(batch_size)
         # y_critic[:batch_size // 2] = 0.0  # 0 == fake
@@ -145,17 +146,22 @@ for state in history:
         y_pred = predict(X, params_critic)
         l_wgan = np.mean(-y_critic * y_pred + (1. - y_critic) * y_pred)
 
-        if lambda_gp == 0.0:
+        if lambda_reg == 0.0:
             return l_wgan
 
-        # Gradient penalty
-        eps = rng.rand(batch_size // 2, 1)
-        _X_hat = eps * _X_obs + (1. - eps) * _X_gen
[…]
